---
title: "SDS 375"
description: Data Visualization in R
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Learn more about creating websites with Distill at:
# https://rstudio.github.io/distill/website.html

# Learn more about publishing to GitHub Pages at:
# https://rstudio.github.io/distill/publish_website.html#github-pages

```

This is the home page for SDS 375, Data Visualization in R. All course materials will be posted on this site.

Instructor: [Claus O. Wilke](https://clauswilke.com)  
Meeting times: TTH 3:30pm to 5:00pm  
Venue: UTC 4.110  
Syllabus: [click here](syllabus.html)  
Upcoming lectures and assignments: [click here](schedule.html)  

## Computing requirements

For students enrolled in this course, you only need a working web browser to access the edupod server, located at: https://edupod.cns.utexas.edu/

**If you are using the edupod server, stop reading here. Everything is pre-installed and no further action is needed.**

To run any of the materials locally on your own machine, you will need the following:

- A recent version of R, [download from here.](https://cloud.r-project.org/)
- A recent version of RStudio, [download from here.](https://rstudio.com/products/rstudio/download/#download)
- The following R packages:    
  broom, cluster, colorspace, cowplot, distill, gapminder, GGally, gganimate,  ggiraph, ggdendro, ggdist, ggforce, ggplot2movies, ggrepel, ggridges, ggthemes, gifski, glue, knitr, learnr, naniar, margins, MASS, Matrix, nycflights13, palmerpenguins, patchwork, rgdal, rmarkdown, sf, tidyverse, transformr, umap, xaringan

You can install all required R packages at once by running the following code in the R command line:

```{r eval = FALSE, echo = TRUE}
install.packages(
  c(
    "broom", "cluster", "colorspace", "cowplot", "distill", "gapminder", 
    "GGally", "gganimate", "ggiraph", "ggdendro", "ggdist", "ggforce",
    "ggplot2movies", "ggrepel", "ggridges", "ggthemes", "gifski", "glue",
    "knitr", "learnr", "naniar", "margins", "MASS", "Matrix",
    "nycflights13", "palmerpenguins", "patchwork", "rgdal", "rmarkdown",
    "sf", "tidyverse", "transformr", "umap", "xaringan"
  )
)
```

## Reuse {.appendix}

Text and figures are licensed under Creative Commons Attribution [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). Any computer code (R, HTML, CSS, etc.) in slides and worksheets, including in slide and worksheet sources, is also licensed under [MIT](https://github.com/wilkelab/SDS375/LICENSE.md). Note that figures in slides may be pulled in from external sources and may be licensed under different terms. For such images, image credits are available in the slide notes, accessible via pressing the letter 'p'.
