---
title: "SDS 375"
description: Data Visualization in R
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Learn more about creating websites with Distill at:
# https://rstudio.github.io/distill/website.html

# Learn more about publishing to GitHub Pages at:
# https://rstudio.github.io/distill/publish_website.html#github-pages

```

This is the home page for SDS 375, Data Visualization in R. All course materials will be posted on this site.

Instructor: [Claus O. Wilke](https://clauswilke.com)  
Meeting times: TTH 9:30am to 11:00am  
Venue: online  


## Computing requirements

For students enrolled in this course, you only need a working web browser to access the RStudio class server.

To run any of the materials locally on your own machine, you will need the following:

- A recent version of R, [download from here.](https://cloud.r-project.org/)
- A recent version of RStudio, [download from here.](https://rstudio.com/products/rstudio/download/#download)
- The following R packages:    
  colorspace, cowplot, distill, gganimate, ggforce, ggdist, ggrepel, ggridges, ggtext, learnr, magick, palmerpenguins, patchwork, plotly, recipes, sf, tidyverse
  
  
You can install all required R packages at once by running the following code in the R command line:

```{r eval = FALSE, echo = TRUE}
install.packages(
  c(
    "colorspace", "cowplot", "distill", "gganimate",
    "ggforce", "ggdist", "ggrepel", "ggridges",
    "ggtext", "learnr", "magick", "palmerpenguins",
    "patchwork", "plotly", "recipes", "sf",
    "tidyverse"
  )
)
```

## Reuse {.appendix}

Text and figures are licensed under Creative Commons Attribution [CC BY 4.0](https://creativecommons.org/licenses/by/4.0/). Note that figures in slides may be pulled in from external sources and may be licensed under different terms. For such images, image credits are available in the slide notes, accessible via pressing the letter 'p'.
