---
title: "Aesthetic mappings"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(lubridate)
library(cowplot)
library(colorspace)
library(here)
# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width

# this function should be added to cowplot
align_legend <- function(p, hjust = 0.5)
{
  # extract legend
  g <- cowplot::plot_to_gtable(p)
  grobs <- g$grobs
  legend_index <- which(sapply(grobs, function(x) x$name) == "guide-box")
  legend <- grobs[[legend_index]]

  # extract guides table
  guides_index <- which(sapply(legend$grobs, function(x) x$name) == "layout")

  # there can be multiple guides within one legend box  
  for (gi in guides_index) {
    guides <- legend$grobs[[gi]]
    
    # add extra column for spacing
    # guides$width[5] is the extra spacing from the end of the legend text
    # to the end of the legend title. If we instead distribute it by `hjust:(1-hjust)` on
    # both sides, we get an aligned legend
    spacing <- guides$width[5]
    guides <- gtable::gtable_add_cols(guides, hjust*spacing, 1)
    guides$widths[6] <- (1-hjust)*spacing
    title_index <- guides$layout$name == "title"
    guides$layout$l[title_index] <- 2

    # reconstruct guides and write back
    legend$grobs[[gi]] <- guides
  }
  
  # reconstruct legend and write back
  g$grobs[[legend_index]] <- legend
  g
}
```

class: center middle

## Plots map data onto graphical elements

---

## Dataset:<br>Daily average temperatures for various locations

.small-font[.center[
```{r echo=FALSE, message=FALSE}
temps_long <- read_csv(here("datasets", "tempnormals.csv")) %>%
  mutate(
    location = factor(
      location, levels = c("Death Valley", "Houston", "San Diego", "Chicago")
    )
  )

temps_long %>%
  select(location, day_of_year, month, temperature) %>%
  head(n = 15L) %>%
  knitr::kable("html")
```
]]

---

## Temperatures mapped onto y position

.center[
```{r temp-normals-vs-time, echo=FALSE, message=FALSE, fig.width = 7, fig.asp = 0.618, dev="svg"}

ggplot(temps_long, aes(x = date, y = temperature, color = location)) +
  geom_line(size = 1) +
  scale_x_date(
    name = "month",
    limits = c(ymd("0000-01-01"), ymd("0001-01-04")),
    breaks = c(
      ymd("0000-01-01"), ymd("0000-04-01"), ymd("0000-07-01"),
      ymd("0000-10-01"), ymd("0001-01-01")
    ),
    labels = c("Jan", "Apr", "Jul", "Oct", "Jan"), expand = c(1/366, 0)
  ) + 
  scale_y_continuous(
    limits = c(19.9, 107),
    breaks = seq(20, 100, by = 20),
    name = "temperature (°F)"
  ) +
  scale_color_manual(
    name = NULL,
    values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7")
  ) +
  theme_minimal_grid(16)
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Temperatures mapped onto color

<br>

.center[
```{r four-locations-temps-by-month, echo = FALSE, message = FALSE, fig.width = 8.5, fig.asp = .3, dev = "svg"}

mean_temps <- temps_long %>%
  group_by(location, month_name) %>%
  summarize(mean = mean(temperature)) %>%
  mutate(
    month_name = factor(
      month_name,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    )
  )

p <- ggplot(mean_temps, aes(x = month_name, y = location, fill = mean)) + 
  geom_tile(width = .95, height = 0.95) + 
  scale_fill_viridis_c(
    option = "B", begin = 0.15, end = 0.98,
    name = "temperature (°F)"
  ) + 
  scale_x_discrete(name = "month") +
  scale_y_discrete(name = NULL) +
  coord_fixed(expand = FALSE) +
  theme_half_open(16) +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    legend.title = element_text(size = 14)
  )

# fix legend (make it centered)
ggdraw(align_legend(p))
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Commonly used aesthetics

.center[
<img src = "https://clauswilke.com/dataviz/aesthetic_mapping_files/figure-html/common-aesthetics-1.png", width = 90%></img>
]

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## The same data values can be mapped to different aesthetics

.center[
<img src = "https://clauswilke.com/dataviz/aesthetic_mapping_files/figure-html/basic-scales-example-1.png", width = 90%></img>
]

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## We can use many different aesthetics at once

<br>

.center.move-up-1em[
```{r mtcars-five-scale, echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = .8, dev = "svg"}
p_mtcars <- ggplot(mtcars, aes(disp, mpg, fill = hp, shape = factor(cyl), size = wt)) + 
  geom_point(color = "white") +
  scale_shape_manual(values = c(23, 24, 21), name = "cylinders") +
  scale_fill_continuous_sequential(
    palette = "Emrld", name = "power (hp)",
    breaks = c(100, 200, 300),
    rev = FALSE
  ) +
  xlab("displacement (cu. in.)") +
  ylab("fuel efficiency (mpg)") +
  guides(
    shape = guide_legend(override.aes = list(size = 4, fill = "#329D84")),
    size = guide_legend(override.aes = list(shape = 21, fill = "#329D84"),
    title = "weight (1000 lbs)")
  ) +
  theme_half_open(14) + background_grid() +
  theme(
    legend.box.background = element_rect(fill = "white", color = "white"),
    legend.position = "top",
    legend.direction = "vertical",
    legend.justification = "center",
    legend.box.margin = margin(7, 7, 7, 7)
  )
legend <- get_legend(align_legend(p_mtcars))
ggdraw() + 
  draw_plot(p_mtcars + theme(legend.position = "none")) + 
  draw_grob(
    legend,
    vjust = 1, hjust = 1, valign = 1, halign = 1,
    x = 1, y = 1, width = .65, height = .3
  )
```
]

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

[//]: # "segment ends here"

---
class: center middle


## Creating aesthetic mappings in ggplot

---

## We define the mapping with `aes()`

```{r echo=FALSE}
temperatures <- temps_long %>%
  select(location, station_id, day_of_year, month, temperature)
```

.small-font[
```{r ggplot-temps-example, eval = FALSE}
ggplot(
  data = temperatures,
  mapping = aes(x = day_of_year, y = temperature, color = location)
) + geom_line()
```
]

.center[
```{r ggplot-temps-example-out, ref.label="ggplot-temps-example", echo=FALSE, fig.width = 5.5, fig.asp = 0.618, dev="svg"}
```
]


---

## We define the mapping with `aes()`

.small-font[
```{r ggplot-temps-example2, eval = FALSE}
ggplot(
  data = temperatures,
  mapping = aes(x = day_of_year, y = location, color = temperature)
) + geom_point(size = 5)
```
]

.center[
```{r ggplot-temps-example2-out, ref.label="ggplot-temps-example2", echo=FALSE, fig.width = 5.5*6/4.2, fig.asp = .3, dev="svg"}
```
]


---

## We frequently omit argument names

Long form, all arguments are named:

.small-font[
```{r ggplot-temps-example3, eval = FALSE}
ggplot(
  data = temperatures,
  mapping = aes(x = day_of_year, y = location, color = temperature)
) + geom_point(size = 5)
```
]

--

Abbreviated form, common arguments remain unnamed:

.small-font[
```{r ggplot-temps-example4, eval = FALSE}
ggplot(temperatures, aes(day_of_year, location, color = temperature)) + 
  geom_point(size = 5)
```
]

---

## The geom determines how the data is shown

.small-font[
```{r ggplot-temps-example5, eval = FALSE}
ggplot(temperatures, aes(day_of_year, temperature, color = location)) + 
  geom_line()
```
]

.center[
```{r ggplot-temps-example5-out, ref.label="ggplot-temps-example5", echo=FALSE, fig.width = 5.5, fig.asp = 0.618, dev="svg"}
```
]


---

## The geom determines how the data is shown

.small-font[
```{r ggplot-temps-example6, eval = FALSE}
ggplot(temperatures, aes(day_of_year, location, color = temperature)) +
  geom_point(size = 5)
```
]

.center[
```{r ggplot-temps-example6-out, ref.label="ggplot-temps-example6", echo=FALSE, fig.width = 5.5*6/4.2, fig.asp = .3, dev="svg"}
```
]

---

## The geom determines how the data is shown

.small-font[
```{r ggplot-temps-example7, eval = FALSE}
ggplot(temperatures, aes(month, temperature, color = location)) + 
  geom_boxplot()
```
]


.center[
```{r ggplot-temps-example7-out, ref.label="ggplot-temps-example7", echo=FALSE, fig.width = 8, fig.asp = 0.5, dev="svg"}
```
]

---

## The geom determines how the data is shown

.small-font[
```{r ggplot-temps-example8, eval = FALSE}
ggplot(temperatures, aes(month, temperature, fill = location)) + 
  geom_violin() + 
  facet_wrap(vars(location)) # make separate panel per location
```
]


.center[
```{r ggplot-temps-example8-out, ref.label="ggplot-temps-example8", echo=FALSE, fig.width = 9, fig.asp = 0.44, dev="svg"}
```
]

[//]: # "segment ends here"

---
class: center middle


## Important: `color` and `fill` apply to different elements

---

## `color` and `fill` apply to different elements

`color`<br>
Applies color to points, lines, text, borders

--

`fill`<br>
Applies color to any filled areas 

---

## Many geoms have both `color` and `fill` aesthetics

.tiny-font[
```{r ggplot-color-fill, eval = FALSE}
ggplot(temperatures, aes(month, temperature, color = location)) + 
  geom_boxplot()
```
]


.center[
```{r ggplot-color-fill-out, ref.label="ggplot-color-fill", echo=FALSE, fig.width = 8, fig.asp = 0.5, dev="svg"}
```
]

---

## Many geoms have both `color` and `fill` aesthetics

.tiny-font[
```{r ggplot-color-fill2, eval = FALSE}
ggplot(temperatures, aes(month, temperature, fill = location)) + 
  geom_boxplot()
```
]


.center[
```{r ggplot-color-fill2-out, ref.label="ggplot-color-fill2", echo=FALSE, fig.width = 8, fig.asp = 0.5, dev="svg"}
```
]

---

## Many geoms have both `color` and `fill` aesthetics

.tiny-font[
```{r ggplot-color-fill3, eval = FALSE}
ggplot(temperatures, aes(month, temperature, color = location, fill = location)) + 
  geom_boxplot()
```
]


.center[
```{r ggplot-color-fill3-out, ref.label="ggplot-color-fill3", echo=FALSE, fig.width = 8, fig.asp = 0.5, dev="svg"}
```
]

---

## Aesthetics can also be used as parameters in geoms

.tiny-font[
```{r ggplot-color-fill-params, eval = FALSE}
ggplot(temperatures, aes(month, temperature, fill = location)) + 
  geom_boxplot(color = "steelblue")
```
]


.center[
```{r ggplot-color-fill-params-out, ref.label="ggplot-color-fill-params", echo=FALSE, fig.width = 8, fig.asp = 0.5, dev="svg"}
```
]


---

## Aesthetics can also be used as parameters in geoms

.tiny-font[
```{r ggplot-color-fill-params2, eval = FALSE}
ggplot(temperatures, aes(month, temperature, color = location)) + 
  geom_boxplot(fill = "steelblue")
```
]


.center[
```{r ggplot-color-fill-params2-out, ref.label="ggplot-color-fill-params2", echo=FALSE, fig.width = 8, fig.asp = 0.5, dev="svg"}
```
]

[//]: # "segment ends here"

---

## Further reading

- Fundamentals of Data Visualization: [Chapter 2: Visualizing data](https://clauswilke.com/dataviz/aesthetic-mapping.html)
- Data Visualization—A Practical Introduction: [Chapter 3: Make a plot](https://socviz.co/makeplot.html#makeplot)
- [**ggplot2** reference documentation](https://ggplot2.tidyverse.org/reference/index.html)
- [**ggplot2** book](https://ggplot2-book.org/)
- [**ggplot2** cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf)
