---
title: "Aesthetic mappings"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(lubridate)
library(cowplot)
library(here)
# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width
```


## Plots map data onto graphical elements

Dataset: Daily average temperatures for various locations

.small-font[.center[
```{r echo=FALSE, message=FALSE}
temps_long <- read_csv(here("datasets", "tempnormals.csv")) %>%
  mutate(
    location = factor(
      location, levels = c("Death Valley", "Houston", "San Diego", "Chicago")
    )
  )

temps_long %>%
  select(location, station_id, day_of_year, temperature) %>%
  head(n = 15L) %>%
  knitr::kable("html")
```
]]

---

## Plots map data onto graphical elements

Temperatures mapped onto y position:

.center[
```{r temp-normals-vs-time, echo=FALSE, message=FALSE, fig.width = 6, fig.asp = 0.618, dev="svg"}

ggplot(temps_long, aes(x = date, y = temperature, color = location)) +
  geom_line(size = 1) +
  scale_x_date(
    name = "month",
    limits = c(ymd("0000-01-01"), ymd("0001-01-04")),
    breaks = c(
      ymd("0000-01-01"), ymd("0000-04-01"), ymd("0000-07-01"),
      ymd("0000-10-01"), ymd("0001-01-01")
    ),
    labels = c("Jan", "Apr", "Jul", "Oct", "Jan"), expand = c(1/366, 0)
  ) + 
  scale_y_continuous(
    limits = c(19.9, 107),
    breaks = seq(20, 100, by = 20),
    name = "temperature (°F)"
  ) +
  scale_color_manual(
    name = NULL,
    values = c("#E69F00", "#56B4E9", "#009E73", "#CC79A7")
  ) +
  theme_minimal_grid() +
  theme(legend.title.align = 0.5)
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Plots map data onto graphical elements

Temperatures mapped onto color:

.center[
```{r four-locations-temps-by-month, echo = FALSE, message = FALSE, fig.width = 5.5*6/4.2, fig.asp = .3, dev = "svg"}

mean_temps <- temps_long %>%
  group_by(location, month_name) %>%
  summarize(mean = mean(temperature)) %>%
  mutate(
    month_name = factor(
      month_name,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    )
  )

p <- ggplot(mean_temps, aes(x = month_name, y = location, fill = mean)) + 
  geom_tile(width = .95, height = 0.95) + 
  scale_fill_viridis_c(
    option = "B", begin = 0.15, end = 0.98,
    name = "temperature (°F)"
  ) + 
  scale_x_discrete(name = "month") +
  scale_y_discrete(name = NULL) +
  coord_fixed(expand = FALSE) +
  theme_half_open() +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    legend.title = element_text(size = 12)
  )

p
# fix legend (make it centered)
#ggdraw(align_legend(p))
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Commonly used aesthetics

.center[
<img src = "https://clauswilke.com/dataviz/aesthetic_mapping_files/figure-html/common-aesthetics-1.png", width = 90%></img>
]

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## The same data values can be mapped to different aesthetics

.center[
<img src = "https://clauswilke.com/dataviz/aesthetic_mapping_files/figure-html/basic-scales-example-1.png", width = 90%></img>
]

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## In ggplot, we define the mapping with `aes()`

```{r echo=FALSE}
temperatures <- temps_long %>%
  select(location, station_id, day_of_year, month, temperature)
```

.small-font[
```{r ggplot-temps-example, eval = FALSE}
ggplot(
  data = temperatures,
  mapping = aes(x = day_of_year, y = temperature, color = location)
) + geom_line()
```
]

.center[
```{r ggplot-temps-example-out, ref.label="ggplot-temps-example", echo=FALSE, fig.width = 5.5, fig.asp = 0.618, dev="svg"}
```
]


---

## In ggplot, we define the mapping with `aes()`

.small-font[
```{r ggplot-temps-example2, eval = FALSE}
ggplot(
  data = temperatures,
  mapping = aes(x = day_of_year, y = location, color = temperature)
) + geom_point(size = 5)
```
]

.center[
```{r ggplot-temps-example2-out, ref.label="ggplot-temps-example2", echo=FALSE, fig.width = 5.5*6/4.2, fig.asp = .3, dev="svg"}
```
]


---

## We frequently omit argument names

Long form, all arguments are named:

.small-font[
```{r ggplot-temps-example3, eval = FALSE}
ggplot(
  data = temperatures,
  mapping = aes(x = day_of_year, y = location, color = temperature)
) + geom_point(size = 5)
```
]

--

Abbreviated form, common arguments remain unnamed:

.small-font[
```{r ggplot-temps-example4, eval = FALSE}
ggplot(temperatures, aes(day_of_year, location, color = temperature)) + 
  geom_point(size = 5)
```
]

---

## The geom determines how the data is shown

.small-font[
```{r ggplot-temps-example5, eval = FALSE}
ggplot(temperatures, aes(month, temperature, color = location)) + 
  geom_boxplot()
```
]


.center[
```{r ggplot-temps-example5-out, ref.label="ggplot-temps-example5", echo=FALSE, fig.width = 8, fig.asp = 0.5, dev="svg"}
```
]
