---
title: "Animations"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
params:
  cache: TRUE
  nframes: 200
  fps: 20
  res: 150
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(here)
library(cowplot)
library(glue)
library(ggrepel)
library(gganimate)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width

# gdp dataset
gdp_ranked <- read_csv(here("datasets", "gdp_ranked.csv")) %>%
  mutate(rank = fct_rev(factor(rank)))
```

## Animations can create more engaging displays

.center.move-up-1em[
```{r gdp-anim-bar-race, cache = params$cache, echo = FALSE, warning = FALSE, out.width = "50%"}
p <- gdp_ranked %>%
  mutate(
    label = scales::comma(gdp, 1),
    highlight = ifelse(country == "China", "China", "other")
  ) %>%
  ggplot(
    aes(
      gdp, rank, group = country,
      color = highlight, fill = highlight
    )
  ) +
  geom_col(alpha = 0.8) +
  geom_text(
    aes(x = -200, label = country),
    hjust = 1, size = 12/.pt
  ) +
  geom_text(
    aes(x = gdp + 200, label = label),
    hjust = 0, size = 12/.pt
  ) +
  scale_color_manual(
    aesthetics = c("color", "fill"),
    values = c(
      other = "gray35", China = "#AB4A39"
    ),
    guide = "none"
  ) +
  scale_x_continuous(
    name = "GDP (billions USD)",
    limits = 1000*c(-7, 23),
    breaks = 1000*c(0, 5, 10, 15, 20),
    labels = scales::comma
  ) +
  labs(
    title = "China's rise in GDP",
    subtitle = "Year: {closest_state}"
  ) +
  theme_minimal_vgrid(12, rel_small = 1) +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    axis.title.x = element_text(hjust = 1)
  ) +
  #facet_wrap(vars(year)) + NULL
  transition_states(year, transition_length = 5)

animate(
  p,
  width = 6*params$res,
  height = 6*params$res,
  res = params$res,
  nframes = params$nframes,
  fps = params$fps
)
```
]

.absolute-bottom-right.tiny-font[
Data source: World Bank
]

---

## Animations can create more engaging displays

.center[
```{r gdp-anim-lines-race, cache = params$cache, echo = FALSE, warning = FALSE, out.width = "66%"}
p <- gdp_ranked %>%
  filter(
    #TRUE
    country %in% c("China", "Japan", "United States", "Germany", "Brazil")
  ) %>%
  mutate(
    label = glue::glue(" {country}")
  ) %>%
  ggplot(aes(year, gdp, color = country)) +
  geom_line() +
  geom_point(size = 3) +
  geom_text_repel(
    aes(label = label),
    hjust = 0,
    nudge_x = 1,
    direction = "y",
    xlim = c(NA, Inf),
    size = 14/.pt
  ) +
  scale_y_log10(name = "GDP (billion USD)", labels = scales::comma) +
  scale_color_manual(
    values = c("#009E73", "#404040", "#0072B2", "#D55E00", "#CC79A7"),
    guide = "none"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(14, rel_small = 1) +
  theme(plot.margin = margin(7, 100, 7, 7)) +
  transition_reveal(year)

animate(
  p,
  width = 8*params$res,
  height = 0.618*8*params$res,
  res = params$res,
  nframes = params$nframes,
  fps = params$fps
)
```
]

.absolute-bottom-right.tiny-font[
Data source: World Bank
]

---

## We make animations in R with **gganimate**

.width-70[
```r
library(tidyverse)
library(gganimate)

# make fun animations
...
```
]

The **gganimate** package adds powerful animation tools to **ggplot2**

<img src = "https://gganimate.com/reference/figures/logo.png", width = 15%, style = "position:absolute; top: 19%; right: 13%;" />

---
class: center middle

## How should we think about making an animation?

---

## Think of an animation as faceting by time

.center.move-up-1em[
```{r gdp-anim-bar-facets, cache = params$cache, echo = FALSE, warning = FALSE, fig.width = 9, fig.height = 5.33, dev = "svg"}
gdp_ranked %>%
  filter(year > 1985 & year %% 5 == 0) %>%
  mutate(
    label = scales::comma(gdp, 1),
    highlight = ifelse(country == "China", "China", "other")
  ) %>%
  ggplot(
    aes(
      gdp, rank, group = country,
      color = highlight, fill = highlight
    )
  ) +
  geom_col(alpha = 0.8) +
  geom_text(
    aes(x = -200, label = country),
    hjust = 1, size = 8/.pt
  ) +
  geom_text(
    aes(x = gdp + 200, label = label),
    hjust = 0, size = 8/.pt
  ) +
  scale_color_manual(
    aesthetics = c("color", "fill"),
    values = c(
      other = "gray35", China = "#AB4A39"
    ),
    guide = "none"
  ) +
  scale_x_continuous(
    name = "GDP (billions USD)",
    limits = 1000*c(-10, 23),
    breaks = 1000*c(0, 5, 10, 15, 20),
    labels = scales::comma
  ) +
  facet_wrap(vars(year), scales = "free_x") +
  theme_minimal_vgrid(8, rel_small = 1) +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    axis.title.x = element_text(hjust = 1),
    panel.grid.major.x = element_line(size = 0.4)
  )
```
]

---

## We know how to make a faceted plot

.tiny-font[
```{r gdp-facet-demo, eval = FALSE}
gdp_ranked %>%
  filter(year > 1985 & year %% 5 == 0) %>%
  ggplot(aes(gdp, rank)) +
  geom_col(aes(fill = country)) +
  facet_wrap(vars(year))
```
]

.center[
```{r gdp-facet-demo-out, ref.label = "gdp-facet-demo", cache = params$cache, echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.66, dev = "svg"}

```
]

---

## Making an animation is about as complicated

.tiny-font[
```{r eval = FALSE}
gdp_ranked %>%
  # gganimate uses the `group` aesthetic to track objects across frames
  ggplot(aes(gdp, rank, group = country)) + 
  geom_col(aes(fill = country)) +
  transition_states(year, transition_length = 5)
```
]

.center[
```{r gdp-bar-race-demo, cache = params$cache, echo = FALSE, warning = FALSE, out.width = "40%"}
p <- gdp_ranked %>%
  ggplot(aes(gdp, rank, group = country)) +
  geom_col(aes(fill = country)) +
  transition_states(year, transition_length = 5)

animate(
  p,
  width = 6*params$res,
  height = 5*params$res,
  res = params$res,
  nframes = params$nframes,
  fps = params$fps
)
```
]

---


## Adding country names and plot title

.tiny-font.pull-left.width-50[
```{r gdp-anim-fancy, eval = FALSE}
gdp_ranked %>%
  ggplot(aes(gdp, rank, group = country)) +
  geom_col(aes(fill = country)) +
  geom_text(
    aes(x = -200, label = country),
    hjust = 1, size = 14/.pt
  ) +
  xlim(-7000, 23000) +
  labs(title = "year: {closest_state}") +
  theme_minimal_vgrid(14, rel_small = 1) +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank()
  ) + 
  guides(fill = "none") +
  transition_states(year, transition_length = 5)
```
]

.pull-right.width-40[
```{r gdp-anim-fancy-out, cache = params$cache, echo = FALSE, warning = FALSE, out.width = "100%"}
p <- gdp_ranked %>%
  ggplot(aes(gdp, rank, group = country)) +
  geom_col(aes(fill = country)) +
  geom_text(
    aes(x = -200, label = country),
    hjust = 1, size = 14/.pt
  ) +
  xlim(-7000, 23000) +
  labs(title = "year: {closest_state}") +
  theme_minimal_vgrid(14, rel_small = 1) +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank()
  ) + 
  guides(fill = "none") +
  transition_states(year, transition_length = 5)

animate(
  p,
  width = 6*params$res,
  height = 6*params$res,
  res = params$res,
  nframes = params$nframes,
  fps = params$fps
)
```
]

---

## We make time series with `transition_reveal()`

.tiny-font.pull-left.width-50[
```{r gdp-line-race-demo, eval = FALSE}
selected <- c("China", "Japan",
  "United States", "Germany", "Brazil")

gdp_ranked %>%
  filter(country %in% selected) %>%
  ggplot(aes(year, gdp, color = country)) +
  geom_line() +
  geom_point(size = 3) +
  scale_y_log10() +
  transition_reveal(year)
```
]

.pull-right.width-45[
```{r gdp-line-race-demo-out, cache = params$cache, echo = FALSE, warning = FALSE, out.width = "100%"}
selected <- c("China", "Japan",
  "United States", "Germany", "Brazil")

p <- gdp_ranked %>%
  filter(country %in% selected) %>%
  ggplot(aes(year, gdp, color = country)) +
  geom_line() +
  geom_point(size = 3) +
  scale_y_log10() +
  transition_reveal(year)

animate(
  p,
  width = 6*params$res,
  height = 0.618*6*params$res,
  res = params$res,
  nframes = params$nframes,
  fps = params$fps
)
```
]

---

## This works also with **ggrepel** for labeling

.tiny-font.pull-left.width-50[
```{r gdp-line-race-labeled, eval = FALSE}
gdp_ranked %>%
  filter(country %in% selected) %>%
  ggplot(aes(year, gdp, color = country)) +
  geom_line() +
  geom_point(size = 3) +
  geom_text_repel(
    aes(label = country),
    hjust = 0,
    nudge_x = 2,
    direction = "y",
    xlim = c(NA, Inf)
  ) +
  scale_y_log10() +
  guides(color = "none") +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(7, 100, 7, 7)) +
  transition_reveal(year)
```
]

.pull-right.width-45[
```{r gdp-line-race-labeled-out, cache = params$cache, echo = FALSE, warning = FALSE, out.width = "100%"}
p <- gdp_ranked %>%
  filter(country %in% selected) %>%
  ggplot(aes(year, gdp, color = country)) +
  geom_line() +
  geom_point(size = 3) +
  geom_text_repel(
    aes(label = country),
    hjust = 0,
    nudge_x = 2,
    direction = "y",
    xlim = c(NA, Inf)
  ) +
  scale_y_log10() +
  guides(color = "none") +
  coord_cartesian(clip = "off") +
  theme(plot.margin = margin(7, 100, 7, 7)) +
  transition_reveal(year)

animate(
  p,
  width = 6*params$res,
  height = 0.618*6*params$res,
  res = params$res,
  nframes = params$nframes,
  fps = params$fps
)
```
]

---

## Further reading

- **gganimate** documentation: [Getting started](https://gganimate.com/articles/gganimate.html)
- **gganimate** reference documentation: [`transition_states()`](https://gganimate.com/reference/transition_states.html)
- **gganimate** reference documentation: [`transition_reveal()`](https://gganimate.com/reference/transition_reveal.html)


