---
title: "Clustering"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
params:
  cache: TRUE
---

```{r setup, include=FALSE, echo=FALSE, message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(broom)
library(cowplot)
library(ggforce)
library(here)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2.0)) # 2 for full width, 2.5 for half width

# spirals
spirals <- read_csv(here("datasets", "spirals.csv"))

# gaussian clusters
gaussian_clusters <- read_csv(here("datasets", "gaussian_clusters.csv"))

```

```{r kmeans-setup, echo = FALSE}
update_clusters <- function(coords, centroids) {
  dmin <- sqrt((centroids$x[1]-coords$x)^2 + (centroids$y[1]-coords$y)^2)
  cluster <- rep(1, nrow(coords))
  
  for (i in 2:nrow(centroids)) {
    d <- sqrt((centroids$x[i]-coords$x)^2 + (centroids$y[i]-coords$y)^2)
    idx <- d < dmin
    dmin[idx] <- d[idx]
    cluster[idx] <- i
  }
  
  coords$cluster <- factor(cluster)
  coords
}

update_centroids <- function(coords) {
  coords %>%
    group_by(cluster) %>%
    summarize(x = mean(x), y = mean(y)) %>%
    arrange(cluster)
}

random_centroids <- function(coords, n) {
  x <- runif(n, min(coords$x), max(coords$x))
  y <- runif(n, min(coords$y), max(coords$y))
  tibble(x, y, cluster = factor(1:n))
}

make_kmeans_plot <- function(coords, centroids, voronoi_centr, color_points = TRUE, plot_voronoi = TRUE, plot_centroids = TRUE) {
  xlim <- c(min(coords$x), max(coords$x))
  xrange <- diff(range(xlim))
  xlim <- xlim + c(-.1*xrange, .1*xrange)
  ylim <- c(min(coords$y), max(coords$y))
  yrange <- diff(range(ylim))
  ylim <- ylim + c(-.1*yrange, .1*yrange)
  
  if (isTRUE(color_points)) {
    p <- ggplot(coords, aes(x, y, color = cluster))
  } else {
    p <- ggplot(coords, aes(x, y))
  }
  
  if (isTRUE(plot_voronoi)) {
    p <- p +
      geom_voronoi_tile(
        data = voronoi_centr,
        aes(fill = cluster, group = 1L),
        alpha = 0.2,
        color = NA,
        bound = c(xlim, ylim)
      )
  }
  
  if (isTRUE(color_points)) {
    p <- p + geom_point()
  } else {
    p <- p + geom_point(color = "black")
  }
  
  if (isTRUE(plot_centroids)) {
    p <- p + 
      geom_point(
        data = centroids,
        aes(fill = cluster),
        color = "black", size = 6, shape = 21
      )
  }
  
  p +
    coord_cartesian(xlim = xlim, ylim = ylim, expand = FALSE) +
    theme_void() +
    theme(legend.position = "none")
}

```

.center[
```{r three-clusters-viz, echo = FALSE, cache = params$cache, fig.width = 6, fig.asp = .85, dev = "svg"}
centroids <- tibble(
  cluster = factor(1:3),
  x = c(2.014419687, 4.321905953, -0.003265806),
  y = c(-4.1133121, 1.6406162, -0.1237819)
)
coords <- update_clusters(gaussian_clusters, centroids)

make_kmeans_plot(coords, centroids,  centroids, color_points = FALSE, plot_centroids = FALSE, plot_voronoi = FALSE)
```
]

These points correspond to three clusters. 
Can a computer find them automatically?

---


.center[
```{r three-clusters-viz2, echo = FALSE, cache = params$cache, fig.width = 6, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE, plot_centroids = FALSE, plot_voronoi = FALSE)
```
]

These points correspond to three clusters. 
Can a computer find them automatically?


---

## *k*-means clustering

--

1\. Start with *k* randomly chosen means

--

2\. Color data points by the shortest distance to any
mean

--

3\. Move means to centroid position of each group of
points

--

4\. Repeat from step 2 until convergence

---

class: center middle

## Let's try it out


---

class: middle

.center[
```{r three-clusters-kmeans1, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
coords <- gaussian_clusters

centroids <- tibble(
  x = c(5.7353523, -0.4681193, -0.5732993),
  y = c(-5.8405302, 0.3293147, -1.0971190),
  cluster = factor(1:3)
)

coords <- update_clusters(coords, centroids)

make_kmeans_plot(coords, centroids, color_points = FALSE, plot_centroids = FALSE, plot_voronoi = FALSE)
```
]

---

class: middle

.center[
```{r three-clusters-kmeans2, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, color_points = FALSE, plot_centroids = TRUE, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Add means at arbitrary locations
]

---

class: middle

.center[
```{r three-clusters-kmeans3, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE, plot_voronoi = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r three-clusters-kmeans4, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r three-clusters-kmeans5, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---

class: middle

.center[
```{r three-clusters-kmeans6, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r three-clusters-kmeans7, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r three-clusters-kmeans8, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---

class: middle

.center[
```{r three-clusters-kmeans9, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r three-clusters-kmeans10, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r three-clusters-kmeans11, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---

class: middle

.center[
```{r three-clusters-kmeans12, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r three-clusters-kmeans13, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r three-clusters-kmeans14, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---

class: middle

.center[
```{r three-clusters-kmeans15, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r three-clusters-kmeans16, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r three-clusters-kmeans17, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---

class: middle

.center[
```{r three-clusters-kmeans18, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r three-clusters-kmeans19, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r three-clusters-kmeans20, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Final result
]

---
class: middle center

## Now we'll cluster the same dataset with five centroids

---
class: middle

.center[
```{r five-clusters-kmeans1, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
# three centers, 5 clusters
coords <- gaussian_clusters

centroids <- tibble(
  x = c(5.3544626, 3.1688159, 5.4307893, -0.9571097, 4.4852765),
  y = c(1.754802, 2.622277, -4.537906, -4.415216, 3.125914),
  cluster = factor(1:5)
)

coords <- update_clusters(coords, centroids)

make_kmeans_plot(coords, centroids, color_points = FALSE, plot_centroids = FALSE, plot_voronoi = FALSE)
```
]


---

class: middle

.center[
```{r five-clusters-kmeans2, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, color_points = FALSE, plot_centroids = TRUE, plot_voronoi = FALSE)
```
]


.absolute-bottom-left[
Add means at arbitrary locations
]


---

class: middle

.center[
```{r five-clusters-kmeans3, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE, plot_voronoi = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]


---

class: middle

.center[
```{r five-clusters-kmeans4, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r five-clusters-kmeans5, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]


---

class: middle

.center[
```{r five-clusters-kmeans6, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r five-clusters-kmeans7, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r five-clusters-kmeans8, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---

class: middle

.center[
```{r five-clusters-kmeans9, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r five-clusters-kmeans10, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r five-clusters-kmeans11, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---

class: middle

.center[
```{r five-clusters-kmeans12, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r five-clusters-kmeans13, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r five-clusters-kmeans14, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---

class: middle

.center[
```{r five-clusters-kmeans15, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r five-clusters-kmeans16, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r five-clusters-kmeans17, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---
class: center middle

## ... do many more iterations ...

---

class: middle

.center[
```{r five-clusters-kmeans18, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2

for (i in 1:40) {
  coords <- update_clusters(coords, centroids)
  centroids <- update_centroids(coords)
}
make_kmeans_plot(coords, centroids, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Final result
]

---
class: middle center

## Let's try this on the spirals dataset

---
class: middle

.center[
```{r spirals-kmeans1, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
coords <- spirals

centroids <- tibble(
  x = c(5.3544626, 3.1688159, 5.4307893, -0.9571097, 4.4852765),
  y = c(1.754802, 2.622277, -4.537906, -4.415216, 3.125914),
  cluster = factor(1:5)
)

set.seed(3425)
set.seed(3427)
centroids <- random_centroids(coords, 3)

coords <- update_clusters(coords, centroids)

make_kmeans_plot(coords, centroids, color_points = FALSE, plot_centroids = FALSE, plot_voronoi = FALSE)
```
]


---

class: middle

.center[
```{r spirals-kmeans2, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, color_points = FALSE, plot_centroids = TRUE, plot_voronoi = FALSE)
```
]


.absolute-bottom-left[
Add means at arbitrary locations
]


---

class: middle

.center[
```{r spirals-kmeans3, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE, plot_voronoi = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]


---

class: middle

.center[
```{r spirals-kmeans4, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r spirals-kmeans5, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]


---

class: middle

.center[
```{r spirals-kmeans6, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r spirals-kmeans7, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r spirals-kmeans8, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---

class: middle

.center[
```{r spirals-kmeans9, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r spirals-kmeans10, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r spirals-kmeans11, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---

class: middle

.center[
```{r spirals-kmeans12, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r spirals-kmeans13, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r spirals-kmeans14, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---

class: middle

.center[
```{r spirals-kmeans15, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2
coords <- update_clusters(coords, centroids)
make_kmeans_plot(coords, centroids, centroids, color_points = FALSE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r spirals-kmeans16, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
make_kmeans_plot(coords, centroids, centroids, color_points = TRUE)
```
]

.absolute-bottom-left[
Color data points by the shortest distance to any mean
]

---

class: middle

.center[
```{r spirals-kmeans17, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids2 <- update_centroids(coords)
make_kmeans_plot(coords, centroids2, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Move means to centroid position of each group of points
]

---
class: center middle

## ... do many more iterations ...

---

class: middle

.center[
```{r spirals-kmeans18, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = .85, dev = "svg"}
centroids <- centroids2

for (i in 1:100) {
  coords <- update_clusters(coords, centroids)
  centroids <- update_centroids(coords)
}
make_kmeans_plot(coords, centroids, centroids, plot_voronoi = FALSE)
```
]

.absolute-bottom-left[
Final result
]

---
class: center middle

## k-means clustering works best when<br>data forms distinct, compact clusters

---

## Other clustering algorithms

.center[
<img src = "clustering_files/clustering_examples.png", width = 75%></img>
]

.absolute-bottom-right.tiny-font[
From George Seif (2018) [The 5 Clustering Algorithms Data Scientists Need to Know](https://towardsdatascience.com/the-5-clustering-algorithms-data-scientists-need-to-know-a36d136ef68)
]

[//]: # "segment ends here"

---
class: center middle


## Doing k-means clustering in R


---

## Example dataset: `iris`

Measurements on the sepals and petals of three iris species

.small-font[
```{r echo = TRUE}
iris
```
]

---

## Example dataset: `iris`

.small-font[
```{r iris-plot, echo = TRUE, eval = FALSE}
ggplot(iris, aes(Petal.Length, Petal.Width, color = Species)) +
  geom_point()
```
]

.center[
```{r iris-plot-out, ref.label="iris-plot", fig.width=6.5, fig.asp = 0.618, echo=FALSE, dev = "svg"}
```
]


---

## We perform k-means clustering with `kmeans()`

.tiny-font[
```{r echo = TRUE}
km_fit <- iris %>% 
  select(where(is.numeric)) %>%
  kmeans(
    centers = 3,  # number of cluster centers
    nstart = 10   # number of independent restarts of the algorithm
  )

km_fit
```
]


---

.tiny-font[
```{r echo = TRUE}
km_fit
```
]


---

## We perform k-means clustering with `kmeans()`

.pull-left.xtiny-font[
```{r iris-kmeans, echo = TRUE, eval = FALSE}
# run kmeans clustering
km_fit <- iris %>% 
  select(where(is.numeric)) %>%
  kmeans(centers = 3, nstart = 10)

# plot
km_fit %>%
  # combine with original data
  augment(iris) %>%
  ggplot() +
  aes(x = Petal.Length, Petal.Width) +
  geom_point(
    aes(color = .cluster, shape = Species)
  ) +
  geom_point(
    data = tidy(km_fit),
    aes(fill = cluster),
    shape = 21, color = "black", size = 4
  ) +
  guides(color = "none")
```
]

.pull-right.width-50[
```{r iris-kmeans-out, ref.label="iris-kmeans", fig.width=6.5, fig.asp = 0.618, echo=FALSE, dev = "svg"}
```
]

---

## We perform k-means clustering with `kmeans()`

.pull-left.xtiny-font[
```{r echo = TRUE, eval = FALSE}
# run kmeans clustering
km_fit <- iris %>% 
  select(where(is.numeric)) %>%
  kmeans(centers = 3, nstart = 10)

# plot
km_fit %>%
  # combine with original data
  augment(iris) %>%
  ggplot() +
  aes(x = Petal.Length, Petal.Width) +
  geom_point(
    aes(color = .cluster, shape = Species)
  ) +
  geom_point(
    data = tidy(km_fit),
    aes(fill = cluster),
    shape = 21, color = "black", size = 4
  ) +
  guides(color = "none")
```
]

.pull-right.width-50[
![](`r knitr::fig_chunk("iris-kmeans-out", "svg")`)

.small-font[
How do we choose the number of clusters?
]]


---

## We perform k-means clustering with `kmeans()`

.pull-left.xtiny-font[
```{r iris-kmeans2, echo = TRUE, eval = FALSE}
# run kmeans clustering
km_fit <- iris %>% 
  select(where(is.numeric)) %>%
  kmeans(centers = 2, nstart = 10)

# plot
km_fit %>%
  # combine with original data
  augment(iris) %>%
  ggplot() +
  aes(x = Petal.Length, Petal.Width) +
  geom_point(
    aes(color = .cluster, shape = Species)
  ) +
  geom_point(
    data = tidy(km_fit),
    aes(fill = cluster),
    shape = 21, color = "black", size = 4
  ) +
  guides(color = "none")
```
]

.pull-right.width-50[
```{r iris-kmeans2-out, ref.label="iris-kmeans2", fig.width=6.5, fig.asp = 0.618, echo=FALSE, dev = "svg"}
```

.small-font[
How do we choose the number of clusters?
]]

---

## We perform k-means clustering with `kmeans()`

.pull-left.xtiny-font[
```{r iris-kmeans3, echo = TRUE, eval = FALSE}
# run kmeans clustering
km_fit <- iris %>% 
  select(where(is.numeric)) %>%
  kmeans(centers = 5, nstart = 10)

# plot
km_fit %>%
  # combine with original data
  augment(iris) %>%
  ggplot() +
  aes(x = Petal.Length, Petal.Width) +
  geom_point(
    aes(color = .cluster, shape = Species)
  ) +
  geom_point(
    data = tidy(km_fit),
    aes(fill = cluster),
    shape = 21, color = "black", size = 4
  ) +
  guides(color = "none")
```
]

.pull-right.width-50[
```{r iris-kmeans3-out, ref.label="iris-kmeans3", fig.width=6.5, fig.asp = 0.618, echo=FALSE, dev = "svg"}
```

.small-font[
How do we choose the number of clusters?
]]

---

## Look for elbow in scree plot

.pull-left.xtiny-font[
```{r iris-scree, echo = TRUE, eval = FALSE}
# function to calculate within sum squares
calc_withinss <- function(data, centers) {
  km_fit <- select(data, where(is.numeric)) %>%
    kmeans(centers = centers, nstart = 10)
  km_fit$tot.withinss
}

tibble(centers = 1:15) %>%
  mutate(
    within_sum_squares = map_dbl(
      centers, ~calc_withinss(iris, .x)
    )
  ) %>%
  ggplot() +
  aes(centers, within_sum_squares) +
  geom_point() +
  geom_line()
```
]

.pull-right[
```{r iris-scree-out, ref.label="iris-scree", fig.width=6.5, fig.asp = 0.618, echo=FALSE, dev = "svg"}
```

.small-font[
Plot suggests number of clusters of about 3
]]

[//]: # "segment ends here"

---

## Further reading

- Wikipedia: [k-means clustering](https://en.wikipedia.org/wiki/K-means_clustering)
- Naftali Harris blog post: [Interactive k-means demonstration](https://www.naftaliharris.com/blog/visualizing-k-means-clustering/)
- Stackoverflow post: [Determining the appropriate number of clusters in k-means](https://stackoverflow.com/questions/15376075/cluster-analysis-in-r-determine-the-optimal-number-of-clusters/15376462#15376462)
- Medium article: [The 5 Clustering Algorithms Data Scientists Need to Know](https://towardsdatascience.com/the-5-clustering-algorithms-data-scientists-need-to-know-a36d136ef68)
