---
title: "Color scales"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(cowplot)
library(ggthemes)
library(colorspace)
library(ggrepel)
library(here)

## ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(14)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width
```


# Uses of color in data visualization

--

<table style = "border: none; line-height: 2.5;">
<tr style = "background: white;">
<td style = "text-align: left; width: 50%;">
1. Distinguish categories (qualitative)
</td>
<td>
<img src = "color-scales_files/qualitative.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>
</table>

---

# Qualitative scale example

.center.nogap[
```{r popgrowth-vs-popsize-colored, echo = FALSE, message = FALSE, fig.width = 7.5, fig.asp = 0.618, dev = "svg"}
US_census <- read_csv(here("datasets", "US_census.csv"))
US_regions <- read_csv(here("datasets", "US_regions.csv"))

popgrowth <- left_join(US_census, US_regions) %>%
    group_by(region, division, state) %>%
    summarize(
      pop2000 = sum(pop2000, na.rm = TRUE),
      pop2010 = sum(pop2010, na.rm = TRUE),
      popgrowth = (pop2010-pop2000)/pop2000
    ) %>%
    arrange(popgrowth) %>%
    ungroup() %>%
    mutate(
      #state = factor(state, levels = state),
      region = factor(region, levels = c("West", "South", "Midwest", "Northeast"))
    )

region_colors <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442")

labeled_states <- c(
  "Alaska", "Arizona", "California", "Florida", "Wisconsin", 
  "Louisiana", "Nevada", "Michigan", "Montana", "New Mexico",
  "Pennsylvania", "New York", "Oregon", "Rhode Island",
  "Tennessee", "Texas", "Utah", "Vermont"
)

df_repel <- select(popgrowth, x = pop2000, y = popgrowth, state) %>%
  mutate(label = ifelse(state %in% labeled_states, as.character(state), ""))

p <- ggplot(popgrowth, aes(x = pop2000, y = popgrowth, color = region, fill = region)) +
  geom_text_repel(
    data = df_repel,
    aes(x, y, label = label),
    segment.alpha = 0.5, point.padding = 0.25,
    box.padding = .8,
    force = 1,
    min.segment.length = 0.1,
    max.overlaps = 1000,
    size = 10/.pt,
    seed = 7586,
    inherit.aes = FALSE
  ) +
  geom_point(size = 4, color = "white") +
  geom_point(size = 3, shape = 21) +
  scale_x_log10(
    breaks = c(1e6, 3e6, 1e7, 3e7),
    labels = expression(10^6, 3 %*% 10^6, 10^7, 3 %*% 10^7)
  ) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  xlab("population size in 2000") +
  ylab("population growth, 2000 to 2010") +
  theme_minimal_grid(14)

p +  
  scale_fill_manual(values = region_colors) +
  scale_color_manual(values = darken(region_colors, .3))

```
]

Palette name: Okabe-Ito

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Qualitative scale example

.center.nogap[
```{r popgrowth-vs-popsize-colored2, echo = FALSE, message = FALSE, fig.width = 7.5, fig.asp = 0.618, dev = "svg"}

region_colors <- RColorBrewer::brewer.pal(4, "Set1")

p +  
  scale_fill_manual(values = region_colors) +
  scale_color_manual(values = darken(region_colors, .3))

```
]

Palette name: ColorBrewer Set1

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Qualitative scale example

.center.nogap[
```{r popgrowth-vs-popsize-colored3, echo = FALSE, message = FALSE, fig.width = 7.5, fig.asp = 0.618, dev = "svg"}

region_colors <- RColorBrewer::brewer.pal(4, "Set3")

p +  
  scale_fill_manual(values = region_colors) +
  scale_color_manual(values = darken(region_colors, .3))

```
]

Palette name: ColorBrewer Set3

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Uses of color in data visualization

<table style = "border: none; line-height: 2.5;">
<tr style = "background: white;">
<td style = "text-align: left; width: 50%;">
1. Distinguish categories (qualitative)
</td>
<td>
<img src = "color-scales_files/qualitative.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
2. Represent numeric values (sequential)
</td>
<td>
<img src = "color-scales_files/sequential.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>
</table>


---

# Sequential scale example

<br>

.center[
```{r four-locations-temps-by-month, echo = FALSE, message = FALSE, fig.width = 9, fig.asp = .3, dev = "svg"}
temps_months <- read_csv(here("datasets", "tempnormals.csv")) %>%
  group_by(location, month_name) %>%
  summarize(mean = mean(temperature)) %>%
  mutate(
    month = factor(
      month_name,
      levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
    ),
    location = factor(
      location, levels = c("Death Valley", "Houston", "San Diego", "Chicago")
    )
  ) %>%
  select(-month_name)

p <- ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile(width = .95, height = 0.95) + 
  scale_y_discrete(name = NULL) +
  coord_fixed(expand = FALSE) +
  theme_half_open() +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    legend.title = element_text(size = 12)
  )

p + scale_fill_viridis_c(
    option = "D",
    name = "temperature (°F)"
  )
```
]

Palette name: Viridis

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Sequential scale example

<br>

.center[
```{r four-locations-temps-by-month2, echo = FALSE, message = FALSE, fig.width = 9, fig.asp = .3, dev = "svg"}
p + scale_fill_viridis_c(
    option = "B", begin = 0.15, end = 0.98,
    name = "temperature (°F)"
  )
```
]

Palette name: Inferno

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

# Sequential scale example

<br>

.center[
```{r four-locations-temps-by-month3, echo = FALSE, message = FALSE, fig.width = 9, fig.asp = .3, dev = "svg"}
p + scale_fill_viridis_c(
    option = "E",
    name = "temperature (°F)"
  )
```
]

Palette name: Cividis

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

# Uses of color in data visualization

<table style = "border: none; line-height: 2.5;">
<tr style = "background: white;">
<td style = "text-align: left; width: 50%;">
1. Distinguish categories (qualitative)
</td>
<td>
<img src = "color-scales_files/qualitative.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
2. Represent numeric values (sequential)
</td>
<td>
<img src = "color-scales_files/sequential.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
3. Represent numeric values (diverging)
</td>
<td>
<img src = "color-scales_files/diverging.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>
</table>

---

# Diverging scale example

.center.nogap[
```{r forensic-correlations1, echo = FALSE, message = FALSE, fig.width = 4., fig.asp = 1, dev = "svg"}
forensic_glass <- read_csv(here("datasets", "forensic_glass.csv"))

cm <- cor(select(forensic_glass, -type, -RI, -Si))
df_wide <- as.data.frame(cm)
df_long <- stack(df_wide)
names(df_long) <- c("cor", "var1")
df_long <- cbind(df_long, var2 = rep(rownames(cm), length(rownames(cm))))
clust <- hclust(as.dist(1-cm), method="average") 
levels <- clust$labels[clust$order]
df_long$var1 <- factor(df_long$var1, levels = levels)
df_long$var2 <- factor(df_long$var2, levels = levels)

p <- ggplot(filter(df_long, as.integer(var1) < as.integer(var2)),
       aes(var1, var2, fill=cor)) + 
  geom_tile(color = "white", size = 1) + 
  scale_x_discrete(position = "top", name = NULL, expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = c(0, 0)) +
  guides(
    fill = guide_colorbar(
      direction = "horizontal",
      label.position = "bottom",
      title.position = "top",
      barwidth = grid::unit(140, "pt"),
      barheight = grid::unit(17.5, "pt"),
      ticks.linewidth = 1
    )
  ) +
  coord_fixed() +
  theme_half_open(rel_small = 1) +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.ticks.length = grid::unit(3, "pt"),
    legend.position = c(.97, .0),
    legend.justification = c(1, 0),
    legend.title.align = 0.5
  )
p + scale_fill_distiller(
  name = "correlation",
  limits = c(-.5, .5),
  breaks = c(-.5, 0, .5),
  labels = c("–0.5", "0.0", "0.5"),
  type = "div",
  palette = "PiYG",
  direction = 1
)
```
]

Palette name: ColorBrewer PiYG

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Diverging scale example

.center.nogap[
```{r forensic-correlations2, echo = FALSE, message = FALSE, fig.width = 4., fig.asp = 1, dev = "svg"}
p + scale_fill_continuous_divergingx(
  name = "correlation",
  limits = c(-.5, .5),
  breaks = c(-.5, 0, .5),
  labels = c("–0.5", "0.0", "0.5"),
  palette = "Earth",
  rev = FALSE
)
```
]

Palette name: Carto Earth

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Diverging scale example

.center.nogap[
```{r forensic-correlations3, echo = FALSE, message = FALSE, fig.width = 4., fig.asp = 1, dev = "svg"}
p + scale_fill_continuous_diverging(
  name = "correlation",
  limits = c(-.5, .5),
  breaks = c(-.5, 0, .5),
  labels = c("–0.5", "0.0", "0.5"),
  palette = "Blue-Red",
  rev = TRUE
)
```
]

Palette name: Blue-Red

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Uses of color in data visualization

<table style = "border: none; line-height: 2.5;">
<tr style = "background: white;">
<td style = "text-align: left; width: 50%;">
1. Distinguish categories (qualitative)
</td>
<td>
<img src = "color-scales_files/qualitative.png" width = 100% style = "text-align: right; vertical-align: middle;"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
2. Represent numeric values (sequential)
</td>
<td>
<img src = "color-scales_files/sequential.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
3. Represent numeric values (diverging)
</td>
<td>
<img src = "color-scales_files/diverging.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
4. Highlight
</td>
<td>
<img src = "color-scales_files/highlight.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>
</table>

---

# Highlight example

.center[
```{r Aus-athletes-track, echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
male_Aus <- ggridges::Aus_athletes %>%
  filter(sex == "m") %>%
  filter(
    sport %in% c("basketball", "field", "swimming", "track (400m)", "track (sprint)", "water polo")
  ) %>%
  mutate(
    sport = 
      case_when(
        sport == "track (400m)" ~ "track",
        sport == "track (sprint)" ~ "track",
        TRUE ~ sport
      ),
    sport = factor(
      sport,
      levels = c("track", "field", "water polo", "basketball", "swimming")
    )
  )

p <- ggplot(male_Aus, aes(x=height, y=pcBfat, shape=sport, color = sport, fill = sport)) +
  geom_point(size = 3) +
  scale_shape_manual(values = 21:25) +
  xlab("height (cm)") +
  ylab("% body fat") +
  theme_minimal_grid(14)

colors <- c("#BD3828", rep("#808080", 4))
fills <- c(
  alpha(colors[1], .815),
  alpha(colors[2:5], .5)
)

p + 
  scale_color_manual(values = colors) +
  scale_fill_manual(values = fills)
```
]

Palette name: Grays with accents

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Highlight example

.center[
```{r Aus-athletes-track2, echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}

accent_OkabeIto <- c("#E69F00", "#56B4E9", "#CC79A7", "#F0E442", "#0072B2", "#009E73", "#D55E00")
accent_OkabeIto[1:4] <- desaturate(lighten(accent_OkabeIto[1:4], .4), .8)
accent_OkabeIto[5:7] <- darken(accent_OkabeIto[5:7], .3)

colors <- c(accent_OkabeIto[5], darken(accent_OkabeIto[1:4], .2))
fills <- c(
  alpha(accent_OkabeIto[5], .7),
  alpha(accent_OkabeIto[1:4], .7)
)

p + 
  scale_color_manual(values = colors) +
  scale_fill_manual(values = fills)
```
]

Palette name: Okabe-Ito accent

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Highlight example

.center[
```{r Aus-athletes-track3, echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}

accent_Brewer <- RColorBrewer::brewer.pal(7, name = "Accent")[c(7, 1:4)]

colors <- darken(accent_Brewer, .2)
fills <- c(accent_Brewer[1], alpha(accent_Brewer[2:5], .7))

p + 
  scale_color_manual(values = colors) +
  scale_fill_manual(values = fills)
```
]

Palette name: ColorBrewer accent

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

# Uses of color in data visualization

<table style = "border: none; line-height: 2.5;">
<tr style = "background: white;">
<td style = "text-align: left; width: 50%;">
1. Distinguish categories (qualitative)
</td>
<td>
<img src = "color-scales_files/qualitative.png" width = 100% style = "text-align: right; vertical-align: middle;"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
2. Represent numeric values (sequential)
</td>
<td>
<img src = "color-scales_files/sequential.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
3. Represent numeric values (diverging)
</td>
<td>
<img src = "color-scales_files/diverging.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>

<tr style = "background: white;">
<td style = "text-align: left;">
4. Highlight
</td>
<td>
<img src = "color-scales_files/highlight.png" width = 100% style = "text-align: right; vertical-align: middle"></img>
</td>
</tr>
</table>

[//]: # "segment ends here"

---
class: center middle

## Color scales in **ggplot2**

---

## **ggplot2** color scale functions are a bit of a mess

--

.small-font.center[

Scale function            | Aesthetic &nbsp;&nbsp;&nbsp;   | Data type     | Palette type
:-----------              | :----------  | :------------ | :------------
`scale_color_hue()` &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | `color` | discrete  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | qualitative &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
]

---

## **ggplot2** color scale functions are a bit of a mess

.small-font.center[

Scale function            | Aesthetic &nbsp;&nbsp;&nbsp;   | Data type     | Palette type
:-----------              | :----------  | :------------ | :------------
`scale_color_hue()` &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | `color` | discrete  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | qualitative &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
`scale_fill_hue()`        | `fill `      | discrete      | qualitative
]

---

## **ggplot2** color scale functions are a bit of a mess

.small-font.center[

Scale function            | Aesthetic &nbsp;&nbsp;&nbsp;   | Data type     | Palette type
:-----------              | :----------  | :------------ | :------------
`scale_color_hue()` &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | `color` | discrete  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | qualitative &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
`scale_fill_hue()`        | `fill `      | discrete      | qualitative
`scale_color_gradient()`  | `color`      | continuous    | sequential
]

---

## **ggplot2** color scale functions are a bit of a mess

.small-font.center[

Scale function            | Aesthetic &nbsp;&nbsp;&nbsp;   | Data type     | Palette type
:-----------              | :----------  | :------------ | :------------
`scale_color_hue()` &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | `color` | discrete  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | qualitative &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
`scale_fill_hue()`        | `fill `      | discrete      | qualitative
`scale_color_gradient()`  | `color`      | continuous    | sequential
`scale_color_gradient2()` | `color`      | continuous    | diverging
]

---

## **ggplot2** color scale functions are a bit of a mess

.small-font.center[

Scale function            | Aesthetic &nbsp;&nbsp;&nbsp;   | Data type     | Palette type
:-----------              | :----------  | :------------ | :------------
`scale_color_hue()` &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | `color` | discrete  &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | qualitative &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
`scale_fill_hue()`        | `fill `      | discrete      | qualitative
`scale_color_gradient()`  | `color`      | continuous    | sequential
`scale_color_gradient2()` | `color`      | continuous    | diverging
`scale_fill_viridis_c()`   | `color`      | continuous    | sequential
`scale_fill_viridis_d()`   | `fill`       | discrete      | sequential
`scale_color_brewer()`    | `color`      | discrete      | qualitative, diverging, sequential
`scale_fill_brewer()`     | `fill`       | discrete      | qualitative, diverging, sequential
`scale_color_distiller()` | `color`      | continuous    | qualitative, diverging, sequential
]

... and there are many many more

---

## Examples

.small-font[
```{r temps-tiles1, eval = FALSE}
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile(width = 0.95, height = 0.95) + 
  coord_fixed(expand = FALSE) +
  theme_classic()
  # no fill scale defined, default is scale_fill_gradient()
```
]

.center[
```{r temps-tiles1-out, ref.label = "temps-tiles1", echo = FALSE, fig.width = 7, fig.asp = .3, dev = "svg"}
```
]


---

## Examples

.small-font[
```{r temps-tiles2, eval = FALSE}
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile(width = 0.95, height = 0.95) + 
  coord_fixed(expand = FALSE) +
  theme_classic() +
  scale_fill_gradient()
```
]

.center[
```{r temps-tiles2-out, ref.label = "temps-tiles2", echo = FALSE, fig.width = 7, fig.asp = .3, dev = "svg"}
```
]

---

## Examples

.small-font[
```{r temps-tiles3, eval = FALSE}
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile(width = 0.95, height = 0.95) + 
  coord_fixed(expand = FALSE) +
  theme_classic() +
  scale_fill_viridis_c()
```
]

.center[
```{r temps-tiles3-out, ref.label = "temps-tiles3", echo = FALSE, fig.width = 7, fig.asp = .3, dev = "svg"}
```
]


---

## Examples

.small-font[
```{r temps-tiles4, eval = FALSE}
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile(width = 0.95, height = 0.95) + 
  coord_fixed(expand = FALSE) +
  theme_classic() +
  scale_fill_viridis_c(option = "B", begin = 0.15)
```
]

.center[
```{r temps-tiles4-out, ref.label = "temps-tiles4", echo = FALSE, fig.width = 7, fig.asp = .3, dev = "svg"}
```
]


---

## Examples

.small-font[
```{r temps-tiles5, eval = FALSE}
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile(width = 0.95, height = 0.95) + 
  coord_fixed(expand = FALSE) +
  theme_classic() +
  scale_fill_distiller(palette = "YlGnBu")
```
]

.center[
```{r temps-tiles5-out, ref.label = "temps-tiles5", echo = FALSE, fig.width = 7, fig.asp = .3, dev = "svg"}
```
]

---

## The **colorspace** package creates some order

Scale name: `scale_<aesthetic>_<datatype>_<colorscale>()`

--

.small-font[
- `<aesthetic>`: name of the aesthetic (`fill`, `color`, `colour`)
- `<datatype>`: type of variable plotted (`discrete`, `continuous`, `binned`)
- `<colorscale>`: type of the color scale (`qualitative`, `sequential`, `diverging`, `divergingx`)
]

--

.small-font.center[

Scale function                       | Aesthetic &nbsp;&nbsp;&nbsp; | Data type  | Palette type &nbsp;&nbsp;&nbsp;
:-----------                         | :--------  | :--------- | :------------
`scale_color_discrete_qualitative()` | `color`    | discrete   | qualitative
`scale_fill_continuous_sequential()` | `fill`     | continuous | sequential
`scale_colour_continous_divergingx()` | `colour`   | continuous | diverging
]

---

## Examples

.tiny-font[
```{r temps-tiles6, eval = FALSE}
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile(width = 0.95, height = 0.95) + 
  coord_fixed(expand = FALSE) +
  theme_classic() +
  scale_fill_continuous_sequential(palette = "YlGnBu", rev = FALSE)
```
]

.center[
```{r temps-tiles6-out, ref.label = "temps-tiles6", echo = FALSE, fig.width = 7, fig.asp = .3, dev = "svg"}
```
]

---

## Examples

.tiny-font[
```{r temps-tiles7, eval = FALSE}
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile(width = 0.95, height = 0.95) + 
  coord_fixed(expand = FALSE) +
  theme_classic() +
  scale_fill_continuous_sequential(palette = "Viridis", rev = FALSE)
```
]

.center[
```{r temps-tiles7-out, ref.label = "temps-tiles7", echo = FALSE, fig.width = 7, fig.asp = .3, dev = "svg"}
```
]

---

## Examples

.tiny-font[
```{r temps-tiles8, eval = FALSE}
ggplot(temps_months, aes(x = month, y = location, fill = mean)) + 
  geom_tile(width = 0.95, height = 0.95) + 
  coord_fixed(expand = FALSE) +
  theme_classic() +
  scale_fill_continuous_sequential(palette = "Inferno", begin = 0.15, rev = FALSE)
```
]

.center[
```{r temps-tiles8-out, ref.label = "temps-tiles8", echo = FALSE, fig.width = 7, fig.asp = .3, dev = "svg"}
```
]

---

.tiny-font.nogap[
```{r colorspace-palettes-seq, eval = FALSE}
colorspace::hcl_palettes(type = "sequential", plot = TRUE) # all sequential palettes
```

.center[
```{r colorspace-palettes-seq-out, ref.label = 'colorspace-palettes-seq', echo = FALSE, fig.width = 12, fig.asp = .5, dev = "svg"}
```
]
]


---

.tiny-font.nogap[
```{r colorspace-palettes-div, eval = FALSE}
colorspace::hcl_palettes(type = "diverging", plot = TRUE, n = 9) # all diverging palettes
```

.center[
```{r colorspace-palettes-div-out, ref.label = 'colorspace-palettes-div', echo = FALSE, fig.width = 6, fig.asp = 1, dev = "svg"}
```
]
]

---

.tiny-font.nogap[
```{r colorspace-palettes-divx, eval = FALSE}
colorspace::divergingx_palettes(plot = TRUE, n = 9) # all divergingx palettes
```

.center[
```{r colorspace-palettes-divx-out, ref.label = 'colorspace-palettes-divx', echo = FALSE, fig.width = 8, fig.asp = .7, dev = "svg"}
```
]
]


[//]: # "segment ends here"

---
class: center middle

## Setting colors manually for discrete, qualitative scales

---

## Discrete, qualitative scales are best set manually

.tiny-font[
```{r qual-scales-example1, eval = FALSE}
ggplot(popgrowth, aes(x = pop2000, y = popgrowth, color = region)) +
  geom_point() +
  scale_x_log10()
  # no color scale defined, default is scale_color_hue()
```
]

.center[
```{r qual-scales-example1-out, ref.label = "qual-scales-example1", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Discrete, qualitative scales are best set manually

.tiny-font[
```{r qual-scales-example2, eval = FALSE}
ggplot(popgrowth, aes(x = pop2000, y = popgrowth, color = region)) +
  geom_point() +
  scale_x_log10() +
  scale_color_hue()
```
]

.center[
```{r qual-scales-example2-out, ref.label = "qual-scales-example2", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Discrete, qualitative scales are best set manually

.tiny-font[
```{r qual-scales-example3, message = FALSE, warning = FALSE, eval = FALSE}
library(ggthemes)  # for scale_color_colorblind()

ggplot(popgrowth, aes(x = pop2000, y = popgrowth, color = region)) +
  geom_point() +
  scale_x_log10() +
  scale_color_colorblind()  # uses Okabe-Ito colors
```
]

.center[
```{r qual-scales-example3-out, ref.label = "qual-scales-example3", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Discrete, qualitative scales are best set manually

.tiny-font[
```{r qual-scales-example4, eval = FALSE}
ggplot(popgrowth, aes(x = pop2000, y = popgrowth, color = region)) +
  geom_point() +
  scale_x_log10() +
  scale_color_manual(
    values = c(West = "#E69F00", South = "#56B4E9", Midwest = "#009E73", Northeast = "#F0E442")
  )
```
]

.center[
```{r qual-scales-example4-out, ref.label = "qual-scales-example4", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

---

## Okabe-Ito RGB codes

.center[
<img src = "https://clauswilke.com/dataviz/pitfalls_of_color_use_files/figure-html/palette-Okabe-Ito-1.png", width = 100%></img>
]

.tiny-font[
Name           | Hex code &nbsp;&nbsp; | R, G, B (0-255)
:----------    | :-------  | :--------
orange         | #E69F00   | 230, 159, 0
sky blue	     | #56B4E9   | 86, 180, 233
bluish green   | #009E73   | 0, 158, 115
yellow	       | #F0E442   | 240, 228, 66
blue	         | #0072B2   | 0, 114, 178
vermilion	     | #D55E00   | 213, 94, 0
reddish purple | #CC79A7   | 204, 121, 167
black	         | #000000   | 0, 0, 0
]

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

[//]: # "segment ends here"

---

## Further reading

- Fundamentals of Data Visualization: [Chapter 4: Color scales](https://clauswilke.com/dataviz/color-basics.html)
- Fundamentals of Data Visualization: [Figure 19.10: Okabe-Ito color palette](https://clauswilke.com/dataviz/color-pitfalls.html#fig:palette-Okabe-Ito)
- **ggplot2** book: [Colour scales and legends](https://ggplot2-book.org/scale-colour.html)
- **ggplot2** reference documentation: [Scales](https://ggplot2.tidyverse.org/reference/index.html#section-scales)
- **colorspace** package: [HCL-Based Color Scales for ggplot2](https://colorspace.r-forge.r-project.org/articles/ggplot2_color_scales.html)



