---
title: "Color spaces and color-vision deficiency"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
params:
  cache: TRUE
---

```{r setup, include=FALSE, echo=FALSE, message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(colorspace)
library(cowplot)
library(here)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2.0)) # 2 for full width, 2.5 for half width

# color space plot theme
theme_colorspace <- function(size = 12) {
  theme_minimal_grid(size, line_size = 0.2) +
  panel_border(color = "black", size = 0.4) +
  theme(
    axis.ticks.length = grid::unit(0, "pt"),
    axis.ticks = element_line(size = 0.2, color = "black"),
    strip.text = element_text(size = size)
  )
}
```

class: center middle

## How does a computer represent color?

---

## The RGB color space

.pull-left.width-50[
- red R (0-255): amount of red light

- green G (0-255): amount of green light

- blue B (0-255): amount of blue light
]

--

.pull-right.move-down-1em[

R | G | B | hex code | &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;color&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
---| --- | --- | --- | ---
0  | 0 | 0 | #000000 | <div style = "background-color: #000000;">&nbsp;</div>
255  | 0 | 0 | #FF0000 | <div style = "background-color: #FF0000;">&nbsp;</div>
0  | 255 | 255 | #00FFFF | <div style = "background-color: #00FFFF;">&nbsp;</div>
128  | 128 | 128 | #808080 | <div style = "background-color: #808080;">&nbsp;</div>
0 | 158 | 115 | #009E73 | <div style = "background-color: #009E73;">&nbsp;</div>
255 | 255 | 255 | #FFFFFF | <div style = "background-color: #FFFFFF;">&nbsp;</div>
]

---

## The RGB color space

.center[
```{r rgb-viz, echo = FALSE, cache = params$cache, fig.width = 10, fig.asp = .5, dev = "svg"}
rgb_by_b <- expand_grid(
  R = seq(0, 255, by = 2),
  G = seq(0, 255, by = 2),
  B = seq(0, 255, by = 35)
)

rgb_by_b$hex <- hex(sRGB(as.matrix(rgb_by_b)/255))

rgb_by_b %>%
  ggplot(aes(R, G, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_identity() +
  scale_x_continuous(
    name = "red R",
    breaks = (0:5)*50,
    labels = c("0", "", "100", "", "200", "")
  ) +
  scale_y_continuous(
    name = "green G",
    breaks = (0:5)*50,
    labels = c("0", "", "100", "", "200", "")
  ) +
  facet_wrap(
    vars(B),
    ncol = 4,
    labeller = as_labeller(function(x) glue::glue("blue B = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace() +
  theme(axis.ticks.length = grid::unit(3, "pt"))
```
]

---
class: center middle

## Humans cannot reason well about the RGB color space

---

## The HSV color space

.pull-left[

- hue H (0-360):  
  hue of the color

- saturation S (0-1):  
  colorfulness relative to the brightness of the color

- value V (0-1):  
  subjective perception of amount of light emitted
]

.pull-right.move-up-1em[
```{r hsv-viz-single, echo = FALSE, cache = params$cache, fig.width = 5, fig.asp = 1, dev = "svg"}
hsv_by_h <- expand_grid(
  H = seq(0, 330, by = 30),
  S = seq(0, 1, by = .01),
  V = seq(0, 1, by = .01)
)

hsv_by_h$hex <- hex(HSV(as.matrix(hsv_by_h)))

hsv_by_h %>%
  filter(H == 90) %>%
  ggplot(aes(S, V, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_identity() +
  scale_x_continuous(
    name = "saturation S",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "")
  ) +
  scale_y_continuous(
    name = "value V",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "1.0")
  ) +
  facet_wrap(
    vars(H),
    ncol = 6,
    labeller = as_labeller(function(x) glue::glue("hue H = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace(16) +
  theme(axis.ticks.length = grid::unit(3, "pt"))
```
]


---

## The HSV color space

.center[
```{r hsv-viz, echo = FALSE, cache = params$cache, fig.width = 10, fig.asp = .5, dev = "svg"}
hsv_by_h %>%
  ggplot(aes(S, V, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_identity() +
  scale_x_continuous(
    name = "saturation S",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "")
  ) +
  scale_y_continuous(
    name = "value V",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "1.0")
  ) +
  facet_wrap(
    vars(H),
    ncol = 6,
    labeller = as_labeller(function(x) glue::glue("hue H = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace() +
  theme(axis.ticks.length = grid::unit(3, "pt"))
```
]

---

## The HLS color space

.pull-left[
- hue H (0-360):  
  hue of the color

- lightness L (0-1):  
  brightness relative to the brightness of a similarly illuminated white

- saturation S (0-1):  
  colorfulness relative to the brightness of the color
]

.pull-right.move-up-1em[
```{r hls-viz-single, echo = FALSE, cache = params$cache, fig.width = 5, fig.asp = 1, dev = "svg"}
hls_by_h <- expand_grid(
  H = seq(0, 330, by = 30),
  L = seq(0, 1, by = .01),
  S = seq(0, 1, by = .01)
)

hls_by_h$hex <- hex(HLS(as.matrix(hls_by_h)))

hls_by_h %>%
  filter(H == 90) %>%
  ggplot(aes(S, L, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_identity() +
  scale_x_continuous(
    name = "saturation S",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "")
  ) +
  scale_y_continuous(
    name = "lightness L",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "1.0")
  ) +
  facet_wrap(
    vars(H),
    ncol = 6,
    labeller = as_labeller(function(x) glue::glue("hue H = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace(16) +
  theme(axis.ticks.length = grid::unit(3, "pt"))
```

]

---

## The HLS color space

.center[
```{r hls-viz, echo = FALSE, cache = params$cache, fig.width = 10, fig.asp = .5, dev = "svg"}
hls_by_h %>%
  ggplot(aes(S, L, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_identity() +
  scale_x_continuous(
    name = "saturation S",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "")
  ) +
  scale_y_continuous(
    name = "lightness L",
    breaks = (0:4)*.25,
    labels = c("0", "", "0.5", "", "1.0")
  ) +
  facet_wrap(
    vars(H),
    ncol = 6,
    labeller = as_labeller(function(x) glue::glue("hue H = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace() +
  theme(axis.ticks.length = grid::unit(3, "pt"))
```
]

---
class: center middle

## Humans cannot reason well about HSV or HLS either

---

## The Luv color space

.pull-left.width-50[

- luminance L (0-100):  
  amount of light emitted

- green/red axis u (approx. -100 to 100):  
  amount of green vs. red

- blue/yellow axis v (approx. -100 to 100):  
  amount of blue vs. yellow
]

.pull-right.move-up-2em[
```{r luv-viz-single, echo = FALSE, cache = params$cache, fig.width = 5, fig.asp = 1, dev = "svg"}
luv_by_l <- expand_grid(
  L = seq(20, 90, by = 10),
  U = seq(-80, 170, by = 1),
  V = seq(-145, 110, by = 1),
)

luv_by_l$hex <- hex(LUV(as.matrix(luv_by_l)))

luv_by_l %>%
  filter(L == 60) %>%
  ggplot(aes(U, V, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_x_continuous(
    name = "green/red axis u",
    breaks = (-3:3)*50,
    labels = c("", "-100", "", "0", "", "100", "")
  ) +
  scale_y_continuous(
    name = "blue/yellow axis v",
    breaks = (-3:3)*50,
    labels = c("", "-100", "", "0", "", "100", "")
  ) +
  scale_fill_identity() +
  facet_wrap(
    vars(L),
    ncol = 4,
    labeller = as_labeller(function(x) glue::glue("luminance L = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace(16)
```
]


---

## The Luv color space

.center[
```{r luv-viz, echo = FALSE, cache = params$cache, fig.width = 10, fig.asp = .5, dev = "svg"}
luv_by_l %>%
  ggplot(aes(U, V, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_x_continuous(
    name = "green/red axis u",
    breaks = (-3:3)*50,
    labels = c("", "-100", "", "0", "", "100", "")
  ) +
  scale_y_continuous(
    name = "blue/yellow axis v",
    breaks = (-3:3)*50,
    labels = c("", "-100", "", "0", "", "100", "")
  ) +
  scale_fill_identity() +
  facet_wrap(
    vars(L),
    ncol = 4,
    labeller = as_labeller(function(x) glue::glue("luminance L = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace()
```
]

---

## The HCL color space: polar Luv

.center.move-up-1em[
```{r hcl-viz-single, echo = FALSE, cache = params$cache, fig.width = 5.5, fig.asp = 1, dev = "svg"}

rC <- 60
hue_df <- tibble(
  t = seq(0, 2*pi, length.out = 100),
  U = rC*sin(t),
  V = rC*cos(t)
)

chroma_df <- tibble(
  U = 4*rC*sin(pi/3),
  V = 4*rC*cos(pi/3)
)

labels_df <- tibble(
  U = c(1.4*rC*sin(pi/3), 1.01*rC*sin(.8*pi)),
  V = c(1.4*rC*cos(pi/3), 1.01*rC*cos(.8*pi)),
  label = c("chroma C", "hue L")
)

luv_by_l %>%
  filter(L == 60) %>%
  ggplot(aes(U, V)) +
  geom_raster(aes(fill = hex), na.rm = TRUE) +
  geom_segment(
    data = chroma_df,
    aes(x = 0, y = 0, xend = U, yend = V),
    size = 0.2
  ) +
  geom_path(
    data = hue_df,
    size = 0.2
  ) +
  geom_text(
    data = labels_df,
    aes(label = label),
    size = 16*(12/14)/.pt,
    hjust = 0, vjust = 1
  ) +
  scale_x_continuous(
    name = "green/red axis u",
    breaks = (-3:3)*50,
    labels = c("", "-100", "-50", "0", "50", "100", "")
  ) +
  scale_y_continuous(
    name = "blue/yellow axis v",
    breaks = (-3:3)*50,
    labels = c("", "-100", "-50", "0", "50", "100", "")
  ) +
  scale_fill_identity() +
  facet_wrap(
    vars(L),
    ncol = 4,
    labeller = as_labeller(function(x) glue::glue("luminance L = {x}"))
  ) +
  coord_fixed(
    expand = FALSE,
    xlim = c(-65, 145),
    ylim = c(-120, 80)
  ) +
  theme_colorspace(16)
```
]


---

## The HCL color space: polar Luv

.move-up-1em.center[
```{r cl-planes, echo = FALSE, cache = params$cache, fig.width = 9, fig.asp = .6, dev = "svg"}
hcl_by_h <- expand_grid(
  L = seq(1, 99, by = 0.5),
  C = seq(0, 155, by = 1),
  H = seq(0, 330, by = 30)
)

hcl_by_h$hex <- hex(polarLUV(as.matrix(hcl_by_h)))

hcl_by_h %>%
  ggplot(aes(C, L, fill = hex)) +
  geom_raster(na.rm = TRUE) +
  scale_fill_identity() +
  scale_x_continuous(
    name = "chroma C",
    breaks = (0:3)*50,
    labels = c("0", "", "100", "")
  ) +
  scale_y_continuous(
    name = "luminance L",
    limits = c(0, 100),
    breaks = (0:4)*25,
    labels = c("0", "", "50", "", "100")
  ) +
  facet_wrap(
    vars(H),
    ncol = 4,
    labeller = as_labeller(function(x) glue::glue("hue H = {x}"))
  ) +
  coord_fixed(expand = FALSE) +
  theme_colorspace()
```
]

[//]: # "segment ends here"


---
class: center middle

## Explore HCL colors interactively in R

---

## We can explore HCL colors in R

```{r eval = FALSE}
colorspace::choose_color()
```

--

<img src = "color-spaces_files/choose_color1.png", width = 35%, style = "position:absolute; top: 20%; left: 52.5%; box-shadow: 3px 5px 3px 1px #00000080;"></img>

--

<img src = "color-spaces_files/choose_color2.png", width = 35%, style = "position:absolute; top: 35%; left: 12.5%; box-shadow: 3px 5px 3px 1px #00000080;"></img>

--

<img src = "color-spaces_files/choose_color3.png", width = 35%, style = "position:absolute; top: 50%; left: 42.5%; box-shadow: 3px 5px 3px 1px #00000080;"></img>


[//]: # "segment ends here"

---
class: center middle

## A few considerations when choosing colors

---

## 1. Avoid high chroma

--

.pull-left.width-45[
High chroma: Toys

<img src = "color-spaces_files/toys-image.jpg", width = 100%, style = "box-shadow: 3px 5px 3px 1px #00000080;"></img>
]

???

[Photo by Pixabay from Pexels](https://www.pexels.com/photo/super-mario-and-yoshi-plastic-figure-163077/)

--

.pull-right.width-45[
Low chroma: Elegance

<img src = "color-spaces_files/home-interior.jpg", width = 95%, style = "box-shadow: 3px 5px 3px 1px #00000080;"></img>
]

???

[Photo by Saviesa Home from Pexels](https://www.pexels.com/photo/kitchen-island-2089698/)

---

## 2. Be aware of color-vision deficiency

5%–8% of men are color blind!

--

<img src = "https://clauswilke.com/dataviz/pitfalls_of_color_use_files/figure-html/red-green-cvd-sim-1.png", width = 100%></img>

Red-green color-vision deficiency is the most common

---

## 2. Be aware of color-vision deficiency

5%–8% of men are color blind!

<img src = "https://clauswilke.com/dataviz/pitfalls_of_color_use_files/figure-html/blue-green-cvd-sim-1.png", width = 100%></img>

Blue-green color-vision deficiency is rare but does occur

---

## 2. Be aware of color-vision deficiency

Choose colors that can be distinguished with CVD

<img src = "https://clauswilke.com/dataviz/pitfalls_of_color_use_files/figure-html/PiYG-cvd-sim-1.png", width = 100%></img>

---

## Consider using the Okabe-Ito scale

.center[
<img src = "https://clauswilke.com/dataviz/pitfalls_of_color_use_files/figure-html/palette-Okabe-Ito-1.png", width = 100%></img>
]

.tiny-font[
Name           | Hex code &nbsp;&nbsp; | R, G, B (0-255)
:----------    | :-------  | :--------
orange         | #E69F00   | 230, 159, 0
sky blue	     | #56B4E9   | 86, 180, 233
bluish green   | #009E73   | 0, 158, 115
yellow	       | #F0E442   | 240, 228, 66
blue	         | #0072B2   | 0, 114, 178
vermilion	     | #D55E00   | 213, 94, 0
reddish purple | #CC79A7   | 204, 121, 167
black	         | #000000   | 0, 0, 0
]

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## CVD is worse for thin lines and tiny dots

.move-up-1em.center[
<img src = "https://clauswilke.com/dataviz/pitfalls_of_color_use_files/figure-html/colors-thin-lines-1.png", width = 65%></img>
]

---

## When in doubt, run CVD simulations

<div class = "move-up-1em" style = "width: 50%; box-shadow: 3px 5px 3px 1px #00000080; padding: 8px;">
.small-font[Original]<br>
<img src = "https://clauswilke.com/dataviz/redundant_coding_files/figure-html/iris-scatter-three-shapes-1.png", width = 100%></img>
</div>

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## When in doubt, run CVD simulations

<div class = "move-up-1em" style = "width: 50%; box-shadow: 3px 5px 3px 1px #00000080; padding: 8px;">
.small-font[Original]<br>
<img src = "https://clauswilke.com/dataviz/redundant_coding_files/figure-html/iris-scatter-three-shapes-1.png", width = 100%></img>
</div>

<img src = "https://clauswilke.com/dataviz/redundant_coding_files/figure-html/iris-scatter-three-shapes-cvd-1.png", width = 60% style = "position: absolute; top: 22%; left: 30%; box-shadow: 3px 5px 3px 1px #00000080; padding: 16px 8px 8px 16px; background: white"></img>

???

Figures from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

[//]: # "segment ends here"


---

## Further reading

- Fundamentals of Data Visualization: [Chapter 19: Common pitfalls of color use](https://clauswilke.com/dataviz/color-pitfalls.html)
- Wikipedia: [HSL and HSV](https://en.wikipedia.org/wiki/HSL_and_HSV)
- **colorspace** package documentation: [Color Spaces](https://colorspace.r-forge.r-project.org/articles/color_spaces.html)
- **colorspace** package documentation: [Apps for Choosing Colors and Palettes Interactively](https://colorspace.r-forge.r-project.org/articles/hclwizard.html)

