---
title: "Data wrangling 1"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      #slideNumberFormat: '%current%' 
      slideNumberFormat: '' 
      countIncrementalSlides: false
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(palmerpenguins)
```


# Elementary data manipulations

This class:

- Pick rows
- Pick columns
- Sort rows

--

Next class:

- Make new columns
- Count

---

# TODO: explain the pipe

---

background-image: url(data-wrangling-1_files/filter.svg)
background-position: 50% 60%
background-size: 75%

# Pick rows from a table: `filter()`

---

# Filter out penguins of species Gentoo

```{r}
penguins %>%
  filter(species == "Gentoo")
```

---

# Filter out penguins with bill length > 50 mm

```{r}
penguins %>%
  filter(bill_length_mm > 50)
```

---

background-image: url(data-wrangling-1_files/arrange.svg)
background-position: 50% 60%
background-size: 75%

# Sort the rows in a table: `arrange()`

---

# Sort penguins by bill length, ascending

```{r}
penguins %>%
  arrange(bill_length_mm)
```

---

# Sort penguins by bill length, descending

```{r}
penguins %>%
  arrange(desc(bill_length_mm))
```


