---
title: "Dimension reduction 2"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
params:
  cache: TRUE
---

```{r setup, include=FALSE, echo=FALSE, message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(broom)
library(cowplot)
library(Rtsne)
library(umap)
library(kernlab)
library(colorspace)
library(here)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2.0)) # 2 for full width, 2.5 for half width

# blue jays dataset
blue_jays <- read_csv(here("datasets", "blue_jays.csv"))

# spirals
spirals <- read_csv(here("datasets", "spirals.csv"))

outline_cols <- darken(c("#D55E00", "#0072B2", "#009E73"), .2)
fill_cols <- c("#D55E0080", "#0072B280", "#009E7380")
fill_cols_dark <- c("#D55E00C0", "#0072B2C0", "#009E73C0")
```

## What if a rotation cannot disentangle the data?

--

.center[
```{r spirals, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
ggplot(spirals, aes(x, y, color = group, fill = group)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2) +
  scale_x_continuous(
    name = NULL,
    limits = c(-4, 4)
  ) +
  scale_y_continuous(
    name = NULL,
    limits = c(-4, 4)
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols_dark,
    guide = "none"
  ) +
  coord_fixed(ratio = 1) +
  theme_bw(16)
```
]


---

## PCA analysis of intertwined spirals is not useful

.center[
```{r pca-spirals, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
pca_fit <- spirals %>% 
  select(where(is.numeric)) %>% # retain only numeric columns
  scale() %>%       # scale to zero mean and unit variance
  prcomp()          # do PCA

pca_fit %>%
  augment(spirals) %>% # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, color = group, fill = group)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2) +
  scale_x_continuous(
    name = "PC 1", 
    limits = c(-2, 2)
  ) +
  scale_y_continuous(
    name = "PC 2",
    limits = c(-2, 2)
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols_dark,
    guide = "none"
  ) +
  coord_fixed(ratio = 1) +
  theme_bw(16)
```
]

---

## One possible approach: Kernel PCA

--

- Kernel PCA performs PCA in a hypothetical, higher-dimensional space

--

- With more dimensions, data points become more separable

--

- Importantly, the space is never explicitly constructed ([kernel trick](https://en.wikipedia.org/wiki/Kernel_method#Mathematics:_the_kernel_trick))

--

- Results from kernel PCA depend on choice of kernel

---

## Kernel PCA can separate the spirals

.center[
```{r kpca-spirals, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = 0.618, dev = "svg"}
do_kpca <- function(data, sigma, kernel = "rbfdot") {
  data_num <- data %>%
    select(where(is.numeric)) %>%
    mutate(across(everything(), scale))
  
  kpca_fit <- kpca(
    ~., data = data_num, kernel = kernel,
    kpar = list(sigma = sigma),
    features = 2
  )

  kpca_fit %>%
    pcv() %>% # extract principal component vectors
    as.data.frame() %>% # convert to data frame
    mutate(
      kernel = kernel,
      sigma = sigma
    ) %>%
    cbind(select(data, -where(is.numeric)))
}

kpca_data <- 2**(-1:6) %>%
  signif(2) %>%
  map_dfr(~do_kpca(spirals, sigma = .x))

kpca_data %>% filter(sigma == 64.) %>%
  ggplot(aes(V1, V2, color = group, fill = group)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2) +
  scale_x_continuous(
    name = "dimension 1", 
    limits = c(-3, 1)
  ) +
  scale_y_continuous(
    name = "dimension 2",
    limits = c(-2.2, 2.2)
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols,
    guide = "none"
  ) +
  coord_fixed(ratio = 1) +
  theme_bw(16)
```
]

Gaussian kernel, sigma = 64

---

## But we need to choose the right sigma value

.center[
```{r kpca-spirals-grid, echo = FALSE, cache = params$cache, fig.width = 10, fig.asp = 0.5, dev = "svg"}
ggplot(kpca_data, aes(V1, V2, color = group, fill = group)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2) +
  facet_wrap(
    vars(sigma),
    labeller = as_labeller(function(x) glue::glue("sigma = {x}")),
    scales = "free",
    nrow = 2
  ) +
  scale_x_continuous(
    name = "dimension 1",
    limits = c(-3, 1)
  ) +
  scale_y_continuous(
    name = "dimension 2",
    limits = c(-2.2, 2.2)
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols,
    guide = "none"
  ) +
  theme_bw(16)
```
]

---

## Other approaches

--

- t-SNE: t-distributed stochastic neighbor embedding

--

- UMAP: Uniform manifold approximation and projection 

--

Both algorithms look at the local distances between points in the original data space and try to reproduce them in the low-dimensional representation


---

## t-SNE can separate the spirals

.center[
```{r tsne-spirals, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = 0.618, dev = "svg"}
## Run the t-SNE algorithm and store the results into an object called tsne_results
set.seed(1234)
do_tsne <- function(data, perplexity = 5, theta = 0.3, seed = 1234) {
  set.seed(seed)
  tsne_fit <- data %>%
    select(where(is.numeric)) %>%
    scale() %>%
    Rtsne(perplexity = perplexity, theta = theta, check_duplicates = FALSE)

  tsne_fit$Y %>%
    as.data.frame() %>%
    mutate(
      perplexity = perplexity,
      theta = theta,
      seed = seed
    ) %>%
    cbind(select(data, -where(is.numeric)))
}

tsne_data <- c(1, 2, 3, 4, 5, 10, 15, 20) %>%
  map_dfr(~do_tsne(spirals, perplexity = .x))

tsne_data %>%
  filter(perplexity == 5) %>%
  ggplot(aes(V1, V2, color = group, fill = group)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2) +
  scale_x_continuous(
    name = "dimension 1", 
    limits = c(-70, 70)
  ) +
  scale_y_continuous(
    name = "dimension 2",
    limits = c(-70, 70)
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols,
    guide = "none"
  ) +
  coord_fixed(ratio = 1) +
  theme_bw(16)
```
]

---

## t-SNE results depend on the perplexity value

.center[
```{r tsne-spirals-grid, echo = FALSE, cache = params$cache, fig.width = 10, fig.asp = 0.5, dev = "svg"}
ggplot(tsne_data, aes(V1, V2, color = group, fill = group)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2) +
  facet_wrap(
    vars(perplexity),
    labeller = as_labeller(function(x) glue::glue("perplexity = {x}")),
    scales = "free",
    nrow = 2
  ) +
  scale_x_continuous(
    name = "dimension 1",
    limits = c(-100, 100)
  ) +
  scale_y_continuous(
    name = "dimension 2",
    limits = c(-100, 100)
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols,
    guide = "none"
  ) +
  theme_bw(16)
```
]


---

## t-SNE results depend on the random starting point

.center[
```{r tsne-spirals-grid2, echo = FALSE, cache = params$cache, fig.width = 10, fig.asp = 0.5, dev = "svg"}
tsne_data2 <- 1:8 %>%
  map_dfr(~do_tsne(spirals, seed = .x))

ggplot(tsne_data2, aes(V1, V2, color = group, fill = group)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2) +
  facet_wrap(
    vars(seed),
    labeller = as_labeller(function(x) glue::glue("case {x}")),
    scales = "free",
    nrow = 2
  ) +
  scale_x_continuous(
    name = "dimension 1",
    limits = c(-100, 100)
  ) +
  scale_y_continuous(
    name = "dimension 2",
    limits = c(-100, 100)
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols,
    guide = "none"
  ) +
  theme_bw(16)
```
]

---

## UMAP can separate the spirals

.center[
```{r umap-spirals, echo = FALSE, cache = params$cache, fig.width = 7, fig.asp = 0.618, dev = "svg"}
do_umap <- function(data, n_neighbors = 15, seed = 123) {
  custom.config <- umap.defaults
  custom.config$n_neighbors <- n_neighbors
  custom.config$random_state <- seed
  custom.config$n_epochs <- 500 

  umap_fit <- data %>%
    select(where(is.numeric)) %>%
    scale() %>%
    umap(config = custom.config)

  umap_fit$layout %>%
    as.data.frame() %>%
    mutate(
      n_neighbors = n_neighbors,
      seed = seed
    ) %>%
    cbind(select(data, -where(is.numeric)))
}

umap_data <- c(4, 5, 6, 8, 10, 12, 16, 20) %>%
  map_dfr(~do_umap(spirals, n_neighbors = .x))

umap_data %>%
  filter(n_neighbors == 5) %>%
  ggplot(aes(V1, V2, color = group, fill = group)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2) +
  scale_x_continuous(
    name = "dimension 1", 
    limits = c(-30, 31)
  ) +
  scale_y_continuous(
    name = "dimension 2",
    limits = c(-30, 31)
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols,
    guide = "none"
  ) +
  coord_fixed(ratio = 1) +
  theme_bw(16)
```
]

---

## UMAP results depend on the number of neighbors

.center[
```{r umap-spirals-grid, echo = FALSE, cache = params$cache, fig.width = 10, fig.asp = 0.5, dev = "svg"}
ggplot(umap_data, aes(V1, V2, color = group, fill = group)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2) +
  facet_wrap(
    vars(n_neighbors),
    labeller = as_labeller(function(x) glue::glue("neighbors = {x}")),
    scales = "free",
    nrow = 2
  ) +
  scale_x_continuous(
    name = "dimension 1",
    limits = c(-30, 31)
  ) +
  scale_y_continuous(
    name = "dimension 2",
    limits = c(-30, 31)
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols,
    guide = "none"
  ) +
  theme_bw(16)
```
]


---

## Random starting point has some impact on results

.center[
```{r umap-spirals-grid2, echo = FALSE, cache = params$cache, fig.width = 10, fig.asp = 0.5, dev = "svg"}
umap_data2 <- 1:8 %>%
  map_dfr(~do_umap(spirals, n_neighbors = 5, seed = .x))

ggplot(umap_data2, aes(V1, V2, color = group, fill = group)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2) +
  facet_wrap(
    vars(seed),
    labeller = as_labeller(function(x) glue::glue("case {x}")),
    scales = "free",
    nrow = 2
  ) +
  scale_x_continuous(
    name = "dimension 1",
    limits = c(-30, 31)
  ) +
  scale_y_continuous(
    name = "dimension 2",
    limits = c(-30, 31)
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols,
    guide = "none"
  ) +
  theme_bw(16)
```
]

---

## What is the meaning of the tuning parameters?

--

Tuning parameters define when points are close in the original data space

--

This implicitly defines the number of clusters generated

--

These have comparable effects:

- sigma (Gaussian kernel PCA)

- perplexity (t-SNE)

- number of neighbors (UMAP)

---
class: center middle

## How do these methods perform<br>on the blue jays dataset?

---

## UMAP of blue jays

.center[
```{r umap-blue-jays-gray, echo = FALSE, cache = params$cache, fig.width = 6, fig.asp = 0.85, dev = "svg"}
bj_umap_data <- c(16) %>%
  map_dfr(~do_umap(blue_jays, n_neighbors = .x, seed = 1234))

bj_umap_data %>%
  filter(n_neighbors == 16) %>%
  ggplot(aes(V1, V2, color = sex, fill = sex)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2, color = "black", fill = "#505050D0") +
  scale_x_continuous(
    name = "dimension 1" 
  ) +
  scale_y_continuous(
    name = "dimension 2"
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols_dark,
    guide = "none"
  ) +
  theme_bw(16)
```
]

---

## UMAP of blue jays

.center[
```{r umap-blue-jays-sex, echo = FALSE, cache = params$cache, fig.width = 6, fig.asp = 0.85, dev = "svg"}
bj_umap_data %>%
  filter(n_neighbors == 16) %>%
  ggplot(aes(V1, V2, color = sex, fill = sex)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2) +
  scale_x_continuous(
    name = "dimension 1"
  ) +
  scale_y_continuous(
    name = "dimension 2"
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols_dark,
    guide = "none"
  ) +
  theme_bw(16)
```
]

---

## Kernel PCA of blue jays

.center[
```{r kpca-blue-jays, echo = FALSE, cache = params$cache, fig.width = 6, fig.asp = 0.85, dev = "svg"}
bj_kpca_data <- c(.0001, .0003, .001, .0025, .01) %>%
  map_dfr(~do_kpca(blue_jays, sigma = .x))

bj_kpca_data %>%
  filter(sigma == .001) %>%
  ggplot(aes(V1, V2, color = sex, fill = sex)) + 
  geom_point(pch = 21, size = 2, stroke = 0.2) +
  scale_x_continuous(
    name = "dimension 1"
  ) +
  scale_y_continuous(
    name = "dimension 2"
  ) +
  scale_color_manual(
    values = outline_cols,
    guide = "none"
  ) +
  scale_fill_manual(
    values = fill_cols_dark,
    guide = "none"
  ) +
  theme_bw(16)
```
]

---

## Nonlinear methods have important downsides

--

- Results depend on parameter fine tuning

--

- Low-dimensional embedding cannot be interpreted<br>
  (no rotation matrix plot)

--

Use only when linear methods clearly aren't working

[//]: # "segment ends here"


---
class: middle center

## Doing nonlinear dimension reduction in R


---

## Doing nonlinear dimension reduction in R

--

- All these methods require special packages:<br>
  **kernlab** (kernel PCA)<br>
  **Rtsne** (t-SNE)<br>
  **umap** (UMAP)

--

- Code examples are somewhat messy

--

- Will do UMAP as example

---

## Doing UMAP in R

.tiny-font[
```{r}
library(umap)

# set up UMAP parameters
custom.config <- umap.defaults
custom.config$n_neighbors <- 16       # number of neighbors
custom.config$n_epochs <- 500         # number of iterations for convergence
custom.config$random_state <- 1234    # random seed

# calculate UMAP fit object
umap_fit <- blue_jays %>%
  select(where(is.numeric)) %>%       # retain only numeric columns
  scale() %>%                         # scale to zero mean and unit variance
  umap(config = custom.config)        # perform UMAP
```
]

---

## Doing UMAP in R

.pull-left.tiny-font[
```{r umap-ggplot-demo, eval = FALSE}
# extract data and plot
umap_fit$layout %>%
  as.data.frame() %>%
  mutate(sex = blue_jays$sex) %>%
  ggplot(aes(V1, V2, color = sex)) +
  geom_point()
```
]

.pull-right[
```{r umap-ggplot-demo-out, ref.label="umap-ggplot-demo", fig.width=5., fig.asp = 0.618, echo=FALSE, dev = "svg"}
```
]

[//]: # "segment ends here"

---

## Further reading

- Wikipedia: [Nonlinear dimensionality reduction](https://en.wikipedia.org/wiki/Nonlinear_dimensionality_reduction)
- Wikipedia: [t-distributed stochastic neighbor embedding](https://en.wikipedia.org/wiki/T-distributed_stochastic_neighbor_embedding)
- Wikipedia: [Kernel principal component analysis](https://en.wikipedia.org/wiki/Kernel_principal_component_analysis)
- **kernlab** reference documentation (for kernel PCA): [pdf document](https://cran.r-project.org/web/packages/kernlab/kernlab.pdf)
- **Rtsne** reference documentation: [pdf document](https://cran.r-project.org/web/packages/Rtsne/Rtsne.pdf)
- **umap** vignette: [Uniform Manifold Approximation and Projection in R](https://cran.r-project.org/web/packages/umap/vignettes/umap.html)