---
title: "Dimension reduction"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE, message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(broom)
library(cowplot)
library(here)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2.0)) # 2 for full width, 2.5 for half width

# blue jays dataset
blue_jays <- read_csv(here("datasets", "blue_jays.csv"))
```


## Principal Components Analysis (PCA)

.center[
```{r blue-jays-PCA, echo = FALSE, fig.width = 11, fig.asp = 0.35, dev = "svg"}
p1 <- ggplot(blue_jays, aes(skull_size_mm, head_length_mm, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2, stroke = 0.2) +
  scale_x_continuous(
    name = "skull size (mm)", limits = c(26, 36),
    breaks = c(26, 28, 30, 32, 34, 36)
  ) +
  scale_y_continuous(
    name = "head length (mm)", limits = c(51, 61),
    breaks = c(52, 54, 56, 58, 60)
  ) +
  scale_fill_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 3)
      )
    ) +
  coord_fixed() +
  theme_minimal_grid(12) +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
bj_standard <- data.frame(
  skull = scale(blue_jays$skull_size_mm),
  head = scale(blue_jays$head_length_mm),
  sex = blue_jays$sex
)
df_arrows <- data.frame(
  x = c(-3, 2),
  y = c(-3, -2),
  xend = c(3, -2),
  yend = c(3, 2)
)
df_labels <- data.frame(
  x = c(3.1, -2.1),
  y = c(3.1, 2.1),
  hjust = c(0, 1),
  vjust = c(0, 0),
  label = c("PC 1", "PC 2")
)
p2 <- ggplot(bj_standard, aes(skull, head, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2, stroke = 0.2) +
  geom_segment(
    data = df_arrows,
    aes(x = x, y = y, xend = xend, yend = yend),
    color = "black",
    arrow = arrow(angle = 15, length = grid::unit(9, "pt"), type = "closed"),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = df_labels,
    aes(x = x, y = y, hjust = hjust, vjust = vjust, label = label),
    size = 12/.pt,
    color = "black", inherit.aes = FALSE
  ) +
  scale_x_continuous(
    name = "skull size (scaled)", limits = c(-4, 4),
    breaks = c(-4, -2, 0, 2, 4), labels = c("-4.0", "-2.0", "0.0", "2.0", "4.0")
  ) +
  scale_y_continuous(
    name = "head length (scaled)", limits = c(-4, 4),
    breaks = c(-4, -2, 0, 2, 4), labels = c("-4.0", "-2.0", "0.0", "2.0", "4.0")
  ) +
  scale_fill_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("", ""),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(shape = NA)
      )
    ) +
  coord_fixed(ratio = 1) +
  theme_minimal_grid(12) +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
df_arrows <- data.frame(
  x = c(-5.2, 0),
  y = c(0, -3.5),
  xend = c(5.2, 0),
  yend = c(0, 3.5)
)
df_labels <- data.frame(
  x = c(5.2, 0),
  y = c(0.4, 3.8),
  hjust = c(0.9, 0.5),
  vjust = c(0, 0),
  label = c("PC 1", "PC 2")
)
p3 <- ggplot(bj_standard, aes(skull + head, -skull + head, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2, stroke = 0.2) +
  geom_segment(
    data = df_arrows,
    aes(x = x, y = y, xend = xend, yend = yend),
    color = "black",
    arrow = arrow(angle = 15, length = grid::unit(9, "pt"), type = "closed"),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = df_labels,
    aes(x = x, y = y, hjust = hjust, vjust = vjust, label = label),
    size = 12/.pt,
    color = "black", inherit.aes = FALSE
  ) +
  scale_x_continuous(
    name = "PC 1", limits = c(-5.2, 5.2),
    breaks = c(-5, -2.5, 0, 2.5, 5), labels = c("-5.0", "-2.5", "0.0", "2.5", "5.0")
) +
  scale_y_continuous(
    name = "PC 2", limits = c(-5.2, 5.2),
    breaks = c(-5, -2.5, 0, 2.5, 5), labels = c("-5.0", "-2.5", "0.0", "2.5", "5.0")
  ) +
  scale_fill_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("", ""),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(shape = NA)
      )
    ) +
  coord_fixed(ratio = 1) +
  theme_minimal_grid(12) +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )

aligned <- align_plots(p1, p2, p3, align = "hv")
p1 <- aligned[[1]]
p2 <- aligned[[2]]
p3 <- aligned[[3]]

plot_grid(
  p1, NULL, NULL, NULL, NULL,
  nrow = 1, rel_widths = c(1, .04, 1, .04, 1),
  labels = c("a", "", "", "", ""), label_y = 0.985
)
```
]

???


Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Principal Components Analysis (PCA)

.center[
```{r blue-jays-PCA2, echo = FALSE, fig.width = 11, fig.asp = 0.35, dev = "svg"}
plot_grid(
  p1, NULL, p2, NULL, NULL,
  nrow = 1, rel_widths = c(1, .04, 1, .04, 1),
  labels = c("a", "", "b", "", ""), label_y = 0.985
)
```
]

???


Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Principal Components Analysis (PCA)

.center[
```{r blue-jays-PCA3, echo = FALSE, fig.width = 11, fig.asp = 0.35, dev = "svg"}
plot_grid(
  p1, NULL, p2, NULL, p3,
  nrow = 1, rel_widths = c(1, .04, 1, .04, 1),
  labels = c("a", "", "b", "", "c"), label_y = 0.985
)
```
]

???


Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Scaling of numeric columns

.tiny-font[
```{r blue-jays-scaled, eval = FALSE}
blue_jays %>% 
  mutate(across(where(is.numeric), scale)) %>%  # scale all numeric columns
  ggplot(aes(skull_size_mm, head_length_mm, color = sex)) + geom_point()
```
]

.center[
```{r blue-jays-scaled-out, ref.label="blue-jays-scaled", fig.width = 5, fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

---

## Fully worked PCA example

First we run the PCA and store results as `pca_fit`:

.tiny-font[
```{r}
pca_fit <- blue_jays %>% 
  select(where(is.numeric)) %>% # retain only numeric columns
  scale() %>%       # scale to zero mean and unit variance
  prcomp()          # do PCA
```
]

--

Then we add PC coordinates into original dataset and plot:

.tiny-font[
```{r eval = FALSE}
library(broom)  # for augment(), tidy()

pca_fit %>%
  augment(blue_jays) %>% # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, color = sex)) + geom_point()
```
]

---

## Fully worked PCA example

.tiny-font[
```{r blue-jays-pca-plot, eval = FALSE}
pca_fit %>%
  augment(blue_jays) %>% # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, color = sex)) + geom_point()
```
]

.center[
```{r blue-jays-pca-plot-out, ref.label="blue-jays-pca-plot", fig.width=5., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

---

## Fully worked PCA example

.tiny-font[
```{r blue-jays-pca-plot2, eval = FALSE}
pca_fit %>%
  augment(blue_jays) %>% # add original dataset back in
  ggplot(aes(.fittedPC2, .fittedPC3, color = sex)) + geom_point()
```
]

.center[
```{r blue-jays-pca-plot2-out, ref.label="blue-jays-pca-plot2", fig.width=5., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

---

## Plot the rotation matrix

.tiny-font[
```{r blue-jays-pca-rotation, eval = FALSE}
arrow_style <- arrow(angle = 20, ends = "first", type = "closed", length = grid::unit(8, "pt"))

pca_fit %>%
  tidy(matrix = "rotation") %>% # extract rotation matrix
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") %>%
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0, yend = 0, arrow = arrow_style) +
  geom_text(aes(label = column), hjust = 1) +
  xlim(-1.5, 0.5) + ylim(-1, 1) + coord_fixed()
```
]


---

## Plot the variance explained

.tiny-font[
```{r blue-jays-pca-eigenvalues, eval = FALSE}
pca_fit %>%
  tidy(matrix = "eigenvalues") %>% # extract eigenvalues
  ggplot(aes(PC, 100*percent)) + geom_col() + ylab("percent")
```
]

.center[
```{r blue-jays-pca-eigenvalues-out, ref.label="blue-jays-pca-eigenvalues", fig.width=5., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

