---
title: "Dimension reduction 1"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE, message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(broom)
library(cowplot)
library(GGally)
library(here)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2.0)) # 2 for full width, 2.5 for half width

# blue jays dataset
blue_jays <- read_csv(here("datasets", "blue_jays.csv"))
```

## Problem:<br>How do we visualize datasets with many variables?

---

## Problem:<br>How do we visualize datasets with many variables?

Example: blue jays dataset

.center.tiny-font[
```{r echo = FALSE}
knitr::kable(blue_jays)
```
]

---

## All-by-all scatter plots are hard to read 

.center[
```{r blue-jays-scattermatrix, echo = FALSE, message = FALSE, cache = TRUE, fig.width = 10, fig.asp = 0.5, dev = "svg"}
blue_jays %>% 
  rename(
    `bill depth` = "bill_depth_mm", 
    `bill width` = "bill_width_mm", 
    `bill length` = "bill_length_mm",
    `head length` = "head_length_mm",
    `body mass` = "body_mass_g",
    `skull size` = "skull_size_mm"
  ) %>%
  ggpairs(
    mapping = aes(color = sex),
    columns = c(
      "bill depth", "bill width", "bill length",
      "head length", "body mass", "skull size"
    ),
    lower = list(
      continuous = wrap("points", size = 1, alpha = .6)
    ),
    upper = list(
      continuous = wrap("points", size = 1, alpha = .6)
    ),
    diag = list(
      continuous = wrap("densityDiag", alpha = 0.6, size = .1)
    )
  ) +
  theme_bw(9) +
  scale_fill_manual(values = c(F = "#D55E00", M = "#0072B2")) +
  scale_color_manual(values = c(F = "#D55E00", M = "#0072B2"))
```
]

---

## Solution: Dimension reduction

--

- We project the data into a lower-dimensional space

--

- Commonly-used method:<br>Principal Components Analysis (PCA)

--

- PCA is a rotation of the coordinate system

---

## Principal Components Analysis (PCA)

.center[
```{r blue-jays-PCA, echo = FALSE, fig.width = 11, fig.asp = 0.35, dev = "svg"}
p1 <- ggplot(blue_jays, aes(skull_size_mm, head_length_mm, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2, stroke = 0.2) +
  scale_x_continuous(
    name = "skull size (mm)", limits = c(26, 36),
    breaks = c(26, 28, 30, 32, 34, 36)
  ) +
  scale_y_continuous(
    name = "head length (mm)", limits = c(51, 61),
    breaks = c(52, 54, 56, 58, 60)
  ) +
  scale_fill_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 3)
      )
    ) +
  coord_fixed() +
  theme_minimal_grid(12) +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
bj_standard <- data.frame(
  skull = scale(blue_jays$skull_size_mm),
  head = scale(blue_jays$head_length_mm),
  sex = blue_jays$sex
)
df_arrows <- data.frame(
  x = c(-3, 2),
  y = c(-3, -2),
  xend = c(3, -2),
  yend = c(3, 2)
)
df_labels <- data.frame(
  x = c(3.1, -2.1),
  y = c(3.1, 2.1),
  hjust = c(0, 1),
  vjust = c(0, 0),
  label = c("PC 1", "PC 2")
)
p2 <- ggplot(bj_standard, aes(skull, head, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2, stroke = 0.2) +
  geom_segment(
    data = df_arrows,
    aes(x = x, y = y, xend = xend, yend = yend),
    color = "black",
    arrow = arrow(angle = 15, length = grid::unit(9, "pt"), type = "closed"),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = df_labels,
    aes(x = x, y = y, hjust = hjust, vjust = vjust, label = label),
    size = 12/.pt,
    color = "black", inherit.aes = FALSE
  ) +
  scale_x_continuous(
    name = "skull size (scaled)", limits = c(-4, 4),
    breaks = c(-4, -2, 0, 2, 4), labels = c("-4.0", "-2.0", "0.0", "2.0", "4.0")
  ) +
  scale_y_continuous(
    name = "head length (scaled)", limits = c(-4, 4),
    breaks = c(-4, -2, 0, 2, 4), labels = c("-4.0", "-2.0", "0.0", "2.0", "4.0")
  ) +
  scale_fill_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("", ""),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(shape = NA)
      )
    ) +
  coord_fixed(ratio = 1) +
  theme_minimal_grid(12) +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
df_arrows <- data.frame(
  x = c(-5.2, 0),
  y = c(0, -3.5),
  xend = c(5.2, 0),
  yend = c(0, 3.5)
)
df_labels <- data.frame(
  x = c(5.2, 0),
  y = c(0.4, 3.8),
  hjust = c(0.9, 0.5),
  vjust = c(0, 0),
  label = c("PC 1", "PC 2")
)
p3 <- ggplot(bj_standard, aes(skull + head, -skull + head, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2, stroke = 0.2) +
  geom_segment(
    data = df_arrows,
    aes(x = x, y = y, xend = xend, yend = yend),
    color = "black",
    arrow = arrow(angle = 15, length = grid::unit(9, "pt"), type = "closed"),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = df_labels,
    aes(x = x, y = y, hjust = hjust, vjust = vjust, label = label),
    size = 12/.pt,
    color = "black", inherit.aes = FALSE
  ) +
  scale_x_continuous(
    name = "PC 1", limits = c(-5.2, 5.2),
    breaks = c(-5, -2.5, 0, 2.5, 5), labels = c("-5.0", "-2.5", "0.0", "2.5", "5.0")
) +
  scale_y_continuous(
    name = "PC 2", limits = c(-5.2, 5.2),
    breaks = c(-5, -2.5, 0, 2.5, 5), labels = c("-5.0", "-2.5", "0.0", "2.5", "5.0")
  ) +
  scale_fill_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("", ""),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(shape = NA)
      )
    ) +
  coord_fixed(ratio = 1) +
  theme_minimal_grid(12) +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )

aligned <- align_plots(p1, p2, p3, align = "hv")
p1 <- aligned[[1]]
p2 <- aligned[[2]]
p3 <- aligned[[3]]

plot_grid(
  p1, NULL, NULL, NULL, NULL,
  nrow = 1, rel_widths = c(1, .04, 1, .04, 1),
  labels = c("", "", "", "", ""), label_y = 0.985
)
```
]

???


Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Principal Components Analysis (PCA)

.center[
```{r blue-jays-PCA2, echo = FALSE, fig.width = 11, fig.asp = 0.35, dev = "svg"}
plot_grid(
  p1, NULL, p2, NULL, NULL,
  nrow = 1, rel_widths = c(1, .04, 1, .04, 1),
  labels = c("", "", "", "", ""), label_y = 0.985
)
```
]

???


Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Principal Components Analysis (PCA)

.center[
```{r blue-jays-PCA3, echo = FALSE, fig.width = 11, fig.asp = 0.35, dev = "svg"}
plot_grid(
  p1, NULL, p2, NULL, p3,
  nrow = 1, rel_widths = c(1, .04, 1, .04, 1),
  labels = c("", "", "", "", ""), label_y = 0.985
)
```
]

--

PCA aligns the major axes with directions of maximum variation in the data

???


Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## PCA analysis of the entire blue jays dataset

.center[
```{r pca-scatter, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
pca_fit <- blue_jays %>% 
  select(where(is.numeric)) %>% # retain only numeric columns
  scale() %>%       # scale to zero mean and unit variance
  prcomp()          # do PCA

pca_fit %>%
  augment(blue_jays) %>% # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 3, stroke = 0.2) +
  scale_x_continuous(
    name = "PC 1", 
    limits = c(-5.2, 5.2),
    breaks = c(-5, -2.5, 0, 2.5, 5),
    labels = c("-5.0", "-2.5", "0.0", "2.5", "5.0")
  ) +
  scale_y_continuous(
    name = "PC 2",
    limits = c(-3.2, 4.2),
    breaks = c(-2, 0, 2, 4),
    labels = c("-2.0", "0.0", "2.0", "4.0")
  ) +
  scale_fill_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("M", "F"),
    labels = c("male birds   ", "female birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(shape = 21)
    )
  ) +
  coord_fixed(ratio = 1) +
  theme_minimal_grid(16) +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

--

.small-font[
Male and female birds separate along PC1
]

---

## The rotation matrix allows us to interpret the PCs

.center[
```{r pca-rotation, fig.width=7., fig.asp = .618, echo=FALSE, message = FALSE, warning=FALSE, dev = "svg"}
arrow_style <- arrow(
  angle = 20, length = grid::unit(8, "pt"),
  ends = "first", type = "closed"
)

nudge_df <- tibble(
  column = c(
    "bill_depth_mm", "bill_width_mm", "bill_length_mm",
    "head_length_mm", "body_mass_g", "skull_size_mm"
  ),
  nudge_y = c(0, 0.03, 0, -0.02, 0, 0),
  name = c(
    "bill depth", "bill width", "bill length",
    "head length", "body mass", "skull size"
  )
)

pca_fit %>%
  # extract rotation matrix
  tidy(matrix = "rotation") %>%
  pivot_wider(
    names_from = "PC", values_from = "value",
    names_prefix = "PC"
  ) %>%
  left_join(nudge_df) %>%
  ggplot(aes(PC1, PC2)) +
  geom_segment(
    xend = 0, yend = 0,
    arrow = arrow_style
  ) +
  geom_text(
    aes(y = PC2 + nudge_y, label = name),
    hjust = 1,
    nudge_x = -0.03,
    size = 14/.pt
    
  ) +
  scale_x_continuous(
    name = "PC 1",
    limits = c(-1.5, .5)
  ) +
  scale_y_continuous(
    name = "PC 2",
    limits = c(-1., 1)
  ) +
  coord_fixed() +
  theme_minimal_grid(16)
```
]

--

.small-font[
All variables contribute negatively to PC1; it represents the overall size of the bird
]

---

## The rotation matrix allows us to interpret the PCs

.center[
![](`r knitr::fig_chunk("pca-rotation", "svg")`)
]

.small-font[
PC2 represents the difference between bill length and skull size
]

---

## The rotation matrix allows us to interpret the PCs

.pull-left.width-50[
![](`r knitr::fig_chunk("pca-scatter", "svg")`)
]

.pull-right.width-50[
![](`r knitr::fig_chunk("pca-rotation", "svg")`)
]

- Male birds are larger than female birds

- Both male and female birds have long and short bills relative to their overall size

---

## We also plot the variance explained by each PC

.center[
```{r pca-eigenvalues, fig.width=7., fig.asp = .618, echo=FALSE, message = FALSE, warning=FALSE, dev = "svg"}
pca_fit %>%
  tidy(matrix = "eigenvalues") %>% # extract eigenvalues
  ggplot(aes(PC, percent)) + 
  geom_col(
    fill = "#56B4E9",
    alpha = 0.9
  ) + 
  scale_x_continuous(
    name = "principal component",
    breaks = 1:6,
    expand = expansion(mult = 0, add = 0.1)
  ) +
  scale_y_continuous(
    name = "variance explained",
    label = scales::label_percent(accuracy = 1),
    expand = expansion(mult = c(0, 0.05))
  ) +
  theme_minimal_hgrid(16)
```
]

.medium-font[
PC 1 captures over 50% of the total variance in the dataset
]

---

## We also plot the variance explained by each PC

.center[
![](`r knitr::fig_chunk("pca-eigenvalues", "svg")`)
]


.medium-font[
Overall bird size explains >50% of the variation in the various measurements
]

[//]: # "segment ends here"

---
class: center middle

## The mathematics of principal components analysis

---

## The mathematics of PCA

Variance of one variable $X$:

$$\text{Var}(X) = \frac{1}{n}\sum_j (\bar x - x_j)^2 = \sigma_X^2$$

--

Covariance of two variables $X$ and $Y$:

$$\text{Cov}(X, Y) = \frac{1}{n}\sum_j (\bar x - x_j)(\bar y - y_j) = \sigma_{XY}$$

---

## The mathematics of PCA

Covariance matrix of $n$ variables $X_1 \dots X_n$:

$$\text{C} = \left(
\begin{array}{cccc}
  \sigma_{11}^2 & \sigma_{12} & \dots & \sigma_{1n}  \\
  \sigma_{21} & \sigma_{22}^2 & \dots & \sigma_{2n} \\
  \vdots & \vdots & \ddots & \vdots \\
  \sigma_{n1} & \sigma_{n2} & \dots & \sigma_{nn}^2  \\
\end{array}
\right)$$

---

## The mathematics of PCA

PCA diagonalizes the covariance matrix $\text{C}$:

$$\begin{eqnarray}
\text{C} & = & \text{U}\text{D}\text{U}^\text{T}\\
& = & \text{U}\left(
\begin{array}{cccc}
  \lambda_{1}^2 & 0 & \dots & 0  \\
  0 & \lambda_{2}^2 & \dots & 0  \\
  \vdots & \vdots & \ddots & \vdots \\
  0 & 0 & \dots & \lambda_{n}^2  \\
\end{array}
\right)\text{U}^\text{T}
\end{eqnarray}$$

--

$\text{U}$: rotation matrix

--

$\text{D}$: diagonal matrix

--

$\lambda_j^2$: eigenvalues (= variance explained by each component)

---

## The mathematics of PCA

PCA diagonalizes the covariance matrix $\text{C}$:

$$\begin{eqnarray}
\text{C} & = & \text{U}\text{D}\text{U}^\text{T}\\
& = & \text{U}\left(
\begin{array}{cccc}
  \lambda_{1}^2 & 0 & \dots & 0  \\
  0 & \lambda_{2}^2 & \dots & 0  \\
  \vdots & \vdots & \ddots & \vdots \\
  0 & 0 & \dots & \lambda_{n}^2  \\
\end{array}
\right)\text{U}^\text{T}
\end{eqnarray}$$

The covariances between components are all 0

Components are uncorrelated

---

## The components are uncorrelated

.center[
![](`r knitr::fig_chunk("pca-scatter", "svg")`)
]

[//]: # "segment ends here"

---
class: center middle

## Doing a PCA in R

---

## Prerequisite: Scaling of numeric columns

.pull-left.width-50[

.medium-font[
Plot without scaling
]

.tiny-font[
```{r blue-jays-unscaled, eval = FALSE}
blue_jays %>% 
  ggplot() +
  aes(skull_size_mm, head_length_mm) + 
  geom_point(aes(color = sex))
```
]]

.pull-right[
```{r blue-jays-unscaled-out, ref.label="blue-jays-unscaled", fig.width = 5, fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]


---

## Prerequisite: Scaling of numeric columns

.pull-left.width-50[

.medium-font[
Plot with scaling
]

.tiny-font[
```{r blue-jays-scaled, eval = FALSE}
blue_jays %>% 
  # scale all numeric columns
  mutate(across(where(is.numeric), scale)) %>%
  ggplot() +
  aes(skull_size_mm, head_length_mm) + 
  geom_point(aes(color = sex))
```
]]

.pull-right[
```{r blue-jays-scaled-out, ref.label="blue-jays-scaled", fig.width = 5, fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

---

## We perform a PCA with `prcomp()`

.tiny-font[
```{r}
blue_jays %>% 
  select(where(is.numeric)) %>% # retain only numeric columns
  scale() %>%                   # scale to zero mean and unit variance
  prcomp()                      # do PCA
```
]


---

## Calculate the components and plot

First we run the PCA and store results as `pca_fit`:

.tiny-font[
```{r}
pca_fit <- blue_jays %>% 
  select(where(is.numeric)) %>% # retain only numeric columns
  scale() %>%                   # scale to zero mean and unit variance
  prcomp()                      # do PCA
```
]

--

Then we add PC coordinates into original dataset and plot:

.tiny-font[
```{r eval = FALSE}
library(broom)  # for augment(), tidy()

pca_fit %>%
  # add PCs to the original dataset
  augment(blue_jays) %>%
  ggplot(aes(.fittedPC1, .fittedPC2)) +
  geom_point(aes(color = sex))
```
]

---


## Calculate the components and plot

.pull-left[
.medium-font[
Plot PC 2 against PC 1
]

.tiny-font[
```{r blue-jays-pca-plot, eval = FALSE}
pca_fit %>%
  # add PCs to the original dataset
  augment(blue_jays) %>%
  ggplot(aes(.fittedPC1, .fittedPC2)) +
  geom_point(aes(color = sex))
```
]]

.pull-right[
```{r blue-jays-pca-plot-out, ref.label="blue-jays-pca-plot", fig.width=5., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

---

## Calculate the components and plot

.pull-left[
.medium-font[
Plot PC 3 against PC 2
]


.tiny-font[
```{r blue-jays-pca-plot2, eval = FALSE}
pca_fit %>%
  # add PCs to the original dataset
  augment(blue_jays) %>%
  ggplot(aes(.fittedPC2, .fittedPC3)) +
  geom_point(aes(color = sex))
```
]]

.pull-right[
```{r blue-jays-pca-plot2-out, ref.label="blue-jays-pca-plot2", fig.width=5., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

---

## Plot the rotation matrix

.tiny-font.pull-left.width-50[
```{r blue-jays-pca-rotation, eval = FALSE}
arrow_style <- arrow(
  angle = 20, length = grid::unit(8, "pt"),
  ends = "first", type = "closed"
)

pca_fit %>%
  # extract rotation matrix
  tidy(matrix = "rotation") %>%
  pivot_wider(
    names_from = "PC", values_from = "value",
    names_prefix = "PC"
  ) %>%
  ggplot(aes(PC1, PC2)) +
  geom_segment(
    xend = 0, yend = 0,
    arrow = arrow_style
  ) +
  geom_text(aes(label = column), hjust = 1) +
  xlim(-1.5, 0.5) + ylim(-1, 1) + 
  coord_fixed()
```
]
.pull-right.width-50[
```{r blue-jays-pca-rotation-out, ref.label="blue-jays-pca-rotation", fig.width=5., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]


---

## Plot the variance explained

.pull-left.tiny-font.width-50[
```{r blue-jays-pca-eigenvalues, eval = FALSE}
pca_fit %>%
  # extract eigenvalues
  tidy(matrix = "eigenvalues") %>%
  ggplot(aes(PC, percent)) + 
  geom_col() + 
  scale_x_continuous(
    # create one axis tick per PC
    breaks = 1:6
  ) +
  scale_y_continuous(
    name = "variance explained",
    # format y axis ticks as percent values
    label = scales::label_percent(accuracy = 1)
  )
```
]

.pull-right[
```{r blue-jays-pca-eigenvalues-out, ref.label="blue-jays-pca-eigenvalues", fig.width=5., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

[//]: # "segment ends here"


---

## Further reading

- Fundamentals of Data Visualization: [Chapter 12.3: Dimension reduction](https://clauswilke.com/dataviz/visualizing-associations.html#dimension-reduction)
- Blogpost on PCA: [Principal Component Analysis
Explained Visually](https://setosa.io/ev/principal-component-analysis/)
- PCA tutorial with mathematical background: [A Tutorial on Principal Component Analysis](https://arxiv.org/pdf/1404.1100.pdf)

