---
title: "Figure design"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(cowplot)
library(colorspace)
library(ggridges)
library(here)
# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width

lincoln_temps <- readRDS(here("datasets", "lincoln_temps.rds"))
```


.pull-left[
.title-font[How do you go from this ...]

.center[
```{r lincoln-ridgeline-raw, echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
ggplot(lincoln_temps, aes(x = mean_temp, y = month_long)) +
  geom_density_ridges()
```
]]

--

.pull-right[
.title-font[... to this?]


.center[
```{r lincoln-ridgeline-polished, echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}

ggplot(lincoln_temps, aes(x = mean_temp, y = month_long)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#56B4E9", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)#, breaks = c(0, 25, 50, 75)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(vjust = 0)
  )
```
]]

--

.small-font[
Requires coordinated modification of multiple elements:
]

--

.small-font[
- geoms (via arguments to geoms)
- scales (via `scale_*()` functions)
- plot appearance (via themes)
]

---

## The starting point, a rough draft

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression1, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges()
```
]

.pull-right[
```{r lincoln-ridgeline-progression1-out, ref.label = "lincoln-ridgeline-progression1", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Set `scale` and `bandwidth` to shape ridgelines

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression2, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4
  )
```
]

.pull-right[
```{r lincoln-ridgeline-progression2-out, ref.label = "lincoln-ridgeline-progression2", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Set `rel_min_height` to cut ridgelines near zero

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression3, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01
  )
```
]

.pull-right[
```{r lincoln-ridgeline-progression3-out, ref.label = "lincoln-ridgeline-progression3", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Use `scale_*()` functions to specify axis labels

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression4, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01,
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)"
  ) +
  scale_y_discrete(
    name = NULL  # NULL means no label
  )
```
]

.pull-right[
```{r lincoln-ridgeline-progression4-out, ref.label = "lincoln-ridgeline-progression4", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Specify scale expansion

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression5, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  )
```
]

.pull-right[
```{r lincoln-ridgeline-progression5-out, ref.label = "lincoln-ridgeline-progression5", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Set overall plot theme

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression6, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_minimal_grid()  # from cowplot
```
]

.pull-right[
```{r lincoln-ridgeline-progression6-out, ref.label = "lincoln-ridgeline-progression6", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Align y axis labels to grid lines

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression7, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(vjust = 0)
  )
```
]

.pull-right[
```{r lincoln-ridgeline-progression7-out, ref.label = "lincoln-ridgeline-progression7", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Change fill color from default gray to blue

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression8, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01,
    fill = "#7DCCFF"
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(vjust = 0)
  )
```
]

.pull-right[
```{r lincoln-ridgeline-progression8-out, ref.label = "lincoln-ridgeline-progression8", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Draw lines in white instead of black

.pull-left.tiny-font[
```{r lincoln-ridgeline-final, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01,
    fill = "#7DCCFF",
    color = "white"
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(vjust = 0)
  )
```
]

.pull-right[
```{r lincoln-ridgeline-final-out, ref.label = "lincoln-ridgeline-final", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]



---
## Further reading

- Fundamentals of Data Visualization: [Chapter 23: Balance the data and the context](https://clauswilke.com/dataviz/balance-data-context.html)
- Data Visualization—A Practical Introduction: [4.6 Histograms and density plots](https://socviz.co/refineplots.html#change-the-appearance-of-plots-with-themes)
- ggplot2 reference documentation: [Complete themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- ggplot2 reference documentation: [Modify components of a theme](https://ggplot2.tidyverse.org/reference/theme.html)




