---
title: "Figure design"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(cowplot)
library(ggthemes)
library(colorspace)
library(ggridges)
library(palmerpenguins)
library(here)
# ggplot2 settings so plots scale well for slide size 
#theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width

lincoln_temps <- readRDS(here("datasets", "lincoln_temps.rds"))
```


.pull-left[
.title-font[How do you go from this ...]

.center[
```{r lincoln-ridgeline-raw, echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
ggplot(lincoln_temps, aes(x = mean_temp, y = month_long)) +
  geom_density_ridges()
```
]]

--

.pull-right[
.title-font[... to this?]


.center[
```{r lincoln-ridgeline-polished, echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}

ggplot(lincoln_temps, aes(x = mean_temp, y = month_long)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#56B4E9", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)#, breaks = c(0, 25, 50, 75)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(vjust = 0)
  )
```
]]

--

.small-font[
Requires coordinated modification of multiple elements:
]

--

.small-font[
- geoms (via arguments to geoms)
- scales (via `scale_*()` functions)
- plot appearance (via themes)
]

---

## The starting point, a rough draft

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression1, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges()
```
]

.pull-right[
```{r lincoln-ridgeline-progression1-out, ref.label = "lincoln-ridgeline-progression1", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

--

.tiny-font[
You can download the dataset using this code:
```{r eval = FALSE}
lincoln_temps <- readRDS(
  url("https://wilkelab.org/DSC385/datasets/lincoln_temps.rds")
)
```
]

---

## Set `scale` and `bandwidth` to shape ridgelines

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression2, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4
  )
```
]

.pull-right[
```{r lincoln-ridgeline-progression2-out, ref.label = "lincoln-ridgeline-progression2", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Set `rel_min_height` to cut ridgelines near zero

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression3, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01
  )
```
]

.pull-right[
```{r lincoln-ridgeline-progression3-out, ref.label = "lincoln-ridgeline-progression3", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Use `scale_*()` functions to specify axis labels

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression4, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01,
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)"
  ) +
  scale_y_discrete(
    name = NULL  # NULL means no label
  )
```
]

.pull-right[
```{r lincoln-ridgeline-progression4-out, ref.label = "lincoln-ridgeline-progression4", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Specify scale expansion

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression5, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  )
```
]

.pull-right[
```{r lincoln-ridgeline-progression5-out, ref.label = "lincoln-ridgeline-progression5", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Set overall plot theme

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression6, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_minimal_grid()  # from cowplot
```
]

.pull-right[
```{r lincoln-ridgeline-progression6-out, ref.label = "lincoln-ridgeline-progression6", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Align y axis labels to grid lines

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression7, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(vjust = 0)
  )
```
]

.pull-right[
```{r lincoln-ridgeline-progression7-out, ref.label = "lincoln-ridgeline-progression7", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Change fill color from default gray to blue

.pull-left.tiny-font[
```{r lincoln-ridgeline-progression8, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01,
    fill = "#7DCCFF"
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(vjust = 0)
  )
```
]

.pull-right[
```{r lincoln-ridgeline-progression8-out, ref.label = "lincoln-ridgeline-progression8", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Draw lines in white instead of black

.pull-left.tiny-font[
```{r lincoln-ridgeline-final, eval = FALSE}
ggplot(lincoln_temps) +
  aes(x = mean_temp, y = month_long) +
  geom_density_ridges(
    scale = 3, bandwidth = 3.4,
    rel_min_height = 0.01,
    fill = "#7DCCFF",
    color = "white"
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_discrete(
    name = NULL,
    expand = expansion(add = c(0.2, 2.6))
  ) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(vjust = 0)
  )
```
]

.pull-right[
```{r lincoln-ridgeline-final-out, ref.label = "lincoln-ridgeline-final", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


[//]: # "segment ends here"


---
class: center middle

## Working with ggplot themes

---

## Using ready-made themes

.tiny-font[
```{r penguins-complete-themes, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point()
  # default theme is theme_gray()
```
]

.center[
```{r penguins-complete-themes-out, ref.label = "penguins-complete-themes", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Using ready-made themes

.tiny-font[
```{r penguins-complete-themes2, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  theme_gray()
```
]

.center[
```{r penguins-complete-themes2-out, ref.label = "penguins-complete-themes2", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Using ready-made themes

.tiny-font[
```{r penguins-complete-themes3, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  theme_gray(14) # most themes take a font-size argument to scale text size
```
]

.center[
```{r penguins-complete-themes3-out, ref.label = "penguins-complete-themes3", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Using ready-made themes

.tiny-font[
```{r penguins-complete-themes4, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  theme_bw(14)
```
]

.center[
```{r penguins-complete-themes4-out, ref.label = "penguins-complete-themes4", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Using ready-made themes

.tiny-font[
```{r penguins-complete-themes5, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  theme_minimal(14)
```
]

.center[
```{r penguins-complete-themes5-out, ref.label = "penguins-complete-themes5", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Using ready-made themes

.tiny-font[
```{r penguins-complete-themes6, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  theme_classic(14)
```
]

.center[
```{r penguins-complete-themes6-out, ref.label = "penguins-complete-themes6", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Using ready-made themes

.tiny-font[
```{r penguins-complete-themes7, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  theme_half_open()  # from package cowplot
```
]

.center[
```{r penguins-complete-themes7-out, ref.label = "penguins-complete-themes7", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Using ready-made themes

.tiny-font[
```{r penguins-complete-themes8, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  theme_minimal_grid()  # from package cowplot
```
]

.center[
```{r penguins-complete-themes8-out, ref.label = "penguins-complete-themes8", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Using ready-made themes

.tiny-font[
```{r penguins-complete-themes9, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  theme_minimal_hgrid()  # from package cowplot
```
]

.center[
```{r penguins-complete-themes9-out, ref.label = "penguins-complete-themes9", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Using ready-made themes

.tiny-font[
```{r penguins-complete-themes10, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  theme_minimal_vgrid()  # from package cowplot
```
]

.center[
```{r penguins-complete-themes10-out, ref.label = "penguins-complete-themes10", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Using ready-made themes


.tiny-font[
```{r penguins-complete-themes11, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  theme_economist(14)       # from package ggthemes
```
]

.center[
```{r penguins-complete-themes11-out, ref.label = "penguins-complete-themes11", echo = FALSE, warning = FALSE, fig.width = 4.944, fig.asp = 0.75, dev = "svg"}
```
]



---

## Using ready-made themes

.tiny-font[
```{r penguins-complete-themes12, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  theme_economist(14) + scale_color_economist() # from package ggthemes
```
]

.center[
```{r penguins-complete-themes12-out, ref.label = "penguins-complete-themes12", echo = FALSE, warning = FALSE, fig.width = 4.944, fig.asp = 0.75, dev = "svg"}
```
]


---

## Using ready-made themes

.tiny-font[
```{r penguins-complete-themes13, eval = FALSE}
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point() +
  theme_fivethirtyeight(14) + scale_color_fivethirtyeight() # from package ggthemes
```
]

.center[
```{r penguins-complete-themes13-out, ref.label = "penguins-complete-themes13", echo = FALSE, warning = FALSE, fig.width = 4.944, fig.asp = 0.75, dev = "svg"}
```
]


---

## Customizing theme elements

.pull-left.tiny-font[
```{r penguins-theme-elements1, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid()
```
]

.pull-right[
```{r penguins-theme-elements1-out, ref.label = "penguins-theme-elements1", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Customizing theme elements

.pull-left.tiny-font[
```{r penguins-theme-elements2, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    # change overall font family
    # (requires font to be available)
    text = element_text(
      family = "Comic Sans MS"
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements2-out, ref.label = "penguins-theme-elements2", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Customizing theme elements

.pull-left.tiny-font[
```{r penguins-theme-elements3, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    # change color of axis titles
    axis.title = element_text(
      color = "royalblue2"
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements3-out, ref.label = "penguins-theme-elements3", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Customizing theme elements

.pull-left.tiny-font[
```{r penguins-theme-elements4, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    # change color of only the x axis title
    axis.title.x = element_text(
      color = "royalblue2"
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements4-out, ref.label = "penguins-theme-elements4", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Customizing theme elements

.pull-left.tiny-font[
```{r penguins-theme-elements5, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    # change all text colors?
    # why does it not work?
    text = element_text(color = "royalblue2")
  )
```
]

.pull-right[
```{r penguins-theme-elements5-out, ref.label = "penguins-theme-elements5", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Customizing theme elements

.pull-left.tiny-font[
```{r penguins-theme-elements6, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    text = element_text(color = "royalblue2"),
    axis.text = element_text(
      color = "royalblue2"
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements6-out, ref.label = "penguins-theme-elements6", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

--

.small-font[
The element `axis.text` has its own color set in the theme. Therefore it doesn't inherit from `text`.
]

---

## Horizontal and vertical alignment

.pull-left.tiny-font[
```{r penguins-theme-elements7, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    axis.title.x = element_text(
      # horizontal justification
      # (0 = left)
      hjust = 0
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements7-out, ref.label = "penguins-theme-elements7", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Horizontal and vertical alignment

.pull-left.tiny-font[
```{r penguins-theme-elements8, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    axis.title.x = element_text(
      # horizontal justification
      # (0.5 = center)
      hjust = 0.5
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements8-out, ref.label = "penguins-theme-elements8", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Horizontal and vertical alignment

.pull-left.tiny-font[
```{r penguins-theme-elements9, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    axis.title.x = element_text(
      # horizontal justification
      # (1 = right)
      hjust = 1
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements9-out, ref.label = "penguins-theme-elements9", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Horizontal and vertical alignment

.pull-left.tiny-font[
```{r penguins-theme-elements10, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(
      # vertical justification
      # (0 = bottom)
      vjust = 0
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements10-out, ref.label = "penguins-theme-elements10", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Horizontal and vertical alignment

.pull-left.tiny-font[
```{r penguins-theme-elements11, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(
      # vertical justification
      # (0.5 = center)
      vjust = 0.5
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements11-out, ref.label = "penguins-theme-elements11", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Horizontal and vertical alignment

.pull-left.tiny-font[
```{r penguins-theme-elements12, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    axis.text.y = element_text(
      # vertical justification
      # (1 = top)
      vjust = 1
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements12-out, ref.label = "penguins-theme-elements12", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Remove elements entirely: `element_blank()`

.pull-left.tiny-font[
```{r penguins-theme-elements13, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    # all text gone
    text = element_blank()
  )
```
]

.pull-right[
```{r penguins-theme-elements13-out, ref.label = "penguins-theme-elements13", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Remove elements entirely: `element_blank()`

.pull-left.tiny-font[
```{r penguins-theme-elements14, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    # no axis titles
    axis.title = element_blank()
  )
```
]

.pull-right[
```{r penguins-theme-elements14-out, ref.label = "penguins-theme-elements14", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Set background color: `element_rect()`

.pull-left.tiny-font[
```{r penguins-theme-elements15, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    plot.background = element_rect(
      fill = "aliceblue"
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements15-out, ref.label = "penguins-theme-elements15", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Set background color: `element_rect()`

.pull-left.tiny-font[
```{r penguins-theme-elements16, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    panel.background = element_rect(
      fill = "aliceblue"
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements16-out, ref.label = "penguins-theme-elements16", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Set background color: `element_rect()`

.pull-left.tiny-font[
```{r penguins-theme-elements17, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    legend.box.background = element_rect(
      fill = "aliceblue",
      color = "steelblue4" # line color
    )
  )
```
]

.pull-right[
```{r penguins-theme-elements17-out, ref.label = "penguins-theme-elements17", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Set background color: `element_rect()`

.pull-left.tiny-font[
```{r penguins-theme-elements18, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    legend.box.background = element_rect(
      fill = "aliceblue",
      color = "steelblue4" # line color
    ),
    legend.box.margin = margin(7, 7, 7, 7)
  )
```
]

.pull-right[
```{r penguins-theme-elements18-out, ref.label = "penguins-theme-elements18", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Move the legend: `legend.position`

.pull-left.tiny-font[
```{r penguins-theme-elements19, eval = FALSE}
ggplot(penguins) +
  aes(flipper_length_mm, body_mass_g) +
  geom_point(aes(color = species)) +
  theme_minimal_grid() +
  theme(
    legend.box.background = element_rect(
      fill = "aliceblue",
      color = "steelblue4" # line color
    ),
    legend.box.margin = margin(7, 7, 7, 7),
    # relative position inside plot panel
    legend.position = c(1, 0),
    # justification relative to position
    legend.justification = c(1, 0)
  )
```
]

.pull-right[
```{r penguins-theme-elements19-out, ref.label = "penguins-theme-elements19", echo = FALSE, warning = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

[//]: # "segment ends here"

---
## Further reading

- Fundamentals of Data Visualization: [Chapter 23: Balance the data and the context](https://clauswilke.com/dataviz/balance-data-context.html)
- Data Visualization—A Practical Introduction: [Chapter 8.3: Change the appearance of plots with themes](https://socviz.co/refineplots.html#change-the-appearance-of-plots-with-themes)
- ggplot2 reference documentation: [Complete themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- ggplot2 reference documentation: [Modify components of a theme](https://ggplot2.tidyverse.org/reference/theme.html)




