---
title: "Functions and functional programming"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
params:
  cache: TRUE
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(here)
library(palmerpenguins)
library(cowplot)
library(glue)
# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width
```

## We often have to run similar code multiple times

--

.tiny-font.pull-left.width-50[
```{r penguins-plot-Adelie, eval = FALSE}
penguins %>%
  filter(species == "Adelie") %>%
  ggplot() +
  aes(bill_length_mm, body_mass_g) +
  geom_point() +
  ggtitle("Species: Adelie") +
  xlab("bill length (mm)") +
  ylab("body mass (g)") +
  theme_minimal_grid() +
  theme(plot.title.position = "plot")
```
]

--

.pull-right.width-45.move-up-1em[
```{r penguins-plot-Adelie-out, ref.label = "penguins-plot-Adelie", cache = params$cache, echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}

```
]

---

## We often have to run similar code multiple times


.tiny-font.pull-left.width-50[
```{r penguins-plot-Chinstrap, eval = FALSE}
penguins %>%
  filter(species == "Chinstrap") %>%
  ggplot() +
  aes(bill_length_mm, body_mass_g) +
  geom_point() +
  ggtitle("Species: Chinstrap") +
  xlab("bill length (mm)") +
  ylab("body mass (g)") +
  theme_minimal_grid() +
  theme(plot.title.position = "plot")
```
]

.pull-right.width-45.move-up-1em[
```{r penguins-plot-Chinstrap-out, ref.label = "penguins-plot-Chinstrap", cache = params$cache, echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}

```
]

---

## We often have to run similar code multiple times


.tiny-font.pull-left.width-50[
```{r penguins-plot-Gentoo, eval = FALSE}
penguins %>%
  filter(species == "Gentoo") %>%
  ggplot() +
  aes(bill_length_mm, body_mass_g) +
  geom_point() +
  ggtitle("Species: Gentoo") +
  xlab("bill length (mm)") +
  ylab("body mass (g)") +
  theme_minimal_grid() +
  theme(plot.title.position = "plot")
```
]

.pull-right.width-45.move-up-1em[
```{r penguins-plot-Gentoo-out, ref.label = "penguins-plot-Gentoo", cache = params$cache, echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}

```
]

--

How can we make our life simpler and avoid massive code duplication?

---

## Step 1: Avoid hard-coding specific values

--

.tiny-font.pull-left.width-50[
```{r penguins-plot-Adelie2, eval = FALSE}
species <- "Adelie" #<<

penguins %>%
  filter(.data$species == .env$species) %>% #<<
  ggplot() +
  aes(bill_length_mm, body_mass_g) +
  geom_point() +
  ggtitle(glue("Species: {species}")) + #<<
  xlab("bill length (mm)") +
  ylab("body mass (g)") +
  theme_minimal_grid() +
  theme(plot.title.position = "plot")
```
]

.pull-right.width-45.move-up-1em[
```{r penguins-plot-Adelie2-out, ref.label = "penguins-plot-Adelie2", cache = params$cache, echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}

```
]


---

## A quick aside: the pronouns `.data` and `.env`

We can use pronouns to distinguish data columns from variables:
```r
species <- "Adelie"

penguins %>%
  filter(.data$species == .env$species)
```

--

`.data$species` is a column in the data frame

--

`.env$species` is a variable in the local environment



---

## A quick aside: the pronouns `.data` and `.env`

Alternatively we would have to make sure the names don't clash:

```r
species_choice <- "Adelie"

penguins %>%
  filter(species == species_choice)
```


---

## Step 1: Avoid hard-coding specific values

.tiny-font.pull-left.width-50[
```{r penguins-plot-Adelie2a, eval = FALSE}
species <- "Adelie" #<<

penguins %>%
  filter(.data$species == .env$species) %>%
  ggplot() +
  aes(bill_length_mm, body_mass_g) +
  geom_point() +
  ggtitle(glue("Species: {species}")) +
  xlab("bill length (mm)") +
  ylab("body mass (g)") +
  theme_minimal_grid() +
  theme(plot.title.position = "plot")
```
]

.pull-right.width-45.move-up-1em[
![](`r knitr::fig_chunk("penguins-plot-Adelie2-out", "svg")`)
]


---

## Step 1: Avoid hard-coding specific values

.tiny-font.pull-left.width-50[
```{r penguins-plot-Chinstrap2, eval = FALSE}
species <- "Chinstrap" #<<

penguins %>%
  filter(.data$species == .env$species) %>%
  ggplot() +
  aes(bill_length_mm, body_mass_g) +
  geom_point() +
  ggtitle(glue("Species: {species}")) +
  xlab("bill length (mm)") +
  ylab("body mass (g)") +
  theme_minimal_grid() +
  theme(plot.title.position = "plot")
```
]

.pull-right.width-45.move-up-1em[
```{r penguins-plot-Chinstrap2-out, ref.label = "penguins-plot-Chinstrap2", cache = params$cache, echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}

```
]

---

## Step 1: Avoid hard-coding specific values

.tiny-font.pull-left.width-50[
```{r penguins-plot-Gentoo2, eval = FALSE}
species <- "Gentoo" #<<

penguins %>%
  filter(.data$species == .env$species) %>%
  ggplot() +
  aes(bill_length_mm, body_mass_g) +
  geom_point() +
  ggtitle(glue("Species: {species}")) +
  xlab("bill length (mm)") +
  ylab("body mass (g)") +
  theme_minimal_grid() +
  theme(plot.title.position = "plot")
```
]

.pull-right.width-45.move-up-1em[
```{r penguins-plot-Gentoo2-out, ref.label = "penguins-plot-Gentoo2", cache = params$cache, echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}

```
]

--

This concept is also called: .highlight[Avoiding magic numbers]


---

## Step 2: Define a function

.tiny-font.pull-left.width-50[
```{r eval = FALSE}
make_plot <- function(species) { #<<
  penguins %>%
    filter(.data$species == .env$species) %>%
    ggplot() +
    aes(bill_length_mm, body_mass_g) +
    geom_point() +
    ggtitle(glue("Species: {species}")) +
    xlab("bill length (mm)") +
    ylab("body mass (g)") +
    theme_minimal_grid() +
    theme(plot.title.position = "plot")
}
```
]
---

## Step 2: Define a function

.tiny-font.pull-left.width-50[
```{r penguins-plot-Adelie3, eval = FALSE}
make_plot <- function(species) {
  penguins %>%
    filter(.data$species == .env$species) %>%
    ggplot() +
    aes(bill_length_mm, body_mass_g) +
    geom_point() +
    ggtitle(glue("Species: {species}")) +
    xlab("bill length (mm)") +
    ylab("body mass (g)") +
    theme_minimal_grid() +
    theme(plot.title.position = "plot")
}

make_plot("Adelie") #<<
```
]

.pull-right.width-45.move-up-1em[
```{r penguins-plot-Adelie3-out, ref.label = "penguins-plot-Adelie3", cache = params$cache, echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}

```
]

---

## Step 2: Define a function

.small-font[
```{r penguins-plot-all, eval = FALSE}
make_plot("Adelie")
make_plot("Chinstrap")
make_plot("Gentoo")
```
]

```{r penguins-plot-all-out, ref.label = "penguins-plot-all", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, out.width = "32%", dev = "svg"}

```

---

## Rules of thumb about functions

--

- You can never write too many functions

--

- When you find yourself writing the same code 2-3 times, put it into a function

--

- A function should be no longer than 20-40 lines

--

- If a function is getting too long, break it into smaller functions


---

## Step 3: Automate calling the function

.small-font[
```{r}
species <- c("Adelie", "Chinstrap", "Gentoo")
plots <- map(species, make_plot)
```
]

--

`map` takes each element of the vector `species` and uses it as input for `make_plot()`

--

It returns a list of created plots:

.small-font[
```{r penguins-plot-map-return, warning = FALSE, fig.width = 6, fig.asp = 0.75, out.width = "32%", dev = "svg"}
plots[[1]]
```
]

---

## Step 3: Automate calling the function

.small-font[
```{r eval = FALSE}
species <- c("Adelie", "Chinstrap", "Gentoo")
plots <- map(species, make_plot)
```
]

`map` takes each element of the vector `species` and uses it as input for `make_plot()`

It returns a list of created plots:

.small-font[
```{r penguins-plot-map-return2, warning = FALSE, fig.width = 6, fig.asp = 0.75, out.width = "32%", dev = "svg"}
plots[[2]]
```
]

---

## Step 3: Automate calling the function

.small-font[
```{r eval = FALSE}
species <- c("Adelie", "Chinstrap", "Gentoo")
plots <- map(species, make_plot)
```
]

`map` takes each element of the vector `species` and uses it as input for `make_plot()`

It returns a list of created plots:

.small-font[
```{r penguins-plot-map-return3, warning = FALSE, fig.width = 6, fig.asp = 0.75, out.width = "32%", dev = "svg"}
plots[[3]]
```
]


---

## Step 3: Automate calling the function

.small-font[
```{r penguins-plot-map, eval = FALSE}
species <- c("Adelie", "Chinstrap", "Gentoo")
plots <- map(species, make_plot)

plots[[1]]
plots[[2]]
plots[[3]]
```
]

```{r penguins-plot-map-out, ref.label = "penguins-plot-map", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, out.width = "32%", dev = "svg"}
```

---

## Step 3: Automate calling the function

.small-font[
```{r penguins-plot-walk, eval = FALSE}
species <- c("Adelie", "Chinstrap", "Gentoo")
plots <- map(species, make_plot)

# `walk()` is like `map()` but doesn't return a value
# we use it only for side effects (such as printing)
walk(plots, print)
```
]

```{r penguins-plot-walk-out, ref.label = "penguins-plot-walk", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, out.width = "32%", dev = "svg"}
```

---

## Step 4: Write a more general function

--

.tiny-font.pull-left.width-50[
```{r eval = FALSE}
make_plot <- function(species) {
  penguins %>% # hard-coded dataset! #<<
    filter(.data$species == .env$species) %>%
    ggplot() +
    aes(bill_length_mm, body_mass_g) +
    geom_point() +
    ggtitle(glue("Species: {species}")) +
    xlab("bill length (mm)") +
    ylab("body mass (g)") +
    theme_minimal_grid() +
    theme(plot.title.position = "plot")
}
```
]

---

## Step 4: Write a more general function

.tiny-font.pull-left.width-50[
```{r eval = FALSE}
make_plot2 <- function(data, species) { #<<
  data %>% #<<
    # filter no longer needed
    ggplot() +
    aes(bill_length_mm, body_mass_g) +
    geom_point() +
    ggtitle(glue("Species: {species}")) +
    xlab("bill length (mm)") +
    ylab("body mass (g)") +
    theme_minimal_grid() +
    theme(plot.title.position = "plot")
}
```
]

---

## Step 4: Write a more general function

.tiny-font.pull-left.width-50[
```{r penguins-plot-generic, eval = FALSE}
make_plot2 <- function(data, species) {
  data %>%
    # filter no longer needed
    ggplot() +
    aes(bill_length_mm, body_mass_g) +
    geom_point() +
    ggtitle(glue("Species: {species}")) +
    xlab("bill length (mm)") +
    ylab("body mass (g)") +
    theme_minimal_grid() +
    theme(plot.title.position = "plot")
}

data_adelie <- penguins %>%
  filter(species == "Adelie")

make_plot2(data_adelie, species = "Adelie")
```
]

.pull-right.width-45[
```{r penguins-plot-generic-out, ref.label = "penguins-plot-generic", cache = params$cache, echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}

```
]


---

## Step 5: Use these concepts in a tidy pipeline

.tiny-font[
```{r}
penguins %>%
  nest(data = -species) %>%
  mutate(plots = map(species, make_plot))
```
]


---

## Step 5: Use these concepts in a tidy pipeline

.tiny-font[
```{r penguins-plot-walk-tidy, eval = FALSE}
penguins %>%
  nest(data = -species) %>%
  mutate(plots = map(species, make_plot)) %>%
  pull(plots) %>%
  walk(print)
```
]

```{r penguins-plot-walk-tidy-out, ref.label = "penguins-plot-walk-tidy", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, out.width = "32%", dev = "svg"}
```

---

## Step 5: Use these concepts in a tidy pipeline

.tiny-font[
```{r penguins-plot-walk-tidy2, eval = FALSE}
penguins %>%
  nest(data = -species) %>%
  mutate(plots = map2(data, species, make_plot2)) %>%
  pull(plots) %>%
  walk(print)
```
]

```{r penguins-plot-walk-tidy2-out, ref.label = "penguins-plot-walk-tidy2", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, out.width = "32%", dev = "svg"}
```

--

.small-font[
`map2()` is like `map()` but for functions with 2 arguments
]

---

## Step 5: Use these concepts in a tidy pipeline

.tiny-font[
```{r penguins-plot-walk-tidy3, eval = FALSE}
penguins %>%
  nest(data = -species) %>%
  mutate(plots = map2(data, species, make_plot2)) %>%
  pull(plots) %>%
  walk(print)
```
]

```{r penguins-plot-walk-tidy3-out, ref.label = "penguins-plot-walk-tidy3", echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.75, out.width = "32%", dev = "svg"}
```

.small-font[
Note: This pipeline automatically processes all species in the dataset, whatever they are called
]

---

## Why no `for` loops?

--

- They often require us to think about data logistics (indexing)

--

- They encourage writing long, monolithic blocks of code

--

- They encourage iterative thinking over conceptual thinking

--

- They cannot easily be parallelized or otherwise optimized

--

- Most modern programming languages are moving away from `for` loops<br>
  (examples: Python, Rust, JavaScript)

[//]: # "segment ends here"

---

## Further reading

- R for Data Science: [Chapter 19: Functions](https://r4ds.had.co.nz/functions.html)
- R for Data Science: [Chapter 21.5: The map functions](https://r4ds.had.co.nz/iteration.html#the-map-functions)
- **purrr** reference documentation: [Apply a function to each element of a list or atomic vector](https://purrr.tidyverse.org/reference/map.html)

