---
title: "Getting things into the right order"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(palmerpenguins)
# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width
```


## Remember from "Visualizing amounts"

.tiny-font[
```{r penguins-ordered, eval = FALSE}
ggplot(penguins, aes(y = fct_relevel(species, "Chinstrap", "Gentoo", "Adelie"))) +
  geom_bar()
```
]

.center[
```{r penguins-ordered-out, ref.label = "penguins-ordered", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

---

## Title

.tiny-font[
```{r penguins-ordered-pipe, eval = FALSE}
penguins %>%
  mutate(species = fct_relevel(species, "Chinstrap", "Gentoo", "Adelie")) %>%
  ggplot(aes(y = species)) +
  geom_bar()
```
]

.center[
```{r penguins-ordered-pipe-out, ref.label = "penguins-ordered-pipe", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

---

## Title

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_relevel(species)) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```
]


---

## Title

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_relevel(species, "Gentoo")) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```
]


---

## Title

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_relevel(species, "Chinstrap", "Gentoo")) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```

]

---

## Title

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_relevel(species, "Chinstrap", "Adelie", "Gentoo")) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```

]

---

## Title

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_relevel(species, "Gentoo", "Chinstrap", "Adelie")) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```

]

---

## Title

.tiny-font[
```{r eval = FALSE}
penguins %>%
  count(species)
```

```{r echo = FALSE}
# remove row names
x <- penguins %>%
  count(species)

row.names(x) <- NULL
x
```


]

---

## Title

.tiny-font[
```{r}
penguins %>%
  count(species) %>%
  mutate(species = fct_reorder(species, n)) %>%
  pull(species)     # pull out just the `species` column
```

]

---

## Title

.tiny-font[
```{r penguins-ordered-pipe2, eval = FALSE}
penguins %>%
  count(species) %>%
  mutate(species = fct_reorder(species, n)) %>%
  ggplot(aes(n, species)) + geom_col()
```

]

.center[
```{r penguins-ordered-pipe2-out, ref.label = "penguins-ordered-pipe2", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]


---

## Title

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_reorder(species, species, length)) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```

]


---

## Title

.tiny-font[
```{r penguins-ordered-pipe3, eval = FALSE}
penguins %>%
  mutate(species = fct_reorder(species, species, length)) %>%
  ggplot(aes(y = species)) + geom_bar()
```

]

.center[
```{r penguins-ordered-pipe3-out, ref.label = "penguins-ordered-pipe3", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

---

## Compare the difference

.xtiny-font.pull-left[
```{r eval = FALSE}
penguins %>%
  count(species) %>%
  mutate(species = fct_reorder(species, n))
```

```{r echo = FALSE}
# remove row names
x <- penguins %>%
  count(species) %>%
  mutate(species = fct_reorder(species, n))

row.names(x) <- NULL
x
```

]

.xtiny-font.pull-right[
```{r}
penguins %>% 
  mutate(species = 
    fct_reorder(species, species, length))
```
]

---

## Compare the difference

.xtiny-font.pull-left[
```{r penguins-geom-col, fig.width = 6, fig.asp = 0.618, dev = "svg"}
penguins %>%
  count(species) %>%
  mutate(species = fct_reorder(species, n)) %>%
  ggplot(aes(n, species)) + geom_col()
```

]

.xtiny-font.pull-right[
```{r penguins-geom-bar, fig.width = 6, fig.asp = 0.618, dev = "svg"}
penguins %>% 
  mutate(species = 
    fct_reorder(species, species, length)) %>%
  ggplot(aes(y = species)) + geom_bar()
```
]

---

## Other examples

```{r include = FALSE}
library(ggridges)
library(nycflights13)

flights %>%
  group_by(carrier) %>% summarize(mean_delay = mean(arr_delay, na.rm = TRUE)) %>%
  left_join(airlines) %>%
  ggplot(aes(mean_delay, fct_reorder(name, mean_delay))) + geom_col()


flights %>%
  left_join(airlines) %>%
  ggplot(aes(y = fct_rev(fct_infreq(name)))) + geom_bar()

flights %>%
  left_join(airlines) %>%
  ggplot(aes(y = fct_rev(fct_infreq(fct_lump(name, 9))))) + geom_bar()

flights %>%
  left_join(airlines) %>%
  mutate(
    name = fct_reorder(name, arr_delay, na.rm = TRUE)
  ) %>%
  ggplot(aes(arr_delay)) + 
  geom_density(fill = "cornsilk") +
  facet_wrap(vars(name)) +
  xlim(-50, 200)

flights %>%
  left_join(airlines) %>%
  mutate(
    name = fct_reorder(name, arr_delay, na.rm = TRUE)
  ) %>%
  ggplot(aes(arr_delay, name)) + 
  geom_density_ridges(fill = "cornsilk") +
  xlim(-50, 200)
```

---

## Further reading

- **forcats** reference documentation: [Change order of levels](https://forcats.tidyverse.org/reference/index.html#section-change-order-of-levels)




