---
title: "Getting things into the right order"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(palmerpenguins)
library(cowplot)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width
```


## Remember from "Visualizing amounts"

.small-font[
We can use `fct_relevel()` to manually order the bars in a bar plot
]

--

.tiny-font[
```{r penguins-ordered, eval = FALSE}
ggplot(penguins, aes(y = fct_relevel(species, "Chinstrap", "Gentoo", "Adelie"))) +
  geom_bar()
```
]

.center[
```{r penguins-ordered-out, ref.label = "penguins-ordered", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

---

## Somewhat cleaner: mutate first, then plot

.tiny-font[
```{r penguins-ordered-pipe, eval = FALSE}
penguins %>%
  mutate(species = fct_relevel(species, "Chinstrap", "Gentoo", "Adelie")) %>%
  ggplot(aes(y = species)) +
  geom_bar()
```
]

.center[
```{r penguins-ordered-pipe-out, ref.label = "penguins-ordered-pipe", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

---

## We order things in ggplot with factors

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_relevel(species, "Chinstrap", "Gentoo", "Adelie")) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```
]

--

.small-font[
- The column `species` is a factor
]

--

.small-font[
- A factor is a categorical variable with defined categories called levels
]

--

.small-font[
- For factors, ggplot generally places visual elements in the order defined by the levels
]

---

## We order things in ggplot with factors

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_relevel(species, "Chinstrap", "Gentoo", "Adelie")) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column

# the order of factor levels is independent of the order of values in the table
penguins %>% 
  mutate(species = fct_relevel(species, "Chinstrap", "Gentoo", "Adelie"))
```
]


---

## Manual ordering of factor levels: `fct_relevel()`

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_relevel(species)) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```
]

.small-font[
Default: alphabetic order
]
---

## Manual ordering of factor levels: `fct_relevel()`

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_relevel(species, "Gentoo")) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```
]

.small-font[
Move `"Gentoo"` in front, rest alphabetic
]

---

## Manual ordering of factor levels: `fct_relevel()`

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_relevel(species, "Chinstrap", "Gentoo")) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```

]

.small-font[
Move `"Chinstrap"` in front, then `"Gentoo"`, rest alphabetic
]

---

## Manual ordering of factor levels: `fct_relevel()`

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_relevel(species, "Chinstrap", "Adelie", "Gentoo")) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```

]

.small-font[
Use order `"Chinstrap"`, `"Adelie"`, `"Gentoo"`
]

---

## Manual ordering of factor levels: `fct_relevel()`

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_relevel(species, "Gentoo", "Chinstrap", "Adelie")) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```

]

.small-font[
Use order `"Gentoo"`, `"Chinstrap"`, `"Adelie"`
]

---

## The order of the y axis is from bottom to top

.tiny-font[
```{r penguins-ordered-pipe2, eval = FALSE}
penguins %>%
  mutate(species = fct_relevel(species, "Chinstrap", "Gentoo", "Adelie")) %>%
  ggplot(aes(y = species)) +
  geom_bar()
```
]

.center[
```{r penguins-ordered-pipe2-out, ref.label = "penguins-ordered-pipe2", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

---

## Reorder based on frequency: `fct_infreq()`

.tiny-font[
```{r}
penguins %>%
  mutate(species = fct_infreq(species)) %>%
  slice(1:30) %>%   # get first 30 rows
  pull(species)     # pull out just the `species` column
```

]

--

.small-font[
- Use the order defined by the number of penguins of different species
]

--

.small-font[
- The order is descending, from most frequent to least frequent
]

---

## Reorder based on frequency: `fct_infreq()`

.tiny-font[
```{r penguins-infreq, eval = FALSE}
penguins %>%
  mutate(species = fct_infreq(species)) %>%
  ggplot(aes(y = species)) + geom_bar()
```

]

.center[
```{r penguins-infreq-out, ref.label = "penguins-infreq", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

---

## Reverse order: `fct_rev()`

.tiny-font[
```{r penguins-infreq-rev, eval = FALSE}
penguins %>%
  mutate(species = fct_rev(fct_infreq(species))) %>%
  ggplot(aes(y = species)) + geom_bar()
```

]

.center[
```{r penguins-infreq-rev-out, ref.label = "penguins-infreq-rev", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

---

## Reorder based on numeric values: `fct_reorder()`

.tiny-font[
```{r eval = FALSE}
penguins %>%
  count(species)
```

```{r echo = FALSE}
# remove row names
x <- penguins %>%
  count(species)

row.names(x) <- NULL
x
```

]

--

.tiny-font[
```{r}
penguins %>%
  count(species) %>%
  mutate(species = fct_reorder(species, n)) %>%
  pull(species)     # pull out just the `species` column
```

]

--

.small-font[
The order is ascending, from smallest to largest value
]

---

## Reorder based on numeric values: `fct_reorder()`

.tiny-font[
```{r penguins-reorder, eval = FALSE}
penguins %>%
  count(species) %>%
  mutate(species = fct_reorder(species, n)) %>%
  ggplot(aes(n, species)) + geom_col()
```

]

.center[
```{r penguins-reorder-out, ref.label = "penguins-reorder", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

---

## Compare to see the difference

.xtiny-font.pull-left[
```{r eval = FALSE}
penguins %>%
  count(species) %>% # summarize data
  mutate(species = fct_reorder(species, n))
```

```{r echo = FALSE}
# remove row names
x <- penguins %>%
  count(species) %>%
  mutate(species = fct_rev(fct_reorder(species, n)))

row.names(x) <- NULL
x
```

]

--

.xtiny-font.pull-right[
```{r}
penguins %>% 
  # modify the original dataset, no summary
  mutate(species = fct_infreq(species))
```
]


---

## Compare to see the difference

.xtiny-font.pull-left[
```{r penguins-geom-col, fig.width = 6, fig.asp = 0.618, dev = "svg"}
penguins %>%
  count(species) %>% # summarize data
  mutate(species = fct_reorder(species, n)) %>%
  ggplot(aes(n, species)) + geom_col()
```

]

.xtiny-font.pull-right[
```{r penguins-geom-bar, fig.width = 6, fig.asp = 0.618, dev = "svg"}
penguins %>% 
  # modify the original dataset, no summary
  mutate(species = fct_infreq(species)) %>%
  ggplot(aes(y = species)) + geom_bar()
```
]

---

## Compare to see the difference

.xtiny-font.pull-left[
```{r penguins-geom-col2, fig.width = 6, fig.asp = 0.618, dev = "svg"}
penguins %>%
  count(species) %>% # summarize data
  mutate(species = fct_reorder(species, n)) %>%
  ggplot(aes(n, species)) + geom_col()
```

]

.xtiny-font.pull-right[
```{r penguins-geom-bar2, fig.width = 6, fig.asp = 0.618, dev = "svg"}
penguins %>% 
  # modify the original dataset, no summary
  mutate(species = fct_infreq(species)) %>%
  ggplot(aes(y = fct_rev(species))) + geom_bar()
```
]

[//]: # "segment ends here"

---

class: middle, center

# Ordering other plot elements

---

## The gapminder dataset: Life expectancy data

.tiny-font[
```{r}
library(gapminder)

gapminder
```
]

---

## Life expectancy in the Americas in 2007

.tiny-font.pull-left[
```{r gapminder-unordered, eval = FALSE}
gapminder %>%
  filter(
    year == 2007,
    continent == "Americas"
  ) %>%
  ggplot(aes(lifeExp, country)) + 
  geom_point()
```
]

.pull-right[
```{r gapminder-unordered-out, ref.label = "gapminder-unordered", echo = FALSE, fig.width = 5, fig.asp = 0.9, dev = "svg"}
```
]



---

## Life expectancy in the Americas in 2007

.pull-left[
.tiny-font[
```{r gapminder-unordered2, eval = FALSE}
gapminder %>%
  filter(
    year == 2007,
    continent == "Americas"
  ) %>%
  ggplot(aes(lifeExp, country)) + 
  geom_point()
```
]

.small-font[
Reminder:  
Default order is alphabetic, from bottom to top
]]

.pull-right[
```{r gapminder-unordered2-out, ref.label = "gapminder-unordered2", echo = FALSE, fig.width = 5, fig.asp = 0.9, dev = "svg"}
```
]

---

## Life expectancy, ordered from highest to lowest

.pull-left[.tiny-font[
```{r gapminder-ordered, eval = FALSE}
gapminder %>%
  filter(
    year == 2007,
    continent == "Americas"
  ) %>%
  mutate(
    country = fct_reorder(country, lifeExp)
  ) %>%
  ggplot(aes(lifeExp, country)) + 
  geom_point()
```
]

.small-font[
Order is ascending from bottom to top
]]


.pull-right[
```{r gapminder-ordered-out, ref.label = "gapminder-ordered", echo = FALSE, fig.width = 5, fig.asp = 0.9, dev = "svg"}
```
]


---

## We can also order facets

.tiny-font[
```{r gapminder-europe-unordered, eval = FALSE}
gapminder %>%
  filter(country %in% c("Norway", "Portugal", "Spain", "Austria")) %>%
  ggplot(aes(year, lifeExp)) + geom_line() +
  facet_wrap(vars(country), nrow = 1)
```
]

.center[
```{r gapminder-europe-unordered-out, ref.label = "gapminder-europe-unordered", echo = FALSE, fig.width = 12, fig.asp = 0.25, dev = "svg"}
```
]

--

.small-font[
- Default ordering is alphabetic; there's no good reason for this ordering
]

---

## We can also order facets

.tiny-font[
```{r gapminder-europe-unordered2, eval = FALSE}
gapminder %>%
  filter(country %in% c("Norway", "Portugal", "Spain", "Austria")) %>%
  ggplot(aes(year, lifeExp)) + geom_line() +
  facet_wrap(vars(country), nrow = 1)
```
]

.center[
```{r gapminder-europe-unordered2-out, ref.label = "gapminder-europe-unordered2", echo = FALSE, fig.width = 12, fig.asp = 0.25, dev = "svg"}
```
]

.small-font[
- Let's apply `fct_reorder()` and see what happens
]


---

## We can also order facets

.tiny-font[
```{r gapminder-europe1, eval = FALSE}
gapminder %>%
  filter(country %in% c("Norway", "Portugal", "Spain", "Austria")) %>%
  mutate(country = fct_reorder(country, lifeExp)) %>% # default: order by median
  ggplot(aes(year, lifeExp)) + geom_line() +
  facet_wrap(vars(country), nrow = 1)
```
]

.center[
```{r gapminder-europe1-out, ref.label = "gapminder-europe1", echo = FALSE, fig.width = 12, fig.asp = 0.25, dev = "svg"}
```
]

--

.small-font[
- When the levels of a factor occur more than once, `fct_reorder()` applies a summary function
]


---

## We can also order facets

.tiny-font[
```{r gapminder-europe2, eval = FALSE}
gapminder %>%
  filter(country %in% c("Norway", "Portugal", "Spain", "Austria")) %>%
  mutate(country = fct_reorder(country, lifeExp)) %>% # default: order by median
  ggplot(aes(year, lifeExp)) + geom_line() +
  facet_wrap(vars(country), nrow = 1)
```
]

.center[
```{r gapminder-europe2-out, ref.label = "gapminder-europe2", echo = FALSE, fig.width = 12, fig.asp = 0.25, dev = "svg"}
```
]

.small-font[
- The default summary function is `median()`
]


---

## We can also order facets

.tiny-font[
```{r gapminder-europe3, eval = FALSE}
gapminder %>%
  filter(country %in% c("Norway", "Portugal", "Spain", "Austria")) %>%
  mutate(country = fct_reorder(country, lifeExp, median)) %>% # order by median
  ggplot(aes(year, lifeExp)) + geom_line() +
  facet_wrap(vars(country), nrow = 1)
```
]

.center[
```{r gapminder-europe3-out, ref.label = "gapminder-europe3", echo = FALSE, fig.width = 12, fig.asp = 0.25, dev = "svg"}
```
]

.small-font[
- We can also set the summary function explicitly
]


---

## Alternative orderings: By smallest value per facet

.tiny-font[
```{r gapminder-europe4, eval = FALSE}
gapminder %>%
  filter(country %in% c("Norway", "Portugal", "Spain", "Austria")) %>%
  mutate(country = fct_reorder(country, lifeExp, min)) %>% # order by minimum
  ggplot(aes(year, lifeExp)) + geom_line() +
  facet_wrap(vars(country), nrow = 1)
```
]

.center[
```{r gapminder-europe4-out, ref.label = "gapminder-europe4", echo = FALSE, fig.width = 12, fig.asp = 0.25, dev = "svg"}
```
]


---

## Alternative orderings: By largest value per facet

.tiny-font[
```{r gapminder-europe5, eval = FALSE}
gapminder %>%
  filter(country %in% c("Norway", "Portugal", "Spain", "Austria")) %>%
  mutate(country = fct_reorder(country, lifeExp, max)) %>% # order by maximum
  ggplot(aes(year, lifeExp)) + geom_line() +
  facet_wrap(vars(country), nrow = 1)
```
]

.center[
```{r gapminder-europe5-out, ref.label = "gapminder-europe5", echo = FALSE, fig.width = 12, fig.asp = 0.25, dev = "svg"}
```
]

---

## Alternative orderings: By smallest difference

.tiny-font[
```{r gapminder-europe6, eval = FALSE}
gapminder %>%
  filter(country %in% c("Norway", "Portugal", "Spain", "Austria")) %>%
  # order by custom function: here, difference between max and min
  mutate(country = fct_reorder(country, lifeExp, function(x) { max(x) - min(x) })) %>%
  ggplot(aes(year, lifeExp)) + geom_line() +
  facet_wrap(vars(country), nrow = 1)
```
]

.center[
```{r gapminder-europe6-out, ref.label = "gapminder-europe6", echo = FALSE, fig.width = 12, fig.asp = 0.25, dev = "svg"}
```
]

---

## Alternative orderings: By largest difference

.tiny-font[
```{r gapminder-europe7, eval = FALSE}
gapminder %>%
  filter(country %in% c("Norway", "Portugal", "Spain", "Austria")) %>%
  # order by custom function: here, difference between min and max
  mutate(country = fct_reorder(country, lifeExp, function(x) { min(x) - max(x) })) %>%
  ggplot(aes(year, lifeExp)) + geom_line() +
  facet_wrap(vars(country), nrow = 1)
```
]

.center[
```{r gapminder-europe7-out, ref.label = "gapminder-europe7", echo = FALSE, fig.width = 12, fig.asp = 0.25, dev = "svg"}
```
]

---

## Final example: Lumping factor levels together

--

.small-font[
Dataset: Flights out of New York City in 2013
]

.tiny-font[
```{r}
library(nycflights13)

flight_data <- flights %>% # take data on individual flights
  left_join(airlines) %>%  # add in full-length airline names
  select(name, carrier, flight, year, month, day, origin, dest) # pick columns of interest
```
]

--

.tiny-font[
```{r}
flight_data
```
]

---

## Flights out of New York City in 2013

.pull-left.tiny-font[
```{r flights-unordered, eval = FALSE}
flight_data %>%
  ggplot(aes(y = name)) + 
  geom_bar()
```
]

.pull-right[
```{r flights-unordered-out, ref.label = "flights-unordered", echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
```
]

--

.small-font[
As (almost) always, the default alphabetic ordering is terrible
]



---

## Flights out of New York City in 2013

.pull-left.tiny-font[
```{r flights-ordered, eval = FALSE}
flight_data %>%
  mutate(
    name = fct_infreq(name)
  ) %>%
  ggplot(aes(y = fct_rev(name))) + 
  geom_bar()
```
]


.pull-right[
```{r flights-ordered-out, ref.label = "flights-ordered", echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
```
]

--

.small-font[
Ordering by frequency is better, but do we need to show all airlines?
]

---

## Flights out of New York City in 2013, with lumping

.pull-left.tiny-font[
```{r flights-lumped, eval = FALSE}
flight_data %>%
  mutate(
    # keep only the 7 most common airlines
    name = fct_lump_n(name, 7)
  ) %>%
  ggplot(aes(y = fct_rev(name))) + 
  geom_bar()
```
]


.pull-right[
```{r flights-lumped-out, ref.label = "flights-lumped", echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
```
]

--

.small-font[
Now the ordering is again alphabetic...
]


---

## Flights out of New York City in 2013, with lumping

.pull-left.tiny-font[
```{r flights-lumped-ordered, eval = FALSE}
flight_data %>%
  mutate(
    # order after lumping
    name = fct_infreq(fct_lump_n(name, 7))
  ) %>%
  ggplot(aes(y = fct_rev(name))) + 
  geom_bar()
```
]

.pull-right[
```{r flights-lumped-ordered-out, ref.label = "flights-lumped-ordered", echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
```
]

---

## Flights out of New York City in 2013, with lumping

.pull-left.tiny-font[
```{r flights-lumped-ordered2, eval = FALSE}
flight_data %>%
  mutate(
    # order before lumping
    name = fct_lump_n(fct_infreq(name), 7)
  ) %>%
  ggplot(aes(y = fct_rev(name))) + 
  geom_bar()
```
]


.pull-right[
```{r flights-lumped-ordered2-out, ref.label = "flights-lumped-ordered2", echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
```
]

--

.small-font[
In most cases, you will want to order before lumping
]


---

## Flights out of New York City in 2013, with lumping

.pull-left.tiny-font[
```{r flights-lumped-ordered3, eval = FALSE}
flight_data %>%
  mutate(
    # order before lumping
    name = fct_lump_n(fct_infreq(name), 7)
  ) %>%
  ggplot(aes(y = fct_rev(name))) + 
  geom_bar()
```
]


.pull-right[
```{r flights-lumped-ordered3-out, ref.label = "flights-lumped-ordered3", echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
```
]

.small-font[
Can we visually separate the "Other" category?
]

---

## Flights out of New York City in 2013, with lumping

.pull-left.tiny-font[
```{r flights-highlight, eval = FALSE}
flight_data %>%
  mutate(
    name = fct_lump_n(fct_infreq(name), 7),
    # Use `fct_other()` to manually lump all
    # levels not called "Other" into "Named"
    highlight = fct_other(
      name,
      keep = "Other", other_level = "Named"
    )
  ) %>%
  ggplot() +
  aes(
    y = fct_rev(name),
    fill = highlight
  ) + 
  geom_bar()
```
]


.pull-right[
```{r flights-highlight-out, ref.label = "flights-highlight", echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
```
]

--

.small-font[
One annoying issue: The legend is in the wrong order
]


---

## Flights out of New York City in 2013, with lumping

.pull-left.tiny-font[
```{r flights-highlight2, eval = FALSE}
flight_data %>%
  mutate(
    name = fct_lump_n(fct_infreq(name), 7),
    # Use `fct_other()` to manually lump all
    # levels not called "Other" into "Named"
    highlight = fct_other(
      name,
      keep = "Other", other_level = "Named"
    )
  ) %>%
  ggplot() +
  aes(
    y = fct_rev(name),
    # reverse fill aesthetic
    fill = fct_rev(highlight)
  ) + 
  geom_bar()
```
]


.pull-right[
```{r flights-highlight2-out, ref.label = "flights-highlight2", echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
```
]


---

## Flights out of New York City in 2013, final tweaks

.pull-left.xtiny-font[
```{r flights-final, eval = FALSE}
flight_data %>%
  mutate(
    name = fct_lump_n(fct_infreq(name), 7),
    highlight = fct_other(
      name, keep = "Other", other_level = "Named"
    )
  ) %>%
  ggplot() +
  aes(y = fct_rev(name), fill = highlight) + 
  geom_bar() +
  scale_x_continuous(
    name = "Number of flights",
    expand = expansion(mult = c(0, 0.07))
  ) +
  scale_y_discrete(name = NULL) +
  scale_fill_manual(
    values = c(
      Named = "gray50", Other = "#98545F"
    ),
    guide = "none"
  ) +
  theme_minimal_vgrid()

```
]


.pull-right[
```{r flights-final-out, ref.label = "flights-final", echo = FALSE, fig.width = 5, fig.asp = 0.75, dev = "svg"}
```
]

---

## Summary of key factor manipulation functions

.small-font.center[

Function           | Use case        | Documentation
:-----------       | :----------     | :----------:
`fct_relevel()`    | Change order of factor levels manually | [click here](https://forcats.tidyverse.org/reference/fct_relevel.html)
`fct_infreq()`     | Put levels in descending order of how frequently each level occurs in the data | [click here](https://forcats.tidyverse.org/reference/fct_inorder.html)
`fct_rev()`        | Reverse the order of factor levels | [click here](https://forcats.tidyverse.org/reference/fct_rev.html)
`fct_reorder()`    | Put levels in ascending order determined by a numeric variable or summary function | [click here](https://forcats.tidyverse.org/reference/fct_reorder.html)
`fct_lump_n()`     | Retain the *n* most frequent levels and lump all others into `"Other"` | [click here](https://forcats.tidyverse.org/reference/fct_lump.html)
`fct_other()`      | Manually group some factor levels into `"Other"` | [click here](https://forcats.tidyverse.org/reference/fct_other.html)

]

For more options, check out the [reference documentation](https://forcats.tidyverse.org/reference/index.html) of the **forcats** package

[//]: # "segment ends here"

---

## Further reading

- Fundamentals of Data Visualization: [Chapter 6: Visualizing amounts](https://clauswilke.com/dataviz/visualizing-amounts.html)
- **forcats** documentation: [Introduction to forcats](https://forcats.tidyverse.org/articles/forcats.html)
- **forcats** reference documentation: [Change order of levels](https://forcats.tidyverse.org/reference/index.html#section-change-order-of-levels)




