---
title: "Hierarchical clustering"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(here)
library(ggdendro)
library(cowplot)
library(colorspace)
library(kableExtra)
library(ggforce)
library(broom)

US_regions <- read_csv(here("datasets", "US_regions.csv"))
US_state_stats <- read_csv(here("datasets", "US_state_stats.csv")) %>%
  filter(state != "District of Columbia")

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width
```

## Example data: State level statistics for US

.tiny-font[
```{r}
US_state_stats
```
]


---

## PCA reveals broad trends

.center[
```{r US_states_pca, echo = FALSE, message = FALSE, fig.width = 8., fig.asp = .618, dev = "svg"}
pca_fit <- US_state_stats %>%
  select(-state, -murder, -robbery, -agg_assault, -larceny, -motor_theft) %>%
  scale() %>%
  prcomp()

pca_fit %>%
  # add PCs to the original dataset
  augment(US_state_stats) %>%
  left_join(US_regions) %>%
  ggplot(aes(.fittedPC1, .fittedPC2)) +
  geom_point(aes(color = region, fill = region, shape = region), size = 3) +
  scale_shape_manual(
    values = c(21, 22, 23, 24),
    name = NULL
  ) +
  scale_color_manual(
    values = darken(c("#56B4E9", "#E69F00", "#009E73", "#F0E442"), 0.3),
    name = NULL
  ) +
  scale_fill_manual(
    values = c("#56B4E9C0", "#E69F00C0", "#009E73C0", "#F0E442C0"),
    name = NULL
  ) +
  scale_x_continuous(
    name = "PC 1",
    limits = c(-5, 6),
    breaks = c(-4, -2, 0, 2, 4, 6)
    
  ) +
  scale_y_continuous(
    name = "PC 2",
    limits = c(-4, 3)
  ) +
  coord_fixed() +
  theme_minimal_grid(rel_small = 1)
```
]

---
class: middle, center

## Can we get a more detailed picture?

---


## Hierarchical clustering


.center[
```{r US_states_dendro, echo = FALSE, fig.width = 12, fig.asp = .45, dev = "svg"}
hc <- US_state_stats %>%
  column_to_rownames(var = "state") %>%
  scale() %>%
  dist(method = "euclidean") %>%
  hclust(method = "ward.D2")

ddata <- dendro_data(as.dendrogram(hc), type = "rectangle")
segments <- segment(ddata)
labels <- label(ddata) %>%
  left_join(US_regions, by = c("label" = "state"))

ggplot() + 
  geom_segment(data = segments, aes(x, -y, xend = xend, yend = -yend)) + 
  geom_text(
    data = labels,
    aes(x, -y+.5, label = label, color = region),
    hjust = 0,
    angle = 90,
    key_glyph = "point",
    size = 4
  ) +
  scale_color_manual(
    values = darken(c("#56B4E9", "#E69F00", "#009E73", "#F0E442"), 0.3),
    name = NULL
  ) +
  scale_x_continuous(name = NULL) +
  scale_y_continuous(name = NULL, limits = c(-20, 6), expand = c(0, 0)) +
  guides(
    color = guide_legend(
      override.aes = list(color = c("#56B4E9", "#E69F00", "#009E73", "#F0E442"))
    )
  ) +
  theme_void(16)
```
]

--

This visualization is called a dendrogram

---

## Cut the dendrogram to obtain defined clusters

```{r US_states_dendro_cut, echo = FALSE, fig.width = 12, fig.asp = .45, dev = "svg"}
cut_height <- 14.7
clusters <- cutree(hc, h = cut_height)
clustered_labels <- left_join(
  labels,
  tibble(label = names(clusters), cluster = factor(clusters)),
  by = "label"
) 

ggplot() + 
  geom_segment(data = segments, aes(x, -y, xend = xend, yend = -yend)) + 
  geom_text(
    data = clustered_labels,
    aes(x, -y+.5, label = label, color = cluster),
    hjust = 0,
    angle = 90,
    key_glyph = "point",
    size = 4
  ) +
  geom_hline(yintercept = -cut_height, linetype = 2) +
  scale_color_viridis_d(option = "B", begin = 0.2, end = 0.8) +
  scale_x_continuous(name = NULL) +
  scale_y_continuous(name = NULL, limits = c(-20, 6), expand = c(0, 0)) +
  theme_void(16)
```

---

## Cut the dendrogram to obtain defined clusters

```{r US_states_dendro_cut2, echo = FALSE, fig.width = 12, fig.asp = .45, dev = "svg"}
cut_height <- 10.9
clusters <- cutree(hc, h = cut_height)
clustered_labels <- left_join(
  labels,
  tibble(label = names(clusters), cluster = factor(clusters)),
  by = "label"
) 

ggplot() + 
  geom_segment(data = segments, aes(x, -y, xend = xend, yend = -yend)) + 
  geom_text(
    data = clustered_labels,
    aes(x, -y+.5, label = label, color = cluster),
    hjust = 0,
    angle = 90,
    key_glyph = "point",
    size = 4
  ) +
  geom_hline(yintercept = -cut_height, linetype = 2) +
  scale_color_viridis_d(option = "B", begin = 0.2, end = 0.8) +
  scale_x_continuous(name = NULL) +
  scale_y_continuous(name = NULL, limits = c(-20, 6), expand = c(0, 0)) +
  theme_void(16)
```

---

## Cut the dendrogram to obtain defined clusters

```{r US_states_dendro_cut3, echo = FALSE, fig.width = 12, fig.asp = .45, dev = "svg"}
cut_height <- 7.5
clusters <- cutree(hc, h = cut_height)
clustered_labels <- left_join(
  labels,
  tibble(label = names(clusters), cluster = factor(clusters)),
  by = "label"
) 

ggplot() + 
  geom_segment(data = segments, aes(x, -y, xend = xend, yend = -yend)) + 
  geom_text(
    data = clustered_labels,
    aes(x, -y+.5, label = label, color = cluster),
    hjust = 0,
    angle = 90,
    key_glyph = "point",
    size = 4
  ) +
  geom_hline(yintercept = -cut_height, linetype = 2) +
  scale_color_viridis_d(option = "B", begin = 0.05, end = 0.95) +
  scale_x_continuous(name = NULL) +
  scale_y_continuous(name = NULL, limits = c(-20, 6), expand = c(0, 0)) +
  theme_void(16)
```

---

## Cut the dendrogram to obtain defined clusters

```{r US_states_dendro_cut4, echo = FALSE, fig.width = 12, fig.asp = .45, dev = "svg"}
cut_height <- 5.0
clusters <- cutree(hc, h = cut_height)
clustered_labels <- left_join(
  labels,
  tibble(label = names(clusters), cluster = factor(clusters)),
  by = "label"
) 

ggplot() + 
  geom_segment(data = segments, aes(x, -y, xend = xend, yend = -yend)) + 
  geom_text(
    data = clustered_labels,
    aes(x, -y+.5, label = label, color = cluster),
    hjust = 0,
    angle = 90,
    key_glyph = "point",
    size = 4
  ) +
  geom_hline(yintercept = -cut_height, linetype = 2) +
  scale_color_viridis_d(option = "B", begin = 0.05, end = 0.95) +
  scale_x_continuous(name = NULL) +
  scale_y_continuous(name = NULL, limits = c(-20, 6), expand = c(0, 0)) +
  theme_void(16)
```

---

## Cut the dendrogram to obtain defined clusters

```{r US_states_dendro_cut5, echo = FALSE, fig.width = 12, fig.asp = .45, dev = "svg"}
cut_height <- 10.9
clusters <- cutree(hc, h = cut_height)
clustered_labels <- left_join(
  labels,
  tibble(label = names(clusters), cluster = factor(clusters)),
  by = "label"
) 

ggplot() + 
  geom_segment(data = segments, aes(x, -y, xend = xend, yend = -yend)) + 
  geom_text(
    data = clustered_labels,
    aes(x, -y+.5, label = label, color = cluster),
    hjust = 0,
    angle = 90,
    key_glyph = "point",
    size = 4
  ) +
  geom_hline(yintercept = -cut_height, linetype = 2) +
  scale_color_viridis_d(option = "B", begin = 0.2, end = 0.8) +
  scale_x_continuous(name = NULL) +
  scale_y_continuous(name = NULL, limits = c(-20, 6), expand = c(0, 0)) +
  theme_void(16)
```

---

## Display clusters in PCA space

.center[
```{r US_states_pca_cluster, echo = FALSE, message = FALSE, fig.width = 8., fig.asp = .618, dev = "svg"}
pca_fit %>%
  # add PCs to the original dataset
  augment(US_state_stats) %>%
  left_join(
    tibble(state = names(clusters), cluster = factor(clusters)),
    by = "state"
  ) %>%
  left_join(US_regions, by = "state") %>%
  ggplot(aes(.fittedPC1, .fittedPC2)) +
  geom_point(aes(color = cluster, fill = cluster, shape = region), size = 3) +
  scale_shape_manual(
    values = c(21, 22, 23, 24),
    name = NULL
  ) +
  scale_color_viridis_d(
    name = "cluster",
    option = "B", begin = 0, end = 0.8
  ) +
  scale_fill_viridis_d(
    name = "cluster",
    option = "B", begin = 0.4, end = 1
  ) +
  scale_x_continuous(
    name = "PC 1",
    limits = c(-5, 6),
    breaks = c(-4, -2, 0, 2, 4, 6)
  ) +
  scale_y_continuous(
    name = "PC 2",
    limits = c(-4, 3)
  ) +
  guides(
    color = guide_legend(
      order = 1,
      override.aes = list(shape = 21)
    ),
    fill = guide_legend(
      order = 1
    ),
    shape = guide_legend(
      order = 2,
      override.aes = list(fill = "gray80")
    )
  ) +
  coord_fixed() +
  theme_minimal_grid(rel_small = 1)
```
]

---

## Hierarchical clustering can be useful with heatmaps

.pull-left[
Arbitrary ordering:
```{r forensic-correlations1, echo = FALSE, message = FALSE, fig.width = 4., fig.asp = 1, dev = "svg"}
forensic_glass <- read_csv(here("datasets", "forensic_glass.csv"))

cm <- cor(select(forensic_glass, -type, -RI, -Si))
df_wide <- as.data.frame(cm)
df_long <- stack(df_wide)
names(df_long) <- c("cor", "var1")
df_long <- cbind(df_long, var2 = rep(rownames(cm), length(rownames(cm))))
clust <- hclust(as.dist(1-cm), method="average") 
levels <- clust$labels[clust$order]
df_long$var1 <- factor(df_long$var1, levels = sort(levels))
df_long$var2 <- factor(df_long$var2, levels = sort(levels))

ggplot(filter(df_long, as.integer(var1) < as.integer(var2)),
       aes(var1, var2, fill=cor)) + 
  geom_tile(color = "white", size = 1) + 
  scale_x_discrete(position = "top", name = NULL, expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = c(0, 0)) +
  scale_fill_continuous_diverging(
    name = "correlation",
    limits = c(-.5, .5),
    breaks = c(-.5, 0, .5),
    labels = c("–0.5", "0.0", "0.5"),
    palette = "Blue-Red",
    rev = TRUE
  ) +
  guides(
    fill = guide_colorbar(
      direction = "horizontal",
      label.position = "bottom",
      title.position = "top",
      barwidth = grid::unit(140, "pt"),
      barheight = grid::unit(17.5, "pt"),
      ticks.linewidth = 1
    )
  ) +
  coord_fixed() +
  theme_half_open(rel_small = 1) +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.ticks.length = grid::unit(3, "pt"),
    legend.position = c(.97, .0),
    legend.justification = c(1, 0),
    legend.title.align = 0.5
  )
```
]

--

.pull-right[
Ordering by clustering similarity:
```{r forensic-correlations2, echo = FALSE, message = FALSE, fig.width = 4., fig.asp = 1, dev = "svg"}
df_long$var1 <- factor(df_long$var1, levels = levels)
df_long$var2 <- factor(df_long$var2, levels = levels)

ggplot(filter(df_long, as.integer(var1) < as.integer(var2)),
       aes(var1, var2, fill=cor)) + 
  geom_tile(color = "white", size = 1) + 
  scale_x_discrete(position = "top", name = NULL, expand = c(0, 0)) +
  scale_y_discrete(name = NULL, expand = c(0, 0)) +
  scale_fill_continuous_diverging(
    name = "correlation",
    limits = c(-.5, .5),
    breaks = c(-.5, 0, .5),
    labels = c("–0.5", "0.0", "0.5"),
    palette = "Blue-Red",
    rev = TRUE
  ) +
  guides(
    fill = guide_colorbar(
      direction = "horizontal",
      label.position = "bottom",
      title.position = "top",
      barwidth = grid::unit(140, "pt"),
      barheight = grid::unit(17.5, "pt"),
      ticks.linewidth = 1
    )
  ) +
  coord_fixed() +
  theme_half_open(rel_small = 1) +
  theme(
    axis.line = element_blank(),
    axis.ticks = element_blank(),
    axis.ticks.length = grid::unit(3, "pt"),
    legend.position = c(.97, .0),
    legend.justification = c(1, 0),
    legend.title.align = 0.5
  )
```
]

```{r eval = FALSE, echo = FALSE}
## delete?

df_long %>%
  filter(
    as.integer(factor(as.character(var1))) <
      as.integer(factor(as.character(var2)))
  ) %>%
  arrange(as.character(var1), as.character(var2)) %>%
  mutate(id = n():1) %>%
  ggplot(aes(1, id, fill=cor)) + 
  geom_tile(color = "white", size = 1) + 
  geom_text(
    aes(x = 1.75, label = var1), size = 12/.pt, hjust = 0
  ) +
  geom_text(
    aes(x = 2.5, label = var2), size = 12/.pt, hjust = 0
  ) +
  scale_x_continuous(limits = c(0, 8)) +
  scale_fill_continuous_diverging(
    name = "correlation",
    limits = c(-.5, .5),
    breaks = c(-.5, 0, .5),
    labels = c("–0.5", "0.0", "0.5"),
    palette = "Blue-Red",
    rev = TRUE
  ) +
  guides(
    fill = guide_colorbar(
      direction = "horizontal",
      label.position = "bottom",
      title.position = "top",
      barwidth = grid::unit(120, "pt"),
      barheight = grid::unit(12, "pt"),
      ticks.linewidth = 1
    )
  ) +
  theme_map(12, rel_small = 1) +
  theme(
    legend.position = c(.97, .1),
    legend.justification = c(1, .5),
    legend.title.align = 0.5
  )
```

```{r eval = FALSE, echo = FALSE}
df_long %>%
  filter(as.integer(var1) < as.integer(var2)) %>%
  arrange(desc(var1), desc(var2)) %>%
  mutate(id = n():1) %>%
  ggplot(aes(1, id, fill=cor)) + 
  geom_tile(color = "white", size = 1) + 
  geom_text(
    aes(x = 1.75, label = var1), size = 12/.pt, hjust = 0
  ) +
  geom_text(
    aes(x = 2.5, label = var2), size = 12/.pt, hjust = 0
  ) +
  scale_x_continuous(limits = c(0, 8)) +
  scale_fill_continuous_diverging(
    name = "correlation",
    limits = c(-.5, .5),
    breaks = c(-.5, 0, .5),
    labels = c("–0.5", "0.0", "0.5"),
    palette = "Blue-Red",
    rev = TRUE
  ) +
  guides(
    fill = guide_colorbar(
      direction = "horizontal",
      label.position = "bottom",
      title.position = "top",
      barwidth = grid::unit(120, "pt"),
      barheight = grid::unit(12, "pt"),
      ticks.linewidth = 1
    )
  ) +
  theme_map(12, rel_small = 1) +
  theme(
    legend.position = c(.97, .1),
    legend.justification = c(1, .5),
    legend.title.align = 0.5
  )
```

[//]: # "segment ends here"

---
class: center middle


## Hierarchical clustering via UPGMA


---

## Hierarchical clustering methods require two ingredients

--

1\. Distance metric

Examples: Euclidean distance, maximum distance

<br>
--

2\. Linkage criteria

Examples:
- Unweighted average linkage clustering  
   (unweighted pair group method with arithmetic mean, UPGMA)
- Complete linkage clustering
- Minimum energy clustering

(See: https://en.wikipedia.org/wiki/Hierarchical_clustering)
 

---

## Hierarchical clustering via UPGMA

.pull-left.move-up-1em[
```{r UPGMA-step1, echo = FALSE, message = FALSE, fig.width = 4.5, fig.asp = 1, dev = "svg"}
data <- tibble(
  x = c(0.50, .5, 3.75, 1.70, 3.30),
  y = c(4.2, 1.6, 3.75, 1, 3.1),
  label = c(LETTERS[1:5]),
  fill = c("white", "white", "white", "white", "white"),
  cluster = c(3, 2, 1, 2, 1)
)

ggplot(data, aes(x, y)) +
  geom_point(aes(fill = fill), size = 10, shape = 21, stroke = 1) +
  geom_text(aes(label = label), size = 14/.pt, fontface = "bold") +
  coord_fixed(clip = "off", xlim = c(-.1, 4.1)) +
  scale_fill_identity() +
  theme_minimal_grid(16, rel_small = 1)
```
]

--

.pull-right[
Distance matrix:

```{r echo = FALSE}
dt <- data %>%
  column_to_rownames("label") %>%
  select(-fill) %>%
  as.matrix() %>%
  dist()

dt %>%
  as.matrix() %>%
  signif(3) %>%
  kbl() %>%
  kable_paper(full_width = F, html_font = "sans-serif") %>%
  row_spec(0, bold = TRUE, background = "white", color = "black") %>%
  column_spec(1, bold = TRUE, color = "black", background = "white") %>%
  column_spec(2, color = "black", background = "white") %>%
  column_spec(3, color = "black", background = "white") %>%
  column_spec(4, color = "black", background = "white") %>%
  column_spec(5, color = "black", background = "white") %>%
  column_spec(6, color = "black", background = "white")
```
]

---

## Hierarchical clustering via UPGMA

.pull-left.move-up-1em[
```{r UPGMA-step2, echo = FALSE, message = FALSE, fig.width = 4.5, fig.asp = 1, dev = "svg"}
data$fill <- c("white", "white", "yellow", "white", "yellow")

ggplot(data, aes(x, y)) +
  geom_mark_hull(
    aes(filter = cluster < 2),
    expand = unit(5, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_point(aes(fill = fill), size = 10, shape = 21, stroke = 1) +
  geom_text(aes(label = label), size = 14/.pt, fontface = "bold") +
  coord_fixed(clip = "off", xlim = c(-.1, 4.1)) +
  scale_fill_identity() +
  theme_minimal_grid(16, rel_small = 1)
```
]

.pull-right[
Distance matrix:

```{r echo = FALSE}
dt %>%
  as.matrix() %>%
  signif(3) %>%
  kbl() %>%
  kable_paper(full_width = F, html_font = "sans-serif") %>%
  row_spec(0, bold = TRUE, background = "white", color = "black") %>%
  column_spec(1, bold = TRUE, color = "black", background = "white") %>%
  column_spec(2, color = "black", background = "white") %>%
  column_spec(3, color = "black", background = "white") %>%
  column_spec(4, color = "black", background = c("white", "white", "white", "white", "yellow")) %>%
  column_spec(5, color = "black", background = "white") %>%
  column_spec(6, color = "black", background = c("white", "white", "yellow", "white", "white"))
```

Join points with minimum distance
]

---

## Hierarchical clustering via UPGMA

.pull-left.move-up-1em[
```{r UPGMA-step3, echo = FALSE, message = FALSE, fig.width = 4.5, fig.asp = 1, dev = "svg"}
data$fill <- c("white", "white", "gray90", "white", "gray90")

ggplot(data, aes(x, y)) +
  geom_mark_hull(
    aes(filter = cluster < 2),
    expand = unit(5, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_point(aes(fill = fill), size = 10, shape = 21, stroke = 1) +
  geom_text(aes(label = label), size = 14/.pt, fontface = "bold") +
  coord_fixed(clip = "off", xlim = c(-.1, 4.1)) +
  scale_fill_identity() +
  theme_minimal_grid(16, rel_small = 1)
```
]

.pull-right[
Update distances by averaging:
$$d_{A:(C, E)} = (d_{A:C} + d_{A:E})/2,$$

similarly for points B and D
]
---

## Hierarchical clustering via UPGMA

.pull-left.move-up-1em[
![](`r knitr::fig_chunk("UPGMA-step3", "svg")`)
]

.pull-right[
Updated distance matrix:

```{r echo = FALSE}
UPGMA_step <- function(m, i1, i2, n1, n2) {
  labels <- rownames(m)
  label_new <- glue::glue("({labels[i1]}, {labels[i2]})")
  row_new <- (1/(n1+n2))*(n1*m[i1, ] + n2*m[i2, ])[c(-i1, -i2)]
  names_new <- c(names(row_new), label_new)
  row_new_padded <- c(row_new, 0)
  m2 <- cbind(rbind(m[c(-i1, -i2), c(-i1, -i2)], row_new), row_new_padded)
  colnames(m2) <- names_new
  rownames(m2) <- names_new
  m2
}

m2 <- dt %>%
  as.matrix() %>%
  UPGMA_step(3, 5, 1, 1)  

m2 %>%
  signif(3) %>%
  kbl() %>%
  kable_paper(full_width = F, html_font = "sans-serif") %>%
  row_spec(0, bold = TRUE, background = "white", color = "black") %>%
  column_spec(1, bold = TRUE, color = "black", background = "white") %>%
  column_spec(2, color = "black", background = "white") %>%
  column_spec(3, color = "black", background = "white") %>%
  column_spec(4, color = "black", background = "white") %>%
  column_spec(5, color = "black", background = "white")
```
]

---

## Hierarchical clustering via UPGMA

.pull-left.move-up-1em[
```{r UPGMA-step4, echo = FALSE, message = FALSE, fig.width = 4.5, fig.asp = 1, dev = "svg"}
data$fill <- c("white", "yellow", "gray90", "yellow", "gray90")

ggplot(data, aes(x, y)) +
  geom_mark_hull(
    aes(filter = cluster == 1),
    expand = unit(5, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_mark_hull(
    aes(filter = cluster == 2),
    expand = unit(5, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_point(aes(fill = fill), size = 10, shape = 21, stroke = 1) +
  geom_text(aes(label = label), size = 14/.pt, fontface = "bold") +
  coord_fixed(clip = "off", xlim = c(-.1, 4.1)) +
  scale_fill_identity() +
  theme_minimal_grid(16, rel_small = 1)
```
]

.pull-right[
Updated distance matrix:

```{r echo = FALSE}
m2 %>%
  signif(3) %>%
  kbl() %>%
  kable_paper(full_width = F, html_font = "sans-serif") %>%
  row_spec(0, bold = TRUE, background = "white", color = "black") %>%
  column_spec(1, bold = TRUE, color = "black", background = "white") %>%
  column_spec(2, color = "black", background = "white") %>%
  column_spec(3, color = "black", background = c("white", "white", "yellow", "white")) %>%
  column_spec(4, color = "black", background = c("white", "yellow", "white", "white")) %>%
  column_spec(5, color = "black", background = "white")
```

Continue joining points with minimum distance
]

---

## Hierarchical clustering via UPGMA

.pull-left.move-up-1em[
```{r UPGMA-step5, echo = FALSE, message = FALSE, fig.width = 4.5, fig.asp = 1, dev = "svg"}
data$fill = c("white", "gray90", "gray90", "gray90", "gray90")

ggplot(data, aes(x, y)) +
  geom_mark_hull(
    aes(filter = cluster == 1),
    expand = unit(5, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_mark_hull(
    aes(filter = cluster == 2),
    expand = unit(5, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_point(aes(fill = fill), size = 10, shape = 21, stroke = 1) +
  geom_text(aes(label = label), size = 14/.pt, fontface = "bold") +
  coord_fixed(clip = "off", xlim = c(-.1, 4.1)) +
  scale_fill_identity() +
  theme_minimal_grid(16, rel_small = 1)
```
]

.pull-right[
Updated distance matrix:

```{r echo = FALSE}
m3 <- m2 %>%
  UPGMA_step(2, 3, 1, 1)  

m3 %>%
  signif(3) %>%
  kbl() %>%
  kable_paper(full_width = F, html_font = "sans-serif") %>%
  row_spec(0, bold = TRUE, background = "white", color = "black") %>%
  column_spec(1, bold = TRUE, color = "black", background = "white") %>%
  column_spec(2, color = "black", background = "white") %>%
  column_spec(3, color = "black", background = "white") %>%
  column_spec(4, color = "black", background = "white")
```
]

---

## Hierarchical clustering via UPGMA

.pull-left.move-up-1em[
```{r UPGMA-step6, echo = FALSE, message = FALSE, fig.width = 4.5, fig.asp = 1, dev = "svg"}
data$fill <- c("yellow", "gray90", "gray90", "gray90", "gray90")

ggplot(data, aes(x, y)) +
  geom_mark_hull(
    aes(filter = cluster %in% c(2, 3)),
    expand = unit(6, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_mark_hull(
    aes(filter = cluster == 2),
    expand = unit(5, "mm"),
    radius = unit(5, "mm"),
    fill = "yellow",
    alpha = .7
  ) +
  geom_mark_hull(
    aes(filter = cluster == 1),
    expand = unit(5, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_point(aes(fill = fill), size = 10, shape = 21, stroke = 1) +
  geom_text(aes(label = label), size = 14/.pt, fontface = "bold") +
  coord_fixed(clip = "off", xlim = c(-.1, 4.1)) +
  scale_fill_identity() +
  theme_minimal_grid(16, rel_small = 1)
```
]

.pull-right[
Updated distance matrix:

```{r echo = FALSE}
m3 %>%
  as.matrix() %>%
  signif(3) %>%
  kbl() %>%
  kable_paper(full_width = F, html_font = "sans-serif") %>%
  row_spec(0, bold = TRUE, background = "white", color = "black") %>%
  column_spec(1, bold = TRUE, color = "black", background = "white") %>%
  column_spec(2, color = "black", background = c("white", "white", "yellow")) %>%
  column_spec(3, color = "black", background = "white") %>%
  column_spec(4, color = "black", background = c("yellow", "white", "white"))
```
]

---

## Hierarchical clustering via UPGMA

.pull-left.move-up-1em[
![](`r knitr::fig_chunk("UPGMA-step6", "svg")`)
]

.pull-right[
In general, distances are weighted by # of data points added to the new cluster:
$$\begin{eqnarray}
  d_{(C, E): (A, (B, D))} &=& \\
   \qquad\qquad(d_{(C, E): A} &+& 2 d_{(C, E): (B, D)})/(1+2)
  \end{eqnarray}$$
]

---

## Hierarchical clustering via UPGMA

.pull-left.move-up-1em[
```{r UPGMA-step7, echo = FALSE, message = FALSE, fig.width = 4.5, fig.asp = 1, dev = "svg"}
data$fill <- c("gray90", "gray90", "gray90", "gray90", "gray90")
  
ggplot(data, aes(x, y)) +
  geom_mark_hull(
    aes(filter = cluster %in% c(2, 3)),
    expand = unit(6, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_mark_hull(
    aes(filter = cluster == 2),
    expand = unit(5, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_mark_hull(
    aes(filter = cluster == 1),
    expand = unit(5, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_point(aes(fill = fill), size = 10, shape = 21, stroke = 1) +
  geom_text(aes(label = label), size = 14/.pt, fontface = "bold") +
  coord_fixed(clip = "off", xlim = c(-.1, 4.1)) +
  scale_fill_identity() +
  theme_minimal_grid(16, rel_small = 1)
```
]


.pull-right[
Distance matrix:

```{r echo = FALSE}
m4 <- m3 %>%
  UPGMA_step(1, 3, 1, 2)  

m4 %>%
  signif(3) %>%
  kbl() %>%
  kable_paper(full_width = F, html_font = "sans-serif") %>%
  row_spec(0, bold = TRUE, background = "white", color = "black") %>%
  column_spec(1, bold = TRUE, color = "black", background = "white") %>%
  column_spec(2, color = "black", background = "white") %>%
  column_spec(3, color = "black", background = "white")
```
]


---

## Hierarchical clustering via UPGMA

.pull-left.move-up-1em[
```{r UPGMA-step8, echo = FALSE, message = FALSE, fig.width = 4.5, fig.asp = 1, dev = "svg"}
data$fill = c("gray90", "gray90", "gray90", "gray90", "gray90")

ggplot(data, aes(x, y)) +
  geom_mark_hull(
    aes(filter = cluster < 5),
    expand = unit(7, "mm"),
    radius = unit(5, "mm")
  ) +
   geom_mark_hull(
    aes(filter = cluster %in% c(2, 3)),
    expand = unit(6, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_mark_hull(
    aes(filter = cluster == 2),
    expand = unit(5, "mm"),
    radius = unit(5, "mm")
  ) +
  geom_mark_hull(
    aes(filter = cluster == 1),
    expand = unit(5, "mm"),
    radius = unit(5, "mm")
  ) +
 geom_point(aes(fill = fill), size = 10, shape = 21, stroke = 1) +
  geom_text(aes(label = label), size = 14/.pt, fontface = "bold") +
  coord_fixed(clip = "off", xlim = c(-.1, 4.1)) +
  scale_fill_identity() +
  theme_minimal_grid(16, rel_small = 1)
```
]

.pull-right[
Final clustering: <br><br>
((A, (B, D)), (C, E))
]


---

## Hierarchical clustering via UPGMA

.pull-left.move-up-1em[
![](`r knitr::fig_chunk("UPGMA-step8", "svg")`)
]

.pull-right.move-up-1em[
Final clustering:

```{r UPGMA-dendro, echo = FALSE, message = FALSE, fig.width = 4.5, fig.asp = 1, dev = "svg"}
dt %>%
  hclust(method = "average") %>%
  ggdendrogram(rotate = TRUE, theme_dendro = FALSE) +
  coord_flip(expand = FALSE) +
  theme_half_open(20) +
  ylab("distance") +
  theme(
    axis.line = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank()
  )
```

]

[//]: # "segment ends here"

---
class: center middle

## Doing hierarchical clustering in R

---

## Doing hierarchical clustering in R

We'll cluster the US state stats dataset

.tiny-font[
```{r}
US_state_stats 
```
]

---
## Step 1: Calculate the distance matrix

.tiny-font[
```{r echo = FALSE}
options(width = 200)
```

```{r}
US_state_stats %>%
  column_to_rownames(var = "state") %>%
  scale() %>%
  dist(method = "euclidean")
```
]

---
## Step 1: Calculate the distance matrix

.tiny-font[
```{r}
US_state_stats %>%
  column_to_rownames(var = "state") %>%
  scale() %>%
  dist(method = "manhattan")
```
]

---

## Step 1: Calculate the distance matrix

.tiny-font[
```{r}
US_state_stats %>%
  column_to_rownames(var = "state") %>%
  scale() %>%
  dist(method = "maximum")
```
]

---
## Step 1: Calculate the distance matrix

.tiny-font[
```{r}
US_state_stats %>%
  column_to_rownames(var = "state") %>%
  scale() %>%
  dist(method = "euclidean")
```
]

---
## Step 2: Cluster and plot

.tiny-font.pull-left[
```{r US_states_simple_dendro, eval = FALSE}
dist_out <- US_state_stats %>%
  column_to_rownames(var = "state") %>%
  scale() %>%
  dist(method = "euclidean")

hc_out <- hclust(
  dist_out, method = "average"
)

library(ggdendro)
ggdendrogram(hc_out, rotate = TRUE)
```

`method = "average"`: UPGMA
]

.pull-right.width-40.move-up-6em[
```{r US_states_simple_dendro-out, ref.label="US_states_simple_dendro", fig.width = 3.5, fig.asp = 2, echo=FALSE, warning=FALSE, dev = "svg"}
```
]
---
## Step 2: Cluster and plot

.tiny-font.pull-left[
```{r US_states_simple_dendro2, eval = FALSE}
dist_out <- US_state_stats %>%
  column_to_rownames(var = "state") %>%
  scale() %>%
  dist(method = "euclidean")

hc_out <- hclust(
  dist_out, method = "ward.D2"
)

library(ggdendro)
ggdendrogram(hc_out, rotate = TRUE)
```
`method = "ward.D2"`: Ward's minimum variance method 

]

.pull-right.width-40.move-up-6em[
```{r US_states_simple_dendro2-out, ref.label="US_states_simple_dendro2", fig.width = 3.5, fig.asp = 2, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

---
## Step 2: Cluster and plot

.tiny-font.pull-left[
```{r US_states_simple_dendro3, eval = FALSE}
dist_out <- US_state_stats %>%
  column_to_rownames(var = "state") %>%
  scale() %>%
  dist(method = "euclidean")

hc_out <- hclust(
  dist_out, method = "complete"
)

library(ggdendro)
ggdendrogram(hc_out, rotate = TRUE)
```

`method = "complete"`: complete linkage method 

]

.pull-right.width-40.move-up-6em[
```{r US_states_simple_dendro3-out, ref.label="US_states_simple_dendro3", fig.width = 3.5, fig.asp = 2, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

---
## We can also plot manually

.tiny-font.pull-left[
```{r US_states_dendro_manual, eval = FALSE}
ddata <- dendro_data(
  hc_out,
  type = "rectangle"
)
segments <- segment(ddata)
labels <- label(ddata)

ggplot() + 
  geom_segment(
    data = segments,
    aes(x, y, xend = xend, yend = yend)
  ) + 
  geom_text(
    data = labels,
    aes(x, y - .5, label = label),
    hjust = 1,
    size = 8/.pt
  ) +
  coord_flip(ylim = c(-5, 12))
```
]

.pull-right.width-45.move-up-3em[
```{r US_states_dendro_manual-out, ref.label="US_states_dendro_manual", fig.width = 3.7, fig.asp = 1.6, echo=FALSE, warning=FALSE, dev = "svg"}
```
]


---
## Assign clusters by cutting the dendrogram

.tiny-font[
```{r echo = FALSE}
options(width = 100)
```

```{r}
# cut dendrogram at height 5
cutree(hc_out, h = 5) 
```
]

---
## Assign clusters by cutting the dendrogram

.tiny-font[
```{r}
# cut dendrogram at height 8
cutree(hc_out, h = 8) 
```
]

---
## Assign clusters by cutting the dendrogram

.tiny-font[
```{r}
# cut dendrogram so there are 4 clusters
cutree(hc_out, k = 4) 
```
]

---

## Add cluster info into scatterplot

.pull-left.tiny-font[
```{r US_states_cluster_scatter, eval = FALSE}
# cut dendrogram so there are 4 clusters
cluster <- cutree(hc_out, k = 4)

US_state_stats %>%
  left_join(
    tibble(
      state = names(cluster),
      cluster = factor(cluster)
    )
  ) %>%
  ggplot(aes(income, tr_deaths)) +
  geom_point(aes(color = cluster))
```
]


.pull-right[
```{r US_states_cluster_scatter-out, ref.label="US_states_cluster_scatter", fig.width = 5.5, fig.asp = .618, echo=FALSE, warning=FALSE, message = FALSE, dev = "svg"}
```
]

[//]: # "segment ends here"

---

## Further reading

- Wikipedia: [Hierarchical clustering](https://en.wikipedia.org/wiki/Hierarchical_clustering)

- **ggdendro** documentation: [Introduction to ggdendro](https://cran.r-project.org/web/packages/ggdendro/vignettes/ggdendro.html)

- **ggdendro** documentation: [reference documentation (pdf)](https://cran.r-project.org/web/packages/ggdendro/ggdendro.pdf)




