---
title: "Interactive plots"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE, message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(colorspace)
library(colorblindr)
library(cowplot)
library(ggrepel)
library(ggiraph)
library(lubridate)
library(sf)
library(here)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2.0)) # 2 for full width, 2.5 for half width

# US_states dataset
US_states <- readRDS(here("datasets", "US_states.rds"))

# tx_counties dataset
US_census <- read_csv(here("datasets", "US_census.csv"))

#tx_counties <- US_census %>% 
#  filter(state == "Texas") %>%
#  select(name, pop2010) %>%
#  extract(name, "county", regex = "(.+) County")

texas_income <- readRDS(url("https://wilkelab.org/SDS375/datasets/Texas_income.rds")) %>%
  select(-county)

tx_census <- US_census %>% 
  filter(state == "Texas") %>%
  select(FIPS, name, pop2010, area) %>%
  extract(name, "county", regex = "(.+) County") %>%
  mutate(
    FIPS = as.character(FIPS),
    popratio = pop2010/median(pop2010),
  )

tx_counties <- left_join(texas_income, tx_census, by = "FIPS")

# CSS for the standard tool tips we'll use
tooltip_css <- "font-size:20px;padding:2px 4px 2px 4px;background:black;color:white;border-radius:2px 2px 2px 2px;"

```

## Interactivity can make a plot more informative

.center[
```{r iris-scatter-three-shapes, echo = FALSE}
breaks <- c("setosa", "virginica", "versicolor")
labels <- paste0("Iris ", breaks)

iris_scatter <- iris %>%
  mutate(
    tooltip = glue::glue("Sepal length: {Sepal.Length}; Sepal width: {Sepal.Width}")
  ) %>%
  ggplot(
    aes(Sepal.Length, Sepal.Width, shape = Species, fill = Species, color = Species)
  ) +     
  scale_shape_manual(
    values = c(21, 22, 23),
    breaks = breaks,
    labels = labels,
    name = NULL
  ) +
  scale_color_manual(
    values = darken(c("#56B4E9", "#E69F00", "#009E73"), 0.3),
    breaks = breaks,
    labels = labels,
    name = NULL
  ) +
  scale_fill_manual(
    values = c("#56B4E980", "#E69F0080", "#009E7380"),
    breaks = breaks,
    labels = labels,
    name = NULL
  ) +
  scale_x_continuous(
    limits = c(3.95, 8.2), expand = c(0, 0),
    labels = c("4.0", "5.0", "6.0", "7.0", "8.0"),
    name = "Sepal length"
  ) +
  scale_y_continuous(
    limits = c(1.9, 4.6), expand = c(0, 0),
    name = "Sepal width"
  ) + 
  geom_point_interactive(
    aes(tooltip = tooltip),
    size=2.5, stroke = 0.5,
    position = position_jitter(
      width = 0.01 * diff(range(iris$Sepal.Length)),
      height = 0.01 * diff(range(iris$Sepal.Width)),
      seed = 3942
    )
  ) +
  theme_minimal_grid() +
  theme(
    legend.title.align = 0.5,
    legend.text = element_text(face = "italic"),
    legend.spacing.y = unit(3.5, "pt"),
    plot.margin = margin(7, 7, 3, 1.5)
  )

girafe(
  ggobj = iris_scatter,
  width_svg = 6,
  height_svg = 6*.618,
  options = list(
    opts_tooltip(css = tooltip_css, delay_mouseover = 0, delay_mouseout = 0),
    opts_sizing(rescale = TRUE, width = 0.8)
  )
)
```
]

.absolute-bottom-right.small-font[
hover over the data points
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## You probably want to highlight the selected points ...

.center[
```{r iris-scatter-three-shapes-hover, echo = FALSE}
breaks <- c("setosa", "virginica", "versicolor")
labels <- paste0("Iris ", breaks)

iris_scatter <- iris %>%
  mutate(
    tooltip = glue::glue("Sepal length: {Sepal.Length}; Sepal width: {Sepal.Width}")
  ) %>%
  ggplot(
    aes(Sepal.Length, Sepal.Width, shape = Species, fill = Species, color = Species)
  ) +     
  scale_shape_manual(
    values = c(21, 22, 23),
    breaks = breaks,
    labels = labels,
    name = NULL
  ) +
  scale_color_manual(
    values = darken(c("#56B4E9", "#E69F00", "#009E73"), 0.3),
    breaks = breaks,
    labels = labels,
    name = NULL
  ) +
  scale_fill_manual(
    values = c("#56B4E980", "#E69F0080", "#009E7380"),
    breaks = breaks,
    labels = labels,
    name = NULL
  ) +
  scale_x_continuous(
    limits = c(3.95, 8.2), expand = c(0, 0),
    labels = c("4.0", "5.0", "6.0", "7.0", "8.0"),
    name = "Sepal length"
  ) +
  scale_y_continuous(
    limits = c(1.9, 4.6), expand = c(0, 0),
    name = "Sepal width"
  ) + 
  geom_point_interactive(
    aes(data_id = tooltip, tooltip = tooltip),
    size=2.5, stroke = 0.5,
    position = position_jitter(
      width = 0.01 * diff(range(iris$Sepal.Length)),
      height = 0.01 * diff(range(iris$Sepal.Width)),
      seed = 3942
    )
  ) +
  theme_minimal_grid() +
  theme(
    legend.title.align = 0.5,
    legend.text = element_text(face = "italic"),
    legend.spacing.y = unit(3.5, "pt"),
    plot.margin = margin(7, 7, 3, 1.5)
  )

girafe(
  ggobj = iris_scatter,
  width_svg = 6,
  height_svg = 6*.618,
  options = list(
    opts_tooltip(css = tooltip_css, delay_mouseover = 0, delay_mouseout = 0),
    opts_hover(css = "stroke:#000000;fill-opacity:1.0;stroke-opacity:1.0"),
    opts_sizing(rescale = TRUE, width = 0.8)
  )
)
```
]

.absolute-bottom-right.small-font[
hover over the data points
]


???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## ... and/or de-emphasize all others

.center[
```{r iris-scatter-three-shapes-hover2, echo = FALSE}
breaks <- c("setosa", "virginica", "versicolor")
labels <- paste0("Iris ", breaks)

iris_scatter <- iris %>%
  mutate(
    tooltip = glue::glue("Sepal length: {Sepal.Length}; Sepal width: {Sepal.Width}")
  ) %>%
  ggplot(
    aes(Sepal.Length, Sepal.Width, shape = Species, fill = Species, color = Species)
  ) +     
  scale_shape_manual(
    values = c(21, 22, 23),
    breaks = breaks,
    labels = labels,
    name = NULL
  ) +
  scale_color_manual(
    values = darken(c("#56B4E9", "#E69F00", "#009E73"), 0.3),
    breaks = breaks,
    labels = labels,
    name = NULL
  ) +
  scale_fill_manual(
    values = c("#56B4E980", "#E69F0080", "#009E7380"),
    breaks = breaks,
    labels = labels,
    name = NULL
  ) +
  scale_x_continuous(
    limits = c(3.95, 8.2), expand = c(0, 0),
    labels = c("4.0", "5.0", "6.0", "7.0", "8.0"),
    name = "Sepal length"
  ) +
  scale_y_continuous(
    limits = c(1.9, 4.6), expand = c(0, 0),
    name = "Sepal width"
  ) + 
  geom_point_interactive(
    aes(data_id = Species, tooltip = tooltip),
    size=2.5, stroke = 0.5,
    position = position_jitter(
      width = 0.01 * diff(range(iris$Sepal.Length)),
      height = 0.01 * diff(range(iris$Sepal.Width)),
      seed = 3942
    )
  ) +
  theme_minimal_grid() +
  theme(
    legend.title.align = 0.5,
    legend.text = element_text(face = "italic"),
    legend.spacing.y = unit(3.5, "pt"),
    plot.margin = margin(7, 7, 3, 1.5)
  )

girafe(
  ggobj = iris_scatter,
  width_svg = 6,
  height_svg = 6*.618,
  options = list(
    opts_tooltip(
      css = tooltip_css,
      delay_mouseover = 0, delay_mouseout = 0
    ),
    opts_hover(
      css = "stroke:#000000;fill-opacity:1.0;stroke-opacity:1.0"
    ),
    opts_hover_inv(css = "opacity:0.5;"),
    opts_sizing(rescale = TRUE, width = 0.8)
  )
)
```
]

.absolute-bottom-right.small-font[
hover over the data points
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## You can combine interactivity with explicit labels

.center[
```{r texas-counties-log, echo = FALSE, message = FALSE, fig.width = 8, fig.asp = 0.6, dev = "svg"}
set.seed(3878)

tx_counties_labels <- tx_counties %>% 
  mutate(popratio = pop2010/median(pop2010)) %>%
  arrange(desc(popratio)) %>%
  mutate(
    index = 1:n(),
    label = ifelse(index <= 3 | index > n()-3 | runif(n()) < .04, county, ""),
    label_large = ifelse(index <= 6, county, "")
  )

tx_counties_plot <- ggplot(tx_counties_labels, aes(x = index, y = popratio)) +
  geom_hline(yintercept = 1, linetype = 2, color = "grey40") +
  geom_point_interactive(
    aes(tooltip = county, data_id = county),
    size = 0.5,
    color = "#0072B2"
  ) +
  geom_text_repel(
    aes(label = label),
    size = 10/.pt,
    point.padding = .4, color = "black",
    min.segment.length = 0,
    max.overlaps = 1000
  ) +
  scale_y_log10(
    breaks = c(.01, .1, 1, 10, 100),
    name = "population number / median",
    labels = expression(10^-2, 10^-1, 10^0, 10^1, 10^2)
  ) +
  scale_x_continuous(
    limits = c(.5, nrow(tx_counties) + .5), expand = c(0, 0),
    breaks = NULL, #c(1, 50*(1:5)),
    name = "Texas counties, from most to least populous"
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(14) +
  theme(axis.line = element_blank())

girafe(
  ggobj = tx_counties_plot,
  width_svg = 8,
  height_svg = 8*.6,
  options = list(
    opts_tooltip(css = tooltip_css, delay_mouseover = 0, delay_mouseout = 0),
    opts_hover(css = "stroke-width: 3; stroke: #E69F00;"),
    opts_sizing(rescale = TRUE, width = 0.8)
  )
)

```
]

.absolute-bottom-right.small-font[
hover over the data points
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Also consider linking to further info

.center[
```{r US_states_clickthrough, echo = FALSE}
US_map <- US_states %>%
  mutate(
    onclick = glue::glue(
'window.open("https://en.wikipedia.org/wiki/{name}")'
)
 ) %>%
 ggplot() +
  geom_sf_interactive(
    aes(data_id = name, tooltip = name, onclick = onclick),
    fill = "#EDBE8A", color = "black", size = 0.5/.pt
  ) +
  theme_void()

girafe(
  ggobj = US_map,
  width_svg = 9,
  height_svg = 9*.618,
  options = list(
    opts_tooltip(css = tooltip_css, delay_mouseover = 0, delay_mouseout = 0),
    opts_hover(css = "fill:#E69F00;"),
    opts_sizing(rescale = TRUE, width = 0.8)
  )
)

```
]

.absolute-bottom-right.small-font[
click on one of the states
]

---

## Highlight across two plots for added context

.center[
```{r tx_income_combined, echo = FALSE}
texas_scatter <- tx_counties %>%
  ggplot(aes(pop2010, median_income)) +
  geom_point_interactive(
    aes(
      tooltip = county,
      data_id = county
    ),
    na.rm = TRUE, size = 4,
    shape = 21,
    color = "#0072B2", fill = "#0072B280"
  ) +
  scale_y_continuous(
    name = "Median income (USD)",
    limits = c(21000, 97000),
    breaks = c(40000, 60000, 80000),
    labels = scales::dollar_format(),
    expand = c(0, 0)
  ) +
  scale_x_log10(
    name = "Number of inhabitants",
    limits = c(300, 4100000),
    breaks = c(1e2, 1e3, 1e4, 1e5, 1e6),
    labels = expression(10^2, 10^3, 10^4, 10^5, 10^6)
  ) +
  theme_minimal_grid()

texas_county_map <- tx_counties %>%
  mutate(
    onclick = glue::glue('window.open("https://en.wikipedia.org/wiki/{county} County, Texas")')
  ) %>%
  ggplot() +
  geom_sf_interactive(
    aes(
      tooltip = county,
      data_id = county,
      onclick = onclick
    ),
    fill = "#0072B240", color = "black", size = 0.5/.pt
  ) +
  coord_sf(crs = 3083) +
  theme_void()

girafe(
  ggobj = plot_grid(texas_scatter, texas_county_map, rel_widths = c(1.2, 1)),
  width_svg = 8,
  height_svg = 4,
  options = list(
    opts_tooltip(css = tooltip_css, delay_mouseover = 0, delay_mouseout = 0),
    opts_hover(css = "fill: #E69F00; fill-opacity: 1.0; stroke: #E69F00;")
  )
)
```
]

[//]: # "segment ends here"

---

class: center middle

## Interactive plots in R

---

## We can make plots interactive with the **ggiraph** package

.tiny-font.pull-left.width-50[
```{r iris-no-girafe-demo, eval = FALSE}
#

iris_scatter <- ggplot(iris) + 
  aes(
    Sepal.Length, Sepal.Width,
    color = Species
  ) +
  geom_point()

iris_scatter
```
]

.pull-right.move-up-1em[
```{r iris-no-girafe-demo-out, ref.label="iris-no-girafe-demo", echo=FALSE, warning=FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```

.small-font[
regular **ggplot2** plot:  
hovering does nothing
]
]

---

## We can make plots interactive with the **ggiraph** package

.tiny-font.pull-left.width-50[
```{r iris-girafe-demo, eval = FALSE}
library(ggiraph)                #<<

iris_scatter <- ggplot(iris) + 
  aes(
    Sepal.Length, Sepal.Width,
    color = Species
  ) +
  geom_point_interactive(      #<<
    aes(tooltip = Species)     #<<
  )

girafe(                        #<<
  ggobj = iris_scatter,        #<<
  width_svg = 6,               #<<
  height_svg = 6*0.618          #<<
)                              #<<
```
]

.pull-right[
```{r iris-girafe-demo-out, ref.label="iris-girafe-demo", echo=FALSE, warning=FALSE}
```

.small-font[
**ggiraph** version:  
hovering displays species names
]
]

---

## Styling happens via Cascading Style Sheets (CSS)

.tiny-font.pull-left.width-50[
```{r iris-girafe-demo2, eval = FALSE}
library(ggiraph)

iris_scatter <- ggplot(iris) + 
  aes(
    Sepal.Length, Sepal.Width,
    color = Species
  ) +
  geom_point_interactive(
    aes(tooltip = Species)
  )

girafe(
  ggobj = iris_scatter,
  width_svg = 6,
  height_svg = 6*0.618,
  options = list( #<<
    opts_tooltip( #<<
css = "background: #F5F5F5; color: #191970;" #<<
    ) #<<
  ) #<<
)
```
]

.pull-right[
```{r iris-girafe-demo2-out, ref.label="iris-girafe-demo2", echo=FALSE, warning=FALSE}
```

.small-font[
**ggiraph** version:  
hovering displays species names
]

]

---

## Select multiple points at once with `data_id` aesthetic

.tiny-font.pull-left.width-50[
```{r iris-girafe-demo3, eval = FALSE}
library(ggiraph)

iris_scatter <- ggplot(iris) + 
  aes(
    Sepal.Length, Sepal.Width,
    color = Species
  ) +
  geom_point_interactive(
    aes(data_id = Species), #<<
    size = 2
  )

girafe(
  ggobj = iris_scatter,
  width_svg = 6,
  height_svg = 6*0.618
)
```
]

.pull-right[
```{r iris-girafe-demo3-out, ref.label="iris-girafe-demo3", echo=FALSE, warning=FALSE}
```
]


---

## Select multiple points at once with `data_id` aesthetic

.tiny-font.pull-left.width-50[
```{r iris-girafe-demo4, eval = FALSE}
library(ggiraph)

iris_scatter <- ggplot(iris) + 
  aes(
    Sepal.Length, Sepal.Width,
    color = Species
  ) +
  geom_point_interactive(
    aes(data_id = Species),
    size = 2
  )

girafe(
  ggobj = iris_scatter,
  width_svg = 6,
  height_svg = 6*0.618,
  options = list(
    opts_hover(css = "fill: #202020;"),  #<<
    opts_hover_inv(css = "opacity: 0.2;")  #<<
  )
)
```
]

.pull-right[
```{r iris-girafe-demo4-out, ref.label="iris-girafe-demo4", echo=FALSE, warning=FALSE}
```

Again, styling via CSS

]

---

## Interactive map example

.tiny-font[
```{r eval = FALSE, echo = TRUE}
# load data
US_states <- readRDS(url("https://wilkelab.org/SDS375/datasets/US_states.rds"))
```
]

---

## Interactive map example

.tiny-font[
```{r eval = FALSE, echo = TRUE}
# load data
US_states <- readRDS(url("https://wilkelab.org/SDS375/datasets/US_states.rds"))

# print
US_states
```

```{r eval = TRUE, echo = FALSE}
US_states <- readRDS(url("https://wilkelab.org/SDS375/datasets/US_states.rds"))

print(US_states, width = 200)
```
]

---

## Interactive map example

.tiny-font.pull-left.width-50[
```{r US_states_highlight-demo, eval = FALSE}
US_map <- US_states %>%
  ggplot() +
  geom_sf_interactive( #<<
    aes(data_id = name, tooltip = name)  #<<
  ) + #<<
  theme_void()

girafe(
  ggobj = US_map,
  width_svg = 6,
  height_svg = 6*0.618
)

```
]

.pull-right.width-50[
```{r US_states_clickthrough-demo-out, ref.label="US_states_highlight-demo", echo=FALSE, warning=FALSE}
```
]

---

## Interactive map example

.tiny-font.pull-left.width-50[
```{r US_states_clickthrough-demo, eval = FALSE}
US_map <- US_states %>%
  mutate( # JavaScript call to open website 
    onclick = glue::glue( #<<
'window.open(
"https://en.wikipedia.org/wiki/{name}")')
  ) %>%
  ggplot() +
  geom_sf_interactive(
    aes(
      data_id = name, tooltip = name,
      onclick = onclick  #<<
    )
  ) +
  theme_void()

girafe(
  ggobj = US_map,
  width_svg = 6,
  height_svg = 6*0.618
)

```
]

.pull-right.width-50[
```{r US_states_clickthrough-demo-out, ref.label="US_states_clickthrough-demo", echo=FALSE, warning=FALSE}
```

.small-font[
<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Click to open a state's wikipedia page
]
]

[//]: # "segment ends here"

---
class: center middle

## A (very) brief intro to CSS

---

## A (very) brief intro to CSS

--

CSS is the language used to style web pages

--

Many online tutorials, e.g.: https://www.w3schools.com/css/default.asp


[//]: # "segment ends here"

---

## Further reading

- **ggiraph** documentation: [Using **ggiraph**](https://davidgohel.github.io/ggiraph/articles/offcran/using_ggiraph.html)
- **ggiraph** documentation: [Function reference](https://davidgohel.github.io/ggiraph/reference/index.html)
- MDN Web Docs: [Learn to style HTML using CSS
](https://developer.mozilla.org/en-US/docs/Learn/CSS)
- w3schools.com: [CSS Tutorial](https://www.w3schools.com/css/default.asp)