---
title: "Getting to know your data 1"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(here)
library(cowplot)

#NCbirths <- read_csv(here("datasets", "NCbirths.csv"), na = c("", "NA", "#N/A")) %>%
#  mutate(
#    Weeks = if_else(Weeks < 0, NA_real_, Weeks)
#  )

# increase number of characters per line
options(width = 74)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width
```

## Any data analysis should start with basic quality control

1. Assess data relevance<br>
  (Can the data answer your question in principle?)

--

2. Assess data providence<br>
  (Where does the data come from? Is it reliable?)

--

3. Clean data if required<br>
  (In practice, this is 70-80% of data analysis work)
  
--

4. Perform descriptive analysis and sanity checks<br>
  (Inspect summary statistics, distributions, scatter plots, etc.)

---

## Components of data quality

.pull-left.width-35[
- Accuracy

- Precision

- Usability

- Completeness

- Consistency

- Accessibility

- Relevance
]

---

## Components of data quality

.pull-left.width-35[
- **Accuracy**

- Precision

- Usability

- Completeness

- Consistency

- Accessibility

- Relevance
]

.pull-right.width-65[
Does the data contain any systematic errors or biases?
]

---

## Components of data quality

.pull-left.width-35[
- Accuracy

- **Precision**

- Usability

- Completeness

- Consistency

- Accessibility

- Relevance
]

.pull-right.width-65[
Does the data contain any systematic errors or biases?

How affected are observations by random errors/noise?
]

---

## Accuracy means low bias, precision means low noise

<img src = "know-your-data-1_files/accuracy-precision.svg", width = 52%, style = "position: absolute;  top: 21%; left: 21%;"></img>


---

## Components of data quality

.pull-left.width-35[
- Accuracy

- Precision

- **Usability**

- Completeness

- Consistency

- Accessibility

- Relevance
]

.pull-right.width-65[
Does the data contain any systematic errors or biases?

How affected are observations by random errors/noise?

Is the data documented? Is it machine-readable?
]

---

## Usability requires a data dictionary

**Data Dictionary**<br>
A "centralized repository of information about data such as meaning, relationships to other data, origin, usage, and format" ([Wikipedia](https://en.wikipedia.org/wiki/Data_dictionary))

<br>
Example: [Palmer penguins reference page](https://allisonhorst.github.io/palmerpenguins/reference/penguins.html)

---

## Components of data quality

.pull-left.width-35[
- Accuracy

- Precision

- Usability

- **Completeness**

- Consistency

- Accessibility

- Relevance
]

.pull-right.width-65[
Does the data contain any systematic errors or biases?

How affected are observations by random errors/noise?

Is the data documented? Is it machine-readable?

Are entire records missing? Are observations missing?
]

---

## Components of data quality

.pull-left.width-35[
- Accuracy

- Precision

- Usability

- Completeness

- **Consistency**

- Accessibility

- Relevance
]

.pull-right.width-65[
Does the data contain any systematic errors or biases?

How affected are observations by random errors/noise?

Is the data documented? Is it machine-readable?

Are entire records missing? Are observations missing?

Does data vary among sources or over time?
]

---

## Components of data quality

.pull-left.width-35[
- Accuracy

- Precision

- Usability

- Completeness

- Consistency

- **Accessibility**

- Relevance
]

.pull-right.width-65[
Does the data contain any systematic errors or biases?

How affected are observations by random errors/noise?

Is the data documented? Is it machine-readable?

Are entire records missing? Are observations missing?

Does data vary among sources or over time?

Do you have access to the data? Can you obtain it?
]

---

## Components of data quality

.pull-left.width-35[
- Accuracy

- Precision

- Usability

- Completeness

- Consistency

- Accessibility

- **Relevance**
]

.pull-right.width-65[
Does the data contain any systematic errors or biases?

How affected are observations by random errors/noise?

Is the data documented? Is it machine-readable?

Are entire records missing? Are observations missing?

Does data vary among sources or over time?

Do you have access to the data? Can you obtain it?

Can you use the data to address your question?
]

[//]: # "segment ends here"


---
class: center middle

## Inspecting and cleaning a dataset in R

---

```{css echo = FALSE}
.move-up-8em {
  margin-top: -8em;
}
```

## The dataset: Births in NC

--

.pull-left.small-font.width-45.move-up-1em[
**Details and Source**<br>
This dataset contains data on a sample of 1450 birth records from 2001 that statistician John Holcomb at Cleveland State University selected from the North Carolina State Center for Health and Environmental Statistics.

The dataset has 1450 observations on 15 variables.
]

--

.pull-right.xtiny-font.move-up-8em.width-50[
Variable       | Description
----------:    | :-----------
`ID`           | Patient ID code
`Plural`       | `1`=single birth, `2`=twins, `3`=triplets
`Sex`          | Sex of the baby: `1`=male `2`=female
`MomAge`	     | Mother's age (in years)
`Weeks`        | Completed weeks of gestation
`Marital`      | Marital status: `1`=married or `2`=not married
`RaceMom`	     | Mother's race: `1`=white, `2`=black, `3`=American Indian, `4`=Chinese, `5`=Japanese, `6`=Hawaiian, `7`=Filipino, or `8`=Other Asian or Pacific Islander
`HispMom`	     | Hispanic origin of mother: `C`=Cuban, `M`=Mexican, `N`=not Hispanic, `O`=Other Hispanic, `P`=Puerto Rico, `S`=Central/South America
`Gained`       | Weight gained during pregnancy (in pounds)
`Smoke`        | Smoker mom? `1`=yes or `0`=no
`BirthWeightOz` |	Birth weight in ounces
`BirthWeightGm` | Birth weight in grams
`Low`           |	Indicator for low birth weight: `1`=2500 grams or less
`Premie`       |	Indicator for premature birth: `1`=36 weeks or sooner
`MomRace`      |	Mother's race: `black`, `hispanic`, `other`, or `white`
]

---

## Start with a cursory inspection

```{r echo = FALSE}
options(width = 95)
```

.tiny-font[
```{r message = FALSE}
NCbirths <- read_csv("https://wilkelab.org/SDS375/datasets/NCbirths.csv")
```
]

--

.tiny-font[
```{r}
NCbirths
```
]

```{r echo = FALSE}
options(width = 74)
```

---

## Start with a cursory inspection

What's going on with the `Gained` column?

```{r echo = FALSE}
options(width = 95)
```

.tiny-font[
```{r message = FALSE}
NCbirths %>%
  pull(Gained)
```
]

```{r echo = FALSE}
options(width = 74)
```

---

## Use the `na` argument to specify allowed `NA` values

```{r echo = FALSE}
options(width = 95)
```

.tiny-font[
```{r message = FALSE}
NCbirths <- read_csv(
  "https://wilkelab.org/DSC385/datasets/NCbirths.csv",
  na = c("", "NA", "#N/A")
)
```
]

--

.tiny-font[
```{r}
NCbirths  # check entire data frame
```
]

```{r echo = FALSE}
options(width = 74)
```

---

## Use the `na` argument to specify allowed `NA` values

```{r echo = FALSE}
options(width = 95)
```

.tiny-font[
```{r eval = FALSE}
NCbirths <- read_csv(
  "https://wilkelab.org/DSC385/datasets/NCbirths.csv",
  na = c("", "NA", "#N/A")
)
```
]

.tiny-font[
```{r message = FALSE}
NCbirths %>%   # check just the column `Gained`
  pull(Gained)
```
]


```{r echo = FALSE}
options(width = 74)
```


---

## Detailed quality checks: Look at all column names

.small-font[
```{r}
names(NCbirths)
```
]

--

What's the difference between `RaceMom` and `MomRace`?

---

## Detailed quality checks: Look at all column names

From the data dictionary:

- `RaceMom`<br>Mother's race: `1`=white, `2`=black, `3`=American Indian, `4`=Chinese,
`5`=Japanese, `6`=Hawaiian, `7`=Filipino, or `8`=Other Asian or Pacific Islander

- `MomRace`<br>Mother's race: `black`, `hispanic`, `other`, or `white`

--

These are similar but not the same. Important not to confuse them!

---

## Detailed quality checks: Look at every data column

.tiny-font.pull-left.width-50[
```{r}
summary(NCbirths$ID) # five number summary
```
]
---

## Detailed quality checks: Look at every data column

.tiny-font.pull-left.width-50[
```{r}
summary(NCbirths$ID) # five number summary
```

```{r echo = FALSE}
options(width = 50)
```

```{r}
unique(NCbirths$ID)  # all unique values
```

```{r echo = FALSE}
options(width = 74)
```
]

--

.xtiny-font.pull-right[
```{r ID-histogram, fig.width=6., fig.asp = 0.618, dev = "svg"}
ggplot(NCbirths, aes(ID)) +
  geom_histogram() # histogram
```
]

---

## Detailed quality checks: Look at every data column

.tiny-font.pull-left.width-50[
```{r}
summary(NCbirths$Plural) # five number summary
```
]

---

## Detailed quality checks: Look at every data column

.tiny-font.pull-left.width-50[
```{r}
summary(NCbirths$Plural) # five number summary
```

```{r}
unique(NCbirths$Plural)  # all unique values
```
]

--

.xtiny-font.pull-right[
```{r plural-births-histogram, fig.width=6., fig.asp = 0.618, dev = "svg"}
ggplot(NCbirths, aes(Plural)) +
  geom_histogram() # histogram
```
]


---

## Detailed quality checks: Look at every data column

```{r echo = FALSE}
options(width = 50)
```

.tiny-font.pull-left.width-50[
```{r}
summary(NCbirths$Weeks) # five number summary
```
]

---

## Detailed quality checks: Look at every data column

.tiny-font.pull-left.width-50[
```{r}
summary(NCbirths$Weeks) # five number summary
```

```{r}
unique(NCbirths$Weeks)  # all unique values
```

```{r echo = FALSE}
options(width = 74)
```
]

--

.xtiny-font.pull-right[
```{r weeks-histogram, fig.width=6., fig.asp = 0.618, dev = "svg"}
ggplot(NCbirths, aes(Weeks)) +
  geom_histogram() # histogram
```
]

--

What is the meaning of -999 weeks?

---

## In some cases, nonsensical values indicate missingness

--

We need to make sure to catch those cases:

.tiny-font[
```{r message = FALSE}
library(naniar) # library for handling missing values

NCbirths <- read_csv(
  "https://wilkelab.org/DSC385/datasets/NCbirths.csv",
  na = c("", "NA", "#N/A")
) %>% replace_with_na_at("Weeks", ~.x < 0) # only change column "Weeks" to be safe
```
]


--

.tiny-font[
```{r}
summary(NCbirths$Weeks)
```
]

--

.tiny-font[
```{r}
unique(NCbirths$Weeks)
```
]



---

## In some cases, nonsensical values indicate missingness

.pull-left[
Before recoding -999 to `NA`:
.xtiny-font[
```{r eval = FALSE}
ggplot(NCbirths, aes(Weeks)) +
  geom_histogram()
```

```{r weeks-histo-flawed, echo = FALSE, message = FALSE, warning = FALSE, fig.width=6., fig.asp = 0.618, dev = "svg"}
read_csv("https://wilkelab.org/DSC385/datasets/NCbirths.csv") %>%
  ggplot(aes(Weeks)) +
  geom_histogram()
```
]]

.pull-right[
After recoding -999 to `NA`:
.xtiny-font[
```{r weeks-histo-fixed, message = FALSE, warning = FALSE, fig.width=6., fig.asp = 0.618, dev = "svg"}
ggplot(NCbirths, aes(Weeks)) +
  geom_histogram()
```
]]


---

## Consider recoding categorical variables

Sex of the baby is encoded as 1/2:

.small-font[
```{r}
NCbirths %>%
  pull(Sex)
```
]

---

## Consider recoding categorical variables

Recode as `"male"`/`"female"`:

.small-font[
```{r}
NCbirths %>%
  mutate(
    Sex = if_else(Sex == 1, "male", "female")
  ) %>%
  pull(Sex)
```
]

---

## Recode multiple categories with `case_when()`

The mother's race is encoded as integers from 1 through 8:

.small-font[
```{r}
NCbirths %>%
  pull(RaceMom)
```
]

---

## Recode multiple categories with `case_when()`

The mother's race is encoded as integers from 1 through 8:

.small-font[
```{r}
NCbirths %>%
  pull(RaceMom) %>%
  unique() # list all the unique values (ordered by first occurrence)
```
]

---

## Recode multiple categories with `case_when()`

```{r echo = FALSE}
options(width = 60)
```

We use `case_when()` to recode multiple values all at once:

.xtiny-font.pull-left[
```{r eval = FALSE}
NCbirths %>%
  mutate(
    RaceMom = case_when(
      RaceMom == 1 ~ "white",
      RaceMom == 2 ~ "black",
      RaceMom == 3 ~ "American Indian",
      RaceMom == 4 ~ "Chinese",
      RaceMom == 5 ~ "Japanese",
      RaceMom == 6 ~ "Hawaiian",
      RaceMom == 7 ~ "Filipino",
      RaceMom == 8 ~ 
        "Other Asian or Pacific Islander",
      TRUE ~ NA_character_ # should never reach
    )
  )
```
]

```{r echo = FALSE}
options(width = 74)
```


---

## Recode multiple categories with `case_when()`


We use `case_when()` to recode multiple values all at once:

.xtiny-font.pull-left[
```{r case-when-example, eval = FALSE}
NCbirths %>%
  mutate(
    RaceMom = case_when(
      RaceMom == 1 ~ "white",
      RaceMom == 2 ~ "black",
      RaceMom == 3 ~ "American Indian",
      RaceMom == 4 ~ "Chinese",
      RaceMom == 5 ~ "Japanese",
      RaceMom == 6 ~ "Hawaiian",
      RaceMom == 7 ~ "Filipino",
      RaceMom == 8 ~ 
        "Other Asian or Pacific Islander",
      TRUE ~ NA_character_ # should never reach
    )
  ) %>%
  pull(RaceMom)
```
]

.xtiny-font.pull-right.width-50[
```{r echo=FALSE}
options(width = 60)

NCbirths %>%
  mutate(
    RaceMom = case_when(
      RaceMom == 1 ~ "white",
      RaceMom == 2 ~ "black",
      RaceMom == 3 ~ "American I",
      RaceMom == 4 ~ "Chinese",
      RaceMom == 5 ~ "Japanese",
      RaceMom == 6 ~ "Hawaiian",
      RaceMom == 7 ~ "Filipino",
      RaceMom == 8 ~ "Other A/PI",
      TRUE ~ NA_character_ # should never reach
    )
  ) %>%
  pull(RaceMom)
```
]

```{r echo = FALSE}
options(width = 74)
```

---

## Make pairwise scatter plots

--

```{r echo = FALSE}
NCbirths <- NCbirths %>%
  mutate(
    Sex = if_else(Sex == 1, "male", "female")
  )
```

.tiny-font[
```{r NCbirths-pairs, eval = FALSE}
ggplot(NCbirths, aes(Plural, Sex, color = Sex)) + geom_point(position = "jitter", size = .5)
ggplot(NCbirths, aes(Plural, MomAge, color = Sex)) + geom_point(position = "jitter", size = .5)
```
]

.center[
```{r NCbirths-pairs-out, ref.label='NCbirths-pairs', echo = FALSE, warning = FALSE, fig.width=4.8, fig.asp = .618, dev = "svg"}
```
]

---

## Make pairwise scatter plots

.tiny-font[
```{r NCbirths-pairs2, eval = FALSE}
ggplot(NCbirths, aes(Weeks, Plural, color = Sex)) + geom_point(position = "jitter", size = .5)
ggplot(NCbirths, aes(Weeks, Gained, color = Sex)) + geom_point(position = "jitter", size = .5)
```
]

.center[
```{r NCbirths-pairs2-out, ref.label='NCbirths-pairs2', echo = FALSE, warning = FALSE, fig.width=4.8, fig.asp = .618, dev = "svg"}
```
]

--

If possible, make scatter plots of every variable against every other variable


[//]: # "segment ends here"


---

## Further reading

- Wikipedia: [Data dictionary](https://en.wikipedia.org/wiki/Data_dictionary)
- **dplyr** reference documentation: [if_else()](https://dplyr.tidyverse.org/reference/if_else.html)
- **dplyr** reference documentation: [case_when()](https://dplyr.tidyverse.org/reference/case_when.html)
- **naniar** documentation: [Getting started with naniar](http://naniar.njtierney.com/articles/getting-started-w-naniar.html)
- **naniar** reference documentation: [replace_with_na_at()](http://naniar.njtierney.com/reference/replace_with_na_at.html)


