---
title: "Dealing with issues of overplotting"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
params:
  cache: true
---

```{r setup, include=FALSE, echo=FALSE, message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(cowplot)
library(colorspace)
library(here)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2.0)) # 2 for full width, 2.5 for half width

blue_jays <- read_csv(here("datasets", "blue_jays.csv"))
```

## Be aware of points plotted exactly on top of one another

.center[
```{r  mpg-cty-displ-solid, cache = params$cache, echo = FALSE, message = FALSE, fig.width = 6, fig.asp=.7416, dev = "svg"}
p_mpg_solid <- ggplot(mpg, aes(y = cty, x = displ, color = drv, fill = drv)) +
  geom_point(size = 3, shape = 21) + 
  scale_x_continuous(
    name = "displacement (l)",
    limits = c(1.4, 7.1),
    breaks = 2:7
  ) +
  scale_y_continuous(
    name = "fuel economy (mpg)",
    limits = c(5.3, 37),
    breaks = 5*(2:7),
    expand = c(0, 0)
  ) +
  scale_color_manual(
    values = c("#202020", "#E69F00", "#56B4E9"), 
    name = "drive train",
    breaks = c("f", "r", "4"),
    labels = c("FWD", "RWD", "4WD")
  ) +
  scale_fill_manual(
    values = c("#202020", "#E69F00", "#56B4E9"), 
    name = "drive train",
    breaks = c("f", "r", "4"),
    labels = c("FWD", "RWD", "4WD")
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.7, .8),
    plot.margin = margin(3, 7, 3, 1.5)
  )

stamp_bad(p_mpg_solid)
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

--

Technical term for this problem: overplotting

---

## Partial transparency helps highlight overlapping points

.center[
```{r mpg-cty-displ-transp, cache = params$cache, echo = FALSE, message = FALSE, fig.width = 6, fig.asp=.7416, dev = "svg"}
ggplot(mpg, aes(y = cty, x = displ, color = drv, fill = drv)) +
  geom_point(size = 3, shape = 21) + 
  scale_x_continuous(
    name = "displacement (l)",
    limits = c(1.4, 7.1),
    breaks = 2:7
  ) +
  scale_y_continuous(
    name = "fuel economy (mpg)",
    limits = c(5.3, 37),
    breaks = 5*(2:7),
    expand = c(0, 0)
  ) +
  scale_color_manual(
    values = c("#202020", "#E69F00", "#56B4E9"), 
    name = "drive train",
    breaks = c("f", "r", "4"),
    labels = c("FWD", "RWD", "4WD")
  ) +
  scale_fill_manual(
    values = c("#20202080", "#E69F0080", "#56B4E980"),
    name = "drive train",
    breaks = c("f", "r", "4"),
    labels = c("FWD", "RWD", "4WD")
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.7, .8),
    plot.margin = margin(3, 7, 3, 1.5)
  )
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## A little jitter shows overlaps even more clearly

.center[
```{r mpg-cty-displ-jitter, cache = params$cache, echo = FALSE, message = FALSE, fig.width = 6, fig.asp=.7416, dev = "svg"}
ggplot(mpg, aes(y = cty, x = displ, color = drv, fill = drv)) +
  geom_point(
    size = 3, shape = 21,
    position = position_jitter(
      width = 0.01 * diff(range(mpg$displ)),
      height = 0.01 * diff(range(mpg$cty)),
      seed = 7384
    )
  ) + 
  scale_x_continuous(
    name = "displacement (l)",
    limits = c(1.4, 7.1),
    breaks = 2:7
  ) +
  scale_y_continuous(
    name = "fuel economy (mpg)",
    limits = c(5.3, 37),
    breaks = 5*(2:7),
    expand = c(0, 0)
  ) +
  scale_color_manual(
    values = c("#202020", "#E69F00", "#56B4E9"), 
    name = "drive train",
    breaks = c("f", "r", "4"),
    labels = c("FWD", "RWD", "4WD")
  ) +
  scale_fill_manual(
    values = c("#20202080", "#E69F0080", "#56B4E980"),
    name = "drive train",
    breaks = c("f", "r", "4"),
    labels = c("FWD", "RWD", "4WD")
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.7, .8),
    plot.margin = margin(3, 7, 3, 1.5)
  )
```
]


???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## But don't jitter too much

.center[
```{r mpg-cty-displ-jitter-extreme, cache = params$cache, echo = FALSE, message = FALSE, fig.width = 6, fig.asp=.7416, dev = "svg"}
p_mpg_jitter_extreme <- ggplot(mpg, aes(y = cty, x = displ, color = drv, fill = drv)) +
  geom_point(
    size = 3, shape = 21,
    position = position_jitter(
      width = 0.1 * diff(range(mpg$displ)),
      height = 0.1 * diff(range(mpg$cty)),
      seed = 7384
    )
  ) + 
  scale_x_continuous(
    name = "displacement (l)",
    limits = c(1.4, 7.1),
    breaks = 2:7
  ) +
  scale_y_continuous(
    name = "fuel economy (mpg)",
    limits = c(5.3, 37),
    breaks = 5*(2:7),
    expand = c(0, 0)
  ) +
  scale_color_manual(
    values = c("#202020", "#E69F00", "#56B4E9"), 
    name = "drive train",
    breaks = c("f", "r", "4"),
    labels = c("FWD", "RWD", "4WD")
  ) +
  scale_fill_manual(
    values = c("#20202080", "#E69F0080", "#56B4E980"),
    name = "drive train",
    breaks = c("f", "r", "4"),
    labels = c("FWD", "RWD", "4WD")
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.7, .8),
    plot.margin = margin(3, 7, 3, 1.5)
  )

stamp_bad(p_mpg_jitter_extreme)
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---
class: center middle

## 2D density plots: Contour lines and contour bands

---

## Contour lines are the 2D version of density plots

.center[
```{r blue-jays-contour, cache = params$cache, echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
blue_jays_base <- ggplot(blue_jays, aes(body_mass_g, head_length_mm)) + 
  scale_x_continuous(
    limits = c(57, 82),
    expand = c(0, 0),
    name = "body mass (g)") +
  scale_y_continuous(
    limits = c(49, 61),
    expand = c(0, 0),
    name = "head length (mm)"
  ) +
  theme_minimal_grid(16) +
  theme(
    legend.position = c(1, 0),
    legend.justification = c(1, 0),
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    axis.ticks.length = unit(0, "pt"),
    axis.ticks = element_blank()
  )

blue_jays_base + 
  geom_density_2d(color = "black", size = 0.3, binwidth = 0.004) +
  geom_point(color = "#202020", size = 2, alpha = 0.7)
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## We can vary shading for added effect

.center[
```{r blue-jays-contour-filled, cache = params$cache, echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
blue_jays_base + 
  geom_density_2d_filled(color = NA, binwidth = 0.004) +
  geom_density_2d(color = "black", size = 0.3, binwidth = 0.004) +
  geom_point(color = "#202020", size = 2, alpha = 0.7) +
  scale_fill_manual(
    values = scales::colour_ramp(c("#F2F2F200", "#80808080"))((0:5)/5),
    guide = "none"
  )
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## What do we do when there are multiple groups?

.center[
```{r blue-jays-no-contour-by-sex, cache = params$cache, echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
blue_jays_base + 
  aes(color = sex) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 2, linetype = 0)
    )
  )
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Colored contour lines can work for 2 to 3 groups

.center[
```{r blue-jays-contour-by-sex, cache = params$cache, echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
blue_jays_base + 
  aes(color = sex) +
  geom_density_2d(size = 0.3, binwidth = 0.006) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 2, linetype = 0)
    )
  ) +
  theme_minimal_grid(16) +
  theme(
    legend.position = c(1, 0),
    legend.justification = c(1, 0),
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    axis.ticks.length = unit(0, "pt"),
    axis.ticks = element_blank()
  )
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## What if there are multiple groups intermingled?

.center[
```{r diamonds-points, cache = params$cache, echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
set.seed(6932)
diamonds_small <- diamonds[sample(1:nrow(diamonds), 6000), ]

diamonds_plot_base <- ggplot(diamonds_small, aes(carat, price, color = cut)) + 
  scale_x_continuous(
    limits = c(-.5, 5)  # set limits larger than coords to not limit contour lines
  ) +
  scale_y_log10(
    name = "price (USD)",
    breaks = c(300, 1000, 3000, 10000),
    labels = c("$300", "$1,000", "$3,000", "$10,000"),
    limits = c(240, 25000)
  ) +
  scale_color_discrete_sequential(
    palette = "Inferno",
    nmax = 6,
    order = 1:5,
    rev = FALSE,
    breaks = c("Ideal", "Premium", "Very Good", "Good", "Fair"),
    labels = c("ideal", "premium", "very good", "good", "fair"),
    guide = guide_legend(
      override.aes = list(size = 2, alpha = 1)
    )
  ) +
  coord_cartesian(
    xlim = c(-.1, 3.2),
    ylim = c(240, 25000),
    expand = FALSE
  ) + 
  theme_minimal_grid() +
  panel_border() +
  theme(
    plot.margin = margin(18, 7, 1, 1.5),
    legend.key.width = unit(6, "pt"),
    legend.spacing.y = unit(3, "pt"),
    legend.title = element_text(hjust = 0, margin = margin(0, 0, 0, 0)),
    legend.position = c(.97, .3),
    legend.justification = c(1, 0.5),
    legend.box.margin = margin(7, 7, 7, 7),
    legend.box.background = element_rect(fill = "#FFFFFFD0", color = NA),
    axis.ticks.length = unit(0, "pt")
  )

diamonds_plot_points <- diamonds_plot_base +
  geom_point(size = .4, alpha = 1/5)
  
diamonds_plot_points
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Don't make plots that look like spaghetti

.center[
```{r diamonds-contours, cache = params$cache, echo = FALSE, fig.width = 6, fig.asp = 0.75, dev = "svg"}
diamonds_plot_contours <- diamonds_plot_base +
  geom_density2d(size = 0.3, binwidth = 0.35)
  
stamp_bad(diamonds_plot_contours)
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Contour lines work well with small multiples (facets)

.center[
```{r diamonds-contour-facets, echo = FALSE, cache = params$cache, fig.width = 8, fig.asp = 0.618, dev = "svg"}
diamonds_facet_base <- ggplot(diamonds_small, aes(carat, price)) + 
  scale_x_continuous(
    limits = c(-.5, 5)  # set limits larger than coords to not limit contour lines
  ) +
  scale_y_log10(
    name = "price (USD)",
    breaks = c(300, 1000, 3000, 10000),
    labels = c("$300", "$1,000", "$3,000", "$10,000"),
    limits = c(240, 25000)
  ) +
  coord_cartesian(
    xlim = c(-.1, 3.2),
    ylim = c(240, 25000),
    expand = FALSE
  ) + 
  facet_wrap(~cut, scales = "free_x", labeller = labeller(cut = tolower)) +
  theme_minimal_grid(12, rel_small = 1) +
  panel_border() +
  theme(
    legend.title = element_text(hjust = 0.5),
    legend.position = c(.95, .05),
    legend.justification = c(1, 0),
    axis.ticks.length = unit(0, "pt")
  )

diamonds_facet_base +
  geom_density2d_filled(binwidth = 0.4) +
  geom_density2d(color = darken("#0072B2", .2), size = .3, binwidth = 0.4) +
  scale_fill_manual(
    values = scales::colour_ramp(c("#ECF0F620", "#8498B3C0"))((0:9)/9),
    guide = "none"
  )

#> desaturate(lighten("#0072B2", .9), .6)
#[1] "#ECF0F6"
#> desaturate(lighten("#0072B2", .3), .6)
#[1] "#8498B3"
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---
class: center middle

## 2D histograms: Rectangular and hex bins

---

## 2D histograms: rectangular bins

.center[
```{r diamonds-bin2d-facets, echo = FALSE, cache = params$cache, fig.width = 8, fig.asp = 0.618, dev = "svg"}
diamonds_facet_base +
  geom_bin2d(binwidth = c(0.2, 0.1)) +
  scale_color_continuous_sequential(
    aesthetics = c("color", "fill"),
    palette = "SunsetDark",
    alpha = 0.7,
    limits = c(0, 280),
    breaks = c(0, 100, 200),
    guide = guide_colourbar(
      direction = "horizontal",
      title.position = "top",
      barwidth = grid::unit(1.8, "in"),
      barheight = grid::unit(0.2, "in"),
      ticks.colour = "black"
    )
  ) +
  theme(
    legend.position = c(0.97, .25),
    legend.justification = c(1, 0.5)
  )
```
]

--

.absolute-bottom-left[
We need to choose a bin size as in regular histograms
]


---

## 2D histograms: rectangular bins

.center[
```{r diamonds-bin2d-facets2, echo = FALSE, cache = params$cache, fig.width = 8, fig.asp = 0.618, dev = "svg"}
diamonds_facet_base +
  geom_bin2d(binwidth = c(0.1, 0.05)) +
  scale_color_continuous_sequential(
    aesthetics = c("color", "fill"),
    palette = "SunsetDark",
    alpha = 0.7,
    limits = c(0, 120),
    breaks = c(0, 25, 50, 75, 100),
    guide = guide_colourbar(
      direction = "horizontal",
      title.position = "top",
      barwidth = grid::unit(1.8, "in"),
      barheight = grid::unit(0.2, "in"),
      ticks.colour = "black"
    )
  ) +
  theme(
    legend.position = c(0.97, .25),
    legend.justification = c(1, 0.5)
  )
```
]

.absolute-bottom-left[
We need to choose a bin size as in regular histograms
]

---

## 2D histograms: hex bins

.center[
```{r diamonds-hexbin-facets, echo = FALSE, cache = params$cache, fig.width = 8, fig.asp = 0.618, dev = "svg"}
diamonds_facet_base +
  geom_hex(aes(colour = stat(count)), bins = 20) +
  scale_color_continuous_sequential(
    aesthetics = c("color", "fill"),
    palette = "SunsetDark",
    alpha = 0.7,
    limits = c(0, 280),
    breaks = c(0, 100, 200),
    guide = guide_colourbar(
      direction = "horizontal",
      title.position = "top",
      barwidth = grid::unit(1.8, "in"),
      barheight = grid::unit(0.2, "in"),
      ticks.colour = "black"
    )
  ) +
  theme(
    legend.position = c(0.97, .25),
    legend.justification = c(1, 0.5)
  )
```
]

---

## 2D histograms: hex bins

.center[
```{r diamonds-hexbin2-facets, echo = FALSE, cache = params$cache, fig.width = 8, fig.asp = 0.618, dev = "svg"}
diamonds_facet_base +
  geom_hex(aes(colour = stat(count)), bins = 50) +
  scale_color_continuous_sequential(
    aesthetics = c("color", "fill"),
    palette = "SunsetDark",
    alpha = 0.7,
    limits = c(0, 100),
    breaks = c(0, 25, 50, 75, 100),
    guide = guide_colourbar(
      direction = "horizontal",
      title.position = "top",
      barwidth = grid::unit(1.8, "in"),
      barheight = grid::unit(0.2, "in"),
      ticks.colour = "black"
    )
  ) +
  theme(
    legend.position = c(0.97, .25),
    legend.justification = c(1, 0.5)
  )
```
]


---

## Choosing the right color scale

.center[
![](`r knitr::fig_chunk("diamonds-hexbin2-facets", "svg")`)
]

.absolute-bottom-left[
palette: SunsetDark
]

---

## Choosing the right color scale

.center[
```{r diamonds-hexbin3-facets, echo = FALSE, cache = params$cache, fig.width = 8, fig.asp = 0.618, dev = "svg"}
diamonds_facet_base +
  geom_hex(aes(colour = stat(count)), bins = 50) +
  scale_color_continuous_sequential(
    aesthetics = c("color", "fill"),
    palette = "Batlow",
    alpha = 0.7,
    limits = c(0, 100),
    breaks = c(0, 25, 50, 75, 100),
    guide = guide_colourbar(
      direction = "horizontal",
      title.position = "top",
      barwidth = grid::unit(1.8, "in"),
      barheight = grid::unit(0.2, "in"),
      ticks.colour = "black"
    )
  ) +
  theme(
    legend.position = c(0.97, .25),
    legend.justification = c(1, 0.5)
  )
```
]

.absolute-bottom-left[
palette: Batlow
]

---

## Choosing the right color scale

.center[
```{r diamonds-hexbin4-facets, echo = FALSE, cache = params$cache, fig.width = 8, fig.asp = 0.618, dev = "svg"}
diamonds_facet_base +
  geom_hex(aes(colour = stat(count)), bins = 50) +
  scale_color_continuous_sequential(
    aesthetics = c("color", "fill"),
    palette = "YlOrRd",
    alpha = 0.7,
    limits = c(0, 100),
    breaks = c(0, 25, 50, 75, 100),
    guide = guide_colourbar(
      direction = "horizontal",
      title.position = "top",
      barwidth = grid::unit(1.8, "in"),
      barheight = grid::unit(0.2, "in"),
      ticks.colour = "black"
    )
  ) +
  theme(
    legend.position = c(0.97, .25),
    legend.justification = c(1, 0.5)
  )
```
]


.absolute-bottom-left[
palette: YlOrRd
]

---

## Choosing the right color scale

.center[
```{r diamonds-hexbin5-facets, echo = FALSE, cache = params$cache, fig.width = 8, fig.asp = 0.618, dev = "svg"}
diamonds_facet_base +
  geom_hex(aes(colour = stat(count)), bins = 50) +
  scale_color_continuous_sequential(
    aesthetics = c("color", "fill"),
    palette = "BluYl",
    alpha = 0.7,
    limits = c(0, 100),
    breaks = c(0, 25, 50, 75, 100),
    guide = guide_colourbar(
      direction = "horizontal",
      title.position = "top",
      barwidth = grid::unit(1.8, "in"),
      barheight = grid::unit(0.2, "in"),
      ticks.colour = "black"
    )
  ) +
  theme(
    legend.position = c(0.97, .25),
    legend.justification = c(1, 0.5)
  )
```
]


.absolute-bottom-left[
palette: BluYl
]

---

## Choosing the right color scale

.center[
```{r diamonds-hexbin6-facets, echo = FALSE, cache = params$cache, fig.width = 8, fig.asp = 0.618, dev = "svg"}
diamonds_facet_base +
  geom_hex(aes(colour = stat(count)), bins = 50) +
  scale_color_continuous_sequential(
    aesthetics = c("color", "fill"),
    palette = "Heat",
    alpha = 0.7,
    limits = c(0, 100),
    breaks = c(0, 25, 50, 75, 100),
    guide = guide_colourbar(
      direction = "horizontal",
      title.position = "top",
      barwidth = grid::unit(1.8, "in"),
      barheight = grid::unit(0.2, "in"),
      ticks.colour = "black"
    )
  ) +
  theme(
    legend.position = c(0.97, .25),
    legend.justification = c(1, 0.5)
  )
```
]

.absolute-bottom-left[
palette: Heat
]

---

## Choosing the right color scale

.center[
```{r diamonds-hexbin7-facets, echo = FALSE, cache = params$cache, fig.width = 8, fig.asp = 0.618, dev = "svg"}
diamonds_facet_base +
  geom_hex(aes(colour = stat(count)), bins = 50) +
  scale_color_continuous_sequential(
    aesthetics = c("color", "fill"),
    palette = "ag_GrnYl",
    alpha = 0.7,
    limits = c(0, 100),
    breaks = c(0, 25, 50, 75, 100),
    guide = guide_colourbar(
      direction = "horizontal",
      title.position = "top",
      barwidth = grid::unit(1.8, "in"),
      barheight = grid::unit(0.2, "in"),
      ticks.colour = "black"
    )
  ) +
  theme(
    legend.position = c(0.97, .25),
    legend.justification = c(1, 0.5)
  )
```
]


.absolute-bottom-left[
palette: ag_GrnYl
]

---

## Choosing the right color scale

.center[
![](`r knitr::fig_chunk("diamonds-hexbin2-facets", "svg")`)
]

.absolute-bottom-left[
palette: SunsetDark
]


[//]: # "segment ends here"

---
class: center middle

## Creating 2D density plots and histograms in R

---

## Contour lines

Data preparation:

.tiny-font[
```{r blue-jays-data}
blue_jays <- read_csv("https://wilkelab.org/SDS375/datasets/blue_jays.csv")
```
]

---

## Contour lines

.tiny-font.pull-left[
```{r scatter-demo, eval = FALSE}
blue_jays %>%
  ggplot(aes(body_mass_g, head_length_mm)) +
  geom_point() +
  theme_bw(14)
```
]

.pull-right[
```{r scatter-demo-out, cache = params$cache, ref.label="scatter-demo", fig.width=6., fig.asp = 0.75, echo=FALSE, dev = "svg"}
```
]



---

## Contour lines

.tiny-font.pull-left[
```{r contour-lines-demo, eval = FALSE}
blue_jays %>%
  ggplot(aes(body_mass_g, head_length_mm)) +
  geom_density_2d() +  #<<
  geom_point() +
  theme_bw(14)
```
]

.pull-right[
```{r contour-lines-demo-out, cache = params$cache, ref.label="contour-lines-demo", fig.width=6., fig.asp = 0.75, echo=FALSE, dev = "svg"}
```
]


---

## Contour lines

.tiny-font.pull-left[
```{r contour-lines-demo2, eval = FALSE}
blue_jays %>%
  ggplot(aes(body_mass_g, head_length_mm)) +
  geom_density_2d(bins = 5) + #<<
  geom_point() +
  theme_bw(14)
```
]

.pull-right[
```{r contour-lines-demo2-out, cache = params$cache, ref.label="contour-lines-demo2", fig.width=6., fig.asp = 0.75, echo=FALSE, dev = "svg"}
```
]


---

## Contour bands

.tiny-font.width-70.move-up-1em[
```{r contour-bands-demo, eval = FALSE}
ggplot(blue_jays, aes(body_mass_g, head_length_mm)) +
  geom_density_2d_filled(bins = 5) + #<<
  #
  geom_point() +
  theme_bw(14)
```
]

.center[
```{r contour-bands-demo-out, cache = params$cache, ref.label="contour-bands-demo", fig.width=7., fig.asp = 0.55, echo=FALSE, dev = "svg"}
```
]


---

## Contour bands

.tiny-font.width-70.move-up-1em[
```{r contour-bands-demo2, eval = FALSE}
ggplot(blue_jays, aes(body_mass_g, head_length_mm)) +
  geom_density_2d_filled(bins = 5, alpha = 0.5) + #<<
  #
  geom_point() +
  theme_bw(14)
```
]

.center[
```{r contour-bands-demo2-out, cache = params$cache, ref.label="contour-bands-demo2", fig.width=7., fig.asp = 0.55, echo=FALSE, dev = "svg"}
```
]


---

## Contour bands

.tiny-font.width-70.move-up-1em[
```{r contour-bands-demo3, eval = FALSE}
ggplot(blue_jays, aes(body_mass_g, head_length_mm)) +
  geom_density_2d_filled(bins = 5, alpha = 0.5) +
  geom_density_2d(bins = 5, color = "black", size = 0.2) + #<<
  geom_point() +
  theme_bw(14)
```
]

.center[
```{r contour-bands-demo3-out, cache = params$cache, ref.label="contour-bands-demo3", fig.width=7., fig.asp = 0.55, echo=FALSE, dev = "svg"}
```
]


---

## 2D histograms

.tiny-font.width-70.move-up-1em[
```{r bins2d-demo, eval = FALSE}
ggplot(blue_jays, aes(body_mass_g, head_length_mm)) +
  geom_bin2d() + #<<
  theme_bw(14)
```
]

.center[
```{r bins2d-demo-out, cache = params$cache, ref.label="bins2d-demo", fig.width=7., fig.asp = 0.618, echo=FALSE, dev = "svg"}
```
]

---

## 2D histograms

.tiny-font.width-70.move-up-1em[
```{r bins2d-demo2, eval = FALSE}
ggplot(blue_jays, aes(body_mass_g, head_length_mm)) +
  geom_bin2d(binwidth = c(3, 3)) + #<<
  theme_bw(14)
```
]

.center[
```{r bins2d-demo2-out, cache = params$cache, ref.label="bins2d-demo2", fig.width=7., fig.asp = 0.618, echo=FALSE, dev = "svg"}
```
]


---

## 2D histograms

.tiny-font.width-70.move-up-1em[
```{r bins2d-demo3, eval = FALSE}
ggplot(blue_jays, aes(body_mass_g, head_length_mm)) +
  geom_bin2d(binwidth = c(1, 5)) + #<<
  theme_bw(14)
```
]

.center[
```{r bins2d-demo3-out, cache = params$cache, ref.label="bins2d-demo3", fig.width=7., fig.asp = 0.618, echo=FALSE, dev = "svg"}
```
]


---

## 2D histograms

.tiny-font.width-70.move-up-1em[
```{r bins2d-demo4, eval = FALSE}
ggplot(blue_jays, aes(body_mass_g, head_length_mm)) +
  geom_bin2d(binwidth = c(5, 1)) + #<<
  theme_bw(14)
```
]

.center[
```{r bins2d-demo4-out, cache = params$cache, ref.label="bins2d-demo4", fig.width=7., fig.asp = 0.618, echo=FALSE, dev = "svg"}
```
]


---

## Hex bins

.tiny-font.width-70.move-up-1em[
```{r hex-demo, eval = FALSE}
ggplot(blue_jays, aes(body_mass_g, head_length_mm)) +
  geom_hex() + #<<
  theme_bw(14)
```
]

.center[
```{r hex-demo-out, cache = params$cache, ref.label="hex-demo", fig.width=7., fig.asp = 0.618, echo=FALSE, dev = "svg"}
```
]


---

## Hex bins

.tiny-font.width-70.move-up-1em[
```{r hex-demo2, eval = FALSE}
ggplot(blue_jays, aes(body_mass_g, head_length_mm)) +
  geom_hex(bins = 15) + #<<
  theme_bw(14)
```
]

.center[
```{r hex-demo2-out, cache = params$cache, ref.label="hex-demo2", fig.width=7., fig.asp = 0.618, echo=FALSE, dev = "svg"}
```
]


---

## Hex bins

.tiny-font.width-70.move-up-1em[
```{r hex-demo3, eval = FALSE}
ggplot(blue_jays, aes(body_mass_g, head_length_mm)) +
  geom_hex(bins = 10) + #<<
  theme_bw(14)
```
]

.center[
```{r hex-demo3-out, cache = params$cache, ref.label="hex-demo3", fig.width=7., fig.asp = 0.618, echo=FALSE, dev = "svg"}
```
]

[//]: # "segment ends here"

---

## Further reading

- Fundamentals of Data Visualization: [Chapter 18: Handling overlapping points](https://clauswilke.com/dataviz/overlapping-points.html)
- **ggplot2** reference documentation: [`geom_density_2d()`](https://ggplot2.tidyverse.org/reference/geom_density_2d.html)
- **ggplot2** reference documentation: [`geom_bin2d()`](https://ggplot2.tidyverse.org/reference/geom_bin2d.html)
- **ggplot2** reference documentation: [`geom_hex()`](https://ggplot2.tidyverse.org/reference/geom_hex.html)
