---
title: "Redundant coding, text annotations"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
params:
  cache: TRUE
---

```{r setup, include=FALSE, echo=FALSE, message = FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(colorspace)
library(colorblindr)
library(cowplot)
library(lubridate)
library(ggrepel)
library(here)

# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2.0)) # 2 for full width, 2.5 for half width

cvd_sim2 <- function(p, severity = 1, scale = 0.9, hjust = 0, vjust = 1,
                     label_x = 0, label_y = 1, label_size = 14, label_fontface = "plain")
{
  deut <- function(c) deutan(c, severity)
  p1 <- edit_colors(p, deut)
  
  prot <- function(c) protan(c, severity)
  p2 <- edit_colors(p, prot)
  
  trit <- function(c) tritan(c, severity)
  p3 <- edit_colors(p, trit)
  
  desat <- function(c) desaturate(c, severity)
  p4 <- edit_colors(p, desat)
  
  plot_grid(
    p1, NULL, p2,
    NULL, NULL, NULL,
    p3, NULL, p4,
    scale = 1, hjust = hjust, vjust = vjust,
    labels = c(
      "deuteranomaly", "", "protanomaly",
      "", "", "",
      "tritanomaly", "", "desaturated"),
    rel_widths = c(scale, 1-scale, scale),
    rel_heights = c(scale, 1-scale, scale),
    label_x = label_x, label_y = label_y, label_size = label_size,
    label_fontface = label_fontface
  )
}

# tech stocks dataset
tech_stocks <- read_csv(here("datasets", "tech_stocks.csv")) %>%
  mutate(date = ymd(date)) %>%
  select(company, date, price_indexed)

# tx_counties dataset
US_census <- read_csv(here("datasets", "US_census.csv"))

tx_counties <- US_census %>% 
  filter(state == "Texas") %>%
  select(name, pop2010) %>%
  extract(name, "county", regex = "(.+) County")
```

## Color encoding alone is not always sufficient

.center[
```{r iris-scatter-one-shape, cache = params$cache, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
breaks <- c("setosa", "virginica", "versicolor")
labels <- paste0("Iris ", breaks)

iris_scatter_base <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, fill = Species, color = Species)) + 
    scale_color_manual(
      values = darken(c("#E69F00", "#56B4E9", "#009E73"), 0.3),
      breaks = breaks,
      labels = labels,
      name = NULL
    ) +
    scale_fill_manual(
      values = c("#E69F0080", "#56B4E980", "#009E7380"),
      breaks = breaks,
      labels = labels,
      name = NULL
    ) +
    scale_x_continuous(
      limits = c(3.95, 8.2), expand = c(0, 0),
      labels = c("4.0", "5.0", "6.0", "7.0", "8.0"),
      name = "Sepal length"
    ) +
    scale_y_continuous(
      limits = c(1.9, 4.6), expand = c(0, 0),
      name = "Sepal width"
    )

iris_scatter <- iris_scatter_base +
  geom_point(
    size = 2.5, shape = 21, stroke = 0.5,
    position = position_jitter(
      width = 0.01 * diff(range(iris$Sepal.Length)),
      height = 0.01 * diff(range(iris$Sepal.Width)),
      seed = 3942
    )
  ) +
  theme_minimal_grid() +
  theme(
    legend.title.align = 0.5,
    legend.text = element_text(face = "italic"),
    legend.spacing.y = unit(3.5, "pt"),
    plot.margin = margin(7, 7, 3, 1.5)
  )

stamp_bad(iris_scatter)
```
]

Iris virginica and Iris versicolor are difficult to tell apart


???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

.center[
```{r iris-scatter-one-shape-cvd, cache = params$cache, echo = FALSE, fig.width = 9, fig.asp = 0.66, dev = "svg"}
iris_scatter_small <- iris_scatter_base +
  geom_point(
    size=.655*2.5, shape=21, stroke = .655*0.5,
    position = position_jitter(
      width = 0.01 * diff(range(iris$Sepal.Length)),
      height = 0.01 * diff(range(iris$Sepal.Width)),
      seed = 3942
    )
  ) +
  theme_minimal_grid(
    .655*14,
    line_size = .85*.5 # make line size a little bigger than mathematically correct
  ) + 
  theme(
    legend.title.align = 0.5,
    legend.text = element_text(face = "italic"),
    legend.spacing.y = grid::unit(.655*3, "pt"),
    plot.margin = margin(18, 1, 1, 1)
  )
cvd_sim2(iris_scatter_small, label_size = 14, label_y = .98, scale = .95)
```
]

.absolute-bottom-left[
CVD simulation of preceding figure
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Using both colors and shapes helps

.center[
```{r iris-scatter-three-shapes, cache = params$cache, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
iris_scatter2_base <- ggplot(
  iris, aes(x = Sepal.Length, y = Sepal.Width, shape = Species, fill = Species, color = Species)
) +     
    scale_shape_manual(
      values = c(21, 22, 23),
      breaks = breaks,
      labels = labels,
      name = NULL
    ) +
    scale_color_manual(
      values = darken(c("#56B4E9", "#E69F00", "#009E73"), 0.3),
      breaks = breaks,
      labels = labels,
      name = NULL
    ) +
    scale_fill_manual(
      values = c("#56B4E980", "#E69F0080", "#009E7380"),
      breaks = breaks,
      labels = labels,
      name = NULL
    ) +
    scale_x_continuous(
      limits = c(3.95, 8.2), expand = c(0, 0),
      labels = c("4.0", "5.0", "6.0", "7.0", "8.0"),
      name = "Sepal length"
    ) +
    scale_y_continuous(
      limits = c(1.9, 4.6), expand = c(0, 0),
      name = "Sepal width"
    )
iris_scatter2 <- iris_scatter2_base +
  geom_point(
    size=2.5, stroke = 0.5,
    position = position_jitter(
      width = 0.01 * diff(range(iris$Sepal.Length)),
      height = 0.01 * diff(range(iris$Sepal.Width)),
      seed = 3942)
  ) +
  theme_minimal_grid() +
  theme(
    legend.title.align = 0.5,
    legend.text = element_text(face = "italic"),
    legend.spacing.y = unit(3.5, "pt"),
    plot.margin = margin(7, 7, 3, 1.5)
  )
iris_scatter2
```
]

Notice also that color assignment has been switched

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

.center[
```{r iris-scatter-three-shapes-cvd, cache = params$cache, echo = FALSE, fig.width = 9, fig.asp = 0.66, dev = "svg"}
iris_scatter2_small <- iris_scatter2_base +
  geom_point(
    size=.655*2.5, stroke = .655*0.5,
    position = position_jitter(
      width = 0.01 * diff(range(iris$Sepal.Length)),
      height = 0.01 * diff(range(iris$Sepal.Width)),
      seed = 3942)
  ) +
  theme_minimal_grid(
   .655*14,
    line_size = .85*.5 # make line size a little bigger than mathematically correct
  ) + 
  theme(
    legend.title.align = 0.5,
    legend.text = element_text(face = "italic"),
    legend.spacing.y = grid::unit(.655*3, "pt"),
    plot.margin = margin(18, 1, 1, 1)
  )
cvd_sim2(iris_scatter2_small, label_size = 14, label_y = .98, scale = .95)
```
]

.absolute-bottom-left[
CVD simulation of preceding figure
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Pay attention to legend order

.center[
```{r tech-stocks-bad-legend, cache = params$cache, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
price_plot_base <- 
  ggplot(tech_stocks, aes(x = date, y = price_indexed, color = company)) +
  geom_line(size = 0.66, na.rm = TRUE) +
  scale_color_manual(
    values = c(
      Facebook = "#000000", 
      Alphabet = "#E69F00", 
      Microsoft = "#56B4E9",
      Apple = "#009E73"
    ),
    name = "",
    breaks = c("Alphabet", "Apple", "Facebook", "Microsoft")
  ) +
  scale_x_date(
    name = "year",
    limits = c(ymd("2012-06-01"), ymd("2017-05-31")),
    expand = c(0,0)
  ) + 
  scale_y_continuous(
    name = "stock price, indexed",
    limits = c(0, 560),
    expand = c(0,0)
  )
stamp_bad(
  price_plot_base + 
    theme_minimal_hgrid() + 
    theme(plot.margin = margin(3, 7, 3, 1.5))
)
```
]

Legend order **does not** match visual order of the lines

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Pay attention to legend order

.center[
```{r tech-stocks-good-legend, cache = params$cache, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
price_plot_base_good <- 
  ggplot(tech_stocks, aes(x = date, y = price_indexed, color = company)) +
  scale_color_manual(
    values = c(
      Facebook = "#000000", 
      Alphabet = "#E69F00", 
      Microsoft = "#56B4E9",
      Apple = "#009E73"
    ),
    name = "",
    breaks = c("Facebook", "Alphabet", "Microsoft", "Apple")
  ) +
  scale_x_date(
    name = "year",
    limits = c(ymd("2012-06-01"), ymd("2017-05-31")),
    expand = c(0,0)
  ) + 
  scale_y_continuous(
    name = "stock price, indexed",
    limits = c(0, 560),
    expand = c(0,0)
  )

price_plot_base_good +
  geom_line(size = 0.66, na.rm = TRUE) +
  theme_minimal_hgrid() + 
  theme(plot.margin = margin(3, 7, 3, 1.5))

```
]

Legend order **does** match visual order of the lines

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

.center[
```{r tech-stocks-good-legend-cvd, cache = params$cache, echo = FALSE, fig.width = 9, fig.asp = 0.66, dev = "svg"}
price_plot_good_small <-
  price_plot_base_good + 
  geom_line(
    size = .85*0.66, # make line size a little bigger than mathematically correct
    na.rm = TRUE
  ) +
  theme_minimal_hgrid(
    .655*14,
    line_size = .85*.5 # make line size a little bigger than mathematically correct
  ) + 
  theme(plot.margin = margin(18, 1, 1, 1))

cvd_sim2(price_plot_good_small, label_size = 14, label_y = .98, scale = .95)
```
]

.absolute-bottom-left[
CVD simulation of preceding figure
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Even better: Don't use a legend at all

.center[
```{r tech-stocks-good-no-legend, cache = params$cache, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
tech_stocks_last <- filter(tech_stocks, date == "2017-06-02")

ggplot(tech_stocks, aes(x = date, y = price_indexed, color = company)) +
  scale_color_manual(
    values = c(
      Facebook = "#000000", 
      Alphabet = "#E69F00", 
      Microsoft = "#56B4E9",
      Apple = "#009E73"
    ),
    name = "",
    breaks = c("Facebook", "Alphabet", "Microsoft", "Apple"),
    guide = "none"
  ) +
  scale_x_date(
    name = "year",
    limits = c(ymd("2012-06-01"), ymd("2017-05-31")),
    expand = c(0,0)
  ) + 
  scale_y_continuous(
    name = "stock price, indexed",
    limits = c(0, 560),
    expand = c(0,0),
    sec.axis = dup_axis(
      breaks = tech_stocks_last$price_indexed,
      labels = tech_stocks_last$company,
      name = NULL
    )
  ) +
  geom_line(size = 0.66, na.rm = TRUE) +
  theme_minimal_hgrid() + 
  theme(
    axis.ticks.length.y.right = grid::unit(0, "pt"),
    plot.margin = margin(3, 7, 3, 1.5)
  )
```
]

This technique is called "direct labeling"

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Direct labeling the Iris scatter plot

.center[
```{r iris-scatter-with-ellipses, cache = params$cache, echo = FALSE, fig.width = 5.5, fig.asp = 0.8, dev = "svg"}
iris_labels <- tibble(
  Species = c("setosa", "virginica", "versicolor"),
  label = c("Iris setosa", "Iris virginica", "Iris versicolor"),
  Sepal.Width = c(4.2, 3.76, 2.08),
  Sepal.Length = c(5.7, 7, 5.1),
  hjust = c(0, 0.5, 0),
  vjust = c(0, 0.5, 1)
)

ggplot(
  iris,
  aes(x = Sepal.Length, y = Sepal.Width, color = Species)
) + 
  geom_point(
    aes(shape = Species, fill = Species),
    size = 2.5,
    position = position_jitter(
      width = 0.01 * diff(range(iris$Sepal.Length)),
      height = 0.01 * diff(range(iris$Sepal.Width)),
      seed = 3942)
  ) +
  stat_ellipse(size = 0.5) +
  geom_text(
    data = iris_labels,
    aes(
      x = Sepal.Length, y = Sepal.Width,
      label = label, color = Species,
      hjust = hjust, vjust = vjust
    ),
    size = 14/.pt,
    fontface = "italic",
    inherit.aes = FALSE
  ) +
  scale_shape_manual(
    values = c(21, 22, 23),
    breaks = breaks,
    name = NULL
  ) +
  scale_fill_manual(
    values = c("#56B4E980", "#E69F0080", "#009E7380"),
    breaks = breaks,
    name = NULL
  ) +
  scale_color_manual(
    values = darken(c("#56B4E9", "#E69F00", "#009E73"), 0.3),
    breaks = breaks,
    name = NULL
  ) +
  guides(fill = "none", color = "none", shape = "none") +
  scale_x_continuous(
    limits = c(3.95, 8.2), expand = c(0, 0),
    labels = c("4.0", "5.0", "6.0", "7.0", "8.0"),
    name = "Sepal length"
  ) +
  scale_y_continuous(
    limits = c(1.9, 4.6), expand = c(0, 0),
    name = "Sepal width"
  ) +
  theme_half_open()
```
]

The ellipses help guide the eye

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Label only a subset if there are too many items to label

.center[
```{r texas-counties-log, cache = params$cache, echo = FALSE, message = FALSE, fig.width = 8, fig.asp = 0.6, dev = "svg"}
set.seed(3878)

tx_counties_labels <- tx_counties %>% 
  mutate(popratio = pop2010/median(pop2010)) %>%
  arrange(desc(popratio)) %>%
  mutate(
    index = 1:n(),
    label = ifelse(index <= 3 | index > n()-3 | runif(n()) < .04, county, ""),
    label_large = ifelse(index <= 6, county, "")
  )

ggplot(tx_counties_labels, aes(x = index, y = popratio)) +
  geom_hline(yintercept = 1, linetype = 2, color = "grey40") +
  geom_point(size = 0.5, color = "#0072B2") +
  geom_text_repel(
    aes(label = label),
    size = 10/.pt,
    point.padding = .4, color = "black",
    min.segment.length = 0,
    max.overlaps = Inf
  ) +
  scale_y_log10(
    breaks = c(.01, .1, 1, 10, 100),
    name = "population number / median",
    labels = expression(10^-2, 10^-1, 10^0, 10^1, 10^2)
  ) +
  scale_x_continuous(
    limits = c(.5, nrow(tx_counties) + .5), expand = c(0, 0),
    breaks = NULL, #c(1, 50*(1:5)),
    name = "Texas counties, from most to least populous"
  ) +
  theme_minimal_hgrid(14) +
  theme(axis.line = element_blank())
```
]


???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

[//]: # "segment ends here"

---

class: center middle

## Direct labeling in R

---

## 1. Secondary axis trick

Data preparation:
.tiny-font[
```{r}
tech_stocks <- read_csv("https://wilkelab.org/SDS375/datasets/tech_stocks.csv") %>%
  mutate(date = ymd(date)) %>%
  select(company, date, price_indexed)
```
]

---

## 1. Secondary axis trick

.pull-left[

The raw input data:
.tiny-font[
```{r}
tech_stocks
```
]]

--

.pull-right[

Most recent values only:
.tiny-font[
```{r}
tech_stocks_last <- tech_stocks %>%
  filter(date == max(date))

tech_stocks_last
```
]]


---

## 1. Secondary axis trick

.tiny-font.pull-left.width-50[
```{r tech-stock-sec-axis, eval = FALSE}
ggplot(tech_stocks) +
  aes(x = date, y = price_indexed) +
  geom_line(aes(color = company), na.rm = TRUE) +
  scale_x_date(
    limits = c(
      ymd("2012-06-01"),
      ymd("2017-05-31")
    ),
    expand = c(0, 0)
  ) + 
  scale_y_continuous(
    limits = c(0, 560),
    expand = c(0, 0)
  )
```
]

.pull-right[
```{r tech-stock-sec-axis-out, cache = params$cache, ref.label="tech-stock-sec-axis", fig.width=5., fig.asp = 0.618, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

Side remark: The default legend order doesn't match the data


---

## 1. Secondary axis trick

.tiny-font.pull-left.width-50[
```{r tech-stock-sec-axis2, eval = FALSE}
ggplot(tech_stocks) +
  aes(x = date, y = price_indexed) +
  geom_line(aes(color = company), na.rm = TRUE) +
  scale_x_date(
    limits = c(
      ymd("2012-06-01"),
      ymd("2017-05-31")
    ),
    expand = c(0, 0)
  ) + 
  scale_y_continuous(
    limits = c(0, 560),
    expand = c(0, 0),
    sec.axis = dup_axis(                        #<<
      breaks = tech_stocks_last$price_indexed,  #<<
      labels = tech_stocks_last$company,        #<< 
      name = NULL                               #<<
    )                                           #<<
  ) +
  guides(color = "none")
```
]

.pull-right[
```{r tech-stock-sec-axis2-out, cache = params$cache, ref.label="tech-stock-sec-axis2", fig.width=5., fig.asp = 0.618, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

---

## 2. Manual labeling with `geom_text()`


The raw input data:
.tiny-font[
```{r}
iris
```
]

---

## 2. Manual labeling with `geom_text()`

Manually create table with label positions:

.tiny-font[
```{r}
iris_labels <- tibble(
  Species = c("setosa", "virginica", "versicolor"),
  Sepal.Width = c(4.2, 3.76, 2.08),
  Sepal.Length = c(5.7, 7, 5.1),
  label = c("Iris setosa", "Iris virginica", "Iris versicolor"),
  hjust = c(0, 0.5, 0),
  vjust = c(0, 0.5, 1)
)

iris_labels
```
]


---

## 2. Manual labeling with `geom_text()`

.xtiny-font.pull-left.width-50[
```{r iris-manual, eval = FALSE}
ggplot(iris) +
  aes(Sepal.Length, Sepal.Width, color = Species) +
  geom_point(aes(shape = Species))
```
]

.pull-right[
```{r iris-manual-out, cache = params$cache, ref.label="iris-manual", fig.width=5., fig.asp = 0.618, echo=FALSE, warning=FALSE, dev = "svg"}
```
]


---

## 2. Manual labeling with `geom_text()`

.xtiny-font.pull-left.width-50[
```{r iris-manual2, eval = FALSE}
ggplot(iris) +
  aes(Sepal.Length, Sepal.Width, color = Species) +
  geom_point(aes(shape = Species)) +
  geom_text(                           #<<
    data = iris_labels,                #<<
    aes(                               #<<
      label = label,                   #<<
      hjust = hjust, vjust = vjust     #<<
    ),                                 #<<
    size = 14/.pt # 14pt font          #<<
  ) +                                  #<<
  guides(color = "none", shape = "none")
```
]

.pull-right[
```{r iris-manual2-out, cache = params$cache, ref.label="iris-manual2", fig.width=5., fig.asp = 0.8, echo=FALSE, warning=FALSE, dev = "svg"}
```
]


---

## 2. Manual labeling with `geom_text()`

.xtiny-font.pull-left.width-50[
```{r iris-manual3, eval = FALSE}
ggplot(iris) +
  aes(Sepal.Length, Sepal.Width, color = Species) +
  geom_point(aes(shape = Species)) +
  geom_text(
    data = iris_labels,
    aes(
      label = label,
      hjust = hjust, vjust = vjust
    ),
    size = 14/.pt # 14pt font
  ) +
  stat_ellipse(size = 0.5) +  #<<
  guides(color = "none", shape = "none")
```
]

.pull-right[
```{r iris-manual3-out, cache = params$cache, ref.label="iris-manual3", fig.width=5., fig.asp = 0.8, echo=FALSE, warning=FALSE, dev = "svg"}
```
]


---

## 3. Automatic labeling with `geom_text_repel()`

.tiny-font.pull-left[
```{r}
mtcars_named <- mtcars %>%
  rownames_to_column("car") %>%
  select(car, weight = wt, mpg)

mtcars_named
```
]

---

## 3. Automatic labeling with `geom_text_repel()`

.tiny-font.pull-left[
```{r mtcars-plot1, eval = FALSE}
ggplot(mtcars_named, aes(weight, mpg)) +
  geom_point()
```
]

.pull-right.width-50[
```{r mtcars-plot1-out, cache = params$cache, ref.label="mtcars-plot1", fig.width=6., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

---

## 3. Automatic labeling with `geom_text_repel()`

.tiny-font.pull-left[
```{r mtcars-plot2, eval = FALSE}
ggplot(mtcars_named, aes(weight, mpg)) +
  geom_point() +
  geom_text(aes(label = car))
```
]

.pull-right.width-50[
```{r mtcars-plot2-out, cache = params$cache, ref.label="mtcars-plot2", fig.width=6., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

---

## 3. Automatic labeling with `geom_text_repel()`

.tiny-font.pull-left[
```{r mtcars-plot3, eval = FALSE}
ggplot(mtcars_named, aes(weight, mpg)) +
  geom_point() +
  geom_text(aes(label = car), hjust = 0)
```
]

.pull-right.width-50[
```{r mtcars-plot3-out, cache = params$cache, ref.label="mtcars-plot3", fig.width=6., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]


---

## 3. Automatic labeling with `geom_text_repel()`

.tiny-font.pull-left[
```{r mtcars-plot4, eval = FALSE}
library(ggrepel)  #<<

ggplot(mtcars_named, aes(weight, mpg)) +
  geom_point() +
  geom_text_repel(          #<<
    aes(label = car),       #<<
    max.overlaps = Inf      #<<
  )                         #<<
```
]

.pull-right.width-50[
```{r mtcars-plot4-out, cache = params$cache, ref.label="mtcars-plot4", fig.width=6., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

---

## 3. Automatic labeling with `geom_text_repel()`

.tiny-font.pull-left[
```{r mtcars-plot5, eval = FALSE}
set.seed(6643)    #<<

mtcars_named %>%
  mutate(
    # randomly exclude 50% of the labels    #<<
    car = ifelse(runif(n()) < 0.5, "", car) #<<
  ) %>% 
  ggplot(aes(weight, mpg)) +
  geom_point() +
  geom_text_repel(
    aes(label = car),
    max.overlaps = Inf
  )
```
]

.pull-right.width-50[
```{r mtcars-plot5-out, cache = params$cache, ref.label="mtcars-plot5", fig.width=6., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]


---

## 3. Automatic labeling with `geom_text_repel()`

.tiny-font.pull-left[
```{r mtcars-plot6, eval = FALSE}
set.seed(6644)    #<<

mtcars_named %>%
  mutate(
    # randomly exclude 50% of the labels    #<<
    car = ifelse(runif(n()) < 0.5, "", car) #<<
  ) %>% 
  ggplot(aes(weight, mpg)) +
  geom_point() +
  geom_text_repel(
    aes(label = car),
    max.overlaps = Inf
  )
```
]

.pull-right.width-50[
```{r mtcars-plot6-out, cache = params$cache, ref.label="mtcars-plot6", fig.width=6., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]


---

## 3. Automatic labeling with `geom_text_repel()`

.tiny-font.pull-left[
```{r mtcars-plot7, eval = FALSE}
set.seed(6644)

mtcars_named %>%
  mutate(
    # randomly exclude 50% of the labels
    car = ifelse(runif(n()) < 0.5, "", car)
  ) %>% 
  ggplot(aes(weight, mpg)) +
  geom_point() +
  geom_text_repel(
    aes(label = car),
    max.overlaps = Inf,
    box.padding = 0.7     #<<
  )
```
]

.pull-right.width-50[
```{r mtcars-plot7-out, cache = params$cache, ref.label="mtcars-plot7", fig.width=6., fig.asp = 0.75, echo=FALSE, warning=FALSE, dev = "svg"}
```
]

`box.padding` controls how far labels are placed from data points

[//]: # "segment ends here"

---

## Further reading

- Fundamentals of Data Visualization: [Chapter 20: Redundant coding](https://clauswilke.com/dataviz/redundant-coding.html)
- **ggplot2** reference documentation: [`geom_text()`](https://ggplot2.tidyverse.org/reference/geom_text.html)
- **ggrepel** documentation: [Examples](https://ggrepel.slowkow.com/articles/examples.html#examples-1)
- **ggrepel** reference documentation:
[`geom_text_repel()`](https://ggrepel.slowkow.com/reference/geom_text_repel.html)
