---
title: "Visualizing distributions 2"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(cowplot)
library(colorspace)
library(ggridges)
library(ggforce)
library(here)
# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width

lincoln_temps <- readRDS(here("datasets", "lincoln_temps.rds"))
```

## Reminder: Density estimates visualize distributions

.pull-left.small-font[
Mean temperatures in Lincoln, NE, in January 2016:

.center[
```{r temps_table, echo = FALSE}
lincoln_temps %>%
  filter(month == "Jan") %>%
  select(date, `mean temp` = mean_temp) %>%
  knitr::kable()
```
]]

--

.pull-right[
```{r temps_densities_january, echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
lincoln_temps %>%
  filter(month == "Jan") %>%
  ggplot(aes(mean_temp)) +
  geom_density(fill = "#56B4E9", alpha = 0.7) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
  ggtitle("Temperature distribution") +
  theme_minimal_grid(18) +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(face = "plain")
  )
```
]

---


## Reminder: Density estimates visualize distributions

.pull-left.small-font[
Mean temperatures in Lincoln, NE, in January 2016:

.center[
```{r temps_table2, ref.label = "temps_table", echo = FALSE}
```
]]

.pull-right[
```{r temps_densities_january2, ref.label = "temps_densities_january", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```

How can we compare distributions across months?
]


---

## A bad idea: Many overlapping density plots

.center[
```{r temps_densities_overlapping, echo = FALSE, fig.width = 8, fig.asp = 0.618, dev = "svg"}
p <- lincoln_temps %>%
  mutate(
    month_long = fct_rev(month_long)
  ) %>%
  ggplot(aes(mean_temp, fill = month)) +
  geom_density(alpha = 0.5) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
  scale_fill_viridis_d(option = "D") +
  labs(subtitle = "") + # empty subtitle for some vertical space
  theme_minimal_grid(16)

stamp_bad(p)
```
]

---

## Another bad idea: Stacked density plots

.center[
```{r temps_densities_stacked, echo = FALSE, fig.width = 8, fig.asp = 0.618, dev = "svg"}
p <- lincoln_temps %>%
  mutate(
    month_long = fct_rev(month_long)
  ) %>%
  ggplot(aes(mean_temp, fill = month)) +
  geom_density(alpha = 0.7, position = "stack") +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
  scale_fill_viridis_d(option = "D") +
  labs(subtitle = "") + # empty subtitle for some vertical space
  theme_minimal_grid(16)

stamp_bad(p)
```
]

---

## Somewhat better: Small multiples

.center[
```{r temps_densities, echo = FALSE, fig.width = 8, fig.asp = 0.618, dev = "svg"}
lincoln_temps %>%
  mutate(
    month_long = fct_rev(month_long)
  ) %>%
  ggplot(aes(mean_temp)) +
  geom_density(fill = "#56B4E9") +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
  facet_wrap(vars(month_long)) +
  theme_minimal_grid(16, color = "gray65") +
  theme(
    panel.background = element_rect(fill = "gray95")
  )
```
]

---

## Instead: Show values along y, conditions along x
   
.center[
```{r temps_boxplots, echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
ggplot(lincoln_temps, aes(x = month, y = mean_temp)) +
  geom_boxplot(fill = "gray90") + 
  xlab("month") + 
  ylab("mean temperature (°F)") +
  theme_half_open()
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

--

A boxplot is a crude way of visualizing a distribution.

---

## How to read a boxplot

.center[
```{r boxplot-schematic, echo = FALSE, fig.width = 8, fig.asp = 0.618, dev = "svg"}
set.seed(3423)

y <- c(rnorm(100), 3.4)
s <- boxplot.stats(y)
df <- tibble(
  y = c(s$stats, max(y)),
  x = c(1.03, 1.405, 1.405, 1.405, 1.03, 1.04),
  label = c("minimum", "first quartile", "median", "third quartile", "maximum within upper fence", "outlier")
)

p_points <- ggplot(tibble(y), aes(x = 0, y = y)) + 
  geom_point(position = position_jitter(width = .4, height = 0, seed = 320)) +
  annotate("text", label = "data", x = 0, y = 4, hjust = 0.5, vjust = 1, size = 20/.pt) +
  scale_x_continuous(limits = c(-1.8, .4), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-2.55, 4), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_nothing()

p_boxplot <- ggplot(tibble(y), aes(x = 1, y = y)) + 
  geom_boxplot(fill = "gray90", outlier.size = 2) +
  annotate("text", label = "boxplot", x = 1, y = 4, hjust = 0.5, vjust = 1, size = 20/.pt) +
  geom_text(
    data = df, aes(x, y, label = label), hjust = 0,
    size = 16/.pt
  ) +
  scale_x_continuous(limits = c(0, 3.5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-2.55, 4), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_nothing()

plot_grid(p_points, p_boxplot, rel_widths = c(.65, 1), nrow = 1)
```
]


???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## If you like density plots, consider violins
   
.center[
```{r temps_violins, echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
ggplot(lincoln_temps, aes(x = month, y = mean_temp)) +
  geom_violin(fill = "gray90") + 
  xlab("month") + 
  ylab("mean temperature (°F)") +
  theme_half_open()
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

--

A violin plot is a density plot rotated 90 degrees and then mirrored.

---

## How to read a violin plot

.center[
```{r violin-schematic, echo = FALSE, fig.width = 8, fig.asp = 0.618, dev = "svg"}
set.seed(3423)

y <- c(rnorm(100), 3.4)
d <- density(y)
df <- tibble(
  y = c(min(y), d$x[which(d$y == max(d$y))], max(y)),
  x = c(1.1, 1.5, 1.08),
  label = c("minimum data value", "maximum point density", "maximum data value")
)

p_points <- ggplot(tibble(y), aes(x = 0, y = y)) + 
  geom_point(position = position_jitter(width = .4, height = 0, seed = 320)) +
  annotate("text", label = "data", x = 0, y = 4, hjust = 0.5, vjust = 1, size = 20/.pt) +
  scale_x_continuous(limits = c(-1.8, .4), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-2.55, 4), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_nothing()

p_violin <- ggplot(tibble(y), aes(x = 1, y = y)) + 
  geom_violin(fill = "gray90") +
  annotate("text", label = "violin plot", x = 1, y = 4, hjust = 0.5, vjust = 1, size = 20/.pt) +
  geom_text(
    data = df, aes(x, y, label = label), hjust = 0,
    size = 16/.pt
  ) +
  scale_x_continuous(limits = c(0, 3.5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(-2.55, 4), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_nothing()

plot_grid(p_points, p_violin, rel_widths = c(.65, 1), nrow = 1)
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## For small datasets, you can also use a strip chart

Advantage: Can see raw data points instead of abstract representation.

.center[
```{r temps_stripchart, echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
ggplot(lincoln_temps, aes(x = month, y = mean_temp)) +
  geom_point(
    position = position_jitter(width = .15, height = 0, seed = 320),
    size = 0.75
  ) +
  xlab("month") + 
  ylab("mean temperature (°F)") +
  theme_half_open()
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

--

Horizontal jittering may be necessary to avoid overlapping points.

---

## For small datasets, you can also use a strip chart

Advantage: Can see raw data points instead of abstract representation.

.center[
```{r temps_stripchart2, echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
ggplot(lincoln_temps, aes(x = month, y = mean_temp)) +
  geom_point(size = 0.75) +
  xlab("month") + 
  ylab("mean temperature (°F)") +
  theme_half_open()
```
]

Horizontal jittering may be necessary to avoid overlapping points.

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## For small datasets, you can also use a strip chart

Advantage: Can see raw data points instead of abstract representation.

.center[
```{r temps_stripchart3, ref.label = "temps_stripchart", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

Horizontal jittering may be necessary to avoid overlapping points.


???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## We can also jitter points into violins
   
.center[
```{r temps_sina, echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
ggplot(lincoln_temps, aes(x = month, y = mean_temp)) +
  geom_violin(fill = "gray90", color = NA) + 
  geom_sina(size = 0.75) + 
  xlab("month") + 
  ylab("mean temperature (°F)") +
  theme_half_open()
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

--

Such plots are called sina plots, to honor [Sina Hadi Sohi.](https://clauswilke.com/dataviz/boxplots-violins.html#fig:lincoln-temp-sina)


---

## But maybe there's hope for overlapping density plots?

.center[
```{r temps_densities_overlapping2, echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
lincoln_temps %>%
  mutate(
    month_long = fct_rev(month_long)
  ) %>%
  ggplot(aes(mean_temp, fill = month)) +
  geom_density(alpha = 0.5) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0),
    breaks = c(0, 25, 50, 75)
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.06))) +
  scale_fill_viridis_d(option = "D") +
  theme_minimal_grid(16)
```
]


???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

--

How about we stagger the densities vertically?

---

## Vertically staggered density plots are called ridgelines

.center[
```{r lincoln-ridgeline-polished, echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}

ggplot(lincoln_temps, aes(x = mean_temp, y = month_long)) +
  geom_density_ridges(
    scale = 3,
    rel_min_height = 0.01,
    bandwidth = 3.4,
    fill = lighten("#56B4E9", .3),
    color = "white"
  ) +
  scale_x_continuous(
    name = "mean temperature (°F)",
    expand = c(0, 0)#, breaks = c(0, 25, 50, 75)
  ) +
  scale_y_discrete(name = NULL, expand = expansion(add = c(0.2, 2.6))) +
  theme_minimal_grid(16) +
  theme(
    axis.text.y = element_text(vjust = 0)
  )
```
]


???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

--

Notice the single fill color. More colors would be distracting. 

---
class: center middle

## Making boxplots, violins, etc. in **ggplot2**

---

## Making boxplots, violins, etc. in **ggplot2**

.small-font.center[

Plot type    | Geom                    | Notes
:----------- | :-----------------      | :-------------------------
boxplot      | `geom_boxplot()`        |
violin plot  | `geom_violin()`         |
strip chart  | `geom_point()`          | Jittering requires `position_jitter()`
sina plot    | `geom_sina()`           | From package **ggforce**
ridgeline    | `geom_density_ridges()` | From package **ggridges**

]

---

## Examples: Boxplot

.tiny-font[
```{r temps-examples-boxplot, eval = FALSE}
ggplot(lincoln_temps, aes(x = month, y = mean_temp)) +
  geom_boxplot(fill = "skyblue") 
```
]

.center[
```{r temps-examples-boxplot-out, ref.label = "temps-examples-boxplot", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Examples: Violins

.tiny-font[
```{r temps-examples-violins, eval = FALSE}
ggplot(lincoln_temps, aes(x = month, y = mean_temp)) +
  geom_violin(fill = "skyblue") 
```
]

.center[
```{r temps-examples-violin-out, ref.label = "temps-examples-violins", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Examples: Strip chart (no jitter)

.tiny-font[
```{r temps-examples-stripchart, eval = FALSE}
ggplot(lincoln_temps, aes(x = month, y = mean_temp)) +
  geom_point(size = 0.75)  # reduce point size to minimize overplotting 
```
]

.center[
```{r temps-examples-stripchart-out, ref.label = "temps-examples-stripchart", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Examples: Strip chart (w/ jitter)

.tiny-font[
```{r temps-examples-stripchart-jitter, eval = FALSE}
ggplot(lincoln_temps, aes(x = month, y = mean_temp)) +
  geom_point(size = 0.75,  # reduce point size to minimize overplotting 
    position = position_jitter(
      width = 0.15,  # amount of jitter in horizontal direction
      height = 0     # amount of jitter in vertical direction (0 = none)
    )
  )
```
]

.center[
```{r temps-examples-stripchart-jitter-out, ref.label = "temps-examples-stripchart-jitter", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Examples: Sina plot

.tiny-font[
```{r temps-examples-sina, eval = FALSE}
library(ggforce) # for geom_sina()

ggplot(lincoln_temps, aes(x = month, y = mean_temp)) +
  geom_violin(fill = "skyblue", color = NA) + # violins in background
  geom_sina(size = 0.75) # sina jittered points in foreground
```
]

.center[
```{r temps-examples-sina-out, ref.label = "temps-examples-sina", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Examples: Ridgeline plot

.tiny-font[
```{r temps-examples-ridgeline, eval = FALSE}
library(ggridges) # for geom_density_ridges

ggplot(lincoln_temps, aes(x = mean_temp, y = month_long)) +
  geom_density_ridges()
```
]

.center[
```{r temps-examples-ridgeline-out, ref.label = "temps-examples-ridgeline", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

[//]: # "segment ends here"

---

## Further reading

- Fundamentals of Data Visualization: [Chapter 7: Visualizing many distributions at once](https://clauswilke.com/dataviz/boxplots-violins.html)
- **ggplot2** reference documentation: [`geom_boxplot()`](https://ggplot2.tidyverse.org/reference/geom_histogram), [`geom_violin()`](https://ggplot2.tidyverse.org/reference/geom_violin), [`position_jitter()`](https://ggplot2.tidyverse.org/reference/position_jitter.html)
- **ggforce** reference documentation: [`geom_sina()`](https://ggforce.data-imaginist.com/reference/geom_sina.html)
- **ggridges** reference documentation: [`geom_density_ridges()`](https://wilkelab.org/ggridges/reference/geom_density_ridges.html)




