---
title: "Visualizing proportions"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(here)
library(ggforce)
library(cowplot)
# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width
```


## The archetypal visualization of proportions: pie chart

```{r echo = FALSE}
bundestag <- tibble(
  party = c("CDU/CSU", "SPD", "FDP"),
  seats = c(243, 214, 39),
  colors = c("#4E4E4E", "#B6494A", "#E7D739")
)
```

.small-font.center[
```{r bundestag-pie, echo = FALSE, fig.width = 7, fig.asp = 0.58, dev = "svg"}
bund_pie <- bundestag %>%
  arrange(seats) %>%
  mutate(
    end_angle = 2*pi*cumsum(seats)/sum(seats),   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for the text label
    # horizontal and vertical justifications depend on whether we're to the left/right
    # or top/bottom of the pie
    hjust = ifelse(mid_angle > pi, 1, 0),
    vjust = ifelse(mid_angle < pi/2 | mid_angle > 3*pi/2, 0, 1)
  )

# radius of the pie and radius for outside and inside labels
rpie <- 1
rlabel_out <- 1.05 * rpie
rlabel_in <- 0.6 * rpie

bt_pie_base <- ggplot(bund_pie) +
  geom_arc_bar(
    aes(
      x0 = 0, y0 = 0, r0 = 0, r = rpie,
      start = start_angle, end = end_angle, fill = colors
    ),
    color = "white"
  ) +
  geom_text(
    aes(
      x = rlabel_in * sin(mid_angle),
      y = rlabel_in * cos(mid_angle),
      label = seats
    ),
    size = 14/.pt,
    color = c("black", "white", "white")
  ) +
  scale_x_continuous(
    name = NULL,
    limits = c(-1.7, 1.4),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = NULL,
    limits = c(-1.05, 1.15),
    expand = c(0, 0)
  ) +
  scale_fill_identity() +
  coord_fixed() +
  theme_map()

# save for later with 12 pt font
bt_pie <- bt_pie_base +
  geom_text(
    aes(
      x = rlabel_out * sin(mid_angle),
      y = rlabel_out * cos(mid_angle),
      label = party,
      hjust = hjust, vjust = vjust
    ),
    size = 12/.pt
  ) +
  theme(
    plot.margin = margin(0, 0, 21, 14)
  )

# output now with 14 pt font
bt_pie_base +
  geom_text(
    aes(
      x = rlabel_out * sin(mid_angle),
      y = rlabel_out * cos(mid_angle),
      label = party,
      hjust = hjust, vjust = vjust
    ),
    size = 14/.pt
  )

```

Party composition of the 8th German Bundestag, 1976–1980
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Pie chart vs stacked bars vs side-by-side bars

<br>

```{r bundestag-stacked, echo = FALSE}
bundestag_stacked <- bundestag %>%
  mutate(
    label_y = cumsum(seats) - seats/2,
    colors = factor(colors, levels = rev(colors))
  )

bt_stacked <- 
  ggplot(bundestag_stacked, aes(x = 1, y = seats, fill = colors)) + 
  geom_col(position = "stack", color = "white") + 
  geom_text(
    aes(x = 1., y = label_y, label = seats), 
    size = 14/.pt,
    color = c("white", "white", "black")
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0)),
    breaks = c(0, 100, 200, 300, 400),
    sec.axis = dup_axis(
      breaks = bundestag_stacked$label_y,
      labels = bundestag_stacked$party
    )
  ) +
  scale_x_discrete(expand = c(0, 0), limits = "1", name = NULL) +
  scale_fill_identity() +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid() +
  theme(
    axis.ticks = element_line(color = "gray70"),
    axis.line.y.right = element_blank(),
    axis.ticks.y.right = element_blank(),
    axis.text.y.right = element_text(size = 12),
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_text(color = NA),
    axis.title.y.right = element_blank(),
    plot.margin = margin(7, 14, 7, 14)
  )
```

.center[
```{r bundestag-various, echo = FALSE, fig.width = 10.5, fig.asp = 0.3, dev = "svg"}
bt_bars <- bundestag %>%
  mutate(party = factor(party, levels = party)) %>%
  ggplot(aes(x = party, y = seats, fill = colors)) + 
  geom_col() + 
  geom_text(
    aes(label = seats),
    size = 14/.pt,
    vjust = 2,
    color = c("white", "white", "black")
  ) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.05)),
    breaks = c(0, 50, 100, 150, 200)
  ) +
  scale_fill_identity() + 
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank()
  )

plot_grid(
  bt_pie, bt_stacked, bt_bars,
  nrow = 1, rel_widths = c(1.05, .65, .9)
)
```
]

???

Figures redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Example where side-by-side bars are preferred

.center[
```{r marketshare-side-by-side, echo = FALSE, message = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
marketshare <- read_csv(here("datasets", "marketshare.csv"))

ggplot(marketshare, aes(x = company, y = percent, fill = company)) + 
  geom_col() + 
  facet_wrap(~year) +
  scale_y_continuous(
    name = "market share",
    labels = scales::percent_format(accuracy = 1, scale = 1),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#F0E442"),
    guide = "none"
  ) + 
  theme_half_open(16) +
  theme(strip.background = element_blank())
```
]

.xtiny-font.right[
Inspired by: https://en.wikipedia.org/wiki/File:Piecharts.svg
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Example where side-by-side bars are preferred

.center[
```{r marketshare-stacked, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
ggplot(marketshare, aes(x = year, y = percent, fill = company)) + 
  geom_col(position = "stack") + 
  scale_y_continuous(
    name = "market share",
    labels = scales::percent_format(accuracy = 1, scale = 1),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#F0E442")
  ) + 
  theme_half_open(16)
```
]

.xtiny-font.right[
Inspired by: https://en.wikipedia.org/wiki/File:Piecharts.svg
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Example where side-by-side bars are preferred

.center[
```{r marketshare-pies, echo = FALSE, fig.width = 8.5, fig.asp = 7*.618/8.5, dev = "svg"}
market_pies_df <- marketshare %>%
  group_by(year) %>%
  mutate(
    total = sum(percent),
    end_angle = 2*pi*cumsum(percent)/total,      # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for the text label
    hjust = ifelse(mid_angle>pi, 1, 0),
    vjust = ifelse(mid_angle<pi/2 | mid_angle>3*pi/2, 0, 1)
  )

rpie = 1
rlabel = 1.05 * rpie

ggplot(market_pies_df) + 
  geom_arc_bar(
    aes(
      x0 = 0, y0 = 0, r0 = 0, r = rpie,
      start = start_angle, end = end_angle, fill = company
    ),
    color = NA
  ) +
  geom_text(
    aes(x = rlabel*sin(mid_angle), y = rlabel*cos(mid_angle), label = company, hjust = hjust, vjust = vjust),
    size = 13.7/.pt
  ) +
  coord_fixed() +
  facet_wrap(~year) +
  scale_x_continuous(
    limits = c(-1.2, 1.2), expand = c(0, 0),
    name = NULL, breaks = NULL, labels = NULL
  ) +
  scale_y_continuous(
    limits = c(-1.2, 1.2), expand = c(0, 0),
    name = NULL, breaks = NULL, labels = NULL
  ) +
  scale_fill_manual(
    values = c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#F0E442"),
    guide = "none"
  ) + 
  theme_half_open(16) +
  theme(
    axis.line = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    axis.ticks.length = grid::unit(0, "pt"),
    legend.position = "none",
    strip.background = element_blank(),
    strip.text.x = element_text(size = 16, margin = margin(0, 0, 0.1, 0))
  )
```
]

.xtiny-font.right[
Inspired by: https://en.wikipedia.org/wiki/File:Piecharts.svg
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Example where side-by-side bars are preferred

.center[
```{r marketshare-side-by-side2, echo = FALSE, message = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
ggplot(marketshare, aes(x = company, y = percent, fill = company)) + 
  geom_col() + 
  facet_wrap(~year) +
  scale_y_continuous(
    name = "market share",
    labels = scales::percent_format(accuracy = 1, scale = 1),
    expand = c(0, 0)
  ) +
  scale_fill_manual(
    values = c("#E69F00", "#56B4E9", "#009E73", "#0072B2", "#F0E442"),
    guide = "none"
  ) + 
  theme_half_open(16) +
  theme(strip.background = element_blank())
```
]

.xtiny-font.right[
Inspired by: https://en.wikipedia.org/wiki/File:Piecharts.svg
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Example where stacked bars are preferred

.small-font.center[
```{r women-parliament, echo = FALSE, message = FALSE, fig.width = 7.5, fig.asp = .55, dev = "svg"}
ccode = "RWA" # Rwanda

women_parliaments <- read_csv(here("datasets", "women_parliaments.csv"))

rwanda_data <- women_parliaments %>%
  filter(country_code == ccode & year > 1990) %>%
  mutate(women = perc_women, men = 100 - perc_women) %>%
  select(-perc_women) %>%
  pivot_longer(cols = c(women, men), names_to = "gender", values_to = "percent") %>%
  mutate(gender = factor(gender, levels = c("women", "men")))

# calculate label position
rwanda_labels <- filter(rwanda_data, year == max(year)) %>%
  mutate(pos = 100 - cumsum(percent) + 0.5*percent)

ggplot(rwanda_data, aes(x = year, y = percent, fill = gender)) +
    geom_col(
      position = "stack",
      width = 1, color = "#FFFFFF",
      size = 0.75, alpha = 0.8
    ) +
    geom_hline(
      yintercept = 50,
      color = "#000000FF", size = 0.4, linetype = 2
    ) +
    geom_hline(yintercept = 100, color = "black") +
    scale_x_continuous(expand = c(0, 0)) +
    scale_y_continuous(
      name = "relative proportion",
      labels = scales::percent_format(accuracy = 1, scale = 1),
      expand = c(0, 0),
      sec.axis = dup_axis(
        breaks = rwanda_labels$pos,
        labels = rwanda_labels$gender
      )
    ) +
    scale_fill_manual(values = c("#D55E00E0", "#0072B2E0"), guide = "none") +
    coord_cartesian(clip = "off") +
    theme_half_open(16) + 
    theme(
      axis.line.y = element_blank(),
      axis.ticks.y.right = element_blank(),
      axis.text.y.right = element_text(size = 16),
      axis.title.y.right = element_blank()
    )
```

Change in the gender composition of the Rwandan parliament from 1997 to 2016

]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Pros and cons of different approaches

.small-font.center[
               | &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Pie chart &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | &nbsp;&nbsp;&nbsp; Stacked bars &nbsp;&nbsp;&nbsp; | Side-by-side bars
:----------    | :-------: | :----------: | :---------------:
Allows easy comparison of relative proportions   | ✖ | ✖ | ✔ 
]

---

## Pros and cons of different approaches

.small-font.center[
               | &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Pie chart &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | &nbsp;&nbsp;&nbsp; Stacked bars &nbsp;&nbsp;&nbsp; | Side-by-side bars
:----------    | :-------: | :----------: | :---------------:
Allows easy comparison of relative proportions   | ✖ | ✖ | ✔ 
Shows data as proportions of a whole             | ✔ | ✔ | ✖
]

---

## Pros and cons of different approaches

.small-font.center[
               | &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Pie chart &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | &nbsp;&nbsp;&nbsp; Stacked bars &nbsp;&nbsp;&nbsp; | Side-by-side bars
:----------    | :-------: | :----------: | :---------------:
Allows easy comparison of relative proportions   | ✖ | ✖ | ✔ 
Shows data as proportions of a whole             | ✔ | ✔ | ✖
Emphasizes simple fractions (1/2, 1/3, ...) | ✔ | ✖ | ✖
]

---

## Pros and cons of different approaches

.small-font.center[
               | &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Pie chart &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | &nbsp;&nbsp;&nbsp; Stacked bars &nbsp;&nbsp;&nbsp; | Side-by-side bars
:----------    | :-------: | :----------: | :---------------:
Allows easy comparison of relative proportions   | ✖ | ✖ | ✔ 
Shows data as proportions of a whole             | ✔ | ✔ | ✖
Emphasizes simple fractions (1/2, 1/3, ...) | ✔ | ✖ | ✖
Visually appealing for small datasets | ✔ | ✖ | ✔
]


---

## Pros and cons of different approaches

.small-font.center[
               | &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Pie chart &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | &nbsp;&nbsp;&nbsp; Stacked bars &nbsp;&nbsp;&nbsp; | Side-by-side bars
:----------    | :-------: | :----------: | :---------------:
Allows easy comparison of relative proportions   | ✖ | ✖ | ✔ 
Shows data as proportions of a whole             | ✔ | ✔ | ✖
Emphasizes simple fractions (1/2, 1/3, ...) | ✔ | ✖ | ✖
Visually appealing for small datasets | ✔ | ✖ | ✔
Works well for a large number of subsets | ✖ | ✖ | ✔ 
]

---

## Pros and cons of different approaches

.small-font.center[
               | &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; Pie chart &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; | &nbsp;&nbsp;&nbsp; Stacked bars &nbsp;&nbsp;&nbsp; | Side-by-side bars
:----------    | :-------: | :----------: | :---------------:
Allows easy comparison of relative proportions   | ✖ | ✖ | ✔ 
Shows data as proportions of a whole             | ✔ | ✔ | ✖
Emphasizes simple fractions (1/2, 1/3, ...) | ✔ | ✖ | ✖
Visually appealing for small datasets | ✔ | ✖ | ✔
Works well for a large number of subsets | ✖ | ✖ | ✔ 
Works well for time series and similar | ✖ | ✔ | ✖
]

--

No one visualization fits all scenarios!

[//]: # "segment ends here"

---
class: center middle

## Nested proportions: proportions of proportions

---
## Mosaic plots subdivide data along two dimensions

.center[
<img src = "https://clauswilke.com/dataviz/nested_proportions_files/figure-html/bridges-mosaic-1.png", width = 75%></img>
]

.absolute-bottom-right.tiny-font[
Dataset: Bridges in Pittsburgh by construction material and era of construction
]

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---
## Closely related to mosaic plot: Treemap

.center[
<img src = "https://clauswilke.com/dataviz/nested_proportions_files/figure-html/bridges-treemap-1.png", width = 65%></img>
]

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---
## Treemaps work well for more complex cases

.center.move-up-1em[
<img src = "https://clauswilke.com/dataviz/nested_proportions_files/figure-html/US-states-treemap-1.png", width = 80%></img>
]

.absolute-bottom-right.tiny-font[
Dataset: Land surface area of US states
]


???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---
## We can nest pie charts with clever coloring

.center.move-up-1em[
<img src = "https://clauswilke.com/dataviz/nested_proportions_files/figure-html/bridges-nested-pie2-1.png", width = 70%></img>
]

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---
## Parallel sets can show many subdivisions at once

.center[
<img src = "https://clauswilke.com/dataviz/nested_proportions_files/figure-html/bridges-parallel-sets1-1.png", width = 75%></img>
]

???

Figure from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

[//]: # "segment ends here"

---
class: center middle

## Making pie charts with **ggplot2**


---

## Making pie charts with **ggplot2**

We have three options:

- `geom_bar()`/`geom_col()` with poolar coordinates  
  Pros: simple  
  Cons: hard to customize
  
--

- `geom_arc_bar()` with `stat_pie()`  
  Pros: relatively simple, some customization  
  Cons: requires **ggforce** & some more complex code
  
--


- `geom_arc_bar()` with manual computation  
  Pros: maximum flexibility for customization  
  Cons: requires **ggforce** & much more complex code
  

---

## Making pie charts with **ggplot2**: polar coords

.pull-left.tiny-font[
```{r}
# the data
bundestag <- tibble(
  party = c("CDU/CSU", "SPD", "FDP"),
  seats = c(243, 214, 39)
)

bundestag
```
]

---

## Making pie charts with **ggplot2**: polar coords

.pull-left.tiny-font[
```{r bundestag-coord-polar1, eval = FALSE}
# the data
bundestag <- tibble(
  party = c("CDU/CSU", "SPD", "FDP"),
  seats = c(243, 214, 39)
)

# make bar chart
ggplot(bundestag) +
  aes(seats, "YY", fill = party) + 
  geom_col()
```
]

.pull-right[
```{r bundestag-coord-polar1-out, ref.label = "bundestag-coord-polar1", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Making pie charts with **ggplot2**: polar coords

.pull-left.tiny-font[
```{r bundestag-coord-polar2, eval = FALSE}
# the data
bundestag <- tibble(
  party = c("CDU/CSU", "SPD", "FDP"),
  seats = c(243, 214, 39)
)

# make bar chart in polar coords
ggplot(bundestag) +
  aes(seats, "YY", fill = party) + 
  geom_col() +
  coord_polar()
```
]

.pull-right[
```{r bundestag-coord-polar2-out, ref.label = "bundestag-coord-polar2", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Making pie charts with **ggplot2**: polar coords

.pull-left.tiny-font[
```{r bundestag-coord-polar3, eval = FALSE}
# the data
bundestag <- tibble(
  party = c("CDU/CSU", "SPD", "FDP"),
  seats = c(243, 214, 39)
)

# make bar chart in polar coords
ggplot(bundestag) +
  aes(seats, "YY", fill = party) + 
  geom_col() +
  coord_polar() +
  scale_x_continuous(
    name = NULL, breaks = NULL
  ) +
  scale_y_discrete(
    name = NULL, breaks = NULL
  ) +
  ggtitle("German Bundestag 1976-1980")
```
]

.pull-right[
```{r bundestag-coord-polar3-out, ref.label = "bundestag-coord-polar3", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]



---

## Making pie charts with **ggplot2**: **ggforce** stat pie

.pull-left.tiny-font[
```{r bundestag-stat-pie1, eval = FALSE}
library(ggforce)

ggplot(bundestag) +
  aes(
    x0 = 0, y0 = 0, # position of pie center
    r0 = 0, r = 1,  # inner and outer radius
    amount = seats, # size of pie slices
    fill = party
  ) + 
  geom_arc_bar(stat = "pie")
```
]

.pull-right[
```{r bundestag-stat-pie1-out, ref.label = "bundestag-stat-pie1", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Making pie charts with **ggplot2**: **ggforce** stat pie

.pull-left.tiny-font[
```{r bundestag-stat-pie2, eval = FALSE}
library(ggforce)

ggplot(bundestag) +
  aes(
    x0 = 0, y0 = 0, # position of pie center
    r0 = 0, r = 1,  # inner and outer radius
    amount = seats, # size of pie slices
    fill = party
  ) + 
  geom_arc_bar(stat = "pie") +
  coord_fixed()
```
]

.pull-right[
```{r bundestag-stat-pie2-out, ref.label = "bundestag-stat-pie2", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Making pie charts with **ggplot2**: **ggforce** stat pie

.pull-left.tiny-font[
```{r bundestag-stat-pie3, eval = FALSE}
library(ggforce)

ggplot(bundestag) +
  aes(
    x0 = 0, y0 = 0, # position of pie center
    r0 = 0, r = 1,  # inner and outer radius
    amount = seats, # size of pie slices
    fill = party
  ) + 
  geom_arc_bar(stat = "pie") +
  coord_fixed(
    xlim = c(-1, 3), ylim = c(-1, 3)
  )
```
]

.pull-right[
```{r bundestag-stat-pie3-out, ref.label = "bundestag-stat-pie3", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Making pie charts with **ggplot2**: **ggforce** stat pie

.pull-left.tiny-font[
```{r bundestag-stat-pie4, eval = FALSE}
library(ggforce)

ggplot(bundestag) +
  aes(
    x0 = 1, y0 = 1, # position of pie center
    r0 = 0, r = 1,  # inner and outer radius
    amount = seats, # size of pie slices
    fill = party
  ) + 
  geom_arc_bar(stat = "pie") +
  coord_fixed(
    xlim = c(-1, 3), ylim = c(-1, 3)
  )
```
]

.pull-right[
```{r bundestag-stat-pie4-out, ref.label = "bundestag-stat-pie4", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Making pie charts with **ggplot2**: **ggforce** stat pie

.pull-left.tiny-font[
```{r bundestag-stat-pie5, eval = FALSE}
library(ggforce)

ggplot(bundestag) +
  aes(
    x0 = 1, y0 = 1, # position of pie center
    r0 = 1, r = 2,  # inner and outer radius
    amount = seats, # size of pie slices
    fill = party
  ) + 
  geom_arc_bar(stat = "pie") +
  coord_fixed(
    xlim = c(-1, 3), ylim = c(-1, 3)
  )
```
]

.pull-right[
```{r bundestag-stat-pie5-out, ref.label = "bundestag-stat-pie5", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.tiny-font[
```{r}
# prepare pie data
pie_data <- bundestag %>%
  arrange(seats) # sort so pie slices end up sorted

pie_data
```
]


---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.tiny-font[
```{r}
# prepare pie data
pie_data <- bundestag %>%
  arrange(seats) %>% # sort so pie slices end up sorted
  mutate(
    end_angle = 2*pi*cumsum(seats)/sum(seats)    # ending angle for each pie slice
  )

pie_data
```
]


---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.tiny-font[
```{r}
# prepare pie data
pie_data <- bundestag %>%
  arrange(seats) %>% # sort so pie slices end up sorted
  mutate(
    end_angle = 2*pi*cumsum(seats)/sum(seats),   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0)    # starting angle for each pie slice
  )

pie_data
```
]

---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.tiny-font[
```{r}
# prepare pie data
pie_data <- bundestag %>%
  arrange(seats) %>% # sort so pie slices end up sorted
  mutate(
    end_angle = 2*pi*cumsum(seats)/sum(seats),   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for text labels
  )

pie_data
```
]

---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.tiny-font[
```{r}
# prepare pie data
pie_data <- bundestag %>%
  arrange(seats) %>% # sort so pie slices end up sorted
  mutate(
    end_angle = 2*pi*cumsum(seats)/sum(seats),   # ending angle for each pie slice
    start_angle = lag(end_angle, default = 0),   # starting angle for each pie slice
    mid_angle = 0.5*(start_angle + end_angle),   # middle of each pie slice, for text labels
    # horizontal and vertical justifications for outer labels
    hjust = ifelse(mid_angle > pi, 1, 0),
    vjust = ifelse(mid_angle < pi/2 | mid_angle > 3*pi/2, 0, 1)
  )

pie_data
```
]

---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.pull-left.tiny-font[
```{r bundestag-arc-bar1, eval = FALSE}
ggplot(pie_data) +
  aes(
    x0 = 0, y0 = 0, r0 = 0, r = 1,
    start = start_angle, end = end_angle,
    fill = party
  ) +
  geom_arc_bar() +
  coord_fixed()
```
]

.pull-right[
```{r bundestag-arc-bar1-out, ref.label = "bundestag-arc-bar1", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.pull-left.tiny-font[
```{r bundestag-arc-bar2, eval = FALSE}
ggplot(pie_data) +
  aes(
    x0 = 0, y0 = 0, r0 = 0, r = 1,
    start = start_angle, end = end_angle,
    fill = party
  ) +
  geom_arc_bar() +
  geom_text( # place amounts inside the pie
    aes(
      x = 0.6 * sin(mid_angle),
      y = 0.6 * cos(mid_angle),
      label = seats
    )
  ) +
  coord_fixed()
```
]

.pull-right[
```{r bundestag-arc-bar2-out, ref.label = "bundestag-arc-bar2", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.pull-left.tiny-font[
```{r bundestag-arc-bar3, eval = FALSE}
ggplot(pie_data) +
  aes(
    x0 = 0, y0 = 0, r0 = 0, r = 1,
    start = start_angle, end = end_angle,
    fill = party
  ) +
  geom_arc_bar() +
  geom_text( # place amounts inside the pie
    aes(
      x = 0.2 * sin(mid_angle),
      y = 0.2 * cos(mid_angle),
      label = seats
    )
  ) +
  coord_fixed()
```
]

.pull-right[
```{r bundestag-arc-bar3-out, ref.label = "bundestag-arc-bar3", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.pull-left.tiny-font[
```{r bundestag-arc-bar4, eval = FALSE}
ggplot(pie_data) +
  aes(
    x0 = 0, y0 = 0, r0 = 0, r = 1,
    start = start_angle, end = end_angle,
    fill = party
  ) +
  geom_arc_bar() +
  geom_text( # place amounts inside the pie
    aes(
      x = 0.8 * sin(mid_angle),
      y = 0.8 * cos(mid_angle),
      label = seats
    )
  ) +
  coord_fixed()
```
]

.pull-right[
```{r bundestag-arc-bar4-out, ref.label = "bundestag-arc-bar4", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.pull-left.tiny-font[
```{r bundestag-arc-bar5, eval = FALSE}
ggplot(pie_data) +
  aes(
    x0 = 0, y0 = 0, r0 = 0, r = 1,
    start = start_angle, end = end_angle,
    fill = party
  ) +
  geom_arc_bar() +
  geom_text( # place amounts inside the pie
    aes(
      x = 0.6 * sin(mid_angle),
      y = 0.6 * cos(mid_angle),
      label = seats
    )
  ) +
  coord_fixed()
```
]

.pull-right[
```{r bundestag-arc-bar5-out, ref.label = "bundestag-arc-bar5", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.pull-left.xtiny-font[
```{r bundestag-arc-bar6, eval = FALSE}
ggplot(pie_data) +
  aes(
    x0 = 0, y0 = 0, r0 = 0, r = 1,
    start = start_angle, end = end_angle,
    fill = party
  ) +
  geom_arc_bar() +
  geom_text( # place amounts inside the pie
    aes(
      x = 0.6 * sin(mid_angle),
      y = 0.6 * cos(mid_angle),
      label = seats
    )
  ) +
  geom_text( # place party name outside the pie
    aes(
      x = 1.05 * sin(mid_angle),
      y = 1.05 * cos(mid_angle),
      label = party,
      hjust = hjust, vjust = vjust
    )
  ) +
  coord_fixed()
```
]

.pull-right[
```{r bundestag-arc-bar6-out, ref.label = "bundestag-arc-bar6", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.pull-left.xtiny-font[
```{r bundestag-arc-bar7, eval = FALSE}
ggplot(pie_data) +
  aes(
    x0 = 0, y0 = 0, r0 = 0, r = 1,
    start = start_angle, end = end_angle,
    fill = party
  ) +
  geom_arc_bar() +
  geom_text( # place amounts inside the pie
    aes(
      x = 0.6 * sin(mid_angle),
      y = 0.6 * cos(mid_angle),
      label = seats
    )
  ) +
  geom_text( # place party name outside the pie
    aes(
      x = 1.05 * sin(mid_angle),
      y = 1.05 * cos(mid_angle),
      label = party,
      hjust = hjust, vjust = vjust
    )
  ) +
  coord_fixed(xlim = c(-1.8, 1.3))
```
]

.pull-right[
```{r bundestag-arc-bar7-out, ref.label = "bundestag-arc-bar7", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Making pie charts with **ggplot2**: **ggforce** manual comp.

.pull-left.xtiny-font[
```{r bundestag-arc-bar8, eval = FALSE}
ggplot(pie_data) +
  aes(
    x0 = 0, y0 = 0, r0 = 0.4, r = 1,
    start = start_angle, end = end_angle,
    fill = party
  ) +
  geom_arc_bar() +
  geom_text( # place amounts inside the pie
    aes(
      x = 0.7 * sin(mid_angle),
      y = 0.7 * cos(mid_angle),
      label = seats
    )
  ) +
  geom_text( # place party name outside the pie
    aes(
      x = 1.05 * sin(mid_angle),
      y = 1.05 * cos(mid_angle),
      label = party,
      hjust = hjust, vjust = vjust
    )
  ) +
  coord_fixed(xlim = c(-1.8, 1.3))
```
]

.pull-right[
```{r bundestag-arc-bar8-out, ref.label = "bundestag-arc-bar8", echo = FALSE, message = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

[//]: # "segment ends here"

---

## Further reading

- Fundamentals of Data Visualization: [Chapter 10: Visualizing proportions](https://clauswilke.com/dataviz/visualizing-proportions.html)
- Fundamentals of Data Visualization: [Chapter 11: Visualizing nested proportions](https://clauswilke.com/dataviz/nested-proportions.html)
- **ggplot2** reference documentation: [`position_stack()`, `position_fill()`](https://ggplot2.tidyverse.org/reference/position_stack.html)
- **ggplot2** reference documentation: [`position_dodge()`](https://ggplot2.tidyverse.org/reference/position_dodge.html)
- **ggforce** reference documentation: [`geom_arc_bar()`](https://ggforce.data-imaginist.com/reference/geom_arc_bar.html)




