---
title: "Visualizing distributions 1"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(cowplot)
library(colorspace)
library(here)
# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width

titanic <- read_csv(here("datasets", "titanic.csv")) %>%
  select(age, sex, class, survived) %>%
  arrange(age, sex, class)
```

class: center middle

## Histograms and density plots

---

## Passengers on the Titanic

.center.small-font[
```{r echo = FALSE}
knitr::kable(titanic[1:10,], "html")
```
&nbsp;&nbsp;&nbsp;&nbsp;
```{r echo = FALSE}
knitr::kable(titanic[11:20,], "html")
```
&nbsp;&nbsp;&nbsp;&nbsp;
```{r echo = FALSE}
knitr::kable(titanic[21:30,], "html")
```
]

---

## Histogram: Define bins and count cases

```{r echo = FALSE}
age_counts <- hist(titanic$age, breaks = (0:15) * 5 + .01, plot = FALSE)$counts
age_hist <- tibble(
  `age range` = c("0–5", "6–10", "11–15", "16–20", "21–25", "26–30", "31–35", "36–40", "41–45", "46–50", "51–55", "56–60", "61–65", "66–70", "71–75", "76–80"),
  count = c(age_counts, 0)
)
```

.pull-left.small-font[
```{r echo = FALSE}
knitr::kable(age_hist[1:8,], "html")
```
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;
```{r echo = FALSE}
knitr::kable(age_hist[9:16,], "html")
```
]

--

.pull-right[
```{r titanic-age-hist, echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
age_hist5 <- age_hist %>%
  mutate(age = (1:16) * 5 - 2.5) %>%
  filter(count > 0)

h1 <- ggplot(age_hist5, aes(x = age, y = count)) + 
  geom_col(width = 4.7, fill = "#56B4E9", na.rm = TRUE)  + 
  scale_y_continuous(expand = c(0, 0), breaks = 25 * (0:5)) +
  scale_x_continuous(name = "age (years)", limits = c(0, 75), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(18) +
  theme(
    axis.line.x = element_blank()
  )

h1
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

# Histograms depend on the chosen bin width

.center[
```{r titanic-age-hist-binwidth, echo = FALSE, fig.width = 8, fig.asp = 0.618, dev = "svg"}
age_hist_1 <- tibble(
  age = (1:75) - 0.5, 
  count = hist(titanic$age, breaks = (0:75) + .01, plot = FALSE)$counts
)
age_hist_3 <- tibble(
  age = (1:25) * 3 - 1.5, 
  count = hist(titanic$age, breaks = (0:25) * 3 + .01, plot = FALSE)$counts
)
age_hist_15 <- tibble(
  age = (1:5) * 15 - 7.5, 
  count = hist(titanic$age, breaks = (0:5) * 15 + .01, plot = FALSE)$counts
)


h2 <- ggplot(age_hist_1, aes(x = age, y = count)) + 
  geom_col(width = .85, fill = "#56B4E9")  + 
  scale_y_continuous(expand = c(0, 0), breaks = 10 * (0:5)) +
  scale_x_continuous(name = "age (years)", limits = c(0, 75), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(14) +
  theme(
    axis.line.x = element_blank()
  )

h3 <- ggplot(age_hist_3, aes(x = age, y = count)) + 
  geom_col(width = 2.75, fill = "#56B4E9")  + 
  scale_y_continuous(expand = c(0, 0), breaks = 25 * (0:5)) +
  scale_x_continuous(name = "age (years)", limits = c(0, 75), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(14) +
  theme(
    axis.line.x = element_blank()
  )

h4 <- ggplot(age_hist_15, aes(x = age, y = count)) + 
  geom_col(width = 14.5, fill = "#56B4E9")  + 
  scale_y_continuous(expand = c(0, 0), breaks = 100 * (0:4)) +
  scale_x_continuous(name = "age (years)", limits = c(0, 75), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(14) +
  theme(
    axis.line.x = element_blank()
  )

h1 <- h1 + theme_minimal_hgrid(16) +
  theme(
    axis.line.x = element_blank()
  )

plot_grid(
  h2, NULL, h3,
  NULL, NULL, NULL,
  h1, NULL, h4,
  align = 'hv',
  rel_widths = c(1, .04, 1),
  rel_heights = c(1, .04, 1)
)
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---


## Alternative to histogram: Kernel density estimate (KDE)

.pull-left[
```{r titanic-age-hist2, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
h1 + ggtitle("Histogram") + theme(plot.title.position = "plot")
```
]

--

.pull-right[
```{r titanic-age-kde, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
ggplot(titanic, aes(x = age)) + 
  geom_density(fill = "#56B4E9", color = darken("#56B4E9", 0.5), bw = 2, kernel = "gaussian") + 
  scale_y_continuous(limits = c(0, 0.046), expand = c(0, 0), name = "density") +
  scale_x_continuous(name = "age (years)", limits = c(0, 75), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  ggtitle("Kernel density estimate") +
  theme_minimal_hgrid(18) +
  theme(
    axis.line.x = element_blank(),
    plot.title.position = "plot"
  )
```
]

--

Histograms show raw counts, KDEs show proportions. (Total area = 1)

???

Figures redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---


## KDEs also depend on parameter settings

.center[
```{r titanic-age-kde-grid, echo = FALSE, fig.width = 8, fig.asp = 0.618, dev = "svg"}
pdens1 <- ggplot(titanic, aes(x = age)) + 
  geom_density(fill = "#56B4E9", color = darken("#56B4E9", 0.5), bw = .5, kernel = "gaussian") + 
  scale_y_continuous(limits = c(0, 0.046), expand = c(0, 0), name = "density") +
  scale_x_continuous(name = "age (years)", limits = c(0, 75), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(14) +
  theme(axis.line.x = element_blank())

pdens2 <- ggplot(titanic, aes(x = age)) + 
  geom_density(fill = "#56B4E9", color = darken("#56B4E9", 0.5), bw = 2, kernel = "gaussian") + 
  scale_y_continuous(limits = c(0, 0.046), expand = c(0, 0), name = "density") +
  scale_x_continuous(name = "age (years)", limits = c(0, 75), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(14) +
  theme(axis.line.x = element_blank())

pdens3 <- ggplot(titanic, aes(x = age)) + 
  geom_density(fill = "#56B4E9", color = darken("#56B4E9", 0.5), bw = 5, kernel = "gaussian") + 
  scale_y_continuous(limits = c(0, 0.046), expand = c(0, 0), name = "density") +
  scale_x_continuous(name = "age (years)", limits = c(0, 75), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(14) +
  theme(axis.line.x = element_blank())

pdens4 <- ggplot(titanic, aes(x = age)) + 
  geom_density(fill = "#56B4E9", color = darken("#56B4E9", 0.5), bw = 2, kernel = "rectangular") + 
  scale_y_continuous(limits = c(0, 0.046), expand = c(0, 0), name = "density") +
  scale_x_continuous(name = "age (years)", limits = c(0, 75), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(14) +
  theme(axis.line.x = element_blank())

plot_grid(
  pdens1, NULL, pdens2, 
  NULL, NULL, NULL,
  pdens3, NULL, pdens4,
  align = 'hv',
  rel_widths = c(1, .04, 1),
  rel_heights = c(1, .04, 1)
)
```
]


???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Careful: KDEs can show non-sensical data

.center[
```{r titanic-age-kde-wrong, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
pdens_neg <- ggplot(titanic, aes(x = age)) + 
  geom_density(fill = "#56B4E9", color = darken("#56B4E9", 0.5)) + 
  scale_y_continuous(limits = c(0, 0.046), expand = c(0, 0), name = "density") +
  scale_x_continuous(name = "age (years)", limits = c(-10, 79), expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(16) +
  theme(
    axis.line.x = element_blank()
  )
stamp_wrong(pdens_neg)
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Careful: Are bars stacked or overlapping?

.pull-left[
```{r titanic-age-hist-stacked, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
gender_counts <- tibble(
  age = (1:25)*3 - 1.5, 
  male = hist(filter(titanic, sex == "male")$age, breaks = (0:25)*3 + .01, plot = FALSE)$counts,
  female = hist(filter(titanic, sex == "female")$age, breaks = (0:25)*3 + .01, plot = FALSE)$counts
) %>%
  pivot_longer(-age, names_to = "gender", values_to = "count")

gender_counts$gender <- factor(gender_counts$gender, levels = c("female", "male"))

ggplot(gender_counts, aes(x = age, y = count, fill = gender)) + 
  geom_col(position = "stack") +
  scale_x_continuous(name = "age (years)", limits = c(0, 75), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 89), expand = c(0, 0), name = "count") +
  scale_fill_manual(values = c("#D55E00", "#0072B2")) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(18) +
  theme(
    axis.line.x = element_blank(),
    legend.position = c(.9, .87),
    legend.justification = c("right", "top"),
    legend.box.background = element_rect(fill = "white", color = "white")
  )
```
]

--

.pull-right[
```{r titanic-age-hist-overlap, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
ggplot(gender_counts, aes(x = age, y = count, fill = gender)) + 
  geom_col(position = "identity", alpha = 0.7) +
  scale_x_continuous(name = "age (years)", limits = c(0, 75), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 56), expand = c(0, 0), name = "count") +
  scale_fill_manual(
    values = c("#D55E00", "#0072B2"),
    guide = guide_legend(reverse = TRUE)
  ) +
  coord_cartesian(clip = "off") +
  theme_minimal_hgrid(18) +
  theme(
    axis.line.x = element_blank(),
    legend.position = c(.9, .87),
    legend.justification = c("right", "top"),
    legend.box.background = element_rect(fill = "white", color = "white")
  )
```
]

--

Stacked or overlapping histograms are rarely a good choice.

???

Figures redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Alternatively: Age pyramid

.center[
```{r titanic-age-pyramid, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
ggplot(gender_counts, aes(x = age, y = ifelse(gender == "male", -1, 1)*count, fill = gender)) + 
  geom_col() +
  scale_x_continuous(name = "age (years)", limits = c(0, 75), expand = c(0, 0)) +
  scale_y_continuous(name = "count", breaks = 20*(-2:1), labels = c("40", "20", "0", "20")) +
  scale_fill_manual(values = c("#D55E00", "#0072B2"), guide = "none") +
  draw_text(x = 70, y = -39, "male", hjust = 0) +
  draw_text(x = 70, y = 21, "female", hjust = 0) +
  coord_flip() +
  theme_minimal_grid(16) +
  theme(axis.title.x = element_text(hjust = 0.61))
```
]


???

Figures redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Alternatively: KDEs showing proportions of total

.center[
```{r titanic-age-props, echo = FALSE, fig.width = 10, fig.asp = 0.45, dev = "svg"}
titanic %>%
  mutate(sex = factor(sex, levels = c("male", "female"))) %>%
  ggplot(aes(x = age, y = after_stat(count))) +
  geom_density(
    data = select(titanic, -sex), aes(fill = "all passengers"),
    color = "transparent"
  ) + 
  geom_density(aes(fill = sex), bw = 2, color = "transparent") +
  scale_x_continuous(limits = c(0, 75), name = "passenger age (years)", expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 26), name = "scaled density", expand = c(0, 0)) +
  scale_fill_manual(
    values = c("#b3b3b3a0", "#0072B2", "#D55E00"), 
    breaks = c("all passengers", "male", "female"),
    labels = c("all passengers  ", "males  ", "females"),
    name = NULL,
    guide = guide_legend(direction = "horizontal")
  ) +
  coord_cartesian(clip = "off") +
  facet_wrap(vars(sex), labeller = labeller(sex = function(sex) paste(sex, "passengers"))) +
  theme_minimal_hgrid(16) +
  theme(
    axis.line.x = element_blank(),
    strip.text = element_text(size = 14, margin = margin(0, 0, 0.2, 0, "cm")),
    legend.position = "bottom",
    legend.justification = "right",
    legend.margin = margin(4.5, 0, 1.5, 0, "pt"),
    legend.spacing.x = grid::unit(4.5, "pt"),
    legend.spacing.y = grid::unit(0, "pt"),
    legend.box.spacing = grid::unit(0, "cm")
  )
```
]

???

Figures redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

[//]: # "segment ends here"


---
class: center middle

## Histograms and density plots in **ggplot2**


---

## Making histograms with ggplot: `geom_histogram()`

.small-font[
```{r titanic-hist-ggplot-demo, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_histogram()
```
]

--

.center.small-font[
```{r titanic-hist-ggplot-demo-out, ref.label = "titanic-hist-ggplot-demo", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Setting the bin width

.small-font[
```{r titanic-hist-ggplot-demo2, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_histogram(binwidth = 5)
```
]


.center[
```{r titanic-hist-ggplot-demo2-out, ref.label = "titanic-hist-ggplot-demo2", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

--

Do you like the bin placement?

---

## Always set the center as well

.small-font[
```{r titanic-hist-ggplot-demo3, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_histogram(
    binwidth = 5,  # width of the bins
    center = 2.5   # center of the bin containing that value
  )
```
]


.center[
```{r titanic-hist-ggplot-demo3-out, ref.label = "titanic-hist-ggplot-demo3", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Always set the center as well

.small-font[
```{r titanic-hist-ggplot-demo4, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_histogram(
    binwidth = 5,  # width of the bins
    center = 12.5  # center of the bin containing that value
  )
```
]


.center[
```{r titanic-hist-ggplot-demo4-out, ref.label = "titanic-hist-ggplot-demo4", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Making density plots with ggplot: `geom_density()`

.small-font[
```{r titanic-dens-ggplot-demo, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_density(fill = "skyblue")
```
]

--

.center[
```{r titanic-dens-ggplot-demo-out, ref.label = "titanic-dens-ggplot-demo", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Making density plots with ggplot: `geom_density()`

.small-font[
```{r titanic-dens-ggplot-demo2, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_density() # without fill
```
]

.center[
```{r titanic-dens-ggplot-demo2-out, ref.label = "titanic-dens-ggplot-demo2", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---


## Modifying bandwidth (`bw`) and kernel parameters

.tiny-font[
```{r titanic-dens-ggplot-demo3, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_density(
    fill = "skyblue",
    bw = 0.5,               # a small bandwidth
    kernel = "gaussian"     # Gaussian kernel (the default)
  )

```
]

.center[
```{r titanic-dens-ggplot-demo3-out, ref.label = "titanic-dens-ggplot-demo3", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Modifying bandwidth (`bw`) and kernel parameters

.tiny-font[
```{r titanic-dens-ggplot-demo4, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_density(
    fill = "skyblue",
    bw = 2,                 # a moderate bandwidth
    kernel = "rectangular"  # rectangular kernel
  )

```
]

.center[
```{r titanic-dens-ggplot-demo4-out, ref.label = "titanic-dens-ggplot-demo4", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

[//]: # "segment ends here"


---
class: center middle

## Setting stats explicitly in **ggplot2**

---

## Statistical transformations (stats) can be set explicitly

.tiny-font[
```{r titanic-stat-demo1, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_density(
    stat = "density",    # the default for geom_density()
    fill = "skyblue"
  )
```
]

.center[
```{r titanic-stat-demo1-out, ref.label = "titanic-stat-demo1", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]


---

## Statistical transformations (stats) can be set explicitly

.tiny-font[
```{r titanic-stat-demo2, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_area(  # geom_area() does not normally use stat = "density"
    stat = "density",
    fill = "skyblue"
  )
```
]

.center[
```{r titanic-stat-demo2-out, ref.label = "titanic-stat-demo2", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Statistical transformations (stats) can be set explicitly

.tiny-font[
```{r titanic-stat-demo3, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_line(  # neither does geom_line()
    stat = "density"
  )
```
]

.center[
```{r titanic-stat-demo3-out, ref.label = "titanic-stat-demo3", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Statistical transformations (stats) can be set explicitly

.tiny-font[
```{r titanic-stat-demo4, eval = FALSE}
ggplot(titanic, aes(age)) +
  # we can use multiple geoms on top of each other
  geom_area(stat = "density", fill = "skyblue") +
  geom_line(stat = "density")
```
]

.center[
```{r titanic-stat-demo4-out, ref.label = "titanic-stat-demo4", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]

---

## Parameters are handed through to the stat

.pull-left.tiny-font[
```{r titanic-stat-demo5, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_line(stat = "density", bw = 3)
```

.center[
```{r titanic-stat-demo5-out, ref.label = "titanic-stat-demo5", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]]

.pull-right.tiny-font[
```{r titanic-stat-demo6, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_line(stat = "density", bw = 0.3)
```

.center[
```{r titanic-stat-demo6-out, ref.label = "titanic-stat-demo6", echo = FALSE, fig.width = 5.5, fig.asp = 0.618, dev = "svg"}
```
]]

--

Here, `bw` is a parameter of `stat_density()`, not of `geom_line()`.

---

## We can explicitly map results from stat computations

.tiny-font[
```{r titanic-stat-demo7, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_tile( # geom_tile() draws rectangular colored areas
    aes(
      y = 1, # draw all tiles at the same y location
      fill = after_stat(density)  # use computed density for fill
    ),
    stat = "density",
    n = 20    # number of points calculated by stat_density() 
  ) 
```
]

.center[
```{r titanic-stat-demo7-out, ref.label = "titanic-stat-demo7", echo = FALSE, fig.width = 10, fig.asp = 0.3, dev = "svg"}
```
]


---

## We can explicitly map results from stat computations

.tiny-font[
```{r titanic-stat-demo8, eval = FALSE}
ggplot(titanic, aes(age)) +
  geom_tile( # geom_tile() draws rectangular colored areas
    aes(
      y = 1, # draw all tiles at the same y location
      fill = after_stat(density)  # use computed density for fill
    ),
    stat = "density",
    n = 200   # number of points calculated by stat_density() 
  ) 
```
]

.center[
```{r titanic-stat-demo8-out, ref.label = "titanic-stat-demo8", echo = FALSE, fig.width = 10, fig.asp = 0.3, dev = "svg"}
```
]

[//]: # "segment ends here"

---

## Further reading

- Fundamentals of Data Visualization: [Chapter 7: Visualizing distributions](https://clauswilke.com/dataviz/histograms-density-plots.html)
- Data Visualization—A Practical Introduction: [4.6 Histograms and density plots](https://socviz.co/groupfacettx.html#histograms)
- **ggplot2** reference documentation: [`geom_histogram()`](https://ggplot2.tidyverse.org/reference/geom_histogram)
- **ggplot2** reference documentation: [`geom_density()`](https://ggplot2.tidyverse.org/reference/geom_density)




