---
title: "Visualizing trends"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(here)
library(ggforce)
library(cowplot)
# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width

blue_jays <- read_csv(here("datasets", "blue_jays.csv"))
preprints <- read_csv(here("datasets", "preprints.csv"))
hpi_trends <- read_csv(here("datasets", "hpi_trends.csv"))
cars93 <- read_csv(here("datasets", "cars93.csv"))
```

## We visualize linear trends with regression lines

.center[
```{r blue-jays-scatter, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  ggplot(aes(body_mass_g, head_length_mm)) + 
  geom_point(aes(fill = "blue jay birds"), pch = 21, color = "white", size = 2.5) +
  scale_fill_manual(
    values = c(`blue jay birds` = "black", `regression line` = "black"),
    limits = c("blue jay birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(
        size = 2.5,
        linetype = 0
      )
    )
  ) +
  scale_color_manual(
    values = c(`blue jay birds` = "black", `regression line` = "black"),
    limits = c("blue jay birds"),
    name = NULL
  ) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## We visualize linear trends with regression lines

.center[
```{r blue-jays-scatter-lm, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  ggplot(aes(body_mass_g, head_length_mm)) + 
  geom_point(aes(fill = "blue jay birds"), pch = 21, color = "white", size = 2.5) +
  geom_smooth(aes(color = "regression line"), method = "lm", size = 0.75, se = FALSE, fullrange = TRUE) +
  scale_fill_manual(
    values = c(`blue jay birds` = "black", `regression line` = "black"),
    breaks = c("regression line", "blue jay birds"),
    labels = c("regression line   ", "blue jay birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(
        size = c(0.75, 2.5),
        linetype = c(1, 0)
      )
    )
  ) +
  scale_color_manual(
    values = c(`blue jay birds` = "black", `regression line` = "black"),
    breaks = c("regression line", "blue jay birds"),
    labels = c("regression line   ", "blue jay birds"),
    name = NULL
  ) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## We visualize linear trends with regression lines

.center[
```{r blue-jays-scatter-line-sex, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  ggplot(aes(body_mass_g, head_length_mm)) + 
  geom_point(aes(fill = sex), pch = 21, color = "white", size = 2.5) +
  #geom_smooth(method = "lm", size = 0.75, se = FALSE, color = "black", fullrange = TRUE) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  scale_fill_manual(
    values = c(F = "#D55E00C0", M = "#0072B2C0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 2.5, linetype = 0)
    )
  ) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL
  ) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---


## We visualize linear trends with regression lines

.center[
```{r blue-jays-scatter-female, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  filter(sex == "F") %>%
  ggplot(aes(body_mass_g, head_length_mm, color = sex, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  scale_fill_manual(
    values = c(F = "#D55E00C0", M = "#0072B2C0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 2.5, linetype = 0)
    )
  ) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL
  ) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## We visualize linear trends with regression lines

.center[
```{r blue-jays-scatter-line-female, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  filter(sex == "F") %>%
  ggplot(aes(body_mass_g, head_length_mm, color = sex, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  geom_smooth(method = "lm", size = 0.75, se = FALSE, fullrange = TRUE) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  scale_fill_manual(
    values = c(F = "#D55E00C0", M = "#0072B2C0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 2.5, linetype = 0)
    )
  ) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL
  ) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## We visualize linear trends with regression lines

.center[
```{r blue-jays-scatter-male, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  filter(sex == "M") %>%
  ggplot(aes(body_mass_g, head_length_mm, color = sex, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  scale_fill_manual(
    values = c(F = "#D55E00C0", M = "#0072B2C0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 2.5, linetype = 0)
    )
  ) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL
  ) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## We visualize linear trends with regression lines

.center[
```{r blue-jays-scatter-line-male, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  filter(sex == "M") %>%
  ggplot(aes(body_mass_g, head_length_mm, color = sex, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  geom_smooth(method = "lm", size = 0.75, se = FALSE, fullrange = TRUE) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  scale_fill_manual(
    values = c(F = "#D55E00C0", M = "#0072B2C0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 2.5, linetype = 0)
    )
  ) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL
  ) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## We visualize linear trends with regression lines

.center[
```{r blue-jays-scatter-sex-line, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  ggplot(aes(body_mass_g, head_length_mm, color = sex, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  geom_smooth(method = "lm", size = 0.75, se = FALSE, fullrange = TRUE) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  scale_fill_manual(
    values = c(F = "#D55E00C0", M = "#0072B2C0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 2.5, linetype = 0)
    )
  ) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL
  ) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Exponential trends are linear trends on a log scale

.center[
```{r biorxiv-nofit, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
biorxiv_growth <- preprints %>%
  filter(archive == "bioRxiv", count > 0)

ggplot(biorxiv_growth, aes(date_dec, count)) + 
  geom_line(aes(color = "data"), size = .5) +
  geom_point(aes(fill = "data"), color = "white", shape = 21, size = 2) +
  scale_y_continuous(
    limits = c(0, 1550),
    breaks = c(0, 500, 1000, 1500),
    expand = c(0, 0),
    name = "preprints / month"
  ) + 
  scale_x_continuous(name = NULL) +
  scale_color_manual(
    name = NULL,
    values = "#0072B2",
    breaks = "data",
    labels = "actual counts"
  ) +
  scale_fill_manual(
    name = NULL,
    values = "#0072B2",
    breaks = "data",
    labels = "actual counts"
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.05, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = unit(3, "pt"),
    legend.title = element_blank() 
  )

```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Exponential trends are linear trends on a log scale

.center[
```{r biorxiv-expfit, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
biorxiv_growth <- preprints %>%
  filter(archive == "bioRxiv", count > 0)

expfit.out <- nls(
  count ~ a*exp(b*(date_dec-2014)),
  data = biorxiv_growth,
  start = c(a = 60.004, b = .773)
)
linfit.out <- nls(
  log(count) ~ log(a) + b*(date_dec-2014),
  data = biorxiv_growth,
  start = c(a = 42.576, b = .878)
)
date_seq = seq(min(biorxiv_growth$date_dec), max(biorxiv_growth$date_dec), by = 0.1)
expfit.df <- data.frame(
  date_dec = date_seq,
  count = predict(expfit.out, data.frame(date_dec = date_seq))
)
linfit.df <- data.frame(
  date_dec = date_seq,
  count = exp(predict(linfit.out, data.frame(date_dec = date_seq)))
)

ggplot(biorxiv_growth, aes(date_dec, count)) + 
  geom_line(data = expfit.df, aes(color = "expfit"), size = .5, linetype = 2) +
  geom_point(aes(fill = "expfit"), shape = NA, na.rm = TRUE) + # dummy for legend
  geom_line(aes(color = "data"), size = .5) +
  geom_point(aes(fill = "data"), color = "white", shape = 21, size = 2) +
  scale_y_continuous(
    limits = c(0, 1550),
    breaks = c(0, 500, 1000, 1500),
    expand = c(0, 0),
    name = "preprints / month"
  ) + 
  scale_x_continuous(name = NULL) +
  scale_color_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black"),
    breaks = c("data", "expfit"),
    labels = c("actual counts", "exponential fit"),
    guide = guide_legend(
      override.aes = list(
        color = c("white", "black"),
        shape = c(21, NA),
        size = c(2, 0.5),
        linetype = c(0, 2)
      )
    )
  ) +
  scale_fill_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black"),
    breaks = c("data", "expfit"),
    labels = c("actual counts", "exponential fit")
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.05, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = unit(3, "pt"),
    legend.title = element_blank() 
  )

```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Exponential trends are linear trends on a log scale

.center[
```{r biorxiv-expfit-logscale, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
biorxiv_growth <- preprints %>%
  filter(archive == "bioRxiv", count > 0)

expfit.out <- nls(
  count ~ a*exp(b*(date_dec-2014)),
  data = biorxiv_growth,
  start = c(a = 60.004, b = .773)
)
linfit.out <- nls(
  log(count) ~ log(a) + b*(date_dec-2014),
  data = biorxiv_growth,
  start = c(a = 42.576, b = .878)
)
date_seq = seq(min(biorxiv_growth$date_dec), max(biorxiv_growth$date_dec), by = 0.1)
expfit.df <- data.frame(
  date_dec = date_seq,
  count = predict(expfit.out, data.frame(date_dec = date_seq))
)
linfit.df <- data.frame(
  date_dec = date_seq,
  count = exp(predict(linfit.out, data.frame(date_dec = date_seq)))
)

ggplot(biorxiv_growth, aes(date_dec, count)) + 
  geom_line(data = expfit.df, aes(color = "expfit"), size = .5, linetype = 2) +
  geom_point(aes(fill = "expfit"), shape = NA, na.rm = TRUE) + # dummy for legend
  geom_line(aes(color = "data"), size = .5) +
  geom_point(aes(fill = "data"), color = "white", shape = 21, size = 2) +
  scale_y_log10(
    limits = c(30, 1670),
    breaks = c(10*(3:9), 100*(1:9), 1000*(1:2)),
    labels = c(
      "", "", "50", "", "", "", "", "100", "", "", "", "500",
      "", "", "", "", "1000", ""
    ),
    expand = c(0, 0),
    name = "preprints / month"
  ) + 
  scale_x_continuous(name = NULL) +
  scale_color_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black"),
    breaks = c("data", "expfit"),
    labels = c("actual counts", "exponential fit"),
    guide = guide_legend(
      override.aes = list(
        color = c("white", "black"),
        shape = c(21, NA),
        size = c(2, 0.5),
        linetype = c(0, 2)
      )
    )
  ) +
  scale_fill_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black"),
    breaks = c("data", "expfit"),
    labels = c("actual counts", "exponential fit")
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.05, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = unit(3, "pt"),
    legend.title = element_blank() 
  )

```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Exponential trends are linear trends on a log scale

.center[
```{r biorxiv-logscale, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
ggplot(biorxiv_growth, aes(date_dec, count)) +
  geom_line(data = expfit.df, aes(color = "expfit"), size = .5, linetype = 2) +
  geom_point(aes(fill = "expfit"), shape = NA, na.rm = TRUE) + # dummy for legend
  geom_line(data = linfit.df, aes(color = "linfit"), size = .5) +
  geom_point(aes(fill = "linfit"), shape = NA, na.rm = TRUE) + # dummy for legend
  geom_line(aes(color = "data"), size = .5) +
  geom_point(aes(fill = "data"), color = "white", shape = 21, size = 2) +
  scale_y_log10(
    limits = c(30, 1670),
    breaks = c(10*(3:9), 100*(1:9), 1000*(1:2)),
    labels = c(
      "", "", "50", "", "", "", "", "100", "", "", "", "500",
      "", "", "", "", "1000", ""
    ),
    expand = c(0, 0),
    name = "preprints / month"
  ) + 
  scale_x_continuous(name = NULL) +
  scale_color_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black", linfit = "black"),
    breaks = c("data", "expfit", "linfit"),
    labels = c("actual counts", "exponential fit", "linear fit, log-transformed data"),
    guide = guide_legend(
      override.aes = list(
        color = c("white", "black", "black"),
        shape = c(21, NA, NA),
        size = c(2, 0.5, 0.5),
        linetype = c(0, 2, 1)
      )
    )
  ) +
  scale_fill_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black", linfit = "black"),
    breaks = c("data", "expfit", "linfit"),
    labels = c("actual counts", "exponential fit", "linear fit, log-transformed data")
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.05, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = unit(3, "pt"),
    legend.title = element_blank()
  )

```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Exponential trends are linear trends on a log scale

.center[
```{r biorxiv-logscale-doublefit, echo = FALSE, warning = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
ggplot(biorxiv_growth, aes(date_dec, count)) +
  geom_line(data = expfit.df, aes(color = "expfit"), size = .5, linetype = 2) +
  geom_point(aes(fill = "expfit"), shape = NA, na.rm = TRUE) + # dummy for legend
  geom_line(data = linfit.df, aes(color = "linfit"), size = .5) +
  geom_point(aes(fill = "linfit"), shape = NA, na.rm = TRUE) + # dummy for legend
  geom_line(aes(color = "data"), size = .5) +
  geom_point(aes(fill = "data"), color = "white", shape = 21, size = 2) +
  scale_y_continuous(
    limits = c(0, 1550),
    breaks = c(0, 500, 1000, 1500),
    expand = c(0, 0),
    name = "preprints / month"
  ) + 
  scale_x_continuous(name = NULL) +
  scale_color_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black", linfit = "black"),
    breaks = c("data", "expfit", "linfit"),
    labels = c("actual counts", "exponential fit", "linear fit, log-transformed data"),
    guide = guide_legend(
      override.aes = list(
        color = c("white", "black", "black"),
        shape = c(21, NA, NA),
        size = c(2, 0.5, 0.5),
        linetype = c(0, 2, 1)
      )
    )
  ) +
  scale_fill_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black", linfit = "black"),
    breaks = c("data", "expfit", "linfit"),
    labels = c("actual counts", "exponential fit", "linear fit, log-transformed data")
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.05, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = unit(3, "pt"),
    legend.title = element_blank()
  )

```
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Detrending: Removing the underlying trend

.center[
```{r hpi-no-trendline, echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}

hpi_trends %>%
  filter(state %in% c("California", "Nevada", "West Virginia", "Texas")) %>%
  ggplot(aes(date, hpi)) +
  geom_line(color = "#0072B2", size = 0.75) +
  scale_x_date(name = NULL) +
  scale_y_log10(name = "House Price Index (Dec. 2000 = 100)") +
  facet_wrap(~state, scales = "free_x") +
  theme_minimal_hgrid() +
  theme(
    strip.text = element_text(size = 12),
    strip.background = element_rect(fill = "grey85"),
    axis.line.x = element_line(color = "grey50"),
    axis.ticks.x = element_line(color = "grey50"),
    axis.ticks.y = element_blank(),
    axis.text.y = element_text(margin = margin(0, 0, 0, 0))
  )

```
]

--

.small-font[
Did housing prices in California decline substantially from 1990 to 1998?
]

--

.small-font[
Did housing prices in West Virginia recover by 2017?
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Detrending: Removing the underlying trend

.center[
```{r hpi-trends, echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}

hpi_trends %>%
  filter(state %in% c("California", "Nevada", "West Virginia", "Texas")) %>%
  ggplot(aes(date, hpi)) +
  geom_line(aes(y = hpi_trend), color = "grey50", size = 0.4) +
  geom_line(color = "#0072B2", size = 0.75) +
  scale_x_date(name = NULL) +
  scale_y_log10(name = "House Price Index (Dec. 2000 = 100)") +
  facet_wrap(~state, scales = "free_x") +
  theme_minimal_hgrid() +
  theme(
    strip.text = element_text(size = 12),
    strip.background = element_rect(fill = "grey85"),
    axis.line.x = element_line(color = "grey50"),
    axis.ticks.x = element_line(color = "grey50"),
    axis.ticks.y = element_blank(),
    axis.text.y = element_text(margin = margin(0, 0, 0, 0))
  )

```
]

.small-font[
Did housing prices in California decline substantially from 1990 to 1998?
]

.small-font[
Did housing prices in West Virginia recover by 2017?
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Detrending: Removing the underlying trend

.center[
```{r hpi-detrended, echo = FALSE, warning = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}

hpi_trends %>%
  filter(state %in% c("California", "Nevada", "West Virginia", "Texas")) %>%
  ggplot(aes(date, hpi_detrended)) +
  geom_hline(yintercept = 1, color = "grey50", size = 0.4) +
  geom_line(color = "#0072B2", size = 0.75) +
  scale_x_date(name = NULL) +
  scale_y_log10(
    name = "House Price Index (detrended)",
    breaks = c(0.752, 1, 1.33, 1.77),
    labels = c("0.75", "1.00", "1.33", "1.77")
  ) +
  facet_wrap(~state, scales = "free_x") +
  theme_minimal_hgrid() +
  theme(
    strip.text = element_text(size = 12),
    strip.background = element_rect(fill = "grey85"),
    axis.line.x = element_line(color = "grey50"),
    axis.ticks.x = element_line(color = "grey50"),
    axis.ticks.y = element_blank(),
    axis.text.y = element_text(margin = margin(0, 0, 0, 0))
  )

```
]

.small-font[
Did housing prices in California decline substantially from 1990 to 1998? — yes
]

.small-font[
Did housing prices in West Virginia recover by 2017?  — no
]

???

Figure redrawn after [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

[//]: # "segment ends here"

---
class: middle, center

# Creating trendlines in **ggplot2**


---

## We add trend lines with `geom_smooth()`

.tiny-font.pull-left[
```{r blue-jays-scatter-gg, eval = FALSE}
ggplot(blue_jays) +
  aes(body_mass_g, head_length_mm) + 
  geom_point() + theme_bw(14)
```
]

.xtiny-font.pull-right[
```{r blue-jays-scatter-gg-out, ref.label = "blue-jays-scatter-gg", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

.small-font[
Scatter plot only
]


---

## We add trend lines with `geom_smooth()`

.tiny-font.pull-left[
```{r blue-jays-scatter-gg-smooth, eval = FALSE}
ggplot(blue_jays) +
  aes(body_mass_g, head_length_mm) + 
  geom_point() + theme_bw(14) +
  geom_smooth()
```
]

.xtiny-font.pull-right[
```{r blue-jays-scatter-gg-smooth-out, ref.label = "blue-jays-scatter-gg-smooth", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

.small-font[
Scatter plot with loess smooth
]

---

## We add trend lines with `geom_smooth()`

.tiny-font.pull-left[
```{r blue-jays-scatter-gg-lm, eval = FALSE}
ggplot(blue_jays) +
  aes(body_mass_g, head_length_mm) + 
  geom_point() + theme_bw(14) +
  geom_smooth(
    # smooth using linear model
    method = "lm"
  )
```
]

.xtiny-font.pull-right[
```{r blue-jays-scatter-gg-lm-out, ref.label = "blue-jays-scatter-gg-lm", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

.small-font[
Scatter plot with linear regression
]


---

## We add trend lines with `geom_smooth()`

.tiny-font.pull-left[
```{r blue-jays-scatter-gg-lm-nose, eval = FALSE}
ggplot(blue_jays) +
  aes(body_mass_g, head_length_mm) + 
  geom_point() + theme_bw(14) +
  geom_smooth(
    # smooth using linear model
    method = "lm",
    # suppress confidence band
    se = FALSE
  )
```
]

.xtiny-font.pull-right[
```{r blue-jays-scatter-gg-lm-nose-out, ref.label = "blue-jays-scatter-gg-lm-nose", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

.small-font[
Scatter plot with linear regression, no confidence band
]

---

## We add trend lines with `geom_smooth()`

.tiny-font.pull-left[
```{r blue-jays-scatter-gg-sex, eval = FALSE}
ggplot(blue_jays) +
  aes(
    body_mass_g, head_length_mm,
    color = sex
  ) + 
  geom_point() + theme_bw(14) +
  geom_smooth(
    # smooth using linear model
    method = "lm",
    # suppress confidence band
    se = FALSE
  )
```
]

.xtiny-font.pull-right[
```{r blue-jays-scatter-gg-sex-out, ref.label = "blue-jays-scatter-gg-sex", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

.small-font[
Scatter plot with linear regression by sex
]

---
class: middle, center

# Linear regression can be nonsensical

---

## Example: Fuel-tank capacity versus price in cars

.tiny-font.pull-left[
```{r cars-lm, eval = FALSE}
ggplot(cars93) +
  aes(x = Price, y = Fuel.tank.capacity) +
  geom_point() + theme_bw(14) +
  geom_smooth(method = "lm")
```
]

.xtiny-font.pull-right[
```{r cars-lm-out, ref.label = "cars-lm", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

--

.small-font[
Do more expensive cars have a larger fuel tank?
]

---

## Example: Fuel-tank capacity versus price in cars

.tiny-font.pull-left[
```{r cars-loess, eval = FALSE}
ggplot(cars93) +
  aes(x = Price, y = Fuel.tank.capacity) +
  geom_point() + theme_bw(14) +
  # default: loess smoothing
  geom_smooth(
    se = FALSE
  )
```
]

.xtiny-font.pull-right[
```{r cars-loess-out, ref.label = "cars-loess", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

---

## Example: Fuel-tank capacity versus price in cars

.tiny-font.pull-left[
```{r cars-loess2, eval = FALSE}
ggplot(cars93) +
  aes(x = Price, y = Fuel.tank.capacity) +
  geom_point() + theme_bw(14) +
  # loess smoothing
  geom_smooth(
    se = FALSE,
    method = "loess",
    formula = y ~ x
  )
```
]

.xtiny-font.pull-right[
```{r cars-loess2-out, ref.label = "cars-loess2", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

--

.small-font[
Caution: Exact shape of smoothing line depends on method details
]


---

## Example: Fuel-tank capacity versus price in cars

.tiny-font.pull-left[
```{r cars-loess3, eval = FALSE}
ggplot(cars93) +
  aes(x = Price, y = Fuel.tank.capacity) +
  geom_point() + theme_bw(14) +
  # loess smoothing
  geom_smooth(
    se = FALSE,
    method = "loess",
    formula = y ~ x,
    span = 0.25
  )
```
]

.xtiny-font.pull-right[
```{r cars-loess3-out, ref.label = "cars-loess3", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

.small-font[
Caution: Exact shape of smoothing line depends on method details
]


---

## Example: Fuel-tank capacity versus price in cars

.tiny-font.pull-left[
```{r cars-loess4, eval = FALSE}
ggplot(cars93) +
  aes(x = Price, y = Fuel.tank.capacity) +
  geom_point() + theme_bw(14) +
  # loess smoothing
  geom_smooth(
    se = FALSE,
    method = "loess",
    formula = y ~ x,
    span = 0.75 # default value
  )
```
]

.xtiny-font.pull-right[
```{r cars-loess4-out, ref.label = "cars-loess4", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

.small-font[
Caution: Exact shape of smoothing line depends on method details
]


---

## Example: Fuel-tank capacity versus price in cars

.tiny-font.pull-left[
```{r cars-loess5, eval = FALSE}
ggplot(cars93) +
  aes(x = Price, y = Fuel.tank.capacity) +
  geom_point() + theme_bw(14) +
  # loess smoothing
  geom_smooth(
    se = FALSE,
    method = "loess",
    formula = y ~ x,
    span = 1.0
  )
```
]

.xtiny-font.pull-right[
```{r cars-loess5-out, ref.label = "cars-loess5", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

.small-font[
Caution: Exact shape of smoothing line depends on method details
]

---

## Example: Fuel-tank capacity versus price in cars

.tiny-font.pull-left[
```{r cars-loess6, eval = FALSE}
ggplot(cars93) +
  aes(x = Price, y = Fuel.tank.capacity) +
  geom_point() + theme_bw(14) +
  # loess smoothing
  geom_smooth(
    se = FALSE,
    method = "loess",
    formula = y ~ x,
    span = 1.5
  )
```
]

.xtiny-font.pull-right[
```{r cars-loess6-out, ref.label = "cars-loess6", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

.small-font[
Caution: Exact shape of smoothing line depends on method details
]

---

## Example: Fuel-tank capacity versus price in cars

.tiny-font.pull-left[
```{r cars-gam, eval = FALSE}
ggplot(cars93) +
  aes(x = Price, y = Fuel.tank.capacity) +
  geom_point() + theme_bw(14) +
  # cubic spline, 5 knots
  geom_smooth(
    se = FALSE,
    method = "gam",
    formula = y ~ s(x, k = 5, bs = 'cr')
  )
```
]

.xtiny-font.pull-right[
```{r cars-gam-out, ref.label = "cars-gam", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

.small-font[
Caution: Exact shape of smoothing line depends on method details
]

---

## Example: Fuel-tank capacity versus price in cars

.tiny-font.pull-left[
```{r cars-gam2, eval = FALSE}
ggplot(cars93) +
  aes(x = Price, y = Fuel.tank.capacity) +
  geom_point() + theme_bw(14) +
  # thin-plate spline, 3 knots
  geom_smooth(
    se = FALSE,
    method = "gam",
    formula = y ~ s(x, k = 3)
  )
```
]

.xtiny-font.pull-right[
```{r cars-gam2-out, ref.label = "cars-gam2", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

.small-font[
Caution: Exact shape of smoothing line depends on method details
]


---

## Example: Fuel-tank capacity versus price in cars

.tiny-font.pull-left[
```{r cars-gam3, eval = FALSE}
ggplot(cars93) +
  aes(x = Price, y = Fuel.tank.capacity) +
  geom_point() + theme_bw(14) +
  # Gaussian process spline, 6 knots
  geom_smooth(
    se = FALSE,
    method = "gam",
    formula = y ~ s(x, k = 6, bs = 'gp')
  )
```
]

.xtiny-font.pull-right[
```{r cars-gam3-out, ref.label = "cars-gam3", echo = FALSE, fig.width = 6, fig.asp = 0.618, dev = "svg"}
```
]

.small-font[
Caution: Exact shape of smoothing line depends on method details
]

--

.small-font[
Smoothing lines are particularly unreliable near their endpoints
]

[//]: # "segment ends here"

---

## Further reading

- Fundamentals of Data Visualization: [Chapter 14: Visualizing trends](https://clauswilke.com/dataviz/visualizing-trends.html)
- Data Visualization—A Practical Introduction: [Chapter 6: Work with models](https://socviz.co/modeling.html)
- **ggplot2** reference documentation: [`geom_smooth()`](https://ggplot2.tidyverse.org/reference/geom_smooth.html)
- **mgcv** reference documentation (for gam smoothing): [pdf document](https://cran.r-project.org/web/packages/mgcv/mgcv.pdf)





