---
title: "Visualizing trends"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(here)
library(ggforce)
library(cowplot)
# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width

blue_jays <- read_csv(here("datasets", "blue_jays.csv"))
preprints <- read_csv(here("datasets", "preprints.csv"))
hpi_trends <- read_csv(here("datasets", "hpi_trends.csv"))
```


## Title

.center[
```{r blue-jays-scatter-female, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  filter(sex == "F") %>%
  ggplot(aes(body_mass_g, head_length_mm, color = sex, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  scale_fill_manual(
    values = c(F = "#D55E00C0", M = "#0072B2C0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 3, linetype = 0)
    )
  ) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL
  ) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Title

.center[
```{r blue-jays-scatter-line-female, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  filter(sex == "F") %>%
  ggplot(aes(body_mass_g, head_length_mm, color = sex, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  geom_smooth(method = "lm", size = 0.75, se = FALSE, fullrange = TRUE) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  scale_fill_manual(
    values = c(F = "#D55E00C0", M = "#0072B2C0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 3, linetype = 0)
    )
  ) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL
  ) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Title

.center[
```{r blue-jays-scatter-male, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  filter(sex == "M") %>%
  ggplot(aes(body_mass_g, head_length_mm, color = sex, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  scale_fill_manual(
    values = c(F = "#D55E00C0", M = "#0072B2C0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 3, linetype = 0)
    )
  ) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL
  ) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Title

.center[
```{r blue-jays-scatter-line-male, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  filter(sex == "M") %>%
  ggplot(aes(body_mass_g, head_length_mm, color = sex, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  geom_smooth(method = "lm", size = 0.75, se = FALSE, fullrange = TRUE) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  scale_fill_manual(
    values = c(F = "#D55E00C0", M = "#0072B2C0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 3, linetype = 0)
    )
  ) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL
  ) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Title

.center[
```{r blue-jays-scatter-line, echo = FALSE, message = FALSE, fig.width = 6, fig.asp = 3.2/4, dev = "svg"}
blue_jays %>%
  ggplot(aes(body_mass_g, head_length_mm, color = sex, fill = sex)) + 
  geom_point(pch = 21, color = "white", size = 2.5) +
  geom_smooth(method = "lm", size = 0.75, se = FALSE, fullrange = TRUE) +
  scale_x_continuous(name = "body mass (g)", limits = c(58.7, 81.6)) +
  scale_y_continuous(name = "head length (mm)", limits = c(51.5, 60.8)) +
  scale_fill_manual(
    values = c(F = "#D55E00C0", M = "#0072B2C0"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL,
    guide = guide_legend(
      direction = "horizontal",
      override.aes = list(size = 3, linetype = 0)
    )
  ) +
  scale_color_manual(
    values = c(F = "#D55E00", M = "#0072B2"),
    breaks = c("F", "M"),
    labels = c("female birds   ", "male birds"),
    name = NULL
  ) +
  theme_minimal_grid() +
  theme(
    legend.position = "top",
    legend.justification = "right",
    legend.box.spacing = unit(3.5, "pt"), # distance between legend and plot
    legend.text = element_text(vjust = 0.6),
    legend.spacing.x = unit(2, "pt"),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.width = unit(10, "pt")
  )
```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Title

.center[
```{r biorxiv-nofit, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
biorxiv_growth <- preprints %>%
  filter(archive == "bioRxiv", count > 0)

ggplot(biorxiv_growth, aes(date_dec, count)) + 
  geom_line(aes(color = "data"), size = .5) +
  geom_point(aes(fill = "data"), color = "white", shape = 21, size = 2) +
  scale_y_continuous(
    limits = c(0, 1550),
    breaks = c(0, 500, 1000, 1500),
    expand = c(0, 0),
    name = "preprints / month"
  ) + 
  scale_x_continuous(name = NULL) +
  scale_color_manual(
    name = NULL,
    values = "#0072B2",
    breaks = "data",
    labels = "actual counts"
  ) +
  scale_fill_manual(
    name = NULL,
    values = "#0072B2",
    breaks = "data",
    labels = "actual counts"
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.05, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = unit(3, "pt"),
    legend.title = element_blank() 
  )

```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Title

.center[
```{r biorxiv-expfit, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
biorxiv_growth <- preprints %>%
  filter(archive == "bioRxiv", count > 0)

expfit.out <- nls(
  count ~ a*exp(b*(date_dec-2014)),
  data = biorxiv_growth,
  start = c(a = 60.004, b = .773)
)
linfit.out <- nls(
  log(count) ~ log(a) + b*(date_dec-2014),
  data = biorxiv_growth,
  start = c(a = 42.576, b = .878)
)
date_seq = seq(min(biorxiv_growth$date_dec), max(biorxiv_growth$date_dec), by = 0.1)
expfit.df <- data.frame(
  date_dec = date_seq,
  count = predict(expfit.out, data.frame(date_dec = date_seq))
)
linfit.df <- data.frame(
  date_dec = date_seq,
  count = exp(predict(linfit.out, data.frame(date_dec = date_seq)))
)

ggplot(biorxiv_growth, aes(date_dec, count)) + 
  geom_line(data = expfit.df, aes(color = "expfit"), size = .5, linetype = 2) +
  geom_point(aes(fill = "expfit"), shape = NA, na.rm = TRUE) + # dummy for legend
  geom_line(aes(color = "data"), size = .5) +
  geom_point(aes(fill = "data"), color = "white", shape = 21, size = 2) +
  scale_y_continuous(
    limits = c(0, 1550),
    breaks = c(0, 500, 1000, 1500),
    expand = c(0, 0),
    name = "preprints / month"
  ) + 
  scale_x_continuous(name = NULL) +
  scale_color_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black"),
    breaks = c("data", "expfit"),
    labels = c("actual counts", "exponential fit"),
    guide = guide_legend(
      override.aes = list(
        color = c("white", "black"),
        shape = c(21, NA),
        size = c(2, 0.5),
        linetype = c(0, 2)
      )
    )
  ) +
  scale_fill_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black"),
    breaks = c("data", "expfit"),
    labels = c("actual counts", "exponential fit")
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.05, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = unit(3, "pt"),
    legend.title = element_blank() 
  )

```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Title

.center[
```{r biorxiv-expfit-logscale, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
biorxiv_growth <- preprints %>%
  filter(archive == "bioRxiv", count > 0)

expfit.out <- nls(
  count ~ a*exp(b*(date_dec-2014)),
  data = biorxiv_growth,
  start = c(a = 60.004, b = .773)
)
linfit.out <- nls(
  log(count) ~ log(a) + b*(date_dec-2014),
  data = biorxiv_growth,
  start = c(a = 42.576, b = .878)
)
date_seq = seq(min(biorxiv_growth$date_dec), max(biorxiv_growth$date_dec), by = 0.1)
expfit.df <- data.frame(
  date_dec = date_seq,
  count = predict(expfit.out, data.frame(date_dec = date_seq))
)
linfit.df <- data.frame(
  date_dec = date_seq,
  count = exp(predict(linfit.out, data.frame(date_dec = date_seq)))
)

ggplot(biorxiv_growth, aes(date_dec, count)) + 
  geom_line(data = expfit.df, aes(color = "expfit"), size = .5, linetype = 2) +
  geom_point(aes(fill = "expfit"), shape = NA, na.rm = TRUE) + # dummy for legend
  geom_line(aes(color = "data"), size = .5) +
  geom_point(aes(fill = "data"), color = "white", shape = 21, size = 2) +
  scale_y_log10(
    limits = c(30, 1670),
    breaks = c(10*(3:9), 100*(1:9), 1000*(1:2)),
    labels = c(
      "", "", "50", "", "", "", "", "100", "", "", "", "500",
      "", "", "", "", "1000", ""
    ),
    expand = c(0, 0),
    name = "preprints / month"
  ) + 
  scale_x_continuous(name = NULL) +
  scale_color_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black"),
    breaks = c("data", "expfit"),
    labels = c("actual counts", "exponential fit"),
    guide = guide_legend(
      override.aes = list(
        color = c("white", "black"),
        shape = c(21, NA),
        size = c(2, 0.5),
        linetype = c(0, 2)
      )
    )
  ) +
  scale_fill_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black"),
    breaks = c("data", "expfit"),
    labels = c("actual counts", "exponential fit")
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.05, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = unit(3, "pt"),
    legend.title = element_blank() 
  )

```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Title

.center[
```{r biorxiv-logscale, echo = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
ggplot(biorxiv_growth, aes(date_dec, count)) +
  geom_line(data = expfit.df, aes(color = "expfit"), size = .5, linetype = 2) +
  geom_point(aes(fill = "expfit"), shape = NA, na.rm = TRUE) + # dummy for legend
  geom_line(data = linfit.df, aes(color = "linfit"), size = .5) +
  geom_point(aes(fill = "linfit"), shape = NA, na.rm = TRUE) + # dummy for legend
  geom_line(aes(color = "data"), size = .5) +
  geom_point(aes(fill = "data"), color = "white", shape = 21, size = 2) +
  scale_y_log10(
    limits = c(30, 1670),
    breaks = c(10*(3:9), 100*(1:9), 1000*(1:2)),
    labels = c(
      "", "", "50", "", "", "", "", "100", "", "", "", "500",
      "", "", "", "", "1000", ""
    ),
    expand = c(0, 0),
    name = "preprints / month"
  ) + 
  scale_x_continuous(name = NULL) +
  scale_color_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black", linfit = "black"),
    breaks = c("data", "expfit", "linfit"),
    labels = c("actual counts", "exponential fit", "linear fit, log-transformed data"),
    guide = guide_legend(
      override.aes = list(
        color = c("white", "black", "black"),
        shape = c(21, NA, NA),
        size = c(2, 0.5, 0.5),
        linetype = c(0, 2, 1)
      )
    )
  ) +
  scale_fill_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black", linfit = "black"),
    breaks = c("data", "expfit", "linfit"),
    labels = c("actual counts", "exponential fit", "linear fit, log-transformed data")
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.05, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = unit(3, "pt"),
    legend.title = element_blank()
  )

```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Title

.center[
```{r biorxiv-logscale-doublefit, echo = FALSE, warning = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}
ggplot(biorxiv_growth, aes(date_dec, count)) +
  geom_line(data = expfit.df, aes(color = "expfit"), size = .5, linetype = 2) +
  geom_point(aes(fill = "expfit"), shape = NA, na.rm = TRUE) + # dummy for legend
  geom_line(data = linfit.df, aes(color = "linfit"), size = .5) +
  geom_point(aes(fill = "linfit"), shape = NA, na.rm = TRUE) + # dummy for legend
  geom_line(aes(color = "data"), size = .5) +
  geom_point(aes(fill = "data"), color = "white", shape = 21, size = 2) +
  scale_y_continuous(
    limits = c(0, 1550),
    breaks = c(0, 500, 1000, 1500),
    expand = c(0, 0),
    name = "preprints / month"
  ) + 
  scale_x_continuous(name = NULL) +
  scale_color_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black", linfit = "black"),
    breaks = c("data", "expfit", "linfit"),
    labels = c("actual counts", "exponential fit", "linear fit, log-transformed data"),
    guide = guide_legend(
      override.aes = list(
        color = c("white", "black", "black"),
        shape = c(21, NA, NA),
        size = c(2, 0.5, 0.5),
        linetype = c(0, 2, 1)
      )
    )
  ) +
  scale_fill_manual(
    name = NULL,
    values = c(data = "#0072B2", expfit = "black", linfit = "black"),
    breaks = c("data", "expfit", "linfit"),
    labels = c("actual counts", "exponential fit", "linear fit, log-transformed data")
  ) +
  theme_half_open() +
  theme(
    legend.position = c(.05, 1),
    legend.justification = c(0, 1),
    legend.spacing.x = unit(3, "pt"),
    legend.title = element_blank()
  )

```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Title

.center[
```{r hpi-no-trendline, echo = FALSE, warning = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}

hpi_trends %>%
  filter(state %in% c("California", "Nevada", "West Virginia", "Texas")) %>%
  ggplot(aes(date, hpi)) +
  geom_line(color = "#0072B2", size = 0.75) +
  scale_x_date(name = NULL) +
  scale_y_log10(name = "House Price Index (Dec. 2000 = 100)") +
  facet_wrap(~state, scales = "free_x") +
  theme_minimal_hgrid() +
  theme(
    strip.text = element_text(size = 12),
    strip.background = element_rect(fill = "grey85"),
    axis.line.x = element_line(color = "grey50"),
    axis.ticks.x = element_line(color = "grey50"),
    axis.ticks.y = element_blank(),
    axis.text.y = element_text(margin = margin(0, 0, 0, 0))
  )

```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Title

.center[
```{r hpi-trends, echo = FALSE, warning = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}

hpi_trends %>%
  filter(state %in% c("California", "Nevada", "West Virginia", "Texas")) %>%
  ggplot(aes(date, hpi)) +
  geom_line(aes(y = hpi_trend), color = "grey50", size = 0.4) +
  geom_line(color = "#0072B2", size = 0.75) +
  scale_x_date(name = NULL) +
  scale_y_log10(name = "House Price Index (Dec. 2000 = 100)") +
  facet_wrap(~state, scales = "free_x") +
  theme_minimal_hgrid() +
  theme(
    strip.text = element_text(size = 12),
    strip.background = element_rect(fill = "grey85"),
    axis.line.x = element_line(color = "grey50"),
    axis.ticks.x = element_line(color = "grey50"),
    axis.ticks.y = element_blank(),
    axis.text.y = element_text(margin = margin(0, 0, 0, 0))
  )

```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)


---

## Title

.center[
```{r hpi-detrended, echo = FALSE, warning = FALSE, fig.width = 7, fig.asp = 0.618, dev = "svg"}

hpi_trends %>%
  filter(state %in% c("California", "Nevada", "West Virginia", "Texas")) %>%
  ggplot(aes(date, hpi_detrended)) +
  geom_hline(yintercept = 1, color = "grey50", size = 0.4) +
  geom_line(color = "#0072B2", size = 0.75) +
  scale_x_date(name = NULL) +
  scale_y_log10(
    name = "House Price Index (detrended)",
    breaks = c(0.752, 1, 1.33, 1.77),
    labels = c("0.75", "1.00", "1.33", "1.77")
  ) +
  facet_wrap(~state, scales = "free_x") +
  theme_minimal_hgrid() +
  theme(
    strip.text = element_text(size = 12),
    strip.background = element_rect(fill = "grey85"),
    axis.line.x = element_line(color = "grey50"),
    axis.ticks.x = element_line(color = "grey50"),
    axis.ticks.y = element_blank(),
    axis.text.y = element_text(margin = margin(0, 0, 0, 0))
  )

```
]

???

Figure redrawn from [Claus O. Wilke. Fundamentals of Data Visualization. O'Reilly, 2019.](https://clauswilke.com/dataviz)

---

## Further reading

- Fundamentals of Data Visualization: [Chapter 14: Visualizing trends](https://clauswilke.com/dataviz/visualizing-trends.html)




