---
title: "Working with models"
author: "Claus O. Wilke"
date: "last updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: [default, "Wilke-slides-theme.css"]
    lib_dir: libs
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      slideNumberFormat: ''
      titleSlideClass: [center, middle]
---

```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "")

library(tidyverse)
library(here)
library(palmerpenguins)
library(cowplot)
# ggplot2 settings so plots scale well for slide size 
theme_set(theme_gray(16)) # 16 for full width, 18 for half width 
update_geom_defaults("point", list(size = 2)) # 2 for full width, 2.5 for half width
```

## How do we obtain information about model fits?

.tiny-font[
```{r penguins-fits, message = FALSE, warning = FALSE, fig.width = 10, fig.asp = 0.4, dev = "svg"}
penguins %>%
  ggplot(aes(body_mass_g, flipper_length_mm)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(vars(species))
```
]

---

## We can fit a linear model with `lm()`

--

.tiny-font[
```{r}
penguins_adelie <- filter(penguins, species == "Adelie")

lm_out <- lm(flipper_length_mm ~ body_mass_g, data = penguins_adelie)
summary(lm_out)
```
]


---

## Use `map()` to fit models to groups of data

.small-font[
```{r}
penguins %>%
  nest(data = -species) # nest all data except species column
```
]

---

## Use `map()` to fit models to groups of data

.small-font[
```{r}
penguins %>%
  nest(data = -species) %>%
  mutate(
    # apply linear model to each nested data frame
    fit = map(data, ~lm(flipper_length_mm ~ body_mass_g, data = .x))
  )
```
]

---

## Use `map()` to fit models to groups of data

.small-font[
```{r}
lm_data <- penguins %>%
  nest(data = -species) %>%
  mutate(
    # apply linear model to each nested data frame
    fit = map(data, ~lm(flipper_length_mm ~ body_mass_g, data = .x))
  )

lm_data$fit[[1]]  # first model fit, for Adelie species
```
]


---

.tiny-font[
```{r}
summary(lm_data$fit[[1]]) # summarize the first model, which is for Adelie
```
]

---

.tiny-font[
```{r}
summary(lm_out)
```
]


---

.tiny-font[
```{r}
summary(lm_data$fit[[1]]) # summarize the first model, which is for Adelie
```
]

---

.tiny-font[
```{r}
summary(lm_data$fit[[2]]) # second model, Chinstrap
```
]

---

.tiny-font[
```{r}
summary(lm_data$fit[[3]]) # third model, Gentoo
```
]

---
class: center middle

## How do we get this information into a data table?

---

## The **broom** package cleans model output: `glance()`

.small-font[
`glance()` provides model-wide summary estimates in tidy format
]

--

.small-font[
```{r}
library(broom)

glance(lm_out)
```
]

---

## The **broom** package cleans model output: `tidy()`

.small-font[
`tidy()` provides information about regression coefficients in tidy format
]

--

.small-font[
```{r}
library(broom)

tidy(lm_out)
```
]


---

## Apply these functions to multiple models with `map()`

--

.small-font[
Reminder: This is the nested table with fitted models
```{r}
lm_data
```
]


---

## Apply these functions to multiple models with `map()`

.small-font[
```{r}
lm_data %>%
  mutate(
    glance_out = map(fit, glance)
  )
```
]

---

## Apply these functions to multiple models with `map()`

.small-font[
```{r}
lm_data %>%
  mutate(
    glance_out = map(fit, ~glance(.x)) # same as just `glance`
  )
```
]

---

## Apply these functions to multiple models with `map()`

.small-font[
```{r}
lm_data %>%
  mutate(
    glance_out = map(fit, glance)
  )
```
]

---

## And unnest

.small-font[
```{r}
lm_data %>%
  mutate(
    glance_out = map(fit, glance)
  ) %>%
  select(species, glance_out)
```
]

---

## And unnest

.small-font[
```{r}
lm_data %>%
  mutate(
    glance_out = map(fit, glance)
  ) %>%
  select(species, glance_out) %>%
  unnest(cols = glance_out)
```
]


---

## All in one pipeline

.tiny-font[
```{r}
lm_summary <- penguins %>%
  nest(data = -species) %>%
  mutate(
    fit = map(data, ~lm(flipper_length_mm ~ body_mass_g, data = .x)),
    glance_out = map(fit, glance)
  ) %>%
  select(species, glance_out) %>%
  unnest(cols = glance_out)

lm_summary
```
]


---

## Make label data

.tiny-font[
```{r message = FALSE}
library(glue)  # for easy text formatting

label_data <- lm_summary %>%
  mutate(
    rsqr = signif(r.squared, 2),  # round to 2 significant digits
    pval = signif(p.value, 2),
    label = glue("R^2 = {rsqr}, P = {pval}"),
    body_mass_g = 6400, flipper_length_mm = 175 # label position in plot
  ) %>%
  select(species, label, body_mass_g, flipper_length_mm)

label_data
```
]
---

## And plot

.tiny-font[
```{r penguins-fit-w-labels, eval = FALSE}
ggplot(penguins, aes(body_mass_g, flipper_length_mm)) + geom_point() +
  geom_text(
    data = label_data, aes(label = label),
    size = 10/.pt, hjust = 1  # 10pt, right-justified
  ) +
  geom_smooth(method = "lm", se = FALSE) + facet_wrap(vars(species))
```
]

.center[
```{r penguins-fit-w-labels-out, ref.label = "penguins-fit-w-labels", echo = FALSE, message = FALSE, warning = FALSE, fig.width = 9, fig.asp = 0.4, dev = "svg"}
```
]

[//]: # "segment ends here"


---

## Further reading

- Data Visualization—A Practical Introduction: [Chapter 6.5: Tidy model objects with broom](https://socviz.co/modeling.html#tidy-model-objects-with-broom)
- **broom** reference documentation: https://broom.tidymodels.org/
- Article on using **broom** with **dplyr**: [**broom** and **dplyr**](https://broom.tidymodels.org/articles/broom_and_dplyr.html)



