---
#############################################################
#                                                           #
# Click on "Run Document" in RStudio to run this worksheet. #
#                                                           #
#############################################################
title: "Data wrangling 2"
author: "Claus O. Wilke"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(palmerpenguins)
knitr::opts_chunk$set(echo = FALSE, comment = "")
```

## Introduction

In this worksheet, we will continue with basic data manipulations, now moving on to counting, pipelining, and creating new data columns. 

We will be using two R packages, **tidyverse** for the data manipulation functions `%>%`, `filter()`, `select()`, `arrange()`, `count()`, and `mutate()`, and **palmerpenguins** for the `penguins` dataset.


```{r library-calls, echo = TRUE, eval = FALSE}
# load required libraries
library(tidyverse)
library(palmerpenguins)
```

## Counting

We frequently want to count how many times a particular value or combination of values occurs in a dataset. We do this using the `count()` function. For example, the following code counts how many penguins of the different species there are in the `penguins` dataset.

```{r count-demo, echo=TRUE}
penguins %>%
  count(species)
```

Now try this yourself. Count how many male and female penguins there are.

```{r count-simple, exercise=TRUE}
penguins %>%
  count(___)
```

```{r count-simple-solution}
penguins %>%
  count(sex)
```

We can also perform more fine-grained counts, by providing the `count()` function with more than one column name. See if you can count how many male and female penguins the dataset contains for each species. 

```{r count-complex, exercise=TRUE}
penguins %>%
  count(___, ___)
```

```{r count-complex-solution}
penguins %>%
  count(species, sex)
```

Now count how many penguins of each species the dataset contains for each island. 

```{r count-complex2, exercise=TRUE}
penguins %>%
  ___
```

```{r count-complex2-hint}
penguins %>%
  count(___, ___)
```


```{r count-complex2-solution}
penguins %>%
  count(island, species)
```


## Chaining analysis steps into pipelines

We can chain multiple analysis steps into a pipeline by continuing to add "and then" statements. For example, `data %>% count(...) %>% arrange(...)` would first count and then sort the data.

Try this out by counting the number of penguins of each species and then sorting by the number of penguins.

```{r analysis-chain, exercise=TRUE}
penguins %>%
  count(___) %>%
  arrange(___)
```

```{r analysis-chain-hint}
penguins %>%
  count(species) %>%
  arrange(___)
```

```{r analysis-chain-solution}
penguins %>%
  count(species) %>%
  arrange(n)
```

You may remember from the previous class that the default sorting is from the smallest to the largest value. Repeat the previous exercise but now arrange the penguin species from the most frequent to the least frequent:

```{r analysis-chain2, exercise=TRUE}
penguins %>%
  count(species) %>%
  arrange(___)
```

```{r analysis-chain2-hint}
penguins %>%
  count(species) %>%
  arrange(desc(___))
```

```{r analysis-chain2-solution}
penguins %>%
  count(species) %>%
  arrange(desc(n))
```


## Creating new data columns

The function `mutate()` allows you to add new columns to a data table. For example, `data %>% mutate(xy = x + y)` would create a new column `xy` that is the sum of the columns `x` and `y`:

```{r simple-mutate-example, echo=TRUE, eval=FALSE}
data <- tibble(x = 1:3, y = c(10, 20, 30))
data
```

```{r simple-mutate-example-code, echo=FALSE, eval=TRUE}
data <- tibble(x = 1:3, y = c(10, 20, 30))
print(data)
```

```{r simple-mutate-example2, echo=TRUE, eval=FALSE}
data %>%
  mutate(
    xy = x + y
  )
```

```{r simple-mutate-example2-code, echo=FALSE, eval=TRUE}
data %>%
  mutate(
    xy = x + y
  ) %>%
  print()
```


Note that the part to the left of the equals sign (here, `xy`) is the name of the new column, and the part to the right of the equals sign (here, `x + y`) is an R expression that evaluates to the values in the new column.

Now apply this concept to the `penguins` dataset. Add a new column `bill_ratio` that is the ratio of bill length and bill depth:

```{r mutate-bill-ratio, exercise=TRUE}
penguins %>%
  select(species, bill_length_mm, bill_depth_mm) %>%
  mutate(
    ___ = ___
  )
```

```{r mutate-bill-ratio-hint}
penguins %>%
  select(species, bill_length_mm, bill_depth_mm) %>%
  mutate(
    bill_ratio = ___
  )
```

```{r mutate-bill-ratio-solution}
penguins %>%
  select(species, bill_length_mm, bill_depth_mm) %>%
  mutate(
    bill_ratio = bill_length_mm / bill_depth_mm
  )
```

## Counting with custom conditions

It is quite common that we want to count items that meet a specific condition. For example, let's say we want to count how many penguins of different species have flippers longer than 200mm. To do this efficiently, we first create a new column that indicates whether the condition is met or not, and we then use count with that indicator column.

The easiest way to create indicator columns is via the function `ifelse()`, which takes three arguments: a condition, a result if the condition is met, and a result if the condition is not met. The following example shows how to create an indicator column showing whether a variable is positive or negative:

```{r indicator-example1, echo=TRUE, eval=FALSE}
data <- tibble(x = c(-0.5, 2.3, 50, -1.4))
data
```

```{r indicator-example1-code, echo=FALSE, eval=TRUE}
data <- tibble(x = c(-0.5, 2.3, 50, -1.4))
print(data)
```

```{r indicator-example2, echo=TRUE, eval=FALSE}
data %>%
  mutate(
    sign_of_x = ifelse(x >= 0, "positive", "negative")
  )
```

```{r indicator-example2-code, echo=FALSE, eval=TRUE}
data %>%
  mutate(
    sign_of_x = ifelse(x >= 0, "positive", "negative")
  ) %>%
  print()
```

Now try this yourself. Count how many penguins of different species have flippers longer than 200mm. Then sort your results from most frequent to least frequent.

```{r flipper-length-count, exercise = TRUE}
penguins %>%
  mutate(
    ___
  ) %>%
  count(___) %>%
  arrange(___)
```

```{r flipper-length-count-hint-1}
penguins %>%
  mutate(
    flipper_length = ifelse(___, "greater than 200mm", "less than 200mm")
  ) %>%
  count(___, ___) %>%
  arrange(___)
```

```{r flipper-length-count-hint-2}
penguins %>%
  mutate(
    flipper_length = ifelse(___, "greater than 200mm", "less than 200mm")
  ) %>%
  count(___, flipper_length) %>%
  arrange(___(n))
```

```{r flipper-length-count-solution}
penguins %>%
  mutate(
    flipper_length = ifelse(flipper_length_mm > 200, "greater than 200mm", "less than 200mm")
  ) %>%
  count(species, flipper_length) %>%
  arrange(desc(n))
```