---
#############################################################
#                                                           #
#   Click on "Knit" in RStudio to render this worksheet.    #
#                                                           #
#############################################################
title: "Intro to R Markdown"
author: "Claus O. Wilke"
output: html_document
---

Much of the work in this class will be done via R Markdown documents. R Markdown documents are documents that combine text, R code, and R output, including figures. They are a great way to produce self-contained and documented statistical analyses.

In this first worksheet, you will learn how to do some basic markdown editing. After you have made a change to the document, press "Knit HTML" in R Studio and see what kind of a result you get.

Edit only below this line.

---------------------------------------------

## 1. Basic Markdown editing

Try out basic R Markdown features, as described here: https://rmarkdown.rstudio.com/authoring_basics.html. Write some text that is bold, and some that is in italics. Make a numbered list and a bulleted list. Make a nested list. Try the block-quote feature.

This text is **bold.**

This text is *in italics.*

A numbered list:

1. Item 1
2. Item 2
3. Item 3

A bulleted list:

- Item 1
- Item 2
- Item 3

A nested list:

1. Item 1
    - Item 1.1. Note that 4 spaces are required for the nesting to work properly.
    - Item 1.2
2. Item 2


Block quote:

> "If we knew what it was we were doing, it would not be called research, would it?" --- Albert Einstein


## 2. Embedding R code

R code embedded in R chunks will be executed and the output will be shown.
```{r}
# R code goes here
x <- rnorm(100)  # random sample from normal distribution
dens <- density(x)    # calculate density
dens

plot(dens)            # plot density
```

Now you try it. For example, take the built-in data set `cars`, which lists speed and stopping distance for cars from the 1920. Plot speed vs. distance, and/or perform a correlation analysis. Then write a few sentences describing what you see.

Plot of speed vs. distance:
```{r}
plot(cars$dist, cars$speed)
```

Correlation analysis:
```{r}
cor.test(cars$dist, cars$speed)
```
There is a strong positive correlations. Cars with higher speed have longer stopping distances.


