---
title: "Homework 2"
---

```{r global_options, include=FALSE}
library(knitr)
library(tidyverse)
opts_chunk$set(fig.align="center", fig.height=4, fig.width=5.5)

# data prep:
txhouse <- txhousing %>%
  filter(city %in% c('Austin', 'Houston', 'San Antonio', 'Dallas')) %>%
  filter(year %in% c('2000', '2005', '2010', '2015')) %>%
  group_by(city, year) %>%
  summarize(total_sales = sum(sales))
  
```

*Enter your name and EID here*

**This homework is due on Jan. 31, 2022 at 11:00am. Please submit as a pdf file on Canvas.**


**Problem 1: (3 pts)** We will work with the dataset `txhouse` that has been derived from the `txhousing` dataset provided by **ggplot2**. See here for details of the original dataset: https://ggplot2.tidyverse.org/reference/txhousing.html. `txhouse` contains three columns: `city` (containing four Texas cities), `year` (containing four years between 2000 and 2015) and `total_sales` indicating the total number of sales for the specified year and city.

```{r}
txhouse
```

Use ggplot to make a bar plot of the total housing sales (column `total_sales`) for each `city` and show one panel per `year`. Hint: Use `facet_wrap()`. See slides from Class 2.


```{r}
# your code goes here
```

**Problem 2: (3 pts)** Use ggplot to make a bar plot of the total housing sales (column `total_sales`) for each `year`, color the bar borders "gray34", and fill the bars by `city`. 

```{r}
# your code goes here
```

**Problem 3: (4 pts)** Modify the plot from Problem 2 by placing `city` bars side-by-side, rather than stacked. Next, reorder the bars for each `year` by `total_sales` in descending order. See slides from Class 4.

```{r}
# your code goes here
```




