```{r global_options, include=FALSE}
library(knitr)
library(ggplot2)
opts_chunk$set(fig.align="center", fig.height=4, fig.width=5.5)
```

## Homework 2

*Enter your name and EID here*

**This homework is due on Feb. 1, 2021 at 11:00pm. Please submit as a pdf file on Canvas.**

In this homework you will be working with the `iris` dataset built into R. This data set contains measurements of flowers (sepal length, sepal width, petal length, petal width) for three different *Iris* species (*I. setosa*, *I. versicolor*, *I. virginica*).

```{r}
head(iris)
```


**Problem 1: (6 pts)** Use ggplot to make a histogram of the `Sepal.Length` column. Manually choose appropriate values for `binwidth` and `center`. Explain your choice of values in 2-3 sentences.

```{r}
# Your code goes here.
```

*Your explanation goes here.*

**Problem 2: (4 pts)** Modify the plot from Problem 1 to show one panel per species. Hint: Use `facet_wrap()`. See Slide 14 from Class 2.


```{r}
# Your code goes here.
```
