---
title: "Homework 5"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r global_options, include=FALSE}
library(knitr)
library(tidyverse)
library(colorspace)
opts_chunk$set(fig.align="center", fig.height=4, fig.width=5.5)

# data prep:
ufo_sightings <- 
  read_csv("https://wilkelab.org/classes/SDS348/data_sets/ufo_sightings_clean.csv") %>%
  separate(datetime, into = c("month", "day", "year"), sep = "/") %>%
  separate(year, into = c("year", "time"), sep = " ") %>%
  separate(date_posted, into = c("month_posted", "day_posted", "year_posted"), sep = "/") %>%
  select(-time, -month_posted, -day_posted) %>%
  mutate(
    year = as.numeric(year),
    state = toupper(state)
  ) %>%
  filter(!is.na(country))
```

*Enter your name and EID here*

**This homework is due on Feb. 28, 2022 at 11:00am. Please submit as a pdf file on Canvas.**

**Problem 1:  (4 pts)** We will work with the `ufo_sightings` dataset. 

Since 1990, what are the top 5 cities that have reported the most UFO sightings? Create a new dataframe to answer the question. No plots are necessary.


```{r}
# your code here
```


**Problem 2: (4 pts)** 

Next, we will be looking at how the number of UFO sightings has changed for five states since 1940 for both problems 2 and 3. Please follow these steps:

1. Filter the dataset to keep the following five states: TX, CA, NM, FL, NY
2. Keep only the records from 1940 and onwards. 
3. Find the number of records for each year and state. 
4. Display the new table below your code block

```{r}
# your code here
```

**Problem 3: (2 pts)** 

Use the new dataframe you made in Problem 2 and add an appropriate color scale from the `colorspace` package to the plot below.

```{r eval = FALSE}
new_df %>% # use the dataframe from Problem 2 here, and set eval = TRUE in the chunk header 
 ggplot(aes(x = year, y = count, color = state)) +
 geom_line() +
 xlab("Year") +
 ylab("UFO Sightings (Count)") +
 theme_bw()
```









