---
title: "Homework 7"
output:
  html_document:
    df_print: paged
---

```{r global_options, include=FALSE}
library(knitr)
library(tidyverse)
library(colorspace)
library(naniar)
opts_chunk$set(fig.align="center", fig.height=4, fig.width=5.5)

#data prep:
midwest2 <- midwest %>%
  filter(state != "IN")

#data prep for problem 3:
oceanbuoys$year <- factor(oceanbuoys$year)
oceanbuoys <- na.omit(oceanbuoys)
```

*Enter your name and EID here*

**This homework is due on April 4, 2022 at 11:00am. Please submit as a pdf file on Canvas.**

**Problem 1: (2 pts)** 

Use the color picker app from the **colorspace** package (`colorspace::choose_color()`) to create a qualitative color scale containing four colors. One of the four colors should be `#4D670C`, so you need to find three additional colors that go with this one. Use the function `swatchplot()` to plot your colors. `swatchplot()` takes in a vector of colors. 

```{r}
# your code goes here
```


**Problem 2: (4 pts)** 

For this problem, we will work with the `midwest2` dataset (derived from `midwest`). In the following plot, you may notice that the axis tick labels are smaller than the axis titles, and also in a different color (gray instead of black). 

1. Use the colors you chose in Problem 1 to color the points.
2. Make the axis tick labels the same size (`size = 12`) and give them the color black (`color = "black"`)
3. Set the entire plot background to the color `"#FBF9F6"`. Make sure there are no white areas remaining, such as behind the plot panel or under the legend.

```{r}
ggplot(midwest2, aes(popdensity, percollege, fill = state)) +
  geom_point(shape = 21, size = 3, color = "white", stroke = 0.2) +
  scale_x_log10(name = "population density") +
  scale_y_continuous(name = "percent college educated") +
  # your color choices go here in a scale fucntion. 
  theme_classic(12) +
  theme(
    # your theme customization code goes here
  )
```



**Problem 3: (4 pts)** 

For this problem, we will work with the `oceanbuoys` dataset from the `naniar` library that contains west pacific tropical atmosphere ocean data for 1993 and 1997. 

Write a function that converts temperature from Celcius to Fahrenheit. Then, use this function and any other data wrangling code you learned from class to make a summary table of average sea temperature and air temperature (in Fahrenheit) for 1997 and 1993. The formula for converting Celcius to Fahrenheit is `Fahrenheit = (Celcius*1.8) + 32`. 
```{r}
# your code goes here
```





