---
title: "Homework 8"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r global_options, include=FALSE}
library(knitr)
library(tidyverse)
library(broom)
opts_chunk$set(fig.align="center", fig.height=4, fig.width=5.5)

#data prep:
BA_degrees <- read_csv("https://wilkelab.org/SDS375/datasets/BA_degrees.csv")
BA_degrees
```

*Enter your name and EID here*

**This homework is due on April 4, 2023 at 11:00pm. Please submit as a pdf file on Canvas.**

**Problem 1: (6 pts)** The dataset `BA_degrees` contains information about the proportion of different degrees students receive, as a function of time.

```{r}
head(BA_degrees)
```

Create a subset of the `BA_degrees` dataset that only considers the degree fields "Business", "Education", and "Psychology". Then make a single plot that satisfies these three criteria:

(a) Plot a time series of the proportion of degrees (colum `perc`) in each field over time and create a separate panel per degree field. 
(b) Add a straight line fit to each panel.
(c) Order the panels by the difference between the maximum and the minimum proportion (i.e., the range of the data).


```{r}
# your code goes here
```

**Problem 2: (4 pts)** 
Create a single pipeline that fits a linear model to each of the three fields from Problem 1 and outputs results in a tidy linear model summary table. The first column of the table should be `field` and the remaining columns should contain the linear model summary statistics such as `r.squared` for each field. Display the resulting table below.

```{r}
# your code goes here
```









