---
title: "Homework 9"
---

```{r global_options, include=FALSE}
library(knitr)
library(tidyverse)
library(ggrepel)
library(colorspace)
library(palmerpenguins)

opts_chunk$set(fig.align="center", fig.height=4.326, fig.width=7)
```

*Enter your name and EID here*

**This homework is due on April 26, 2021 at 11:00pm. Please submit as a pdf file on Canvas.**

**Problem 1: (2 pts)** 

Use the color picker app from the **colorspace** package (`colorspace::choose_color()`) to create a qualitative color scale containing four colors. One of the four colors should be `#5626B4`, so you need to find three additional colors that go with this one.

```{r fig.width = 6, fig.height = 3}
# replace "#FFFFFF" with your own colors
colors <- c("#5626B4", "#FFFFFF", "#FFFFFF", "#FFFFFF")

swatchplot(colors)
```


**Problem 2: (4 pts)** Take the following scatter plot of the penguins dataset and make three modifications:

1. Use the colors you chose in Problem 1.
2. Improve the visual appearance by choosing a theme and cleaning up axis labels.
3. Remove the need for a legend by direct-labeling the points.

```{r fig.width = 6, fig.height = 0.75*6}
ggplot(penguins, aes(bill_length_mm, body_mass_g, color = species)) +
  geom_point(size = 2, na.rm = TRUE)
```

**Problem 3: (4 pts)** The following scatter plot shows per-capita income versus number of inhabitants in all Texas counties in 2010. Use `geom_text_repel()` to label a subset of the counties by name. You can choose the counties to subset as you wish. Also, choose a theme and clean up the axis labeling, and make any other improvements to the plot design you consider appropriate. 

**Hint:** If you're not sure how to select a subset of counties to label, check out the examples on the **ggrepel** website for some inspiration:
https://ggrepel.slowkow.com/articles/examples.html#examples-1

```{r fig.width = 5, fig.height = 0.75*5, message = FALSE}
tx_census <- read_csv("https://wilkelab.org/SDS375/datasets/US_census.csv") %>%
  filter(state == "Texas") %>%
  select(county = name, pop2010, per_capita_income)

tx_census %>%
  ggplot(aes(pop2010, per_capita_income)) +
  geom_point(size = 1.5) +
  scale_x_log10()
```

