---
title: "Project 1"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(colorspace)
knitr::opts_chunk$set(echo = TRUE)

olympics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-07-27/olympics.csv')

# data prep:
olympics_top <- olympics %>%
  mutate(team = ifelse(team == "Soviet Union", "USSR/Russia", team)) %>%
  mutate(medal = ifelse(is.na(medal), "no medal", medal)) %>%
  mutate(gold = ifelse(medal == "Gold", "gold medalist", "no gold")) %>%
  mutate(medalist = ifelse(medal != "no medal", "medalist", "no medal")) %>%
  mutate(decade = year - (year %% 10)) %>%
  group_by(team) %>%
  mutate(count = n()) %>%
  filter(count > 5500) %>%
  select(-count) %>%
  ungroup() %>%
  group_by(sport, season) %>%
  mutate(count = n()) %>%
  filter(count > 700) %>%
  select(-count) %>%
  ungroup() %>%
  filter(!sport %in% c("Art Competitions", "Handball", "Short Track Speed Skating", "Modern Pentathlon", "Shooting", "Canoeing", "Athletics", "Hockey", "Water Polo"))
```

We will work with the dataset `olympics_top` that contains data for the Olympic Games from Athens 1896 to Rio 2016 and has been derived from the `olympics` dataset. More information about the dataset can be found at: https://github.com/rfordatascience/tidytuesday/blob/master/data/2021/2021-07-27/readme.md
The dataset, `olympics_top`, contains four new columns: `decade` (the decade during which the Olympics took place), `gold` (whether or not the athlete won a gold medal), `medalist` (whether or not the athlete won any medal) and `medal` (if the athlete won "Gold", "Silver", "Bronze" or received "no medal").

### **Part 1**

**Question:** Which sports have the tallest or shortest athletes? And does the distribution of heights change for the various sports between medalists and non-medalists? 

We recommend you use box plots for the first part of the question and use a ridgeline plot for the second part of the question. 

**Hints:**

- To order boxplots by the median, you may have add the following to your ordering function to remove missing values before ordering: `na.rm = TRUE`

- To trim the tails in your ridgeline plot, you can set 
`rel_min_height = 0.01` inside `geom_density_ridges()`.

**Introduction:** *Your introduction here.*

**Approach:** *Your approach here.*

**Analysis:**

```{r}
# Your R code here
```

```{r}
# Your R code here
```


**Discussion:** *Your discussion of results here.*

### **Part 2**

**Question:** *Your question here.*

**Introduction:** *Your introduction here.*

**Approach:** *Your approach here.*

**Analysis:**
```{r }
# Your R code here
```

```{r }
# Your R code here
```

**Discussion:** *Your discussion of results here.*
