---
title: "Project 2"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```


In this project, you will be working with a dataset about the members of Himalayan expeditions:
```{r message = FALSE}
members <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/members.csv')

members
```

More information about the dataset can be found at https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-22/readme.md and https://www.himalayandatabase.com/.

**Hints:**

- Make sure your two questions are actually questions, and not veiled instructions to perform a particular analysis.

- Remember your code needs to contain at least three data manipulation functions for data wrangling before you plot. You are allowed to put all the data wrangling into the answer for one of the two questions.

- Adjust `fig.width` and `fig.height` in the chunk headers to customize figure sizing and figure aspect ratios.

You can delete these instructions from your project. Please also delete text such as *Your approach here* or `# Q1: Your R code here`.

**Question 1:** *Your question 1 here.*

**Question 2:** *Your question 2 here.*

**Introduction:** *Your introduction here.*

**Approach:** *Your approach here.*

**Analysis:**

```{r fig.width = 5, fig.height = 5}
# Q1: Your R code here
```

```{r fig.width = 5, fig.height = 5}
# Q2: Your R code here
```

**Discussion:** *Your discussion of results here.*
