---
title: "Project 3"
output:
  html_document:
    theme:
      version: 4
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```


In this project, you will be working with a dataset of your own choosing. **Important:** The dataset needs to be picked from the [TidyTuesday project,](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023), and it needs to be one that has been released between May 30, 2023 and December 26, 2023 (both dates inclusive).

**Hints:**

- Read in your data with `readr::read_csv()`, as we have done in prior projects. **Do not use the tidytuesdayR package.** The TidyTuesday site explains for each dataset how it can be read with `readr::read_csv()`, under "Get the data here", part "Or read in the data manually".
  
- Make sure your question is actually a question, and not a veiled instruction to perform a particular analysis.

- Adjust `fig.width` and `fig.height` in the chunk headers to customize figure sizing and figure aspect ratios. These numbers are measured in inches and will usually fall between 4 and 10.

You can delete these instructions from your project. Please also delete text such as *Your approach here* or `# Code for figure 1 here`.

**Introduction:** *Your introduction here.*

**Question:** *Your question here.*

**Approach:** *Your approach here.*

**Analysis:**

```{r}
# Data loading/wrangling/analysis code here
```

```{r fig.width = 5, fig.height = 5}
# Code for figure 1 here
```

```{r fig.width = 5, fig.height = 5}
# Code for figure 2 here
```

**Discussion:** *Your discussion of results here.*
