---
title: "Project 1"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

This is the dataset you will be working with:
```{r message = FALSE}
olympics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-07-27/olympics.csv')

olympics_alpine <- olympics %>% 
  filter(!is.na(weight)) %>%             # only keep athletes with known weight
  filter(sport == "Alpine Skiing") %>%   # keep only alpine skiers
  mutate(
    medalist = case_when(                # add column to 
      is.na(medal) ~ FALSE,              # NA values go to FALSE
      !is.na(medal) ~ TRUE               # non-NA values (Gold, Silver, Bronze) go to TRUE
    )
  )
```

`olympics_alpine` is a subset of `olympics` and contains only the data for alpine skiers. More information about the original `olympics` dataset can be found at https://github.com/rfordatascience/tidytuesday/tree/master/data/2021/2021-07-27/readme.md and https://www.sports-reference.com/olympics.html.

For this project, use `olympics_alpine` to answer the following questions about the weights of alpine skiers:

1. Are there weight differences for male and female Olympic skiers who were successful or not in earning a medal? 
2. Are there weight differences for skiers who competed in different alpine skiing events?
3. How has the weight distribution of alpine skiers changed over the years? 

You should make one plot per question.

**Hints:**

- We recommend you use a violin plot for question 1 and boxplots for questions 2 and 3. However, you are free to use any of the plots we have discussed in class so far.
- For question 3, it may be helpful to consider only a subset of alpine skiers, such as those who competed in a specific event. 
- To make a series of boxplots over time, you will have to add the following to your `aes()` statement: `group = year`. 
- It can be a bit tricky to re-label facets generated with `facet_wrap()`. The trick is to add a `labeller` argument, for example:
```r
 + facet_wrap(
    # your other arguments to facet_wrap() go here
    ...,
    # this replaces "TRUE" with "medaled" and "FALSE" with "did not medal"
    labeller = as_labeller(c(`TRUE` = "medaled", `FALSE` = "did not medal"))
  )
```

**Introduction:** *Your introduction here.*

**Approach:** *Your approach here.*

**Analysis:**

```{r}
# Your R code here
```

```{r}
# Your R code here
```

```{r}
# Your R code here
```

**Discussion:** *Your discussion of results here.*
