---
#############################################################
#                                                           #
# Click on "Run Document" in RStudio to run this worksheet. #
#                                                           #
#############################################################
title: "Data wrangling 1"
author: "Claus O. Wilke"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(palmerpenguins)
knitr::opts_chunk$set(echo = FALSE, comment = "")
```

## Introduction

In this worksheet, we will discuss how to perform basic data manipulations, such as filtering data rows that meet certain conditions, choosing data columns, and arranging data in ascending or descending order.

We will be using two R packages, **tidyverse** for the data manipulation functions `%>%`, `filter()`, `select()`, and `arrange()`, and **palmerpenguins** for the `penguins` dataset.

```{r library-calls, echo = TRUE, eval = FALSE}
# load required libraries
library(tidyverse)
library(palmerpenguins)
```

## The pipe (`%>%`, read: "and then")

When writing complex data analysis pipelines, we frequently use the pipe operator `%>%` to move data from one analysis step to the next. The pipe is pronounced "and then", and it takes the data on its left and uses it as the first argument for the function on its right.

For example, to see the first few lines of a dataset, we often write `head(data)`. Instead, we can write `data %>% head()`.

Try this yourself. Write code that displays the first few lines of the `penguins` dataset, using `%>%` and `head()`:

```{r pipe, exercise=TRUE, exercise.eval=TRUE}
penguins
```

```{r pipe-hint}
penguins ___ head()
```

```{r pipe-solution}
penguins %>% head()
```


## Filtering data

The function `filter()` allows you to find rows in a dataset that meet one or more specific conditions. The syntax is `data %>% filter(condition)`, where condition is a logical condition. For example, `filter(x > 5)` would pick all rows for which the value in column `x` is greater than 5.

As an example, the following code picks all penguins from the island Biscoe:

```{r filter-island-demo, echo=TRUE}
penguins %>%
  filter(island == "Biscoe")
```

Now it's your turn. Pick all penguins from the island Dream:

```{r filter-island, exercise=TRUE}
penguins %>%
  filter(___)
```

```{r filter-island-solution}
penguins %>%
  filter(island == "Dream")
```

### Filtering for multiple conditions

You can also state multiple conditions, separated by a comma. For example, `filter(x > 5, y < 2)` would pick all rows for which the value in the column `x` is greater than 5 and the value in the column `y` is less than 2. Note that the conditions are combined via logical and, both need to be satisfied for the row to be picked.

To try this out, pick all penguins of species Chinstrap from the island Dream:

```{r filter-species-island, exercise=TRUE}
penguins %>%
  filter(___, ___)
```

```{r filter-species-island-hint}
penguins %>%
  filter(species == ___, island == ___)
```

```{r filter-species-island-solution}
penguins %>%
  filter(species == "Chinstrap", island == "Dream")
```

```{r penguin-island, echo=FALSE}
question("Which penguins species can be found on the island Dream?",
  answer("Adelie"),
  answer("Chinstrap"),
  answer("Gentoo"),
  answer("Adelie and Chinstrap", correct = TRUE),
  answer("Adelie and Gentoo")
)
```

## Choosing data columns

The function `select()` allows you to pick specific data columns by name. This is frequently useful when a dataset has many more columns than we are interested in at the time. For example, if we are only interested in the penguins' species, island, and sex, we could select these three columns:

```{r select-example, echo = TRUE}
penguins %>%
  select(species, island, sex)
```

Try this yourself, picking the columns representing the penguin species, the bill length, and then flipper length.

```{r select-exercise, exercise=TRUE}
penguins %>%
  select(___)
```

```{r select-exercise-solution}
penguins %>%
  select(species, bill_length_mm, flipper_length_mm)
```

### Choosing columns for removal

Another situation that arises frequently is one where we want to remove specific columns. We can also do this with `select()`, but now write `select(-column)` to remove one or more columns.

Try this. Remove the column `species`.

```{r select-remove, exercise = TRUE}
penguins %>%
  select(___)
```

```{r select-remove-solution}
penguins %>%
  select(-species)
```


And now remove both `species` and `island`.

```{r select-remove2, exercise = TRUE}
penguins %>%
  select(___)
```

```{r select-remove2-hint}
penguins %>%
  select(-___, -___)
```

```{r select-remove2-solution}
penguins %>%
  select(-species, -island)
```

## Sorting data

The function `arrange()` allows you to sort data by one or more columns. For example, `data %>% arrange(x)` would sort the data by increasing values of `x`, and `data %>% arrange(x, y)` would sort the data first by `x` and then, for ties in `x`, by `y`.

As an example, the following code sorts penguins by their flipper length:

```{r arrange-flipper-demo, echo=TRUE}
penguins %>%
  arrange(flipper_length_mm)
```

Now it's your turn. Sort the penguins by bill length:

```{r arrange-bill-length, exercise=TRUE}
penguins %>%
  arrange(___)
```

```{r arrange-bill-length-solution}
penguins %>%
  arrange(bill_length_mm)
```

### Arranging in descending order

To arrange data in descending order, enclose the data column in `desc()`. For example, `data %>% arrange(desc(x))` would sort the data by decreasing values of `x`. (`desc` stands for "descending".)

Try this out. Sort the penguins by body mass, from largest to smallest:

```{r arrange-body-mass, exercise=TRUE}
penguins %>%
  arrange(___)
```

```{r arrange-body-mass-hint}
penguins %>%
  arrange(___(body_mass_g))
```

```{r arrange-body-mass-solution}
penguins %>%
  arrange(desc(body_mass_g))
```

