---
#############################################################
#                                                           #
# Click on "Run Document" in RStudio to run this worksheet. #
#                                                           #
#############################################################
title: "Getting things into the right order"
author: "Claus O. Wilke"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(palmerpenguins)
library(gapminder)
knitr::opts_chunk$set(echo = FALSE, comment = "")
```

## Introduction

In this worksheet, we will continue with basic data manipulations, now moving on to grouping and summarizing, making data tables wider or longer, and joining data tables.

We will be using two R packages, **tidyverse** for various data manipulation functions and **palmerpenguins** and **gapminder** for the `penguins` and `gapminder` datasets, respectively.


```{r library-calls, echo = TRUE, eval = FALSE}
# load required libraries
library(tidyverse)
library(palmerpenguins)
library(gapminder)
```


We will be working with the dataset `penguins` containing data on individual penguins on Antarctica.
```{r echo = TRUE}
penguins
```


**gapminder**
```{r echo = TRUE}
gapminder
```

## Title

```{r echo = TRUE, eval = FALSE}
gapminder %>%
  filter(continent == "Asia") %>%
  #filter(country %in% c("Norway", "Portugal", "Spain", "Austria")) %>%
  mutate(country = fct_reorder(country, lifeExp, min)) %>%
  ggplot(aes(year, country, fill = lifeExp)) + geom_tile() +
  scale_fill_viridis_c(option = "E")
```
