---
title: "Homework 10"
output:
  html_document:
    df_print: paged
---

```{r global_options, include=FALSE}
library(knitr)
library(tidyverse)
library(ggdendro)
library(sf)
library(colorspace)
opts_chunk$set(fig.align="center", fig.height=4.326, fig.width=7)
```

*Enter your name and EID here*

**This homework is due on April 25, 2022 at 11:00am. Please submit as a pdf file on Canvas.**

**Problem 1: (3 pts)** For Problem 1, we will be using `happiness` taken from the World Happiness Report. You can read more about the dataset here: https://www.kaggle.com/datasets/unsdsn/world-happiness.

```{r message = FALSE}
# data preparation
happiness <- read_csv("https://wilkelab.org/SDS375/datasets/happiness.csv")
head(happiness)
```

a) Perform hierarchical clustering of the countries and calculate the distance matrix. You do not need to display the distance matrix.

b) Display clustering results in a dendrogram.

```{r}
# your code here
```


**Problem 2: (3 pts)** Use the clustering results you found in Problem 1 for Problem 2. 

a) Assign clusters by cutting the dendrogram.

b) Plot a scatter plot for two numeric variables of your choice from `happy_data` and add cluster info into scatterplot.

c) Interpret the plot.

```{r, message=FALSE}
# your code here
```

*your answer here*

**Problem 3: (4 pts)** For Problem 3, we will work with the dataset `texas_income`.

a) Bin the `median_income` column into 3 bins (20K-40K, 40K-60K, 60K-90K). Hint: use `case_when()`.

b) Make a choropleth map of Texas counties colored by median income bin (3 total colors).

c) Use an appropriate color scale and use a theme that shows longitude and latitude (nearly any theme other than `theme_void()` will work).

```{r message = FALSE}
# data preparation
texas_income <- readRDS(url("https://wilkelab.org/SDS375/datasets/Texas_income.rds"))
head(texas_income)
```

```{r}
# your code here
```

