---
title: "Homework 3"
---

```{r global_options, include=FALSE}
library(knitr)
library(tidyverse)
options(scipen = 999)
opts_chunk$set(fig.align="center", fig.height=4, fig.width=5.5)

# data prep:
IL_pop <- midwest %>%
  filter(state == "IN") %>%
  arrange(desc(poptotal)) %>%
  mutate(row = row_number()) %>%
  filter(row %in% c(1, 2, 3, 4, 5, 12, 20, 25, 40, 50, 60, 75)) %>%
  select(c(county, poptotal))

```

*Enter your name and EID here*

**This homework is due on Feb. 7, 2022 at 11:00am. Please submit as a pdf file on Canvas.**


**Problem 1: (2 pts)** For problem 1, we will work with the `diamonds` dataset. See here for details: https://ggplot2.tidyverse.org/reference/diamonds.html.

```{r}
diamonds
```
(a) Use ggplot to make a bar plot of the total diamond count per `color` and show the proportion of each `cut` within each `color` category. 
(b) In two sentences, explain when to use `geom_bar()` instead of `geom_col()`. Which of these functions requires only an `x` or `y` variable?

```{r}
# your code goes here
```

**Problem 2: (4 pts)**
For problems 2 and 3, we will work with the dataset `IL_pop` that contains Illinois state demographics and has been derived from the `midwest` dataset provided by **ggplot2**. See here for details of the original dataset: https://ggplot2.tidyverse.org/reference/midwest.html. `IL_pop` contains two columns: `county` and `poptotal` (the county's total population).

```{r}
IL_pop
```

(a) Use ggplot to make a scatter plot of `county` vs total population (column `poptotal`) and order the counties by increasing population. 
(b) Rename the axes and set appropriate limits, breaks and labels. Note: Do not use `xlab()` or `ylab()` to label the axes.

```{r}
# your code goes here
  
```

**Problem 3: (4 pts)** 

(a) Modify the plot from Problem 2 by changing the scale for `poptotal` to logarithmic.
(b) Adjust the limits, breaks and labels for the logarithmic scale.

```{r}
# your code goes here
```






