---
title: "Homework 4"
---

```{r global_options, include=FALSE}
library(knitr)
library(ggplot2)
library(colorspace)
library(cowplot)

opts_chunk$set(fig.align="center", fig.height=4, fig.width=5.5)
```

*Enter your name and EID here*

**This homework is due on Feb. 15, 2021 at 11:00pm. Please submit as a pdf file on Canvas.**

**Problem 1: (5 pts)** We will work with the `midwest` dataset provided by **ggplot2**. See here for details: https://ggplot2.tidyverse.org/reference/midwest.html

For the following plot, add an appropriate color scale from the **colorspace** package and explicitly set the palette you want to use. Also add an appropriate theme that works well with the color scale you have chosen. Explain in 2-3 sentences your choice of color scale and theme.

```{r}
ggplot(midwest, aes(popdensity, percollege, fill = state)) +
  geom_point(shape = 21, size = 2.5, color = "white", stroke = 0.5) +
  scale_x_log10(name = "population density") +
  scale_y_continuous(name = "percent college educated") +
  NULL # replace NULL with your color scale and theme code
```

*Provide your explanation here.*

**Problem 2: (2 pts)** The following is a plot of the `diamonds` dataset provided by **ggplot2**. See here for details: https://ggplot2.tidyverse.org/reference/diamonds.html

```{r}
ggplot(diamonds, aes(carat, price, color = cut)) +
  geom_point()
```

As you can see, if we don't specify a color scale ggplot chooses the viridis scale by default. Explain in 2-3 sentences why the viridis scale is an appropriate choice here.

*Provide your explanation here.*

**Problem 3: (3 pts)** For this problem, we will go back to the `midwest` dataset. In the following plot, you may notice that the axis tick labels are smaller than the axis titles, and also in a different color (gray instead of black). Make the axis tick labels the same size (`size = 12`) and give them the color black (`color = "black"`). Then, set the entire plot background to the color `"#F3F8FF"`. Make sure there are no white areas remaining, such as behind the plot panel or under the legend.

```{r}
ggplot(midwest, aes(popdensity, percollege, fill = state)) +
  geom_point(shape = 21, size = 2, color = "white", stroke = 0.2) +
  scale_x_log10(name = "population density") +
  scale_y_continuous(name = "percent college educated") +
  theme_classic(12) +
  theme(
    # your theme customization code goes here
  )
```
