---
title: "Homework 4"
---

```{r global_options, include=FALSE}
library(knitr)
library(ggplot2)
library(ggridges)
opts_chunk$set(fig.align="center", fig.height=4, fig.width=5.5)
```

*Enter your name and EID here*

**This homework is due on Feb. 14, 2023 at 11:00pm. Please submit as a pdf file on Canvas.**


**Problem 1: (4 pts)** We will work with the `mpg` dataset provided by **ggplot2**. See here for details: https://ggplot2.tidyverse.org/reference/mpg.html

Make two different strip charts of highway fuel economy (`hwy`) versus number of cylinders (`cyl`), the first one without horizontal jitter and second one with horizontal jitter. In both plots, please replace names of the data columns (`hwy`, `cyl`) along the axes with nice, easily readable lables.

Explain in 1-2 sentences why the plot without jitter is misleading.

Hint: Make sure you do not accidentally apply vertical jitter. This is a common mistake many people make.

```{r}
# your code goes here.
```

*Your explanation goes here.*


**Problem 2: (6 pts)** For this problem, we will continue working with the `mpg` dataset. Visualize the distribution of each car's city fuel economy by class (`class`) and type of drive train (`drv`) with (i) boxplots and (ii) ridgelines. Make one plot per geom and do not use faceting. In both cases, put city mpg on the x axis and class on the y axis. Use color to indicate the car's drive train. As in Problem 1, rename the axis labels.

The boxplot ggplot generates will have a problem. Describe what the problem is. (You do not have to solve it.)

Hint: To change the name of the legend, use `+ labs(fill = "legend name")`

```{r}
# your code goes here.
```

*Your explanation goes here.*









