---
title: "Homework 5"
---

```{r global_options, include=FALSE}
library(knitr)
library(tidyverse)
library(colorspace)
library(cowplot)

opts_chunk$set(fig.align="center", fig.height=4.326, fig.width=7)
```

*Enter your name and EID here*

**This homework is due on March 22, 2021 at 11:00pm. Please submit as a pdf file on Canvas.**

For both problems in this homework, we will work with the `internet` dataset. It contains the number of internet users over time for 20 select countries. Internet users are reported as percentages.

```{r message = FALSE}
internet <- read_csv("https://wilkelab.org/SDS375/datasets/internet.csv")
internet
```

**Problem 1: (5 pts)** 

Take the following plot and make two modifications:

1. Put the countries into a meaningful order
2. Use scale and theme functions to improve the visual design of the plot

Grading rubric: 2 pts for ordering, 3 pts for visual design

```{r}
internet %>%
  ggplot(aes(x = year, y = country, fill = users)) +
  geom_tile()
```

**Problem 2: (5 pts)** Take the plot from the previous problem and make the following modifications:

1. Select a subset of 6 countries, using arbitrary criteria
2. Use `geom_line()` to show internet users over time, and use facets to show the different countries
3. Use a different ordering than you used in Problem 1.
4. Modify the visual design so it is appropriate for your new plot

**Hint:** To get started, see slides 33 to 43 in the class on getting things into the right order: https://wilkelab.org/SDS375/slides/getting-things-in-order.html#33

Grading rubric: 3 pts for making the right plot, 2 pts for visual design

```{r}
# your code goes here
```

