---
title: "Homework 8"
---

```{r global_options, include=FALSE}
library(knitr)
library(tidyverse)
library(sf)

opts_chunk$set(fig.align="center", fig.height=4.326, fig.width=7)
```

*Enter your name and EID here*

**This homework is due on April 19, 2021 at 11:00pm. Please submit as a pdf file on Canvas.**

In this homework, we will work with two datasets, `US_counties` and `US_census`. The dataset `US_counties` contains the geometry of each county in the US and thus can be used for drawing maps. The dataset `US_census` contains numerous pieces of information about US counties obtained from the US census. Both datasets have a column `FIPS` which can be used to uniquely identify each county in each dataset.

```{r message = FALSE}
# data preparation
US_counties <- readRDS(url("https://wilkelab.org/SDS375/datasets/US_counties.rds")) %>%
  rename(FIPS = GEOID)

# workaround for missing rgdal on edupod
st_crs(US_counties) <- NA

US_census <- read_csv(
  "https://wilkelab.org/SDS375/datasets/US_census.csv",
  col_types = cols(FIPS = "c")
)
```

**Problem 1: (6 pts)** Make a choropleth map of the percent home-ownership (column `home_ownership` in `US_census`) for all counties in the US. Choose an appropriate color scale and design for this plot. You may notice that there is one county in Alaska for which home-ownership data is not available. Write data analysis code to identify this county.

**Hints:**

1. Use `theme_void()` as your theme

2. You will have to join `US_counties` and `US_census`. Join them by the `FIPS` column.

3. To make nice percent labels, you can use `label = scales::label_percent(scale = 1)` in your color scale function.

4. To find rows with missing data, you may want to use the function `is.na()`.

Grade breakdown: 2pt for the plot, 2pt for the plot design, and 2pt for identifying the county in Alaska for which home ownership data is not available.

```{r}
# your code goes here
```

**Problem 2: (4 pts)** Make a choropleth map of the percent foreign born (column `foreign_born` in `US_census`) for the counties in Texas only. Use a different color scale than you used for Problem 1 and use a theme that shows longitude and latitude (nearly any theme other than `theme_void()` will work).

Grade breakdown: 3pt for the plot, 1pt for the plot design

```{r}
# your code goes here
```


