---
title: "Project 1"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

*Enter your name and EID here*

This is the dataset you will be working with:
```{r message = FALSE}
members <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/members.csv')

members_everest <- members %>%
  filter(peak_name == "Everest") %>% # only keep expeditions to Everest
  filter(!is.na(age)) %>%     # only keep expedition members with known age
  filter(year >= 1960)        # only keep expeditions since 1960 
```

More information about the dataset can be found at https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-22/readme.md and https://www.himalayandatabase.com/.

### **Part 1**

**Question:** Are there age differences for expedition members who were successful or not in climbing Mt. Everest with or without oxygen, and how has the age distribution changed over the years? 

We recommend you use a violin plot for the first part of the question and faceted boxplots for the second question part of the question. 

**Hints:**

- To make a series of boxplots over time, you will have add the following to your `aes()` statement: `group = year`. 
- It can be a bit tricky to re-label facets generated with `facet_wrap()`. The trick is to add a `labeller` argument, for example:
```r
 + facet_wrap(
    # your other arguments to facet_wrap() go here
    ...,
    # this replaces "TRUE" with "summited" and "FALSE" with "did not summit"
    labeller = as_labeller(c(`TRUE` = "summited", `FALSE` = "did not summit"))
  )
```

**Introduction:** *Your introduction here.*

**Approach:** *Your approach here.*

**Analysis:**
```{r }

# Your R code here

```

```{r }

# Your R code here

```

**Discussion:** *Your discussion of results here.*

### **Part 2**

**Question:** *Your question here.*

**Introduction:** *Your introduction here.*

**Approach:** *Your approach here.*

**Analysis:**
```{r }

# Your R code here

```

```{r }

# Your R code here

```

**Discussion:** *Your discussion of results here.*
