---
title: "Project 2"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

*Enter your name and EID here*

This is the dataset you will be working with:
```{r message = FALSE}
members <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/members.csv')
```

More information about the dataset can be found at https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-22/readme.md and https://www.himalayandatabase.com/.

### **Part 1**

**Question:** Looking only at expeditions to Mt.Everest since 1960, how do deaths in each season break down by the seven most common causes?

To answer this question, create a summary table and one visualization. The summary table should have 4 columns: "death_cause", "Spring", "Summer", "Autumn" and "Winter", where the seasons columns have the raw number of deaths for each cause in the first column. Remember to replace any `NA` values with `0`. 

We recommend you use faceted pie charts for the visualization. The visualization should show the relative proportion of the 7 most common death causes for each season. Include an additional category called "other" for all other death causes. 

Please note that we are not asking you to find the seven most common causes of death separately for each season. Find the seven most common causes of death overall and then perform the analysis by season. 

**Introduction:** *Your introduction here.*

**Approach:** *Your approach here.*

**Analysis:**

```{r}
# your code here
```

```{r}
# your code here
```

**Discussion:** *Your discussion of results here.*

### **Part 2**

**Question:** *Your question here.*

**Introduction:** *Your introduction here.*

**Approach:** *Your approach here.*

**Analysis:**

```{r}
# your code here
```

```{r}
# your code here
```

**Discussion:** *Your discussion of results here.*

